---
title: "Analysis of Screening Logs"
author: "Kai-Lin Jen, Kaelin Loftus and Dave Bridges"
date: "May 1, 2026"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  pdf: default
knitr:
  opts_chunk:
    fig.path: "figures/"
    dev: ["png", "pdf"]  # Remove !expr, just use array syntax
    fig.keep: "all"
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)
library(knitr)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

To quantify screening efforts for the calcium-cholesterol meta-analysis

## Data Entry

```{r data-entry}
log.file <- 'Screening Log for Calcium and Cholesterol Meta-Analysis - Screening Log.csv'
screening.log <- read_csv(log.file,
                          col_types = cols(
  Key = col_character(),
  Title = col_character(),
  DOI = col_character(),
  `Abstract Note` = col_character(),
  DateScreened = col_date(format = "%m/%d/%y"),
  Screener = col_factor(),
  Decision = col_character(),
  FullTextView = col_factor(),
  ExclusionReason = col_factor(),
  Notes = col_character()
))
```

## Analysis

I wanted to identify papers with the words correlation, association in the title or abstract

```{r correlation-finding}
correlation.association.papers <- screening.log %>%
  filter(str_detect(Title, regex("correlation|association", ignore_case = TRUE)) |
           str_detect(`Abstract Note`, regex("correlation|association", ignore_case = TRUE)))

#and papers with just hte word correlation in the title or abstract
correlation.papers <- screening.log %>%
  filter(str_detect(Title, regex("correlation", ignore_case = TRUE)) |
           str_detect(`Abstract Note`, regex("correlation", ignore_case = TRUE)))
```

There were `r nrow(correlation.association.papers)` papers with the words correlation or association in the title or abstract.`. There were `r nrow(correlation.papers)` papers with the word correlation in the title or abstract.

## Session Information

```{r session-information}
sessionInfo()
```
