---
title: "Validation of Pre-Screening of Deduplicated Articles by GPT-5"
author: "Dave Bridges"
date: "2025-09-21"
editor: source
format: 
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
    fig-path: "figures/"
  gfm:
    html-math-method: webtex
theme: journal
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

I used a GPT-based screen to evaluate the deduplicated list to remove meta-analyses, nonhuman studies, and review articles

## GPT Prompt

You are acting as an expert systematic-reviewer following PRISMA guidelines.  
You will classify each article in a deduplicated list by reading its Title and Abstract (fields: `Title`, `Abstract Note`).

Flag any article as:
- "Case report/series" if the title or abstract clearly states “case report”, “case series”, or “a case of …”.
- "Review/Systematic review/Meta-analysis" if the title or abstract clearly states “systematic review”, “meta-analysis”, “review article”, or “narrative review”.
- "Non-human study" only if it is clear the study population is not human and no human data are included.
- "Unsure" if you cannot confidently classify it into one of the above categories.

Be conservative: only flag an article if it clearly matches the criteria.  
Do not exclude any article with both human and non-human data.

Output a CSV file with two columns:  
`Key` (article key from input file) and `Reason` (one of the four categories above).

## Raw Data

Describe your raw data files, including what the columns mean (and what units they are in).

```{r data-input}
library(readr) #loads the readr package
library.file <- 'deduplicated-library.csv'
deduplicated.file <- 'classified_articles.csv'

deduplicatd.library <- read_csv(library.file)
annotation <- read_csv(deduplicated.file)

combined.file <- left_join(deduplicatd.library,annotation, by="Key") |>
  mutate(Key = as.factor(Reason)) |>
  mutate(method_ineligible = ifelse(Reason !="Unsure", 1, 0))
```

These data can be found in `r getwd()` in files named `r library.file` and `r deduplicated.file`.  This input file was most recently updated on `r format(file.info(library.file)$mtime, "%Y-%m-%d")`.  This script was most recently updated on `r date()`.

## Analysis

### Manual Screening

```{r manual-screening}
added.fields <- c('DateScreened','Screener','Decision','FullTextView','ExclusionReason','Notes')
deduplicatd.library |>
  select(Key,Title,DOI,`Abstract Note`) |>
  mutate(!!!set_names(rep(NA, length(added.fields)), added.fields)) |>
  write_csv("Screening Log.csv")
```

This screen flagged `r combined.file |> filter(method_ineligible==1) |> nrow()` out of `r combined.file |> nrow()` articles.

### Annotations by Key

```{r annotations-by-key}
combined.file %>%
  group_by(Reason) |>
  count() -> annotation.counts
library(knitr)
library(kableExtra)
kable(annotation.counts, caption="Annotated articles by keys")

ggplot(annotation.counts |> filter(Reason!="Unsure"),
       aes(x=reorder(Reason,desc(n)),y=n)) +
  geom_bar(stat='identity') +
  theme_classic(base_size=16) +
  labs(y='Removed articles',
       x='') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

## Interpretation

A brief summary of what the interpretation of these results were

## Session Information

```{r session-information}
sessionInfo()
```
