---
title: "Analysis of Screening Logs"
author: "Kai-Lin Jen and Dave Bridges"
date: "October 14, 2025"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  pdf: default
knitr:
  opts_chunk:
    fig.path: "figures/"
    dev: ["png", "pdf"]  # Remove !expr, just use array syntax
    fig.keep: "all"
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)
library(knitr)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

To quantify screening efforts for the calcium-cholesterol meta-analysis

## Data Entry

```{r data-entry}
log.file <- 'Screening Log - Screening Log.csv'
screening.log <- read_csv(log.file,
                          col_types = cols(
  Key = col_character(),
  Title = col_character(),
  DOI = col_character(),
  `Abstract Note` = col_character(),
  DateScreened = col_date(format = "%m/%d/%y"),
  Screener = col_factor(),
  Decision = col_character(),
  FullTextView = col_factor(),
  ExclusionReason = col_factor(),
  Notes = col_character()
))
```

## Analysis

To this point we have screened 

We have decided to review the full text of `r nrow(filter(screening.log, FullTextView == "Yes"))` articles and not to review the full text of `r nrow(filter(screening.log, FullTextView == "No"))` articles.  `r nrow(filter(screening.log, is.na(FullTextView)))` articles have not yet been screened.

### Exclusion Reasons

```{r exclusion-reasons}
screening.log |>
  filter(FullTextView == "No") |>
  group_by(ExclusionReason) |>
  summarize(n = n()) |>
  arrange(desc(n)) |>
  kable()
```

## Session Information

```{r session-information}
sessionInfo()
```
