---
title: "Layer 4c — Outcome ascertainment swap (Task B): the decisive test"
author: "Dave Bridges and Katie Kittell"
date: today
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
knitr:
  opts_chunk:
    fig.path: "figures/"
    dev: ["png", "pdf"]
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| message: false
library(TwoSampleMR)
library(ieugwasr)
library(MendelianRandomization)
library(knitr)
source(here::here("R", "helpers.R"))
source(here::here("R", "run_taskB.R"))
cfg <- load_config()
set.seed(cfg$seed)
```

## Purpose

The primary outcome **Bellenguez 2022** is **proxy-majority** (≈39,106 clinical cases vs
≈46,828 UK Biobank by-proxy cases; proxy fraction ≈ 0.55). Proxy / prevalent sampling biases
AD *risk* alleles downward — toward apparent *protection*. If the protective smoking→AD
signal is a selection/collider artefact of proxy ascertainment, it should **attenuate toward
null** when the outcome is a **de-selected, clinically-ascertained** AD GWAS.

This layer re-estimates the effect against a **gradient of by-proxy fraction**, on a
harmonised per-SD-CPD log-OR scale (fixing the scale mismatch noted in SUMMARY.md), and tests
attenuation formally:

| Outcome | Proxy fraction | UKB / proxy |
|---|---|---|
| Kunkle 2019 (IGAP) | 0.00 | clinical, no UKB |
| Lambert 2013 (IGAP) | 0.00 | clinical, no UKB |
| Bellenguez 2022 (primary) | 0.55 | proxy-majority |
| Schwartzentruber 2021 (family-history GWAX) | 1.00 | pure UKB by-proxy |

The Schwartzentruber GWAX outcome is on a family-history scale; its MR estimate is rescaled
×2 to approximate case-control log-OR (each first-degree relative shares ~50% of the genome).
This rescale is **approximate and flagged**; the decisive contrast — proxy-majority Bellenguez
vs the clinical-only Kunkle/Lambert — needs no rescaling and is fully scale-comparable.

## Effect across the ascertainment gradient

```{r swap}
# live_or_cache: never overwrite verified results with a failed/empty OpenGWAS pull.
live <- if (opengwas_ok()) tryCatch(run_taskB(cfg), error = function(e) NULL) else NULL
res <- live_or_cache(live, here::here("results", "ascertainment_swap.csv"), "b")
res |>
  filter(status == "ok") |>
  transmute(set, outcome, proxy_frac, n = nsnp,
            b = round(b,3), OR = round(or,3),
            `95% CI` = sprintf("[%.3f, %.3f]", ci_lo, ci_hi),
            p = signif(pval,2)) |>
  arrange(set, proxy_frac) |>
  kable(caption = "Task B: smoking→AD effect by outcome proxy fraction (per-SD-CPD log-OR)")
```

## Formal attenuation test

```{r tests}
# Difference: proxy-majority Bellenguez vs the inverse-variance pooled clinical (proxy=0)
# anchor; meta-regression of effect on proxy fraction across the gradient (per instrument set).
tests <- live_or_cache(tryCatch(taskB_tests(res), error = function(e) NULL),
                       here::here("results", "ascertainment_metareg.csv"), "b_clinical")
tests |>
  transmute(set,
            b_clinical = round(b_clinical,3), b_proxymaj = round(b_proxymaj,3),
            z_diff = round(z_diff,2), p_diff = signif(p_diff,2),
            `metareg slope (b~proxy)` = round(metareg_slope,3),
            metareg_p = signif(metareg_p,2)) |>
  kable(caption = "Proxy→clinical difference + meta-regression of effect on proxy fraction")
```

A **negative meta-regression slope** (effect more protective as proxy fraction rises) with a
**significant proxy-vs-clinical difference** is the collider/selection signature. A flat slope
with overlapping CIs across the gradient supports a **genuine** effect.

```{r forest}
plot_df <- res |> filter(status == "ok", !is.na(b))
if (nrow(plot_df) == 0) {
  cat("No plottable rows (ascertainment swap returned no estimates).\n")
} else {
ggplot(plot_df, aes(x = b, y = reorder(outcome, proxy_frac), colour = factor(proxy_frac))) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red") +
  geom_point(size = 2.5) +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi), height = 0.25) +
  facet_wrap(~ set, ncol = 1, scales = "free_y") +
  theme_classic(base_size = 12) +
  labs(x = "smoking→AD effect (per SD CPD, log-OR)", y = "",
       colour = "proxy frac",
       title = "Task B: ascertainment swap — does protection attenuate in clinical-only AD?")
}
```

## Decision

```{r decision}
for (i in seq_len(nrow(tests))) {
  t <- tests[i,]
  if (is.na(t$p_diff)) next
  patt <- if (!is.na(t$metareg_p) && t$metareg_p < 0.05 && !is.na(t$metareg_slope) && t$metareg_slope < 0)
            "ATTENUATES in clinical-only (supports selection/collider bias)"
          else if (!is.na(t$p_diff) && t$p_diff >= 0.05)
            "persists with comparable CIs (supports genuine effect)"
          else "mixed / inconclusive"
  log_decision("L4c", sprintf("%s proxy→clinical (Δ test + meta-reg)", t$set),
               sprintf("Δb z=%.2f p=%.2g; slope=%.3f p=%.2g",
                       t$z_diff, t$p_diff, t$metareg_slope, t$metareg_p),
               patt)
  cat(sprintf("%-18s: %s\n", t$set, patt))
}
```

::: callout-note
**Caveats.** Clinical IGAP outcomes (Kunkle, Lambert) are smaller (≈22k / 17k cases) so CIs
are wider — an apparent "persistence" must be read against power. The GWAX ×2 rescale is
approximate. Incident / younger-onset AD was not reachable in OpenGWAS (gap). The decisive,
scale-clean comparison is Bellenguez (proxy 0.55) vs Kunkle+Lambert (proxy 0).
:::
