---
title: "Layer 4d — Direction negative controls (Task E): does the outcome itself smell selected?"
author: "Dave Bridges and Katie Kittell"
date: today
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
knitr:
  opts_chunk:
    fig.path: "figures/"
    dev: ["png", "pdf"]
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| message: false
library(TwoSampleMR)
library(ieugwasr)
library(knitr)
source(here::here("R", "helpers.R"))
source(here::here("R", "run_taskE.R"))
cfg <- load_config()
set.seed(cfg$seed)
```

## Purpose

This is an **outcome-level diagnostic, independent of any nAChR claim**. If the primary AD
outcome (Bellenguez, proxy-majority) is survival/participation-selected, then unrelated
exposures that affect *survival and study participation* should show paradoxical effects
through it: **lung cancer** and **CAD** appearing "protective" against AD, and **education**
appearing paradoxically protective (the GWAX/proxy signature reported by Jiang et al — the
non-AD component of by-proxy AD correlating with lower CAD / diabetes / higher survival).

We run **EA → AD**, **lung-cancer → AD**, and **CAD → AD** through the same primary outcome.
A negative (protective) lung-cancer→AD or CAD→AD estimate is a positive selection signature.

## Direction negative controls

```{r negcontrols}
live <- if (opengwas_ok()) tryCatch(run_taskE(cfg), error = function(e) NULL) else NULL
res <- live_or_cache(live, here::here("results", "direction_negcontrols.csv"), "b")
res |>
  transmute(exposure, note, n = nsnp, b = round(b,3), OR = round(or,3),
            `95% CI` = sprintf("[%.3f, %.3f]", ci_lo, ci_hi),
            p = signif(pval,2), signature = selection_signature) |>
  kable(caption = "Task E: exposures → primary AD outcome (selection diagnostic)")
```

## Interpretation

```{r decision}
for (i in seq_len(nrow(res))) {
  r <- res[i,]
  log_decision("L4d", sprintf("%s → AD (primary outcome)", r$exposure),
               sprintf("b=%.3f, p=%.2g (n=%d)", r$b, r$pval, r$nsnp),
               r$selection_signature)
}
sig <- res |> filter(grepl("SELECTION SIGNATURE", selection_signature))
if (nrow(sig) > 0)
  cat("Selection signature DETECTED at the outcome level via:",
      paste(sig$exposure, collapse = ", "), "\n",
      "=> the primary AD outcome shows survival/participation-selection structure.\n")
else
  cat("No lung-cancer/CAD 'protective-against-AD' selection signature detected at the",
      "outcome level (survival-collider structure not evident in this diagnostic).\n")
```

::: callout-note
**Caveats.** A clean genetic-correlation version (LDSC of the by-proxy AD residual vs CAD /
longevity, the Jiang signature) needs genome-wide LD-score regression and full summary stats;
this MR-direction diagnostic is a reachable proxy for it. Lung cancer and CAD are themselves
smoking-related, so a non-null estimate has multiple readings — interpret jointly with Task A
(instrument-level) and Task B (ascertainment swap).
:::
