---
title: "Layer 4e — Sample overlap (Task C) + index-event/selection correction (Task D)"
author: "Dave Bridges and Katie Kittell"
date: today
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
knitr:
  opts_chunk:
    fig.path: "figures/"
    dev: ["png", "pdf"]
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| message: false
library(TwoSampleMR)
library(ieugwasr)
library(SlopeHunter)
library(knitr)
source(here::here("R", "helpers.R"))
source(here::here("R", "run_taskCD.R"))
cfg <- load_config()
set.seed(cfg$seed)
```

## Task C — sample overlap (exposure ∩ outcome UK Biobank)

The CPD exposure (`ieu-b-142`, GSCAN/Liu 2019) and the Bellenguez controls both draw heavily
on **UK Biobank**, so the two-sample MR is partly one-sample — biasing the estimate toward
the confounded observational association.

```{r overlap}
tibble::tribble(
  ~component, ~dataset, ~UKB_content,
  "exposure (CPD)", "ieu-b-142 GSCAN/Liu 2019 (N=249,752)", "includes UK Biobank",
  "outcome controls", "Bellenguez ~401,577 controls", "largely UK Biobank",
  "outcome proxy cases", "Bellenguez ~46,828 by-proxy", "UK Biobank by-proxy") |>
  kable(caption = "Exposure–outcome UK Biobank overlap")
```

**No UKB-excluded CPD GWAS is reachable in OpenGWAS** — both `ieu-b-142` (N=249,752) and the
larger `ieu-b-25` (N=337,334) GSCAN CPD releases include UK Biobank, and a published no-UKB
CPD release is not hosted there. Per the project guardrails this is reported as a **gap**, not
silently substituted.

**The overlap is, however, broken from the *outcome* side in Task B.** The clinical-only
outcomes **Kunkle 2019** and **Lambert 2013** contain **zero UK Biobank** and zero by-proxy
cases, so the exposure∩outcome overlap there is negligible — and the protective effect
**persists** (15q25 −0.136 / −0.123; pooled −0.117). So the strongest available overlap check
is already embedded in Task B and the effect is not abolished by removing the shared UKB.

```{r mrlap-note}
cat("MRlap (overlap-aware correction) is installed but requires genome-wide munged summary",
    "statistics + LD scores for both traits, which are not cached and are not bulk-retrievable",
    "via the OpenGWAS API. It is flagged as the concrete next step (download full GSCAN CPD +",
    "Bellenguez sumstats to data/cache/), not run here.\n")
```

## Task D (LEGACY, SUPERSEDED) — original OpenGWAS-tophits run

::: callout-warning
**This section is superseded and does not execute** (`eval: false`). It is retained only as an
audit trail of the original underpowered run (n=10 tophits), whose slope was unidentified. The
**primary, current** Task D is the genome-wide correction in the next section.
:::

We correct the **outcome** SNP→AD effects for the survival/selection axis. The *incidence*
(inducing) trait is **parental lifespan**; the *prognosis* trait is **AD** (Bellenguez).
SlopeHunter estimates the selection-induced slope `b_SH`. Because IVW is linear, the corrected
smoking→AD slope is exact:
$$ \text{corrected}(\text{smk}\to\text{AD}) = (\text{smk}\to\text{AD}) - b_{SH}\,(\text{smk}\to\text{lifespan}). $$

```{r slopehunter}
#| eval: false
d <- if (opengwas_ok()) tryCatch(run_taskD(cfg),
              error = function(e) list(status = "failed", b_SH = NA_real_,
                                       se_SH = NA_real_, n_snp = 0L))
     else list(status = "offline", b_SH = NA_real_, se_SH = NA_real_, n_snp = 0L)
if (is.null(d$b_SH)) d$b_SH <- NA_real_
cat(sprintf("SlopeHunter slope b_SH = %.3f (SE %.3f), fitted on %s selection-axis SNPs\n",
            d$b_SH, d$se_SH %||% NA_real_, d$n_snp %||% 0L))

pooled_ad <- read_csv(here::here("results","cis_smoking_ldaware_pooled.csv"), show_col_types = FALSE) |>
  filter(method == "LD-aware IVW")
per15 <- read_csv(here::here("results","cis_smoking_ldaware_per_locus.csv"), show_col_types = FALSE) |>
  filter(locus == "CHRNA5_A3_B4", method == "LD-aware IVW")
sb <- read_csv(here::here("results","selection_battery.csv"), show_col_types = FALSE) |>
  filter(outcome == "parental_lifespan")

corr <- bind_rows(
  apply_correction(pooled_ad$b, pooled_ad$se, sb$b[sb$set=="pooled_nAChR"],
                   sb$se[sb$set=="pooled_nAChR"], d$b_SH, d$se_SH) |> mutate(set="pooled_nAChR", .before=1),
  apply_correction(per15$b, per15$se, sb$b[sb$set=="15q25_CHRNA5A3B4"],
                   sb$se[sb$set=="15q25_CHRNA5A3B4"], d$b_SH, d$se_SH) |> mutate(set="15q25_CHRNA5A3B4", .before=1))
corr$b_SH <- d$b_SH; corr$n_SH_snp <- d$n_snp
# Don't overwrite the verified correction with a failed (all-NA) SlopeHunter run.
corr <- live_or_cache(corr, here::here("results","indexevent_correction.csv"), "b_corr")

corr |>
  transmute(set, b_uncorr = round(b_uncorr,3), b_corr = round(b_corr,3),
            `corr 95% CI` = sprintf("[%.3f, %.3f]", ci_lo, ci_hi), p_corr = signif(p_corr,2)) |>
  kable(caption = "Index-event-corrected smoking→AD (uncorrected vs corrected)")
```

```{r decision}
#| eval: false
log_decision("L4e", "Task D index-event-corrected nAChR→AD",
             sprintf("pooled %.3f→%.3f; 15q25 %.3f→%.3f (b_SH=%.2f, n=%d)",
                     corr$b_uncorr[1], corr$b_corr[1], corr$b_uncorr[2], corr$b_corr[2],
                     d$b_SH, d$n_snp),
             "correction attenuates the protective effect toward null — implicates survival collider")
cat(sprintf("Corrected pooled p=%.2g; corrected 15q25 p=%.2g\n", corr$p_corr[1], corr$p_corr[2]))
```

## Task D (PRIMARY) — genome-wide index-event correction

The OpenGWAS-tophits fit above is **superseded**. Using **full summary statistics** (Bellenguez
GCST90027158 × Pilling GCST006697, merged on rsID, allele-aligned, palindromic dropped →
148,162 SNPs; `plink2 --clump` r²<0.01/1Mb against 1000G EUR → **2,346 independent SNPs**), the
SlopeHunter slope becomes tightly identified. See `R/prep_indexevent_genomewide.R` and
`R/run_indexevent_genomewide.R`.

```{r genomewide-indexevent}
fits <- read_csv(here::here("results","indexevent_slopehunter_fits.csv"), show_col_types = FALSE)
corr_gw <- read_csv(here::here("results","indexevent_correction_genomewide.csv"), show_col_types = FALSE)

fits |>
  transmute(`SNP set` = set, n_clumped = n_snp, n_fit,
            b_SH = round(b_SH,3), se = round(se_SH,3),
            `95% CI` = sprintf("[%.3f, %.3f]", ci_lo, ci_hi)) |>
  kable(caption = "SlopeHunter selection slope, genome-wide LD-clumped (2,346 SNPs)")

corr_gw |>
  filter(grepl("PRIMARY", slopehunter_set)) |>
  transmute(`nAChR set` = nachr_set, b_SH = round(b_SH,3),
            uncorrected = round(b_uncorr,3), corrected = round(b_corr,3),
            `95% CI` = sprintf("[%.3f, %.3f]", ci_lo, ci_hi), p = signif(p_corr,2)) |>
  kable(caption = "Index-event-corrected smoking→AD (PRIMARY fit, APOE excluded)")
```

```{r coherence}
# Does the collider structure quantitatively predict the observed effect, at every level?
b_SH <- fits$b_SH[grepl("PRIMARY", fits$set)]
sb <- read_csv(here::here("results","selection_battery.csv"), show_col_types = FALSE) |>
  filter(outcome == "parental_lifespan")
# canonical 20-SNP X->S arm computed from the local Pilling sumstats (see chunk in text)
xs20 <- readRDS(here::here("data","cache","canonical20_smk_lifespan.rds"))

xs  <- c(xs20$b, sb$b[sb$set == "pooled_nAChR"], sb$b[sb$set == "15q25_CHRNA5A3B4"])
obs <- c(-0.113, -0.129, -0.105)
tibble::tibble(
  set = c("canonical_20snp", "pooled_nAChR", "15q25_CHRNA5A3B4"),
  `smk→lifespan (X→S)` = xs,
  `predicted induced effect` = b_SH * xs,
  `observed uncorrected` = obs,
  `% of signal explained` = round(100 * (b_SH * xs) / obs)) |>
  mutate(across(where(is.numeric), ~round(.x, 4))) |>
  kable(caption = "Coherence: does b_SH × (smk→lifespan) reproduce the observed effect?")
```

::: callout-important
**Task D verdict, revised.** With the correction properly identified (b_SH = −0.944, 95% CI
[−1.081, −0.806]), the protective effect is **abolished**: pooled −0.129 → **+0.005 (p=0.91)**,
15q25 −0.105 → **+0.019 (p=0.70)**. Robust to pruning method (distance-pruned replicate
−0.876) and to exclusions — removing APOE *strengthens* the slope, removing 15q25 changes
nothing, so **the earlier "APOE/circularity contamination" diagnosis is retracted**: the n=10
failure was purely a power problem. The collider structure also *quantitatively* predicts the
signal (−0.134 predicted vs −0.129 observed). **Residual limitation:** the correction cannot
separate collider-induced association from a genuine effect collinear with the mortality axis,
so this is strong evidence, not proof.
:::

```{r contamination-check-legacy}
#| eval: false
# LEGACY (superseded by the genome-wide fit above) — retained for the audit trail.
# The SlopeHunter slope is unidentified with reachable data. The fitting set (~10-11 lifespan
# tophits) includes APOE (rs429358) — a GENUINE AD-longevity pleiotropic locus, not a collider
# — and a 15q25 SNP, which is circular (it is the locus under test). Dropping both:
#   b_SH: -0.783 (full) -> -0.758 (clean), but 95% CI widens to [-1.79, +0.27], INCLUDING 0.
# So the corrected estimate ranges from "abolished" (at b_SH=-0.78) to "unchanged" (at b_SH=0).
tibble::tribble(
  ~b_SH_scenario,            ~pooled_corrected, ~`15q25_corrected`,
  "point (-0.78, incl APOE)",  -0.018,            -0.002,
  "clean (-0.76, no APOE/15q25)", -0.022,         -0.006,
  "b_SH = 0 (within CI)",      -0.129,            -0.105) |>
  knitr::kable(caption = "Corrected nAChR→AD across the b_SH confidence range (D is unidentified)")
```

::: callout-note
**Audit trail of the legacy reading (superseded).** The n=10 run was interpreted as
"inconclusive — slope unidentified", diagnosing APOE/circularity contamination. The genome-wide
redo shows that diagnosis was **wrong**: excluding APOE *strengthens* the slope
(−0.863 → −0.944) and excluding 15q25 changes it by <0.001. The n=10 failure was purely a
**power** problem, and the original point estimate (−0.78) was close to the identified value
(−0.944). Both the original reading and its retraction are recorded here deliberately; the
current conclusion is in the PRIMARY section above and in `SUMMARY.md`.
:::

## Where this leaves the project

Task D is now the decisive result in the selection battery, and it points to a collider. Task B
(persistence in clinical-only AD) remains valid but is **narrower than it first read** — it
excludes the *proxy-ascertainment* channel only, and says nothing about survival-to-diagnosis
selection, which operates in clinical cohorts too. Consequently:

- The protective smoking/nAChR→AD effect **should not be treated as a causal, druggable signal.**
- The **brain eQTL/pQTL expansion (Layer 3) is gated**, not cancelled — see APPROACH.md §14.
- The single decisive next test is an **incident / younger-onset AD** outcome, which does not
  condition on survival to diagnosis age.
