---
title: "Meta Analysis of Correlations between Calcium and Cholesterol"
author: "Dave Bridges and Kaelin Loftus"
format:
  html:
    toc: true
    toc-location: left
editor: visual
execute:
  keep-md: true
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
```

## Data Souces

Located studies from PubMed searches and checking internal references. Manually re-calculated cholesterol to mM when presented in mg/dL

```{r data-input}
data.sheet <- 'https://docs.google.com/spreadsheets/d/e/2PACX-1vTyvQnc6bLRLGT6QXEMHxiAQVbK_zag_JIAjvYjTMXINcqdkBwglmg_mlj_k9ml9QsrNQl-tZgy8ACl/pub?gid=1100702568&single=true&output=csv'
library(readr)
data <- read_csv(data.sheet)#from a google sheet
```

The data can be found in the google sheet `r data.sheet`. This script can be found in `r getwd()` and was most recently run on `r date()`

## Meta Analysis

Analysed data from mean +/- SD of cases and controls

```{r ca-chol-meta}
library(meta)
library(tidyr)
analysis <- metacor(data=data %>% dplyr::filter(!is.na(`r`)),
                   cor = r, 
                   n = n,
                 studlab = Study,
                 fixed = FALSE,
                 random = TRUE,
                 method.tau = "REML",
                 hakn = TRUE,
                 title="Calcium and Cholesterol")

forest.meta(analysis,
            fontsize=6,,
            test.overall.random=TRUE)
```

# Analysis

There is solid evidence of cross-sectional associations between cholesterol and calcium levels in multiple studies.  The aggregate effect size is `r analysis$TE.random` +/- `r analysis$seTE.random` (p=`r analysis$pval.random`).

# Session Information

```{r session-information, echo=T}
sessionInfo()
```
