---
title: "Meta Analysis of Correlations between Calcium and Cholesterol"
author: "Dave Bridges and Kaelin Loftus"
format:
  html:
    toc: true
    toc-location: left
editor: source
execute:
  keep-md: true
bibliography: bayesian.bib
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
```

## Data Souces

Located studies from PubMed searches and checking internal references. Manually re-calculated cholesterol to mM when presented in mg/dL

```{r data-input}
data.sheet <- 'https://docs.google.com/spreadsheets/d/e/2PACX-1vTyvQnc6bLRLGT6QXEMHxiAQVbK_zag_JIAjvYjTMXINcqdkBwglmg_mlj_k9ml9QsrNQl-tZgy8ACl/pub?gid=1100702568&single=true&output=csv'
library(readr)
data <- read_csv(data.sheet)#from a google sheet
```

The data can be found in a [google sheet](`r data.sheet`). This script can be found in `r getwd()` and was most recently run on `r date()`

## Meta Analysis - Standard Approach

Analysed data from mean +/- SD of cases and controls

```{r ca-chol-meta}
library(meta)
library(tidyr)
analysis <- metacor(data=data %>% dplyr::filter(!is.na(`r`)),
                   cor = r, 
                   n = n,
                 studlab = Study,
                 fixed = FALSE,
                 random = TRUE,
                 method.tau = "REML",
                 hakn = TRUE,
                 title="Calcium and Cholesterol")

forest(analysis,
            fontsize=6,,
            test.overall.random=TRUE)
```

## Bayesian Hierarchical Model

Followed the procedure outlined in @bayesianmetabookdown and @harrer2021doing at [here]( https://bookdown.org/MathiasHarrer/Doing_Meta_Analysis_in_R/bayesian-ma.html).  This uses a framework described by @higgins2008.  Under this approach there is a "true" effect ($\mu$) with a cross-study variance of $\tau$.  Within this each study ($\theta _k$) should be normally distributed defined:

$$\theta _k \sim N(\mu,\tau^2)$$
Each study in turn is a proximate measure of the effect size in that population $\hat{\theta_k}$, with sampling error $\sigma^2$ drawn from another normal distribution

$$\hat\theta _k \sim N(\theta _k,\sigma^2_k)$$

This can be simplified to

$$\hat\theta _k \sim N(\mu,\sigma^2_k + \tau^2)$$

### Defining a Prior Distribution

$$
p \sim (\mu,tau^2)
$$
We took an approach using *weakly informative* priors (recommended in @williamsBayesianMetaAnalysisWeakly2018) of $\mu = N(0,0.2)$ and $\tau=HC(0,0.5)$.  This means we predict the correlations to be around zero but with a standard deviation of $\pm 0.2$.  For the betwen study variance ($\tau$) we used the heavy-tailed only positive Half-Cauchy distribution.

We calculated the standard error of the estimate of r using

$$SE = \sqrt{\frac{(1 - r^2)}{(n - 2)}}$$

```{r se-of-r}
library(dplyr)
data <-
  data %>%
  mutate(se=sqrt((1-r^2)/(n-2)))
```


This was analysed using {brms} (see @) using no U-turn sampling (NUTS) as described in @hoffman14a.

```{r calcium-brms}
#| output: false
library(brms)
priors <- c(prior(normal(0,0.2), class = Intercept),
            prior(cauchy(0,0.5), class = sd))

meta.brm <- brm(r|se(se) ~ 1 + (1|Study),
             data = data,
             prior = priors,
             iter = 4000)
```

### Internpretation of Bayesian Analysis

```{r bayesian-interpretation}
#summary(meta.brm)
#ranef(meta.brm)
fixef(meta.brm) %>% kable(caption="Bayesian estimates of correlation between cholesterol and calcium")
```

#### Posterior Probability Check

```{r post-probability}
pp_check(meta.brm)
```

#### Posterior Probabilities

```{r calcium-posterior-probabilities}
post.samples <- as_draws_df(meta.brm, c("Intercept", "sd_Study__Intercept"))
names(post.samples) <- c("smd", "tau")
library(ggplot2)

ggplot(aes(x = smd), data = post.samples) +
  geom_density(fill = "lightblue",                # set the color
               color = "lightblue", alpha = 0.7) +  
  geom_vline(xintercept = mean(post.samples$smd),lty=2) +
  geom_vline(xintercept = 0,lty=2) +
  labs(x = expression(italic(r)),
       y = element_blank()) +
  theme_classic(base_size=16)

ggplot(aes(x = tau), data = post.samples) +
  geom_density(fill = "lightgreen",               # set the color
               color = "lightgreen", alpha = 0.7) +  
  geom_vline(xintercept= mean(post.samples$tau),lty=2) +        # add point at mean
    labs(x = expression(tau),
       y = element_blank()) +
  theme_classic(base_size=16)
```
#### Forest Plot for Bayesian Meta-Analysis

```{r calcium-bayesian-forest}
library(tidybayes)
library(ggridges)
library(glue)
library(stringr)
library(forcats)

# posterior probabilitieis for each study
study.draws <- spread_draws(meta.brm, r_Study[Study,], b_Intercept) %>% 
  mutate(b_Intercept = r_Study + b_Intercept)

# pooled effect size draws
pooled.effect.draws <- spread_draws(meta.brm, b_Intercept) %>% 
  mutate(Study = "Pooled Effect")

# combined draws
forest.data <- bind_rows(study.draws, 
                         pooled.effect.draws) %>% 
   ungroup() %>%
   mutate(Study = str_replace_all(Study, "[.]", " ")) %>% 
   mutate(Study = reorder(Study, b_Intercept)) #arrange by effect size

# calculate effect size of each study
forest.data.summary <- group_by(forest.data, Study) %>% 
  mean_qi(b_Intercept)

# generate plot
ggplot(aes(b_Intercept, 
           relevel(Study, "Pooled Effect", 
                   after = Inf)), 
       data = forest.data) +
  
  # Add vertical lines for pooled effect and CI
  geom_vline(xintercept = fixef(meta.brm)[1, 1], 
             color = "grey", size = 1) +
  geom_vline(xintercept = fixef(meta.brm)[1, 3:4], 
             color = "grey", linetype = 2) +
  geom_vline(xintercept = 0, color = "black", 
             size = 1) +
  
  # Add densities
  geom_density_ridges(fill = "grey", 
                      rel_min_height = 0.01, 
                      col = NA, scale = 1,
                      alpha = 0.8) +
  geom_pointinterval(data = forest.data.summary, 
                      size = 1,
                     orientation='horizontal',
                     aes(xmin = .lower, xmax = .upper)) +
  
  # Add text and labels
  geom_text(data = mutate_if(forest.data.summary, 
                             is.numeric, round, 2),
    aes(label = glue("{b_Intercept} [{.lower}, {.upper}]"), 
        x = Inf), hjust = "inward") +
  labs(x = "Correlation Coefficient", # summary measure
       y = element_blank()) +
  theme_classic(base_size=16)
```

## Summary

There is solid evidence of cross-sectional associations between cholesterol and calcium levels in multiple studies. The aggregate effect size is `r analysis$TE.random` +/- `r analysis$seTE.random` (p=`r analysis$pval.random`).

## References

::: {#refs}
:::

# Session Information

```{r session-information, echo=T}
sessionInfo()
```
