---
title: "Systematic Review of LDL Changes on a Ketogenic Diet"
author: "Cody Cousineau and Dave Bridges"
date: "June 18, 2021"
output:
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: yes
    toc: yes
  pdf_document:
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```
# Purpose

To evaluate how LDL changes in ketogenic diet studies

# Experimental Details

Evaluated studies where ketogenic diets (<25g/day of CHO) are used and weight and LDL-C are reported as an outcome

# Raw Data

Reviewed data from the Choi *et al* meta-analysis (http://dx.doi.org/10.3390/nu12072005), pulling in data on baseline weight, weight changes, LDL, LDL changes and standard deviations. A systematic literature search of PubMed was then performed to identify other randomized controlled trials (RCTs) and single-arm interventions of patients that evaluated the effects of a ketogenic diet on weight and lipid profile as primary endpoints. All studies using a KD diet that met our inclusion criteria where intake of carbohydrate was less than 25 grams per day were included. This search was most recently updated on `r date()`.

We used a value 130mg/dL of LDL-C at baseline to stratify individuals as being hypercholesterolemic or not.


Correlations in this meta-analysis before and after the administration of the ketogenic diet were analyzed with linear models and results given using pearsons correlation coefficient, statistical significance was defined as below 0.05. 


For all outcomes, we tested sex as a modifier and as a covariate. For outcomes where sex was found to be a significant modifier, these results are reported. 

```{r data-input}
library(readxl) #loads the readr package
filename <- 'LDL Study Summary.xlsx' #make this a separate line, you can use any variable you want

#this loads whatever the file is into a dataframe called exp.data if it exists
exp.data <- read_excel(filename)
eval.data <- exp.data %>% 
  filter(Use=='x') %>%
  mutate(Pct.Wt.Change = `Weight Change`/`Baseline Weight`*100)%>%
  mutate(PCT.BMI.Change = `BMI Change` / `Baseline BMI`*100)%>%
  mutate(Sex.Group = cut(`Percent Male`, breaks = c(0,.1,.9,1), include.lowest = TRUE, labels = c("Mostly Female", "Mixed", "Mostly Male")))
```

These data can be found in **`r getwd()`** in a file named **`r ifelse(filename %in% dir(), filename, "no file found")`**.  This script was most recently updated on **`r date()`**.

# Analysis

# Variance in LDL-C

```{r ldlc-endpoint}
library(ggplot2)

eval.data <-
  eval.data %>%
  mutate(LDL.endpoint = `Baseline LDL`+`Change in LDL-C`)

eval.data %>%
  filter(!(is.na(`Endpoint SD`))) %>%
  ggplot(aes(y=LDL.endpoint,
             ymin=LDL.endpoint-`Endpoint SD`,
             ymax=LDL.endpoint+`Endpoint SD`,
             x=reorder(Study,-`LDL.endpoint` ))) +
  geom_point() +
  geom_errorbar()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))

eval.data %>% 
  summarize_at(.vars=vars(Pct.Wt.Change,`Weight Change`,`Change in LDL-C`,`BMI Change`),
               .funs=funs(mean(.,na.rm=TRUE))) 

#Above is non-weighted average
```

# Meta-Analysis 

```{r meta-analysis}
meta.data <-
  eval.data %>%
  filter(!is.na(`Endpoint SD`)) %>%
  mutate(Pooled.SD=sqrt(`Baseline SD`^2+`Endpoint SD`^2)) %>%
  mutate(SMD=`Change in LDL-C`) %>%
  mutate(SMD.Wt = `Weight Change`) 

library(meta)
ldl.c.meta <- metagen(TE = SMD,
                 seTE = Pooled.SD,
                 n.e=n,
                 n.c=n,
                 studlab = Study,
                 data = meta.data,
                 sm = "SMD",
                 comb.fixed = TRUE,
                 comb.random = FALSE,
                 #method.tau = "REML",
                 hakn = TRUE,
                 title = "LDL-C Changes in Ketogenic Diet Studies")

#forest.meta(ldl.c.meta, 
            #sortvar = SMD,
            #predict = F, 
            #print.tau2 = TRUE,
           # print.I2 = TRUE,
            #leftcols=c('Study'),
            #rightlabs=c('Change','95% CI','Weight'),
           # layout = "RevMan5",
          #  fontsize=8,
          #  leftcols = c("Study",'effect',"seTE",'ci','w.fixed'),
          #  leftlabs = c("Study","Change","SE","95% CI","Weight"),
          #  range = F,
          #  digits=1,
          #  digits.se=1)
```

We evaluated `r ldl.c.meta$k` studies for this meta-analysis. Using the meta-analysis method, we found fasting blood LDL-C levels were increased `r ldl.c.meta$TE.fixed` mg/dL (95% CI: `r ldl.c.meta$lower.fixed` to `r ldl.c.meta$upper.fixed`) after the ketogenic diet intervention compared to pre-intervention levels, with a significant p-value of `r ldl.c.meta$pval.fixed`. Across these studies, the I<sup>2</sup> is `r ldl.c.meta$I2`, the p-value for Q is `r ldl.c.meta$pval.Q`. This is a highly consistent I^2. 

# Average Change in LDL-C

```{r ldl-change}
eval.data %>%
  ggplot(aes(y=`Change in LDL-C`,
             x=reorder(Study, -`Change in LDL-C`))) +
  geom_bar(stat='identity') +
  labs(x="Study")

eval.data %>%
  ggplot(aes(y=`Change in LDL-C`,
             x=reorder(Study, -`Change in LDL-C`))) +
  geom_bar(stat='identity', aes(fill=`Percent Male`)) +
  labs(x="Study")+
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))

eval.data %>%
  ggplot(aes(y=`Change in LDL-C`,
             x=reorder(Study, -`Change in LDL-C`))) +
  geom_bar(stat='identity', aes(fill=`Sex.Group`)) +
  labs(x="Study") +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))
```

## Relative to Weight

```{r ldl-change-vs-weight}
library(ggrepel)
library(ggplot2)
eval.data %>%
  ggplot(aes(y=`Change in LDL-C`,
             x=`Baseline Weight`)) +
  geom_point() +
  geom_smooth(span=1) +
  geom_text_repel(aes(label=Study)) +
  labs(x="Baseline Weight")

eval.data %>%
  ggplot(aes(y=`Change in LDL-C`,
             x=`Baseline BMI`)) +
  geom_point() +
  geom_smooth(span=1) +
  geom_text_repel(aes(label=Study)) +
  labs(x="Baseline BMI")

eval.data %>%
  ggplot(aes(y=`Change in LDL-C`,
             x=`Baseline Weight`)) +
  geom_point() +
  geom_smooth(method='lm') +
  geom_text_repel(aes(label=Study)) +
  labs(x="Baseline Weight")

eval.data %>%
  ggplot(aes(y=`Change in LDL-C`,
             x=`Baseline BMI`)) +
  geom_point() +
  geom_smooth(method='lm') +
  geom_text_repel(aes(label=Study)) +
  labs(x="Baseline BMI")

eval.data %>%
  ggplot(aes(y=`Change in LDL-C`,
             x=`Baseline Weight`,
             col=`Percent Male`)) +
  geom_point() +
  geom_smooth(method='lm') +
  geom_text_repel(aes(label=Study)) +
  labs(x="Baseline Weight")

eval.data %>%
  ggplot(aes(y=`Change in LDL-C`,
             x=`Baseline Weight`,
             col=`Sex.Group`)) +
  geom_point() +
  geom_smooth(method='lm') +
  geom_text_repel(aes(label=Study)) +
  labs(x="Baseline Weight")

library(broom)
bind_rows(`Baseline`=shapiro.test(eval.data$`Baseline Weight`)$p.value,
          `Change`=shapiro.test(eval.data$`Change in LDL-C`)$p.value) %>%
  kable(caption="Shapiro Tests for Correlates")
with(eval.data, cor.test(`Change in LDL-C`,`Baseline Weight`, method="spearman")) %>% tidy %>% kable(caption="Correlation between change in LDL-C and baseline weight")


lm(`Change in LDL-C`~`Baseline Weight`+`Percent Male`, data=eval.data) %>% tidy %>% kable(caption="Linear model between change in LDL-C and baseline weight, including gender")

ldl.weight.baseline.lm <- lm(`Change in LDL-C` ~ `Baseline Weight`, data = eval.data)
ldl.weight.baseline.lm %>% tidy %>% kable

ldl.weightsex.baseline.lm <- lm(`Change in LDL-C` ~ `Baseline Weight` + `Sex.Group`, data = eval.data)
ldl.weightsex.baseline.lm %>% tidy %>% kable

ldl.weightsex.baseline.aov <- aov(`Change in LDL-C` ~ `Baseline Weight` + `Sex.Group`, data = eval.data)
ldl.weightsex.baseline.aov %>% tidy %>% kable

ldl.bmi.baseline.lm <- lm(`Change in LDL-C` ~ `Baseline BMI`, data = eval.data)
ldl.bmi.baseline.lm %>% tidy %>% kable

ldl.bmi.baseline.sex.aov <- aov(`Change in LDL-C` ~ `Baseline BMI` + `Sex.Group`, data = eval.data)
ldl.bmi.baseline.sex.aov %>% tidy %>% kable

ldl.weight.baseline.aov <- aov(`Change in LDL-C` ~ `Baseline Weight` + `Sex.Group`, data = eval.data)
ldl.weight.baseline.aov %>% tidy %>% kable

ldl.bmi.baseline.aov <- aov(`Change in LDL-C` ~ `Baseline BMI` + `Sex.Group`, data = eval.data)
ldl.bmi.baseline.aov %>% tidy %>% kable

ldl.bmi2.baseline.aov <- aov(`Change in LDL-C` ~ `Baseline BMI`*`Sex.Group`, data = eval.data)
ldl.bmi2.baseline.aov %>% tidy %>% kable

ldl.weight.baseline.aov <- aov(`Change in LDL-C` ~ `Baseline Weight`, data = eval.data)
ldl.weight.baseline.aov %>% tidy %>% kable
```

Lower baseline BMI was associated with an increased change in LDL-C after consumption of a ketogenic diet (r<sup>2</sup> = `r ldl.bmi.baseline.lm %>% glance %>% pull(r.squared)`, p-value = `r ldl.bmi.baseline.lm %>% tidy %>% pull(p.value) %>% last`). The association with increased LDL-C was consistent with baseline weight, where a lower baseline weight was associated with an increased change in LDL-C (r<sup>2</sup> = `r ldl.weight.baseline.lm %>% glance %>% pull(r.squared)`, p-value = `r ldl.weight.baseline.lm %>% tidy %>% pull(p.value) %>% last`). 

## Relative to Weight Loss

```{r ldl-change-vs-weight-change}
library(ggplot2)

eval.data %>%
  ggplot(aes(y=`Change in LDL-C`,
             x=Pct.Wt.Change)) +
  geom_point() +
  geom_smooth(method='lm') +
  geom_text_repel(aes(label=Study)) +
  labs(x="Weight Change (%)")

eval.data %>%
  ggplot(aes(y=`Change in LDL-C`,
             x=`BMI Change`)) +
  geom_point() +
  geom_smooth(method='lm') +
  geom_text_repel(aes(label=Study)) +
  labs(x="BMI Change")

eval.data %>%
  ggplot(aes(y=`Change in LDL-C`,
             x=`PCT.BMI.Change`)) +
  geom_point() +
  geom_smooth(method='lm') +
  geom_text_repel(aes(label=Study)) +
  labs(x="% BMI Change")

eval.data %>%
  ggplot(aes(y=`Change in LDL-C`,
             x=Pct.Wt.Change,
             col=`Percent Male`)) +
  geom_point() +
  geom_smooth(method='lm') +
  geom_text_repel(aes(label=Study)) +
  labs(x="Weight Change (%)")

library(broom)
bind_rows(`Weight Change`=shapiro.test(eval.data$Pct.Wt.Change)$p.value,
          `LDL Change`=shapiro.test(eval.data$`Change in LDL-C`)$p.value) %>%
  kable(caption="Shapiro Tests for Correlates")
with(eval.data, cor.test(`Change in LDL-C`,Pct.Wt.Change, method="spearman")) %>% tidy %>% kable(caption="Correlation between change in LDL-C and weight change")

ldl.weight.change.lm <- lm(`Change in LDL-C` ~ `Pct.Wt.Change`, data = eval.data)
ldl.weight.change.lm %>% tidy %>% kable

ldl.weightsex.change.lm <- lm(`Change in LDL-C` ~ `Pct.Wt.Change` + `Sex.Group`, data = eval.data)
ldl.weightsex.change.lm %>% tidy %>% kable

ldl.weightsex.change.aov <- aov(`Change in LDL-C` ~ `Pct.Wt.Change` + `Sex.Group`, data = eval.data)
ldl.weightsex.change.aov %>% tidy %>% kable

ldl.bmi.change.lm <- lm(`Change in LDL-C` ~ `BMI Change`, data = eval.data)
ldl.bmi.change.lm %>% tidy %>% kable

ldl.bmisex.change.lm <- lm(`Change in LDL-C` ~ `BMI Change` + `Sex.Group`, data = eval.data)
ldl.bmisex.change.lm %>% tidy %>% kable

ldl.bmisex.change.aov <- aov(`Change in LDL-C` ~ `BMI Change` + `Sex.Group`, data = eval.data)
ldl.bmisex.change.aov %>% tidy %>% kable

ldl.pctbmi.change.lm <- lm(`Change in LDL-C` ~ `PCT.BMI.Change`, data = eval.data)
ldl.pctbmi.change.lm %>% tidy %>% kable

ldl.pctbmisex.change.lm <- lm(`Change in LDL-C` ~ `PCT.BMI.Change` + `Sex.Group`, data = eval.data)
ldl.pctbmisex.change.lm %>% tidy %>% kable

ldl.weight.change.aov <- aov(`Change in LDL-C` ~ `Pct.Wt.Change` + `Sex.Group`, data = eval.data)
ldl.weight.change.aov %>% tidy %>% kable

ldl.bmi.change.aov <- aov(`Change in LDL-C` ~ `BMI Change`, data = eval.data)
ldl.bmi.change.aov %>% tidy %>% kable

ldl.bmisex.change.aov <- aov(`Change in LDL-C` ~ `BMI Change` + `Sex.Group`, data = eval.data)
ldl.bmisex.change.aov %>% tidy %>% kable

ldl.pctbmi.change.aov <- aov(`Change in LDL-C` ~ `PCT.BMI.Change` + `Sex.Group`, data = eval.data)
ldl.pctbmi.change.aov %>% tidy %>% kable

ldl.weight.change.aov <- aov(`Change in LDL-C` ~ `Pct.Wt.Change`, data = eval.data)
ldl.weight.change.aov %>% tidy %>% kable

```
Greater BMI decreases over the study period were associated with a smaller increase in LDL-C after consumption of a ketogenic diet, though this did not reach significance (r<sup>2</sup> = `r ldl.bmi.change.lm %>% glance %>% pull(r.squared)`, p-value = `r ldl.bmi.change.lm %>% tidy %>% pull(p.value) %>% last`). The association with the change in LDL-C and decrease in BMI was consistent with weight, with change in weight on LDL-C reaching significance (r<sup>2</sup> = `r ldl.weight.change.lm %>% glance %>% pull(r.squared)`, p-value = `r ldl.weight.change.lm %>% tidy %>% pull(p.value) %>% last`), where greater decreases in weight were associated with lower increases in LDL-C after consumption of a ketogenic diet. Looking at percent BMI change to account for baseline BMI, greater percent change decreases were associated with a lower increase in LDL-C on a ketogenic diet, though this was not significant (r<sup>2</sup> = `r ldl.pctbmi.change.lm %>% glance %>% pull(r.squared)`, p-value = `r ldl.pctbmi.change.lm %>% tidy %>% pull(p.value) %>% last`).


## Relative to Baseline LDL-C

```{r ldl-change-vs-baseline}
library(ggplot2)
eval.data %>%
  ggplot(aes(y=`Change in LDL-C`,
             x=`Baseline LDL`)) +
  geom_point() +
  geom_smooth(method='lm') +
  geom_text_repel(aes(label=Study)) +
  labs(x="Baseline LDL-C")

eval.data %>%
  ggplot(aes(y=`Change in LDL-C`,
             x=`Baseline LDL`,
             col=`Percent Male`)) +
  geom_point() +
  geom_smooth(method='lm') +
  geom_text_repel(aes(label=Study)) +
  labs(x="Baseline LDL-C")

eval.data %>%
  ggplot(aes(y=`Change in LDL-C`,
             x=`Baseline LDL`,
             col=`Sex.Group`)) +
  geom_point() +
  geom_smooth(method='lm') +
  geom_text_repel(aes(label=Study)) +
  labs(x="Baseline LDL-C")

library(broom)
bind_rows(`Baseline`=shapiro.test(eval.data$`Baseline LDL`)$p.value,
          `Change`=shapiro.test(eval.data$`Change in LDL-C`)$p.value) %>%
  kable(caption="Shapiro Tests for Correlates")
with(eval.data, cor.test(`Change in LDL-C`,`Baseline LDL`, method="spearman")) %>% tidy %>% kable(caption="Correlation between baseline and change in LDL-C")

ldl.baseline.lm <- lm(`Change in LDL-C` ~ `Baseline LDL`, data = eval.data)
ldl.baseline.lm %>% tidy %>% kable

ldl.baselinesex.lm <- lm(`Change in LDL-C` ~ `Baseline LDL` + `Sex.Group`, data = eval.data)
ldl.baselinesex.lm %>% tidy %>% kable

ldl.baseline.aov <- aov(`Change in LDL-C` ~ `Baseline LDL` + `Sex.Group`, data = eval.data)
ldl.baseline.aov %>% tidy %>% kable

ldl.baseline.aov <- aov(`Change in LDL-C` ~ `Sex.Group`, data = eval.data)
ldl.baseline.aov %>% tidy %>% kable

ldl.baseline.aov <- aov(`Change in LDL-C` ~ `Percent Male`, data = eval.data)
ldl.baseline.aov %>% tidy %>% kable
```

Among individuals, baseline LDL-C was not positively correlated with change in LDL-C after consumption of a ketogenic diet and the relationship was not significant (r<sup>2</sup> = `r ldl.baseline.lm %>% glance %>% pull(r.squared)`, p-value = `r ldl.baseline.lm %>% tidy %>% pull(p.value) %>% last`). 

# Session Information

```{r session-information, echo=T}
sessionInfo()
```