---
title: "Meta Analysis of Case Control Studies of pPTH on Cholesterol"
author: "Dave Bridges and Kaelin Loftus"
format:
  html:
    toc: true
    toc-location: left
editor: visual
execute:
  keep-md: true
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
```

## Data Souces

Located studies from PubMed searches and checking internal references.  Manually re-calculated cholesterol to mM when presented in mg/dL


```{r data-input}
data.sheet <- 'https://docs.google.com/spreadsheets/d/e/2PACX-1vTyvQnc6bLRLGT6QXEMHxiAQVbK_zag_JIAjvYjTMXINcqdkBwglmg_mlj_k9ml9QsrNQl-tZgy8ACl/pub?gid=381583063&single=true&output=csv'
library(readr)
data <- read_csv(data.sheet)#from a google sheet
```

The data can be found in the google sheet `r data.sheet`.  This script can be found in `r getwd()` and was most recently run on `r date()`

## Meta Analysis

Analysed data from mean +/- SD of cases and controls

```{r pth-meta-analysis}
library(meta)
analysis <- metacont(data=data,
                    n.e=PTH_n,
                    n.c=Controls_n,
                    mean.e=PTH_mean,
                    mean.c=Controls_mean,
                    sd.e=PTH_SD,
                    sd.c=Controls_mean,
                    studlab = Study,
                    method.smd="Hedges",
                    sm="SMD",
                    fixed=FALSE,
                    random=TRUE,
                    method.tau="REML",
                    hakn=TRUE,
                    title="pPTH on Cholesterol")

forest.meta(analysis,
            fontsize=6)
```

# Analysis

There is no evidence that primary hyperparathyroidism results in elevated total cholesterol from case-control studies.

# Session Information

```{r session-information, echo=T}
sessionInfo()
```
