---
title: "Associations with Diversity Outbred Calcium and Cholesterol Levels"
author: "Dave Bridges"
date: "November 27, 2022"
output:
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: yes
    toc: yes
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: yes
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```

# Purpose

To analyze associations between other phenotypes and cholesterol and calcium to identify potential measured confounders.

# Experimental Details

This analysis uses the complete dataset (F01-F425 and M01-M425). 

# Raw Data

```{r datafiles}
phenotype.filename <- 'Svenson-183_Svenson_DO-phenotypes.csv'
```


```{r data-input}
library(readr) #loads the readr package


phenotype.data <- read_csv(phenotype.filename)
#set phenotypes of zero or  to na
phenotype.data[phenotype.data < 0] <- NA

library(forcats)
cholesterol.data <-
  phenotype.data %>%
  mutate(Diet = fct_recode(as.factor(diet),
                           "NCD"="chow",
                           "HFHS"="hf")) %>%
  select(-diet) %>%
  mutate(chol.avg = rowMeans(select(., starts_with("chol")), 
                             na.rm = TRUE))
```

# Correlation Matrix

```{r calcium-associations}
library("psych")    
correlation.matrix <- cholesterol.data %>%
  select(-coat_color,-gen,-sex,-Diet,-sample) %>%
  corr.test(method="spearman", use="pairwise",adjust="BH")

correlation.matrix$p %>%
  as.data.frame() %>%
  select(calcium2) %>%
  filter(calcium2<0.05/162) -> calcium.correlates

correlation.matrix$p %>%
  as.data.frame() %>%
  select(chol.avg) %>%
  filter(chol.avg<0.05/162) -> cholesterol.correlates

overlapping.correlates <- intersect(rownames(calcium.correlates),
                                    rownames(cholesterol.correlates))
```

```{r confounders-venn}
library(venneuler)

venneuler(c("Calcium"=dim(calcium.correlates)[1]+1,
            "Cholesterol"=dim(cholesterol.correlates)[1]+1,
            "Calcium&Cholesterol"=length(overlapping.correlates)+1)) %>%
  plot
```

# Testing Moderation by Correlates


```{r chol-calcium-effect-changes}
base.model <- lm(chol2 ~ Diet + sex + calcium2, data=cholesterol.data)

confounders.data <-
  data.frame("Measure"=overlapping.correlates) %>%
  mutate(Calcium.mod = NA)

for (measure in overlapping.correlates) {
  formula <- as.formula(paste("chol2 ~ ", measure, " + Diet + sex + calcium2"))
  test.model = lm(formula,data=cholesterol.data)
  confounders.data <- 
    confounders.data %>%
    add_row(Measure=measure,
            Calcium.mod=coefficients(test.model)['calcium2'])
}

confounders.data <-
  confounders.data %>% 
  mutate(Calcium.base = coefficients(base.model)['calcium2']) %>%
  filter(!is.na(Calcium.mod)) 

library(ggplot2)
confounders.data %>%
  mutate(Estimate.change = Calcium.mod-Calcium.base) %>%
  ggplot(aes(y=Estimate.change,x=reorder(Measure,Calcium.mod))) +
  geom_bar(stat="identity") +
  labs(y="Change in Calcium Beta",
       x="Phenotype") +
  theme_classic() +
  theme(text=element_text(size=16),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

confounders.data %>%
  arrange(Calcium.mod) %>%
  mutate(Estimate.change = Calcium.mod-Calcium.base) %>%
  kable(caption="Base and Moderated Model Estimates")
```

# Potential Mediators

These are defined as phenotypes that associate with both calcium and cholesterol, and *weaken* the sex and diet adjusted association of cholesterol and calcium when included in the model.

## Fat Mass

```{r mediator-fat mass}
library(broom)
complete.data <- 
  cholesterol.data %>%
  filter(!(is.na(calcium2))) %>%
  filter(!(is.na(chol2))) %>%
  filter(!(is.na(fat_mri)))
         


base.lm <- lm(chol2 ~ Diet + sex + calcium2 + fat_mri, data=complete.data)        
fat.lm <- lm(chol2 ~ Diet + sex + calcium2 + fat_mri, data=complete.data)
fat.lm %>% tidy %>% kable(caption="Model including fat mass")

library(mediation)
results <- mediate(fat.lm, base.lm, treat='calcium2', mediator='fat_mri',
                   boot=TRUE, sims=1000)
results %>% summary
```


# Potential Confounders

These are defined as phenotypes that associate with both calcium and cholesterol, and *strengthen* the sex and diet adjusted association of cholesterol and calcium when included in the model.

## Phosphorous

```{r confounder-phosphorous}
phos.lm <- lm(chol2 ~ Diet + sex + calcium2 + phosphorus2, data=cholesterol.data)
library(broom)
phos.lm %>% tidy %>% kable(caption="Model including phosphorus")
#anova(base.model,phos.lm) %>% tidy %>% kable(caption="Chi squared test of models with or without calcium")
```

# Session Information

```{r session-information, echo=T}
sessionInfo()
```

