---
title: "Predictors of Diversity Outbred Cholesterol Levels"
author: "Dave Bridges and Cody Cousineau"
date: "March 25, 2022"
output:
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: yes
    toc: yes
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: yes
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```

# Purpose

To analyze cholesterol levels from diversity outbred mice, identifying predictors of elevated choelsterol.

# Experimental Details

This analysis uses the complete dataset (F01-F425 and M01-M425). 

# Raw Data

```{r datafiles}
phenotype.filename <- 'Svenson_HFD_DO_phenotype_V12.csv'
```


```{r data-input}
library(readr) #loads the readr package


phenotype.data <- read_csv(phenotype.filename)
#set phenotypes of zero or  to na
phenotype.data[phenotype.data < 0] <- NA

library(forcats)
cholesterol.data <-
  phenotype.data %>%
  mutate(Diet = fct_recode(as.factor(diet),
                           "NCD"="chow",
                           "HFHS"="hf")) %>%
  mutate(chol.avg = rowMeans(select(., starts_with("chol")), 
                             na.rm = TRUE))
```

# Analysis

The total number of mice with cholesterol levels in this dataset is `r cholesterol.data %>% distinct(mouse.id) %>% count`
 mice, broken down as follows:
 
```{r sample-size}
cholesterol.data %>%
  group_by(Diet,sex) %>%
  count() %>%
  kable(caption="Number of mice in each group")
```
 
## Cholesterol Levels for NCD Annimals

```{r cholesterol-normality}
cholesterol.data %>%
  group_by(sex,diet) %>%
  summarize_at(.vars=vars(chol1,chol2,chol.avg), .funs=list(~shapiro.test(.)$p.value)) %>%
  kable(caption="Groupwise Shapiro-Wilk normality tests for cholesterol levels", digits=c(0,0,5,5,5))

library(ggplot2)
cholesterol.data %>%
  select(sex,Diet,starts_with('chol')) %>%
  group_by(sex,Diet) %>%
  pivot_longer(cols=starts_with('chol'),
               names_to = "Measure",
               values_to = "Cholesterol") %>%
  ggplot(aes(x=Cholesterol,fill=Measure)) +
  geom_density(alpha=0.5) +
  facet_grid(Diet~sex)
```

### Similarity at Both Time Points

```{r cholesterol-similarity}
  
summary.data.complete <-
  cholesterol.data %>%
  group_by(sex,Diet) %>%
  summarize_at(.vars=vars(chol1,chol2,chol.avg), .funs=list(mean=~mean(., na.rm=T),
                                                            se=se,
                                                            sd=~sd(., na.rm=T)))

kable(summary.data.complete, caption="Cholesterol levels at 11 and 18 weeks")

library(broom)
wilcox.test(cholesterol.data$chol2,
       cholesterol.data$chol1, 
       paired=TRUE) %>% tidy %>%
  kable(caption="Pairwise t-test ofcholesterol levels week 11 and week 18")
  
summary.data.complete %>%
  pivot_longer(cols=starts_with('chol'),
               names_sep="_",
               names_to = c("Measure","Measurement"),
               values_to = "Cholesterol") %>%
  pivot_wider(values_from="Cholesterol",names_from = "Measurement") %>%
  mutate(Value=fct_recode(as.factor(Measure),
                          "Average"="chol.avg",
                          `8 Weeks`="chol1",
                          `19 Weeks`="chol2")) %>%
  ggplot(aes(y=mean,ymin=mean-sd,ymax=mean+sd,x=Value)) +
  geom_bar(stat='identity') +
  geom_errorbar(width=0.5) +
  facet_grid(Diet~sex) +
    labs(y="Cholesterol (mg/dL)",
       x="") +
  scale_fill_grey() +
  scale_color_grey() +
  theme_classic() +
  theme(text=element_text(size=16),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r cholesterol-analysis}
summary.data <-
  cholesterol.data %>%
  group_by(sex,diet) %>%
  summarize_at(.vars=vars(chol2), .funs=list(mean=~mean(., na.rm=T),
                                             se=se,
                                             sd=~sd(., na.rm=T)))

library(broom)
lm(chol2~sex*diet, data=cholesterol.data) %>%
  tidy %>%
  kable(caption="Global interactions between sex and diet")

lm(chol2~sex+diet, data=cholesterol.data) %>%
  tidy %>%
  kable(caption="Global effects of sex and diet, no interaction",
        digits=c(0,2,2,2,99))

cholesterol.data %>%
  group_by(sex,diet) %>%
  filter(!is.na(chol2)) %>%
  count %>%
  kable(caption="Total cholesterol values at 19 weeks for the complete DO dataset")

library(ggplot2)
cholesterol.data %>%
  ggplot(aes(y=chol2,x=sex,
             fill=Diet)) +
  geom_violin() +
  geom_jitter(alpha=0.2,
              position = position_jitterdodge(dodge.width = 0.9,
                                              jitter.width = 0.5,
                                              jitter.height = 0)) +
  labs(y="Cholesterol (mg/dL)",
       x="Sex") +
  scale_fill_grey() +
  scale_color_grey() +
  theme_classic() +
  theme(text=element_text(size=16),
        legend.position = c(0.15,0.8))
  
```

There is no evidence of an interaction between sex and diet, though both covariates were significant individually.  

# Cholesterol Classification

Classified elevated cholesterol as being greater than the mean for the averaged cholesterol data for all mice

```{r cholesterol-class}
cholesterol.data <-
  cholesterol.data %>%
  mutate(High.Chol = chol2 > mean(chol2,na.rm=T)) %>%
  mutate(sex = as.factor(sex),
         diet = as.factor(diet))
```

# Classification Tree

First used only sex and diet to predict using classification trees

```{r sex-diet-tree}
library(tree)

# load libraries
library(rpart)
library(rattle)

tree.sex.diet <- rpart(High.Chol~sex+Diet, data=cholesterol.data)
fancyRpartPlot(tree.sex.diet)
```

Next included fat mass as a predictor

```{r fat-tree}
tree.fat <- rpart(High.Chol~sex+diet+percfat2, data=cholesterol.data)
log.fat <- glm(High.Chol~sex+diet+percfat2, data=cholesterol.data, family='binomial')
log.fat %>% tidy %>% kable(caption="Logistic regression for sex and diet as predictors of above average cholesterol levels")

fancyRpartPlot(tree.sex.diet)
```

```{r full-tree}
#function to figure out if all columns are na, pass to select  
not_all_na <- function(x) any(!is.na(x))  

cholesterol.data %>%
  dplyr::select(where(not_all_na)) %>% #remove columns with all nA
  dplyr::select(-mouse.id,-chol.avg,-chol1, -hdld1,-hdld2,-diet) -> #remove cholesterol and hdl columns
  chol.pred.data.high #testing higher than average cholesterol

cholesterol.data %>%
  select(where(not_all_na)) %>% #remove columns with all nA
  select(-mouse.id,-chol1,-chol.avg,-High.Chol, -hdld1,-hdld2,-diet) -> #remove cholesterol and hdl columns
  chol.pred.data.cont #testing continuous cholesterol levels

 
chol.pred.data.cont %>% 
  rename(`TG 19w`="tg2",
         `Ca 19w`="calcium2",
         `BW 19w`="bw.19") %>%
  rpart(chol2~., data=., method="anova") -> tree.all.cont
chol.pred.data.high %>% rpart(High.Chol~., data=., method="class") -> tree.all.high

fancyRpartPlot(tree.all.cont, main="Full tree, predicting continuous cholesterol levels")
fancyRpartPlot(tree.all.high, main="Full tree, predicting above average cholesterol levels")

#pruning of the continuous model, first showed the complexity parameter table
tree.all.cont$cptable %>% kable(caption="Complexity parameter table, used to idenfiy minumum crossvalidated error rate (xerror)")

prune(tree.all.cont, cp=0.0365) -> tree.all.cont.pruned

fancyRpartPlot(tree.all.cont.pruned, uniform=TRUE, main="Pruned tree predicting continuous cholesterol levels")

rpart.plot::prp(tree.all.cont.pruned, extra=1, 
                main="",
                cex=1.2) 

tree.all.cont.pruned

```

```{r Calcium}
log.calcium <- glm(High.Chol~sex+diet+tg2+calcium2, data=cholesterol.data, family='binomial')
summary(log.calcium)

library(ggplot2)

ggplot(data=cholesterol.data,
       aes(y=chol2,
           x=calcium2,
           col=Diet)) +
  geom_point() +
  facet_grid(.~sex) +
  geom_smooth(method=lm, se=F) +
  labs(y="Cholesterol (mg/dL)",
       x="Calcium (mg/dL)",
       title="DO Strains") +
  scale_fill_grey() +
  scale_color_grey() +
  theme_classic() +
    guides(color=guide_legend(override.aes=list(fill=NA))) +
  theme(text=element_text(size=16),
        legend.position = c(0.15,0.80),
        legend.key=element_blank(),
        legend.background=element_blank())

 lm(chol2~Diet+sex+calcium2, data=cholesterol.data) %>% 
   tidy %>% 
   kable(caption="Diet adjusted association of cholesterol with calcium",
         digits=c(3,3,3,2,99))
 
  lm(chol2~Diet+sex+calcium2, data=cholesterol.data) %>% 
   glance %>% 
   kable(caption="Diet adjusted association of cholesterol with triglycerides",
         digits=c(3,3,3,2,99))
  
cholesterol.data %>%
  group_by(Diet,sex) %>%
  summarize(Estimate = cor.test(chol2,calcium2, method="spearman")$estimate,
            P.value = cor.test(chol2,calcium2, method="spearman")$p.value) %>%
  kable(caption="Spearman's rho estimates for cholesterol and calcium for each subgroup of diet and sex",
        digits=c(0,0,3,99))

lm.calcium.1 <- lm(chol2~calcium2, data=cholesterol.data)
lm.calcium.2 <- lm(chol2~calcium2+sex, data=cholesterol.data)
lm.calcium.3 <- lm(chol2~calcium2+sex+diet, data=cholesterol.data)
lm.calcium.4 <- lm(chol2~calcium2+sex+tg2, data=cholesterol.data)
lm.calcium.5 <- lm(chol2~calcium2+sex+bw.19, data=cholesterol.data)
lm.calcium.6 <- lm(chol2~sex+diet+calcium2, data=cholesterol.data)
lm.calcium.7 <- lm(chol2~sex+diet+calcium2+tg2, data=cholesterol.data)
lm.calcium.8 <- lm(chol2~sex+diet+calcium2+tg2+bw.19, data=cholesterol.data)
anova(lm.calcium.3,lm.calcium.4)
lm.calcium.9 <- lm(chol2~sex+diet*calcium2, data=cholesterol.data)
aov.calcium <- aov(chol2~sex+diet+calcium2, data=cholesterol.data)
summary(lm.calcium.1) %>% tidy %>% kable(caption="Effects of calcium on cholesterol at 18w", digits=50)
summary(lm.calcium.2) %>% tidy %>% kable(caption="Sex adjusted effects of calcium on cholesterol at 18w", digits=50)
summary(lm.calcium.3) %>% tidy %>% kable(caption="Sex and diet adjusted effects of calcium on cholesterol at 18w", digits=50)
summary(lm.calcium.4) %>% tidy %>% kable(caption="Sex and diet and tg adjusted effects of calcium on cholesterol at 18w", digits=50)
summary(lm.calcium.5) %>% tidy %>% kable(caption="Sex and diet and tg adjusted effects of calcium on cholesterol at 18w", digits=50)
summary(lm.calcium.6) %>% tidy %>% kable(caption="Sex and diet and tg adjusted effects of calcium on cholesterol at 18w", digits=50)
summary(lm.calcium.7) %>% tidy %>% kable(caption="Sex and diet and tg adjusted effects of calcium on cholesterol at 18w", digits=50)
summary(lm.calcium.8) %>% tidy %>% kable(caption="Sex and diet and tg adjusted effects of calcium on cholesterol at 18w", digits=50)


lm.calcium.chow <- lm(chol2~sex+calcium2, filter(cholesterol.data, diet == "chow"))
summary(lm.calcium.chow) %>% glance %>% kable

lm.calcium.hf <- lm(chol2~sex+calcium2, filter(cholesterol.data, diet == "hf"))
summary(lm.calcium.hf) %>% glance %>% kable

lm.calcium <- lm(chol2~sex+diet+calcium2, cholesterol.data)
summary(lm.calcium) %>% glance %>% kable(caption="Linear model of calcium on cholesterol adjusting for sex and diet")

library(effectsize)
eta_squared(lm.calcium, partial = TRUE) %>% kable(caption="Partial effect sizes for calcium model")

#moderation by triglycerides
lm.calcium.tg <- lm(chol2~sex+diet+calcium2+tg2, cholesterol.data)
summary(lm.calcium.tg) %>% tidy %>% kable(caption="Linear model of calcium on cholesterol adjusting for sex, diet and triglycerides")

anova(lm.calcium,lm.calcium.tg)%>% 
  kable(caption="Comparason of models with or without triglyceride levels",
        digits=c(0,0,0,0,0,99))

```

## Effects of Diet and Sex on Calcium

```{r calcium-stratification}
summary.data <-
  cholesterol.data %>%
  group_by(sex,diet) %>%
  summarize_at(.vars=vars(calcium2), .funs=list(mean=~mean(., na.rm=T),
                                                se=se,
                                                sd=~sd(., na.rm=T)))

lm(calcium2~sex*diet, data=cholesterol.data) %>%
  tidy %>%
  kable(caption="Global interactions between sex and diet on calcium levels")

lm(calcium2~sex+diet, data=cholesterol.data) %>%
  tidy %>%
  kable(caption="Global effects of sex and diet on calcium levels, no interaction",
        digits=c(0,2,2,2,99))

cholesterol.data %>%
  group_by(sex,diet) %>%
  filter(!is.na(calcium2)) %>%
  count %>%
  kable(caption="Total calcium values for complete DO dataset")

cholesterol.data %>%
  ggplot(aes(y=calcium2,x=sex,
             fill=Diet)) +
  geom_violin() +
  geom_jitter(alpha=0.2,
              position = position_jitterdodge(dodge.width = 0.9,
                                              jitter.width = 0.5,
                                              jitter.height = 0)) +
  labs(y="Calcium (mg/dL)",
       x="Sex") +
  scale_fill_grey() +
  scale_color_grey() +
  theme_classic() +
  theme(text=element_text(size=16),
        legend.position = c(0.85,0.15))
```

# Bone Content and Density

```{r bmd}
ggplot(data=cholesterol.data,
       aes(y=chol2,
           x=bmd2,
           col=Diet)) +
  geom_point() +
  facet_grid(.~sex) +
  geom_smooth(method=lm, se=F) +
  labs(y="Cholesterol (mg/dL)",
       x="Bone Mineral Density (g/cm2)") +
  scale_fill_grey() +
  scale_color_grey() +
  theme_classic() +
    guides(color=guide_legend(override.aes=list(fill=NA))) +
  theme(text=element_text(size=16),
        legend.position = c(0.15,0.80),
        legend.key=element_blank(),
        legend.background=element_blank())

 lm(chol2~Diet+sex+bmd2, data=cholesterol.data) %>% 
   tidy %>% 
   kable(caption="Diet adjusted association of cholesterol with bone mineral density",
         digits=c(0,3,3,2,99))
```

```{r bmc}
ggplot(data=cholesterol.data,
       aes(y=chol2,
           x=bmc2,
           col=Diet)) +
  geom_point() +
  facet_grid(.~sex) +
  geom_smooth(method=lm, se=F) +
  labs(y="Cholesterol (mg/dL)",
       x="Bone Mineral Content (g)") +
  scale_fill_grey() +
  scale_color_grey() +
  theme_classic() +
    guides(color=guide_legend(override.aes=list(fill=NA))) +
  theme(text=element_text(size=16),
        legend.position = c(0.15,0.80),
        legend.key=element_blank(),
        legend.background=element_blank())

 lm(chol2~Diet+sex+bmc2, data=cholesterol.data) %>% 
   tidy %>% 
   kable(caption="Diet adjusted association of cholesterol with bone mineral content",
         digits=c(0,3,3,2,99))
```


```{r TG}
ggplot(data=cholesterol.data,
       aes(y=chol2,
           x=tg2,
           col=Diet)) +
  geom_point(alpha=0.5) +
  facet_grid(.~sex) +
  geom_smooth(method=lm, se=F) +
  labs(y="Cholesterol (mg/dL)",
       x="Triglycerides (mg/dL)",
       title="DO Strains") +
  scale_fill_grey() +
  scale_color_grey() +
  theme_classic() +
    guides(color=guide_legend(override.aes=list(fill=NA))) +
  theme(text=element_text(size=16),
        legend.position = c(0.15,0.80),
        legend.key=element_blank(),
        legend.background=element_blank())

 lm(chol2~Diet+sex+tg2, data=cholesterol.data) %>% 
   tidy %>% 
   kable(caption="Diet adjusted association of cholesterol with triglycerides",
         digits=c(0,3,3,2,99))
 
  lm(chol2~Diet+sex+tg2, data=cholesterol.data) %>% 
   glance %>% 
   kable(caption="Diet adjusted association of cholesterol with triglycerides",
         digits=c(0,3,3,2,99))
lm.tg.1 <- lm(chol2~sex+tg2, data=cholesterol.data)
lm.tg.2 <- lm(chol2~sex+diet+tg2, data=cholesterol.data)
lm.tg.3 <- lm(chol2~sex+calcium2+tg2, data=cholesterol.data)
lm.tg.4 <- lm(chol2~sex+bw.19+tg2, data=cholesterol.data)
lm.tg.5 <- lm(chol2~sex+diet+tg2+calcium2, data=cholesterol.data)
lm.tg.6 <- lm(chol2~sex+diet+tg2+calcium2+bw.19, data=cholesterol.data)
lm.tg.male <- lm(chol2~diet*tg2, dplyr::filter(cholesterol.data, sex == "M"))
lm.tg.female <- lm(chol2~diet*tg2, dplyr::filter(cholesterol.data, sex == "F"))
lm.tg.complicated <- lm(chol2~diet + sex + diet*tg2 + sex*tg2 + sex*diet + sex*diet*tg2, data=cholesterol.data)
aov.tg <- aov(chol2~sex+diet+tg2, data=cholesterol.data)

summary(lm.tg.1) %>% tidy %>% kable(caption="Sex adjusted effects of tg on cholesterol at 18w", digits=50)
summary(lm.tg.2) %>% tidy %>% kable(caption="Sex and diet adjusted effects of tg on cholesterol at 18w", digits=50)
summary(lm.tg.3) %>% tidy %>% kable(caption="Sex and diet adjusted effects of tg on cholesterol at 18w", digits=50)
summary(lm.tg.4) %>% tidy %>% kable(caption="Sex and diet adjusted effects of tg on cholesterol at 18w", digits=50)
summary(lm.tg.5) %>% tidy %>% kable(caption="Sex and diet adjusted effects of tg on cholesterol at 18w", digits=50)
summary(lm.tg.6) %>% tidy %>% kable(caption="Sex and diet adjusted effects of tg on cholesterol at 18w", digits=50)
summary(lm.tg.male) %>% tidy %>% kable(caption="Sex and diet adjusted effects of calcium on cholesterol at 18w for males", digits=50)
summary(lm.tg.female) %>% tidy %>% kable(caption="Sex and diet adjusted effects of calcium on cholesterol at 18w for females", digits=50)
summary(lm.tg.complicated) %>% tidy %>% kable(caption="Sex and diet adjusted effects of calcium on cholesterol at 18w", digits=50)

lm.tg.chow <- lm(chol2~sex+tg2, filter(cholesterol.data, diet == "chow"))
summary(lm.tg.chow) %>% glance %>% kable

lm.tg.hf <- lm(chol2~sex+tg2, filter(cholesterol.data, diet == "hf"))
summary(lm.tg.hf) %>% glance %>% kable

library(effectsize)
omega_squared(aov.calcium, partial = TRUE) %>%
  kable(caption="Partial effect size estimates for predictors of continuous cholesterol levels")

dplyr::filter(cholesterol.data, sex == "M")
```

```{R BW_19}
ggplot(data=cholesterol.data,
       aes(y=chol2,
           x=bw.19,
           col=diet)) +
  geom_point() +
  facet_grid(.~sex) +
  geom_smooth(method=lm)


lm.bw.1 <- lm(chol2~bw.19, data=cholesterol.data)
lm.bw.2 <- lm(chol2~bw.19+sex, data=cholesterol.data)
lm.bw.3 <- lm(chol2~sex+diet+bw.19, data=cholesterol.data)
lm.bw.4 <- lm(chol2~sex+calcium2+bw.19, data=cholesterol.data)
lm.bw.5 <- lm(chol2~sex+tg2+bw.19, data=cholesterol.data)
lm.bw.6 <- lm(chol2~sex+diet+bw.19+tg2, data=cholesterol.data)
lm.bw.7 <- lm(chol2~sex+diet+bw.19+tg2+calcium2, data=cholesterol.data)

lm.bw.6 <- lm(chol2~sex+diet*bw.19, data=cholesterol.data)
aov.calcium <- aov(chol2~sex+diet+bw.19, data=cholesterol.data)
summary(lm.bw.1) %>% tidy %>% kable(caption="Effects of calcium on cholesterol at 18w", digits=50)
summary(lm.bw.2) %>% tidy %>% kable(caption="Sex adjusted effects of calcium on cholesterol at 18w", digits=50)
summary(lm.bw.3) %>% tidy %>% kable(caption="Sex and diet adjusted effects of calcium on cholesterol at 18w", digits=50)
summary(lm.bw.4) %>% tidy %>% kable(caption="Sex and diet and tg adjusted effects of calcium on cholesterol at 18w", digits=50)
summary(lm.bw.5) %>% tidy %>% kable(caption="Sex and diet and tg adjusted effects of calcium on cholesterol at 18w", digits=50)


lm.bw.chow <- lm(chol2~sex+bw.19, filter(cholesterol.data, diet == "chow"))
summary(lm.bw.chow) %>% glance %>% kable

lm.bw.hf <- lm(chol2~sex+bw.19, filter(cholesterol.data, diet == "hf"))
summary(lm.bw.hf) %>% glance %>% kable

library(effectsize)
omega_squared(aov.calcium, partial = TRUE) %>%
  kable(caption="Partial effect size estimates for predictors of continuous cholesterol levels")

dplyr::filter(cholesterol.data, sex == "M")

```



# Random Forests

```{r random-forrest}
library(caret)
library(ipred)

# Specify 10-fold cross validation
ctrl <- trainControl(method = "cv",  number = 10) 

# train bagged model
bagged_cv <- bagging(
  formula = chol2~Diet+calcium2,
  data    = chol.pred.data.cont,
  coob=TRUE
  )

# assess 10-50 bagged trees
ntree <- 10:50

# create empty vector to store OOB RMSE values
rmse <- vector(mode = "numeric", length = length(ntree))

for (i in seq_along(ntree)) {
  # reproducibility
  set.seed(123)
  
  # perform bagged model
  model <- bagging(
  formula = chol2~Diet+sex+calcium2+tg2,
  data    = chol.pred.data.cont,
  coob=TRUE,
  nbagg   = ntree[i]
)
  # get OOB error
  rmse[i] <- model$err
}

plot(ntree, rmse, type = 'l', lwd = 2)
abline(v = 25, col = "red", lty = "dashed")

predict_model<-predict(bagged_cv, chol.pred.data.cont)

ggplot(chol.pred.data.cont, aes(x=predict_model,y=chol2)) +
  geom_point()


library(randomForest)
forest <- randomForest(chol2~Diet+sex+calcium2+tg2, 
             data = chol.pred.data.cont,
             na.action=na.exclude) 
varImpPlot(forest)

predict_model<-predict(forest, chol.pred.data.cont)

```

# Session Information

```{r session-information, echo=T}
sessionInfo()
```

