---
title: "Bayesian inference to identify locus-specific causal genes for cholesterol QTLs"
author: "Dave Bridges"
date: "2025-12-15"
editor: source
format: 
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
    fig-path: "figures/"
theme: journal
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)
library(knitr)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

# Purpose

To set up and execute a causal framework to evaluate causal genes in individual QTLs we have identified for cholesterol.

# Framework

This used chatGPT to help draft this code.  The presumption is that we have several pieces of information

* The presence of a cholesterol QTL either diet specific or not
* Colocalization analyses using coloc/SuSiE

Our goal is to compute $P(Cg​=1∣\text{all observed data})$

Our *prior probablity* for each gene sparse, and is based on $P(Cg​=1)=\pi_o\approx\frac{1}{N_\text{genes in locus}}$

```{r clump-data-entry}
ncd.clump.filename <- 'ld-calculations/Annotated Gene Clumps NCD QTLs.csv'
hfd.clump.filename <- 'ld-calculations/Annotated Gene Clumps HFD QTLs.csv'

library(readr)
ncd.clump.data <- read_csv(ncd.clump.filename) 

ncd.clump.summary <-
  ncd.clump.data |>
  group_by(SNP) |>
  filter(gene_biotype == 'protein_coding') |>
  summarize(BP = first(BP),
            Min.QTL = first(Min_Position),
            Max.QTL = first(Max_Position),
            Genes = paste(external_gene_name, collapse = ';'),
            Gene.Count = length(ensembl_gene_id)) |>
  rename(Clump=SNP) |>
  mutate(Prior = 1 / Gene.Count)


hfd.clump.data <- read_csv(hfd.clump.filename) 

hfd.clump.summary <-
  hfd.clump.data |>
  group_by(SNP) |>
  filter(gene_biotype == 'protein_coding') |>
  summarize(BP = first(BP),
            Min.QTL = first(Min_Position),
            Max.QTL = first(Max_Position),
            Genes = paste(external_gene_name, collapse = ';'),
            Gene.Count = length(ensembl_gene_id)) |>
  rename(Clump=SNP) |>
  mutate(Prior = 1 / Gene.Count)

twas.all.filename <- 'Cholesterol - Expression Associations - Linear Model.csv'
twas.all.data <- read_csv(twas.all.filename)
```

There is an average of `r round(mean(ncd.clump.summary$Gene.Count),1)` genes [`r min(ncd.clump.summary$Gene.Count)` -`r max(ncd.clump.summary$Gene.Count)`] per NCD QTL and `r round(mean(hfd.clump.summary$Gene.Count),1)` genes per HFD QTL [`r min(hfd.clump.summary$Gene.Count)` -`r max(hfd.clump.summary$Gene.Count)`] .

## Formalizing of observed data

### Coloc data

```{r coloc-data-entry}
coloc.summary.filename <- 'Coloc/Coloc results summary.csv'
coloc.data <- read_csv(coloc.summary.filename)

alpha.coloc <- 1

coloc.summary <-
  coloc.data |>
  group_by(Gene, Diet) %>%
  arrange(desc(PP.H4.abf), desc(PP.H3.abf), desc(PP.H2.abf)) %>%
  slice(1) %>%
  ungroup() |>
  mutate(logLR.coloc = alpha.coloc * log(PP.H4.abf/(PP.H2.abf + PP.H3.abf + alpha.coloc))) 
```

From coloc analyses, found in  we have several probabilities:

* $PP.H2$ is the probability of eQTL signal only
* $PP.H3$ is the probability of both traits being associated, but with *different causal* variants
* $PP.H4$ is the probability of both traits being associated and share the *same causal* variant

$LR_{coloc,g}=\frac{P(data∣C_g=0)}{P(data∣C_g=1)}=\alpha \times log(\frac{PP.H4}{PP.H3+PP.H2+\epsilon})$

From this analysis:

- $\epsilon$ is a numerical and conceptual stabilizer, not a biological parameter.  It prevents division by zero, extreme log-odds when posteriors are near zero, and overconfidence from poorly powered regions.  Generally something like $1 \times 10^{-6}$ or more conservatively $1\times 10^{-4}$.
- $\alpha$ is a weighting parameter controlling how strongly coloc evidence influences the final posterior.  Something between $1-1.5$ is reasonable here.  I used `r alpha.coloc`.

### Distance to QTL

Using gene coordinates to define a distance-based prior modifier, rather than a filter.  This will be defined as $log(LR_{dist,g}λ⋅lo\lambda1+dg​)$_ where $\l4ambda$ is a scaling parameter controlling the influence of distance on the final posterior.  $d_g$ is the distance from the gene to the lead SNP for the QTL.

```{r gene-qtl-distance}
lambda <- 1.0
ncd.clump.data %>%
  mutate(Gene_Midpoint = (start_position + end_position) / 2) |>
  mutate(Distance_to_QTL = abs(Gene_Midpoint - BP)/1E6) |> #in Mb
  mutate(logLR_dist = -lambda * log(1+Distance_to_QTL)) |>
  select(SNP,ensembl_gene_id, external_gene_name, Distance_to_QTL, logLR_dist) ->
  ncd.gene.distances
```

We incorporated a weak distance-based prior that modestly favored genes proximal to the QTL peak while allowing strong genetic or transcriptional evidence to outweigh spatial proximity. The distance penalty was calibrated such that a gene 1 Mb from the peak incurred less than `one`r lambda` log-odds unit of penalty.

## Gene Expression Proxy

This is not causal by itself, but informative conditional on genetics.

To do this, first we defined a standardized effect: $Z_{expr,g} = \frac{β_{g}}{SE_{g}}$ where $β_{g}$ is the effect estimate from the linear model of expression vs cholesterol and $SE_{g}$ is the standard error of that estimate.

```{r gene-expression-proxy}
mu_med    <- 1.0
sigma_med <- 1.0

moduleB_logLR <- function(beta, se,
                          max_reward = 1.0,
                          weight = 0.2) {

  Z <- beta / se

  # reward strong associations, but do not penalize weak ones
  logLR <- weight * (Z^2) / 2

  # cap the contribution
  logLR <- pmin(logLR, max_reward)

  return(logLR)
}

twas.all.data %>%
  right_join(hfd.clump.data %>%
               select(SNP,ensembl_gene_id, external_gene_name, CHR, start_position, end_position) %>%
               distinct(),
             by = c('ID'='ensembl_gene_id')) %>%
  mutate(logLR.expression = moduleB_logLR(beta, se)) |>
  select(chr,symbol,logLR.expression) ->
  hfd.gene.expression.proxy
```

Then define a likelihood ratio component as $LR_{expr,g} = logit^{-1}(Z_{expr,g})$.

# Analyses

## Chromosome 18 QTL NCD specific

```{r ncd-chr-18}
twas.all.data %>%
  right_join(ncd.clump.data %>%
               filter(CHR==18) %>%
               select(SNP,ensembl_gene_id, external_gene_name, CHR, start_position, end_position),
             by = c('ID'='ensembl_gene_id')) %>%
  mutate(logLR_expression = moduleB_logLR(beta, se)) |>
  select(SNP,chr,symbol,logLR_expression) |>
  filter(!is.na(symbol)) |>
  left_join(ncd.clump.summary |>
              select(Clump,Prior,Gene.Count),by=c("SNP"="Clump")) |>
  full_join(coloc.summary |> select(Gene,Diet, logLR.coloc), by=c("symbol"="Gene")) |>
  left_join(ncd.gene.distances |> select(logLR_dist,external_gene_name), by=c("symbol"="external_gene_name")) |>
  filter(!is.na(SNP)) |>
  mutate(log_prior = log(Prior)) |>
  mutate(log_odds = log_prior + logLR.coloc + logLR_dist + logLR_expression) |>
  select(symbol, log_odds, log_prior, logLR.coloc, logLR_dist, logLR_expression) |>
  mutate(posterior = plogis(log_odds)) |>
  arrange(-log_odds) ->
  chr18.summary.data

kable(chr18.summary.data, caption="liklihood of each gene in the chromosome 18 QTL")

#df$log_odds <-
#log_prior +
#df$logLR_coloc +
#df$logLR_B +
#df$logLR_dist +
#df$logLR_human

```

# Session Information

```{r session-info}
sessionInfo()
```

