---
title: "Creating Plink files from HFD-DO Viewer Dataset"
author: "Dave Bridges"
date: "2024-06-30"
format: html
---

This script is for generating Plink compatible files, this is necessary for genotype clumping and LD calculations.  THis script can be found in `r getwd()` and was most recently run on `r date()`.

# Load Dataset

```{r load}
filename <- 'core.Svenson_DO_HFD.v12.Rdata'
load(filename)
```

Loaded file at `r filename`.

# Plink2 format

Plink2 uses the binary pgen/pvar/psam formating.  These are binaries that can be generated by `--make pgen` commands which start from a bgen file, a vcf file or a plink1 set of bed/bim/fam files.

## Plink BED file

To make a bed file used the genio package to create bed/bim/fam files

Need a matrix of are m loci and n individuals.

```{r df-1-5}
library(dplyr)
library(tibble)
library(stringr)
library(tidyr)

df.1 <- as.data.frame(genoprobs$`1`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 

df.2 <- as.data.frame(genoprobs$`2`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 

df.3 <- as.data.frame(genoprobs$`3`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 

df.4 <- as.data.frame(genoprobs$`4`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 

df.5 <- as.data.frame(genoprobs$`5`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 
```

```{r df6-10}
df.6 <- as.data.frame(genoprobs$`6`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 

df.7 <- as.data.frame(genoprobs$`7`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 

df.8 <- as.data.frame(genoprobs$`8`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 

df.9 <- as.data.frame(genoprobs$`9`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 

df.10 <- as.data.frame(genoprobs$`10`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 
```

```{r df-11-15}
df.11 <- as.data.frame(genoprobs$`11`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 

df.12 <- as.data.frame(genoprobs$`12`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 

df.13 <- as.data.frame(genoprobs$`13`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 

df.14 <- as.data.frame(genoprobs$`14`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 

df.15 <- as.data.frame(genoprobs$`15`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 
```

```{r df-16-x}
df.16 <- as.data.frame(genoprobs$`16`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 

df.17 <- as.data.frame(genoprobs$`17`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 

df.18 <- as.data.frame(genoprobs$`18`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 

df.19 <- as.data.frame(genoprobs$`19`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 

df.x <- as.data.frame(genoprobs$`X`, stringsAsFactors = FALSE) %>%
  pivot_longer(cols = everything(-1),  # Replace "Sample" with the actual name of your sample identifier
               names_to = "Sample",
               values_to = "Value") %>%
  mutate(Sample=dimnames(K$X)[[1]]) %>% #added sample names which were lost in the previous step
  pivot_longer(-Sample,
               names_to="Genotype.ID",
               values_to="Likelihood") %>%
  separate(Genotype.ID, sep=2, into=c("Allele","Locus")) %>%
  mutate(Allele=str_trunc(Allele,1,side="right", ellipsis="")) %>%
  mutate(LikelihoodRound=round(Likelihood*2)) %>%
  select(Sample,Locus,Allele,LikelihoodRound) %>%
  pivot_wider(names_from=Sample,
              values_from = LikelihoodRound,values_fn=first) 
```

# Combining and Annotating with Major Alleles

```{r combining}
all.df <-
  bind_rows(df.1,
            df.2,
            df.3,
            df.4,
            df.5,
            df.6,
            df.7,
            df.8,
            df.9,
            df.10,
            df.11,
            df.12,
            df.13,
            df.14,
            df.15,
            df.16,
            df.17,
            df.18,
            df.19,
            df.x) 

#now need to find out the major allele for each locus

allele.counts <-
  all.df %>%
  ungroup %>%
  group_by(Locus) %>%
  rowwise() %>%
  mutate(Allele.Count = sum(c_across(where(is.numeric))))

library(readr) 

major.alleles <-
  allele.counts %>%
  ungroup %>%
  group_by(Locus) %>%
  filter(Allele.Count == max(Allele.Count)) %>%
  select(Locus,Allele)

summary(as.factor(major.alleles$Allele))%>% knitr::kable(caption="Most common alleles")
#the common allele is always H?
  
annotated.df <-
  allele.counts %>%
  ungroup %>%
  filter(Allele.Count>836*0.05) %>% #only alleles with greater than 5% occurance
  left_join(major.alleles, by="Locus", suffix = c("_all","_major")) %>%
  filter(Allele_all!=Allele_major) %>% #remove SNPS marked with the common allele
  mutate(ID = paste(Locus,Allele_major,Allele_all,sep="_")) %>% #minor allele first to match bimbam format
  filter(Locus != 'lue') %>%
  select(-Allele_major,-Allele_all,-Allele.Count,-Locus) %>% #dont need allele names or counts anymore
  distinct(ID,.keep_all = T) %>% #need unique row names
  column_to_rownames("ID") %>%
  as.matrix  


pheno <- rnorm(ncol(annotated.df)) #random data

#needs sex for some genotype information
fam <- data.frame(fam=paste("FAM",seq(1,ncol(annotated.df)),sep=""), 
                  id=colnames(annotated.df), 
                  pat=rep(0,ncol(annotated.df)),
                  mat=rep(0,ncol(annotated.df))) %>%
  mutate(sex=case_when(substring(id, 1, 1)=="M"~1,
                       substring(id, 1, 1)=="F"~2,
                       .default=0)) %>% #sex id first letter of mouse id %>%
  mutate(pheno=rep(0,ncol(annotated.df)))

library(genio)
bim.data <- data.frame(ID=rownames(annotated.df)) %>%
  separate(ID, sep="_",
           into=c("chr","pos","ref","alt"),
           remove=F) %>%
  rename(id=ID) %>%
  mutate(posg=as.integer(pos)/1000000) %>%
  select(chr,id,posg,pos,ref,alt)


plink_out <- 'genotypes.plink'
write_plink(plink_out, 
            annotated.df, 
            bim=bim.data,
            pheno = pheno,
            fam=fam)

#map file
map.data <- data.frame(ID=rownames(annotated.df)) %>%
  separate(ID, sep="_",into=c("Chr","Position","Ref","Alt"),remove=F) %>%
  filter(Chr!="lue") %>%
  mutate(cM=as.integer(Position)/1000000) %>%
  select(Chr,ID,cM,Position)

library(readr)
write_tsv(map.data,
          file="genotypes.plink.map",
          col_names = F)
```
 
# Session Information

```{r session-info}
sessionInfo()
```
