---
title: "Serum Calcium Effect — Outcome Robustness: Albumin (assay artefact) & Phosphate (co-regulation)"
author: "Dave Bridges"
date: today
editor: source
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
    fig-path: "figures-calcium-artefact/"
theme: journal
execute:
  echo: true
  warning: false
---

```{r global-options}
#| echo: false
#| message: false
library(tidyverse)
library(knitr)
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

The cholesterol/HMGCR → serum-calcium effect is strong (≈ 0.185, p < 1e-20) and
HMGCR-specific, but it is **not** routed through bone mineral density (every BMD
site → calcium leg screened null). Before interpreting it as a real biological
pathway, this script stress-tests the **outcome** itself with two cheap,
decisive checks:

1. **Albumin (assay / binding artefact).** UK Biobank measures *total* serum
   calcium, ~45 % of which is albumin-bound, and lipaemic samples interfere with
   colorimetric calcium assays. If cholesterol perturbs albumin, part of the
   "calcium" signal is a binding/measurement artefact rather than ionised
   calcium. Tests: does HMGCR/LDL → albumin? does albumin → calcium (positive
   control)? and does adjusting for albumin in MVMR **attenuate** the
   cholesterol → calcium effect?

2. **Phosphate (co-regulation diagnostic).** Calcium and phosphate are jointly
   controlled by PTH, FGF23 and vitamin D. The **sign pattern** of the
   cholesterol effect on Ca vs PO₄ discriminates mechanisms:
   - Ca ↑ **and** PO₄ ↑ → intestinal absorption / vitamin-D axis, or bone
     resorption (both ions released together)
   - Ca ↑ **and** PO₄ ↓ → PTH-like (renal phosphate wasting)
   - Ca ↑ **and** PO₄ ≈ 0 → calcium-specific route, or an albumin/assay artefact
     (phosphate is not albumin-bound, so an assay artefact should spare it)

All UK Biobank, so this is one-sample MR (bias toward the null) — noted, not
avoided; the calcium, albumin and phosphate GWAS come from the *same* UKB
biomarker batch (Barton et al. 2021, PMID 34226706), which keeps the
albumin adjustment clean.

---

## Setup

```{r setup}
#| label: setup
#| message: false
library(TwoSampleMR)
library(MendelianRandomization)
library(ieugwasr)
library(knitr)
library(kableExtra)

GWAS <- list(
  ldl       = "ieu-b-110",           # LDL-C, UKB Neale (all-SNP exposure)
  calcium   = "ebi-a-GCST90025990",  # Serum calcium   (Barton et al. 2021 UKB)
  albumin   = "ebi-a-GCST90025992",  # Serum albumin   (same batch)
  phosphate = "ebi-a-GCST90025948"   # Serum phosphate (same batch)
)

# HMGCR cis window (GRCh37 ±500 kb) and clumping (allele score, Swerdlow 2015)
HMGCR_REGION <- "5:74132993-75157941"
R2_CIS       <- 0.30
R2_CLUMP     <- 0.001
CLUMP_KB     <- 10000
P_CIS        <- 5e-8
P_GW         <- 5e-8

dir.create("results", showWarnings = FALSE)
```

### Confirm the datasets (trait + N)

```{r gwasinfo}
#| label: gwasinfo
#| cache: true

tryCatch(
  ieugwasr::gwasinfo(unlist(GWAS)) %>% as_tibble() %>%
    select(any_of(c("id", "trait", "sample_size", "population", "author", "year"))) %>%
    kable(caption = "Datasets used in the outcome-robustness checks"),
  error = function(e) message("gwasinfo failed: ", e$message)
)
```

---

## Helper functions

```{r helpers}
#| label: helpers

# HMGCR cis exposure in TwoSampleMR format (allele-score clumped at r2<0.30).
hmgcr_cis_exposure <- function(exposure_id = GWAS$ldl) {
  raw <- ieugwasr::associations(variants = HMGCR_REGION, id = exposure_id,
                                proxies = FALSE) %>% as_tibble() %>%
    filter(p <= P_CIS)
  clumped <- if (nrow(raw) >= 2) {
    tryCatch(
      ieugwasr::ld_clump(tibble(rsid = raw$rsid, pval = raw$p, id = exposure_id),
                         clump_r2 = R2_CIS, clump_kb = CLUMP_KB, pop = "EUR") %>%
        inner_join(raw, by = "rsid"),
      error = function(e) { warning("HMGCR clump: ", e$message); raw }
    )
  } else raw
  clumped %>%
    transmute(SNP = rsid, beta.exposure = beta, se.exposure = se,
              effect_allele.exposure = toupper(ea), other_allele.exposure = toupper(nea),
              eaf.exposure = eaf, pval.exposure = p,
              exposure = "HMGCR (cis LDL-C)", id.exposure = exposure_id,
              mr_keep.exposure = TRUE, pval_origin.exposure = "reported",
              data_source.exposure = "igd")
}

# Univariable MR of one exposure -> one outcome, returned as a tidy one-row frame.
run_uni <- function(exp_dat, out_id, exp_label, out_label) {
  if (is.null(exp_dat) || nrow(exp_dat) < 1)
    return(tibble(exposure = exp_label, outcome = out_label, n_snps = 0L,
                  beta = NA_real_, se = NA_real_, ci_lower = NA_real_,
                  ci_upper = NA_real_, pval = NA_real_, egger_p = NA_real_))
  out  <- extract_outcome_data(snps = exp_dat$SNP, outcomes = out_id)
  harm <- harmonise_data(exp_dat, out) %>% filter(mr_keep)
  nsnp <- nrow(harm)
  ml   <- if (nsnp >= 3) c("mr_ivw_mre", "mr_egger_regression", "mr_weighted_median") else
          if (nsnp == 2) "mr_ivw_mre" else "mr_wald_ratio"
  res  <- mr(harm, method_list = ml) %>% slice(1)
  egg  <- tryCatch(mr_pleiotropy_test(harm)$pval, error = function(e) NA_real_)
  tibble(exposure = exp_label, outcome = out_label, n_snps = nsnp,
         beta = res$b, se = res$se,
         ci_lower = res$b - 1.96 * res$se, ci_upper = res$b + 1.96 * res$se,
         pval = res$pval, egger_p = egg)
}

# Difference-method MVMR (IVW-MRE) via MendelianRandomization::mr_mvivw.
mvmr_ivw <- function(mvdat, labels) {
  fit <- MendelianRandomization::mr_mvivw(MendelianRandomization::mr_mvinput(
    bx = mvdat$exposure_beta, bxse = mvdat$exposure_se,
    by = as.numeric(mvdat$outcome_beta), byse = as.numeric(mvdat$outcome_se)))
  tibble(exposure = labels, beta = fit@Estimate, se = fit@StdError,
         ci_lower = fit@CILower, ci_upper = fit@CIUpper, pval = fit@Pvalue)
}

# Delta-method product-of-coefficients indirect effect (a * b).
product_indirect <- function(a, se_a, b, se_b) {
  est <- a * b; se <- sqrt(a^2 * se_b^2 + b^2 * se_a^2)
  tibble(beta = est, se = se, ci_lower = est - 1.96*se, ci_upper = est + 1.96*se,
         pval = 2 * pnorm(-abs(est / se)))
}

# Batched OpenGWAS lookup (avoids silent truncation of long variant lists).
associations_batched <- function(rsids, gwas_id, batch_size = 100) {
  rsids <- unique(rsids)
  split(rsids, ceiling(seq_along(rsids) / batch_size)) %>%
    purrr::map_dfr(function(b) tryCatch(
      ieugwasr::associations(variants = b, id = gwas_id, proxies = FALSE) %>% as_tibble(),
      error = function(e) { warning(gwas_id, " batch: ", e$message); tibble() })) %>%
    distinct(rsid, .keep_all = TRUE)
}

# Manual (cis exposure + genome-wide mediator) MVMR on calcium. mv_extract_exposures
# assumes both exposures share a genome-wide instrument pool, which is false for a
# cis drug target — so we assemble the exposure matrix from the rsID union by hand,
# aligning the mediator and outcome betas to the cholesterol effect allele.
cis_mvmr_on_calcium <- function(cis_exp, med_inst, med_id, cis_label, med_label) {
  union_rsids <- union(cis_exp$SNP, med_inst$SNP)
  chol_eff <- associations_batched(union_rsids, cis_exp$id.exposure[1]) %>%
    transmute(SNP = rsid, ea, nea, bx1 = beta, sx1 = se)
  med_eff  <- associations_batched(union_rsids, med_id) %>%
    transmute(SNP = rsid, bx2 = beta, sx2 = se, ea_m = ea)
  ca_eff   <- associations_batched(union_rsids, GWAS$calcium) %>%
    transmute(SNP = rsid, by = beta, sy = se, ea_ca = ea)
  mv <- chol_eff %>%
    inner_join(med_eff, by = "SNP") %>% inner_join(ca_eff, by = "SNP") %>%
    mutate(bx2 = if_else(toupper(ea_m)  == toupper(ea), bx2, -bx2),
           by  = if_else(toupper(ea_ca) == toupper(ea), by,  -by)) %>%
    filter(!is.na(bx1), !is.na(bx2), !is.na(by))
  mvdat <- list(exposure_beta = as.matrix(mv[, c("bx1", "bx2")]),
                exposure_se   = as.matrix(mv[, c("sx1", "sx2")]),
                outcome_beta  = mv$by, outcome_se = mv$sy)
  colnames(mvdat$exposure_beta) <- colnames(mvdat$exposure_se) <- c(cis_label, med_label)
  list(res = mvmr_ivw(mvdat, c(cis_label, med_label)), n_snps = nrow(mv))
}
```

---

## Instruments

```{r instruments}
#| label: instruments
#| cache: true

# All-SNP LDL-C instruments, HMGCR cis instruments, and mediator instruments.
ldl_inst     <- extract_instruments(GWAS$ldl, p1 = P_GW, clump = TRUE,
                                    r2 = R2_CLUMP, kb = CLUMP_KB)
hmgcr_inst   <- hmgcr_cis_exposure(GWAS$ldl)
alb_inst     <- extract_instruments(GWAS$albumin,   p1 = P_GW, clump = TRUE,
                                    r2 = R2_CLUMP, kb = CLUMP_KB)
phos_inst    <- extract_instruments(GWAS$phosphate, p1 = P_GW, clump = TRUE,
                                    r2 = R2_CLUMP, kb = CLUMP_KB)

tibble(
  instrument_set = c("All LDL-C", "HMGCR cis", "Albumin", "Phosphate"),
  n_snps = c(nrow(ldl_inst), nrow(hmgcr_inst), nrow(alb_inst), nrow(phos_inst))
) %>% kable(caption = "Instrument counts")
```

---

## Univariable panel

Every leg we need, in one table: cholesterol → each biomarker, and each mediator
→ calcium (albumin → calcium is a positive control for the artefact pathway).

```{r uni-panel}
#| label: uni-panel
#| cache: true

uni_panel <- bind_rows(
  # cholesterol -> biomarkers
  run_uni(ldl_inst,   GWAS$albumin,   "All LDL-C",  "Albumin"),
  run_uni(hmgcr_inst, GWAS$albumin,   "HMGCR cis",  "Albumin"),
  run_uni(ldl_inst,   GWAS$phosphate, "All LDL-C",  "Phosphate"),
  run_uni(hmgcr_inst, GWAS$phosphate, "HMGCR cis",  "Phosphate"),
  run_uni(ldl_inst,   GWAS$calcium,   "All LDL-C",  "Calcium"),
  run_uni(hmgcr_inst, GWAS$calcium,   "HMGCR cis",  "Calcium"),
  # mediator -> calcium (second legs)
  run_uni(alb_inst,   GWAS$calcium,   "Albumin",    "Calcium"),
  run_uni(phos_inst,  GWAS$calcium,   "Phosphate",  "Calcium")
)

write_csv(uni_panel, "results/calcium_artefact_univariable.csv")
kable(uni_panel %>% mutate(across(c(beta, se, ci_lower, ci_upper), ~round(.x, 4)),
                           pval = signif(pval, 3), egger_p = signif(egger_p, 3)),
      caption = "Univariable MR panel (per SD exposure)")
```

---

## Step 1 — Albumin adjustment (the artefact test)

If the calcium signal is an albumin-binding artefact, conditioning on albumin in
an MVMR should **collapse** the cholesterol → calcium effect, and the two-step
`LDL → albumin → calcium` indirect effect should account for much of the total.

```{r albumin-mvmr}
#| label: albumin-mvmr
#| cache: true

# All-SNP MVMR: (LDL-C, albumin) -> calcium
mv_exp   <- mv_extract_exposures(id_exposure = c(GWAS$ldl, GWAS$albumin),
                                 clump_r2 = R2_CLUMP, clump_kb = CLUMP_KB)
mv_out   <- extract_outcome_data(snps = unique(mv_exp$SNP), outcomes = GWAS$calcium)
mvdat    <- mv_harmonise_data(mv_exp, mv_out)
labels   <- colnames(mvdat$exposure_beta)
mvmr_res <- mvmr_ivw(mvdat, labels)

# Total LDL-C -> calcium (from the panel above) vs albumin-adjusted direct effect
total_ldl_ca <- uni_panel %>% filter(exposure == "All LDL-C", outcome == "Calcium")
direct_ldl   <- mvmr_res %>% filter(exposure == GWAS$ldl)

# Two-step indirect via albumin: (LDL -> albumin) x (albumin -> calcium)
a_alb <- uni_panel %>% filter(exposure == "All LDL-C", outcome == "Albumin")
b_alb <- uni_panel %>% filter(exposure == "Albumin",   outcome == "Calcium")
indirect_alb <- product_indirect(a_alb$beta, a_alb$se, b_alb$beta, b_alb$se)

albumin_summary <- bind_rows(
  tibble(model = "LDL-C -> calcium (total)",          beta = total_ldl_ca$beta,
         se = total_ldl_ca$se, pval = total_ldl_ca$pval),
  tibble(model = "LDL-C -> calcium (direct | albumin)", beta = direct_ldl$beta,
         se = direct_ldl$se, pval = direct_ldl$pval),
  tibble(model = "Indirect via albumin (LDL->Alb->Ca)", beta = indirect_alb$beta,
         se = indirect_alb$se, pval = indirect_alb$pval)
) %>%
  mutate(ci_lower = beta - 1.96*se, ci_upper = beta + 1.96*se,
         pct_of_total = round(100 * beta / total_ldl_ca$beta, 1))

write_csv(albumin_summary, "results/calcium_artefact_albumin_mvmr.csv")
kable(albumin_summary %>% mutate(across(c(beta, se, ci_lower, ci_upper), ~round(.x, 4)),
                                 pval = signif(pval, 3)),
      caption = "Albumin adjustment — SECONDARY (all-SNP LDL-C: weak & pleiotropic total effect)")
```

The all-SNP LDL-C total effect on calcium is weak (≈0.04, p≈0.1) and shows a
non-zero Egger intercept, so decomposing it is unreliable. The **primary** artefact
test uses the strong, pleiotropy-clean **HMGCR-cis** signal, decomposed with the
**two-step product method** (the reliable estimator for a cis exposure). The
difference-method MVMR is included only as a flagged cross-check: pooling 16 cis
SNPs with ~200 genome-wide albumin instruments contaminates the cis coefficient,
so its "direct" effect should not be read as the answer.

```{r albumin-mvmr-hmgcr}
#| label: albumin-mvmr-hmgcr
#| cache: true

total_hmgcr_ca <- uni_panel %>% filter(exposure == "HMGCR cis", outcome == "Calcium")
a_alb_h        <- uni_panel %>% filter(exposure == "HMGCR cis", outcome == "Albumin")

# PRIMARY estimator = two-step product of coefficients. For a cis exposure this is
# the reliable decomposition: albumin-mediated (indirect) = a*b, and direct =
# total - indirect. (a = HMGCR->albumin, b = albumin->calcium.)
indirect_alb_h <- product_indirect(a_alb_h$beta, a_alb_h$se, b_alb$beta, b_alb$se)
direct_ts_beta <- total_hmgcr_ca$beta - indirect_alb_h$beta
direct_ts_se   <- sqrt(total_hmgcr_ca$se^2 + indirect_alb_h$se^2)   # approximate (treats legs as independent)

# CROSS-CHECK ONLY = difference-method MVMR. mv_extract_exposures cannot build a
# cis MVMR, so we pool the 16 cis SNPs with the genome-wide albumin instruments.
# That mixing makes the "HMGCR/LDL" coefficient a contaminated genome-wide LDL
# estimate rather than the cis effect (same failure mode as the PCSK9 arm in the
# main analysis) — so this row is a flagged cross-check, NOT the headline number.
hmgcr_alb    <- cis_mvmr_on_calcium(hmgcr_inst, alb_inst, GWAS$albumin,
                                    "HMGCR (LDL-C)", "Albumin")
direct_hmgcr <- hmgcr_alb$res %>% filter(exposure == "HMGCR (LDL-C)")

albumin_summary_hmgcr <- bind_rows(
  tibble(model = "HMGCR -> calcium (total)",                         beta = total_hmgcr_ca$beta,
         se = total_hmgcr_ca$se, pval = total_hmgcr_ca$pval),
  tibble(model = "Albumin-mediated / indirect (two-step a*b)",       beta = indirect_alb_h$beta,
         se = indirect_alb_h$se, pval = indirect_alb_h$pval),
  tibble(model = "HMGCR -> calcium DIRECT (two-step: total-indirect)", beta = direct_ts_beta,
         se = direct_ts_se, pval = 2 * pnorm(-abs(direct_ts_beta / direct_ts_se))),
  tibble(model = "HMGCR -> calcium direct (MVMR — UNRELIABLE for cis)", beta = direct_hmgcr$beta,
         se = direct_hmgcr$se, pval = direct_hmgcr$pval)
) %>%
  mutate(ci_lower = beta - 1.96*se, ci_upper = beta + 1.96*se,
         pct_of_total = round(100 * beta / total_hmgcr_ca$beta, 1))

write_csv(albumin_summary_hmgcr, "results/calcium_artefact_albumin_mvmr_hmgcr.csv")
kable(albumin_summary_hmgcr %>% mutate(across(c(beta, se, ci_lower, ci_upper), ~round(.x, 4)),
                                       pval = signif(pval, 3)),
      caption = paste0("Albumin decomposition of the HMGCR-cis calcium effect. ",
                       "Read the two-step rows: albumin-mediated fraction is negative/tiny, so the ",
                       "direct effect is ~100%+ of total ⇒ NOT an albumin artefact. The MVMR row (",
                       hmgcr_alb$n_snps, " SNPs) is contaminated by the genome-wide albumin ",
                       "instruments and is shown only as a flagged cross-check."))
```

---

## Step 2 — Phosphate co-regulation pattern

```{r phosphate-pattern}
#| label: phosphate-pattern

pattern <- uni_panel %>%
  filter(exposure %in% c("All LDL-C", "HMGCR cis"),
         outcome %in% c("Calcium", "Phosphate", "Albumin")) %>%
  select(exposure, outcome, beta, ci_lower, ci_upper, pval) %>%
  arrange(exposure, outcome)

kable(pattern %>% mutate(across(c(beta, ci_lower, ci_upper), ~round(.x, 4)),
                         pval = signif(pval, 3)),
      caption = "Cholesterol effect on calcium vs phosphate vs albumin (sign pattern is diagnostic)")
```

```{r pattern-plot}
#| label: pattern-plot
#| fig-width: 8
#| fig-height: 4.5

pdat <- uni_panel %>%
  filter(exposure %in% c("All LDL-C", "HMGCR cis"),
         outcome %in% c("Calcium", "Phosphate", "Albumin"), !is.na(beta))
ggplot(pdat, aes(x = beta, y = outcome, colour = exposure)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = ci_lower, xmax = ci_upper),
                  position = position_dodge(width = 0.4)) +
  scale_colour_manual(values = color_scheme, name = NULL) +
  labs(title = "Cholesterol effect across calcium, phosphate and albumin",
       subtitle = "Ca↑PO₄↓ = PTH-like · Ca↑PO₄↑ = absorption/resorption · Ca↑PO₄~0 = Ca-specific or albumin artefact",
       x = "Effect on biomarker (SD per SD LDL-C)", y = NULL) +
  theme_minimal(base_size = 12) + theme(legend.position = "top")
```

---

## Interpretation

```{r interpretation}
#| label: interpretation
#| echo: false
#| results: asis

g <- function(exp, out, col) {
  v <- uni_panel %>% filter(exposure == exp, outcome == out) %>% pull(.data[[col]])
  if (length(v) == 0) NA_real_ else v[1]
}
# Primary artefact test keys off the HMGCR-cis signal decomposed by the two-step
# product method (the reliable estimator for a cis exposure).
tot_h  <- total_hmgcr_ca$beta
ind_h  <- indirect_alb_h$beta          # albumin-mediated (two-step a*b)
dir_h  <- direct_ts_beta               # two-step direct = total - indirect
surv_h <- round(100 * dir_h / tot_h)
mvmr_h <- direct_hmgcr$beta            # contaminated MVMR direct (cross-check only)
alb_h <- g("HMGCR cis", "Albumin", "beta");   alb_hp <- g("HMGCR cis", "Albumin", "pval")
pho_h <- g("HMGCR cis", "Phosphate", "beta"); pho_hp <- g("HMGCR cis", "Phosphate", "pval")
ca_h  <- g("HMGCR cis", "Calcium", "beta")
alb_ca <- g("Albumin", "Calcium", "beta"); alb_ca_p <- g("Albumin", "Calcium", "pval")

cat("**Step 1 — albumin (assay artefact): NOT an artefact.**\n\n")
cat("- Albumin → calcium (positive control): β = ", signif(alb_ca, 3),
    " (p = ", signif(alb_ca_p, 2), ") — strongly positive, so the total-calcium ",
    "assay does track albumin binding: the artefact channel is open and the test is sensitive.\n", sep = "")
cat("- HMGCR → albumin: β = ", signif(alb_h, 3), " (p = ", signif(alb_hp, 2),
    "), **negative** — so the albumin-mediated contribution to calcium (a×b = ",
    signif(ind_h, 3), ") is the **opposite sign** to the observed HMGCR→calcium (",
    signif(ca_h, 3), "). Mediation through albumin cannot produce a positive calcium effect.\n", sep = "")
cat("- **Two-step direct effect = ", signif(dir_h, 3), " (", surv_h,
    "% of total)** — albumin adjustment leaves the effect intact (slightly larger), ",
    "so the HMGCR→calcium signal is **not an albumin/binding artefact**.\n", sep = "")
cat("- (Cross-check: the difference-method MVMR gives ", signif(mvmr_h, 3),
    ", but that estimate is contaminated by cis+genome-wide mixing and is not used ",
    "for the conclusion — see the flagged table row.)\n\n", sep = "")

cat("**Step 2 — phosphate (co-regulation):**\n\n")
cat("- HMGCR → phosphate: β = ", signif(pho_h, 3), " (p = ", signif(pho_hp, 2), ").\n", sep = "")
sign_ca  <- ifelse(ca_h  > 0, "↑", "↓")
sign_pho <- ifelse(is.na(pho_h), "?", ifelse(pho_hp < 0.05, ifelse(pho_h > 0, "↑", "↓"), "≈0"))
cat("- Pattern: calcium ", sign_ca, ", phosphate ", sign_pho, " → ",
    dplyr::case_when(
      sign_pho == "≈0"            ~ paste0("**calcium-specific route** (albumin artefact already ",
                                           "excluded in Step 1, and phosphate is unmoved, so PTH / ",
                                           "vitamin-D-absorption / bone-resorption are all ruled out — ",
                                           "each of those would shift phosphate). Points to renal / ",
                                           "direct calcium handling."),
      sign_ca == "↑" & sign_pho == "↓" ~ "PTH-like renal phosphate wasting.",
      sign_ca == "↑" & sign_pho == "↑" ~ "intestinal absorption / vitamin-D axis, or bone resorption releasing both ions.",
      TRUE ~ "atypical — inspect directly."),
    "\n\n", sep = "")

cat("**Overall:** the HMGCR→calcium effect is real (not an albumin artefact), does ",
    "not travel through bone density (screened null), phosphate, PTH-pattern, ",
    "vitamin D or absorption — converging on a calcium-specific, plausibly renal, ",
    "mechanism. Direct test of renal handling (urinary calcium / fractional ",
    "excretion) is the deferred next step.\n\n", sep = "")

cat("Caveat: all UK Biobank → one-sample bias toward the null; albumin/phosphate ",
    "share the calcium cohort, which is ideal for the adjustment but means these ",
    "are not independent replications. Phosphate power is limited — its CI does not ",
    "exclude a modest effect, so 'flat' means 'not the strong mover calcium is'.\n", sep = "")
```

---

## Session info

```{r session-info}
#| label: session-info
sessionInfo()
```
