---
title: "Drug-Target MR Analysis — Bone Outcomes"
author: "Dave Bridges"
date: today
editor: source
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
    fig-path: "figures-drug-target-mr-bmd/"
theme: journal
execute:
  echo: true
  warning: false
---

```{r global_options}
#| echo: false
#| message: false

se <- function(x) sd(x, na.rm = TRUE) / sqrt(length(x))

library(tidyverse)
library(knitr)

color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

This script implements a **drug-target Mendelian Randomization (MR)** analysis
testing whether genetically proxied reductions in LDL-cholesterol (LDL-C) and
total cholesterol causally affect bone mineral density (BMD), fracture risk,
and (as a negative control) vitamin D. The three lipid-lowering drug targets
evaluated are **HMGCR** (statins), **PCSK9** (PCSK9 inhibitors), and
**NPC1L1** (ezetimibe).

This analysis is a parallel extension of the previously completed
drug-target MR for serum calcium. The motivating hypothesis is that the
cholesterol → bone demineralization → calcium release pathway is
**mevalonate-pathway-mediated** rather than LDL-C-mediated. If true, this
predicts:

- **HMGCR cis-instruments**: significant negative effect on BMD (effect on
  bone phenotype is direct via mevalonate-pathway intermediates such as
  farnesyl pyrophosphate and geranylgeranyl pyrophosphate)
- **PCSK9 cis-instruments**: null effect on BMD with tight CI
  (PCSK9 inhibition lowers LDL-C via increased LDLR-mediated uptake without
  engaging the mevalonate pathway in non-hepatic tissues)
- **NPC1L1 cis-instruments**: null effect on BMD with wider CI
  (ezetimibe lowers cholesterol via intestinal absorption inhibition; bypasses
  mevalonate pathway entirely)

The pattern of HMGCR-significant but PCSK9-null effects, if observed, would
provide positive evidence that the cholesterol → BMD pathway is mevalonate-
mediated rather than LDL-C-mediated. This parallels the drug-target MR
finding for serum calcium and motivates the cell-specific HMGCR knockout
experiments proposed in Aim 1A2.

### Outcomes

This script evaluates four outcomes:

- **Heel BMD** (Morris et al. 2019, UK Biobank quantitative ultrasound,
  n = 426,824): primary BMD outcome
- **Femoral neck BMD** (Zheng et al. 2015, GEFOS, n = 32,735): secondary
  lower-powered BMD outcome
- **Fractures** (Dönertaş et al. 2021, UK Biobank, n = 484,598):
  clinical endpoint
- **25-hydroxyvitamin D** (MGI-BioVU LabWAS, n = 12,250): negative control

### Sample Overlap Considerations

The LDL-C exposure GWAS (`ieu-b-110`, UKB Neale Lab, n ≈ 440,546) overlaps
with the UK Biobank-derived outcome GWAS for both heel BMD (Morris 2019,
UKB) and fractures (Dönertaş 2021, UKB). This means the analysis is
effectively a *one-sample* MR for those outcome × exposure combinations.
With strong instruments (F >> 10), one-sample bias is generally toward the
null. The total cholesterol exposure (`ebi-a-GCST90025953`, GLGC
meta-analysis) is non-overlapping with UKB, so estimates from that
exposure are immune to sample overlap concerns. Zheng 2015 femoral neck
BMD (GEFOS) and the MGI-BioVU vitamin D GWAS are non-overlapping with
both exposures.

---

## Setup

```{r setup}
#| label: setup
#| message: false

library(tidyverse)
library(TwoSampleMR)
library(ieugwasr)
library(data.table)
library(knitr)
library(kableExtra)

# ── Drug-target gene windows (GRCh37/hg19 ±500 kb around gene body) ──────────
WINDOW_KB <- 500

gene_windows <- tribble(
  ~gene,    ~chr, ~gene_start,  ~gene_end,
  "HMGCR",  5,    74632993,     74657941,
  "PCSK9",  1,    55505221,     55530525,
  "NPC1L1", 7,    44552971,     44604640
) %>%
  mutate(
    region_start = pmax(0, gene_start - WINDOW_KB * 1000),
    region_end   = gene_end + WINDOW_KB * 1000,
    region_str   = str_glue("{chr}:{region_start}-{region_end}")
  )

kable(gene_windows %>% select(gene, chr, region_start, region_end, region_str),
      caption = "Drug-target gene windows (GRCh37, ±500 kb)")

# ── Outcome GWAS IDs ─────────────────────────────────────────────────────────
outcome_gwas <- tribble(
  ~label,                              ~gwas_id,              ~n,       ~year, ~pmid,    ~kind,
  "Heel BMD (Morris 2019)",            "ebi-a-GCST006979",    426824,   2019,  30598549, "ukb",
  "Femoral neck BMD (Zheng 2015)",     "ieu-a-980",           32735,    2015,  26367794, "ukb",
  "Fractures (Donertas 2021)",         "ebi-a-GCST90038703",  484598,   2021,  34187969, "ukb",
  "Vitamin D (MGI-BioVU)",             "local",               12250,    2020,  32907938, "mgi"
)

kable(outcome_gwas %>% select(label, gwas_id, n, year, pmid),
      caption = "Outcome GWAS for drug-target MR")
```

---

## Exposure: Instrument Extraction

Instruments are drawn from two large GWAS:

- **`ieu-b-110`**: LDL-C from UK Biobank (Neale Lab) — overlaps with UKB outcomes
- **`ebi-a-GCST90025953`**: Total cholesterol from the Global Lipids Genetics
  Consortium meta-analysis — non-overlapping with UKB

```{r extract-regional-snps}
#| label: extract-regional-snps
#| cache: true

exposure_ids <- c(
  "LDL-C"             = "ieu-b-110",
  "Total cholesterol" = "ebi-a-GCST90025953"
)

extract_regional <- function(gwas_id, gene_df) {
  map_dfr(seq_len(nrow(gene_df)), function(i) {
    row <- gene_df[i, ]
    cat("  Querying", gwas_id, "—", row$gene, "(", row$region_str, ")\n")
    tryCatch({
      res <- ieugwasr::associations(
        variants = row$region_str,
        id       = gwas_id,
        proxies  = FALSE
      )
      if (nrow(res) == 0) return(tibble())

      res_tib <- res %>% as_tibble()

      if ("pos" %in% names(res_tib) && !"position" %in% names(res_tib)) {
        res_tib <- res_tib %>% dplyr::rename(position = pos)
      } else if (!"position" %in% names(res_tib)) {
        res_tib <- res_tib %>% mutate(position = NA_integer_)
      }

      res_tib %>%
        mutate(
          position = as.integer(position),
          n        = as.character(n),
          beta     = as.numeric(beta),
          se       = as.numeric(se),
          eaf      = as.numeric(eaf),
          p        = as.numeric(p),
          gene     = row$gene,
          gwas_id  = gwas_id
        ) %>%
        select(-any_of("pos"))

    }, error = function(e) {
      warning("Failed for ", gwas_id, " / ", row$gene, ": ", e$message)
      tibble()
    })
  })
}

cat("Extracting regional SNPs from OpenGWAS...\n\n")
regional_raw <- map_dfr(names(exposure_ids), function(name) {
  cat("== Exposure:", name, "(", exposure_ids[name], ") ==\n")
  extract_regional(exposure_ids[name], gene_windows) %>%
    mutate(exposure_name = name)
})

regional_raw %>%
  count(exposure_name, gene, name = "n_snps_raw") %>%
  kable(caption = "Raw SNP counts per gene window (pre-filtering, pre-clumping)")
```

### Clumping and Instrument Finalisation

Following the same approach as the calcium drug-target MR:

- **HMGCR**: allele score at `r²<0.30` (Swerdlow et al. 2015 *Lancet*)
- **PCSK9**: strict clumping at `r²<0.001`
- **NPC1L1**: strict clumping at `r²<0.001`

```{r finalise-instruments}
#| label: finalise-instruments
#| cache: true

clump_gene <- function(df, gene_name, p_thresh, r2_thresh) {
  df %>%
    filter(gene == gene_name, p <= p_thresh) %>%
    group_by(exposure_name, gene) %>%
    group_modify(~ {
      if (nrow(.x) < 2) return(.x)
      tryCatch(
        ieugwasr::ld_clump(
          tibble(rsid = .x$rsid, pval = .x$p, id = .x$gwas_id),
          clump_r2 = r2_thresh,
          clump_kb = 10000,
          pop      = "EUR"
        ) %>% inner_join(.x, by = "rsid"),
        error = function(e) { warning(e$message); .x }
      )
    }) %>%
    ungroup()
}

instruments_hmgcr  <- clump_gene(regional_raw, "HMGCR",  5e-8, 0.30)
instruments_pcsk9  <- clump_gene(regional_raw, "PCSK9",  5e-8, 0.001)
instruments_npc1l1 <- clump_gene(regional_raw, "NPC1L1", 5e-8, 0.001)

instruments_final <- bind_rows(instruments_hmgcr,
                               instruments_pcsk9,
                               instruments_npc1l1)

instruments_final %>%
  mutate(F_stat = (beta / se)^2) %>%
  group_by(exposure_name, gene) %>%
  summarise(
    n_SNPs   = n(),
    min_F    = round(min(F_stat), 1),
    median_F = round(median(F_stat), 1),
    max_F    = round(max(F_stat), 1),
    min_p    = signif(min(p), 2),
    .groups  = "drop"
  ) %>%
  kable(caption = "Final instrument sets across all three drug targets")
```

---

## Outcomes: Fetching from OpenGWAS

For the three OpenGWAS-hosted outcomes (Heel BMD, Femoral neck BMD,
Fractures), we query directly via the API using the instrument rsIDs.
Vitamin D is loaded from a local PheWeb file.

```{r extract-opengwas-outcomes}
#| label: extract-opengwas-outcomes
#| cache: true

all_rsids <- instruments_final %>% distinct(rsid) %>% pull(rsid)

cat("Total unique instrument SNPs to query:", length(all_rsids), "\n\n")

# Outcomes hosted on OpenGWAS
opengwas_outcomes <- outcome_gwas %>% filter(kind == "ukb", gwas_id != "local")

fetch_opengwas_outcome <- function(gwas_id, label) {
  cat("  Fetching", label, "(", gwas_id, ")...\n")
  res <- tryCatch({
    ieugwasr::associations(
      variants = all_rsids,
      id       = gwas_id,
      proxies  = FALSE
    ) %>% as_tibble()
  }, error = function(e) { cat("    Failed:", e$message, "\n"); tibble() })

  if (nrow(res) == 0) return(tibble())

  if ("pos" %in% names(res) && !"position" %in% names(res)) {
    res <- res %>% dplyr::rename(position = pos)
  }

  res %>%
    mutate(
      position = as.integer(position),
      chr      = as.character(chr),
      beta     = as.numeric(beta),
      se       = as.numeric(se),
      eaf      = as.numeric(eaf),
      p        = as.numeric(p),
      outcome_label = label,
      outcome_gwas_id = gwas_id
    )
}

opengwas_outcome_raw <- pmap_dfr(
  list(opengwas_outcomes$gwas_id, opengwas_outcomes$label),
  fetch_opengwas_outcome
)

# Coverage check
opengwas_outcome_raw %>%
  count(outcome_label, name = "n_snps_returned") %>%
  mutate(n_instruments = length(all_rsids),
         pct_recovered = round(100 * n_snps_returned / n_instruments, 1)) %>%
  kable(caption = "OpenGWAS outcome SNP recovery (direct rsID match)")
```

### Coverage by Gene

```{r opengwas-coverage-by-gene}
#| label: opengwas-coverage-by-gene

opengwas_coverage <- instruments_final %>%
  distinct(rsid, exposure_name, gene) %>%
  left_join(
    opengwas_outcome_raw %>% distinct(rsid, outcome_label) %>% mutate(found = TRUE),
    by = "rsid",
    relationship = "many-to-many"
  ) %>%
  filter(!is.na(outcome_label)) %>%
  count(outcome_label, exposure_name, gene, name = "n_in_outcome")

# Compare to instrument counts
instruments_final %>%
  count(exposure_name, gene, name = "n_instruments") %>%
  left_join(opengwas_coverage,
            by = c("exposure_name", "gene"),
            relationship = "many-to-many") %>%
  pivot_wider(names_from = outcome_label,
              values_from = n_in_outcome,
              values_fill = 0L) %>%
  kable(caption = "OpenGWAS outcome coverage by drug target and exposure")
```

### Standardising OpenGWAS Outcomes

```{r format-opengwas-outcomes}
#| label: format-opengwas-outcomes

opengwas_outcome_std <- opengwas_outcome_raw %>%
  dplyr::rename(
    REF      = nea,
    ALT      = ea,
    pval     = p,
    eaf_out  = eaf
  ) %>%
  mutate(
    CHR = as.integer(chr),
    POS = as.integer(position)
  ) %>%
  filter(!is.na(CHR), !is.na(POS)) %>%
  select(rsid, CHR, POS, REF, ALT,
         beta_out = beta, se_out = se, pval_out = pval, eaf_out,
         outcome_label, outcome_gwas_id)

cat("OpenGWAS outcomes standardised:",
    scales::comma(nrow(opengwas_outcome_std)),
    "SNP-outcome pairs ready for harmonisation\n")
```

---

## Outcome: Vitamin D (MGI-BioVU LabWAS, Negative Control)

The Vitamin D analysis acts as a negative control. We do NOT expect any of
the drug-target instruments to affect vitamin D, since the genome-wide MR
already rejected cholesterol → vitamin D as a mediator of the
cholesterol → calcium effect.

```{r load-vitd-outcome}
#| label: load-vitd-outcome
#| cache: true

cat("Loading MGI-BioVU vitamin D GWAS...\n")

vitd_gwas_raw <- read_tsv(
  gzfile("PheWeb Summary Statistics/phenocode-Vit-D.tsv.gz"),
  show_col_types = FALSE
)

cat("Columns:\n"); print(names(vitd_gwas_raw))

vitd_gwas <- vitd_gwas_raw %>%
  as_tibble() %>%
  dplyr::rename(
    CHR  = chrom,
    POS  = pos,
    REF  = ref,
    ALT  = alt,
    pval = pval,
    beta = beta,
    se   = sebeta,
    eaf  = maf
  ) %>%
  mutate(CHR = as.integer(str_remove(as.character(CHR), "^chr")),
         POS = as.integer(POS)) %>%
  select(CHR, POS, REF, ALT,
         beta_vitd = beta, se_vitd = se, pval_vitd = pval, eaf_vitd = eaf)

cat("\nMGI-BioVU vitamin D GWAS:",
    scales::comma(nrow(vitd_gwas)), "SNPs\n")
```

---

## Outcome Extraction and Harmonisation

We use two helpers that handle the OpenGWAS (rsID-keyed) and MGI-BioVU
(position-keyed) outcomes. The MGI helper includes a ±50 bp positional
window fallback to recover SNPs not exactly matched.

```{r extract-outcome-helpers}
#| label: extract-outcome-helpers

# OpenGWAS outcomes: matched by rsID across multiple outcomes
extract_outcome_opengwas <- function(instruments_df, opengwas_std,
                                     outcome_label_filter) {
  instruments_df %>%
    distinct(rsid, chr, position, ea, nea, eaf, gene, exposure_name,
             beta_exp = beta, se_exp = se) %>%
    inner_join(
      opengwas_std %>% filter(outcome_label == outcome_label_filter),
      by = "rsid"
    )
}

# MGI-BioVU vitamin D outcome: matched by CHR + POS with ±50 bp fallback
extract_outcome_vitd <- function(instruments_df, vitd_df, window_bp = 50) {
  instruments_df %>%
    distinct(rsid, chr, position, ea, nea, eaf, gene, exposure_name,
             beta_exp = beta, se_exp = se) %>%
    mutate(CHR = as.integer(chr), POS = as.integer(position)) %>%
    filter(!is.na(POS)) %>%
    left_join(
      vitd_df %>%
        filter(!is.na(CHR), !is.na(POS)) %>%
        dplyr::rename(beta_out = beta_vitd,
                      se_out   = se_vitd,
                      pval_out = pval_vitd,
                      eaf_out  = eaf_vitd),
      by = c("CHR", "POS")
    ) %>%
    {
      matched   <- filter(., !is.na(beta_out))
      unmatched <- filter(., is.na(beta_out))

      if (nrow(unmatched) > 0 && window_bp > 0) {
        fallback <- map_dfr(seq_len(nrow(unmatched)), function(i) {
          row <- unmatched[i, ]
          hit <- vitd_df %>%
            filter(!is.na(CHR), !is.na(POS),
                   CHR == row$CHR,
                   POS >= row$POS - window_bp,
                   POS <= row$POS + window_bp) %>%
            slice_min(abs(POS - row$POS), n = 1, with_ties = FALSE)

          if (nrow(hit) == 0) return(tibble())

          tibble(
            rsid          = row$rsid,
            chr           = row$chr,
            position      = row$position,
            ea            = row$ea,
            nea           = row$nea,
            eaf           = row$eaf,
            gene          = row$gene,
            exposure_name = row$exposure_name,
            beta_exp      = row$beta_exp,
            se_exp        = row$se_exp,
            CHR           = row$CHR,
            POS           = row$POS,
            REF           = hit$REF,
            ALT           = hit$ALT,
            beta_out      = hit$beta_vitd,
            se_out        = hit$se_vitd,
            pval_out      = hit$pval_vitd,
            eaf_out       = hit$eaf_vitd
          )
        })
        bind_rows(matched, fallback)
      } else {
        matched
      }
    } %>%
    mutate(outcome_label = "Vitamin D (MGI-BioVU)")
}
```

```{r harmonise-helper}
#| label: harmonise-helper

# Strict allele-pair harmonisation — both alleles must match (or both
# strand-flipped).
harmonise_alleles <- function(df) {
  df %>%
    filter(!is.na(beta_out)) %>%
    mutate(
      ea_comp  = chartr("ACGT", "TGCA", ea),
      nea_comp = chartr("ACGT", "TGCA", nea),

      direct_match    = (ea  == ALT & nea  == REF),
      flipped_match   = (ea  == REF & nea  == ALT),
      comp_match      = (ea_comp == ALT & nea_comp == REF),
      comp_flip_match = (ea_comp == REF & nea_comp == ALT),

      ambiguous = (
        (ea == "A" & nea == "T") | (ea == "T" & nea == "A") |
        (ea == "C" & nea == "G") | (ea == "G" & nea == "C")
      ),
      eaf_incompatible = ambiguous & abs(eaf - eaf_out) > 0.3,

      needs_flip = case_when(
        direct_match    ~ FALSE,
        flipped_match   ~ TRUE,
        comp_match      ~ FALSE,
        comp_flip_match ~ TRUE,
        TRUE            ~ NA
      ),
      beta_out_h = case_when(
        is.na(needs_flip) | eaf_incompatible ~ NA_real_,
        needs_flip                           ~ -beta_out,
        TRUE                                 ~ beta_out
      ),
      se_out_h = if_else(is.na(beta_out_h), NA_real_, se_out),
      harmonise_status = case_when(
        is.na(needs_flip)  ~ "incompatible_alleles",
        eaf_incompatible   ~ "ambiguous_eaf_mismatch",
        needs_flip         ~ "flipped",
        TRUE               ~ "direct"
      )
    )
}

build_harmonised <- function(instruments_df, outcome_df, outcome_label,
                              outcome_kind = c("opengwas", "vitd")) {
  outcome_kind <- match.arg(outcome_kind)

  hits <- if (outcome_kind == "opengwas") {
    extract_outcome_opengwas(instruments_df, outcome_df, outcome_label)
  } else {
    extract_outcome_vitd(instruments_df, outcome_df, window_bp = 50)
  }

  cat("\n=== Harmonising for", outcome_label, "===\n")
  cat("Instruments matched in outcome:", nrow(hits), "\n")

  harm <- harmonise_alleles(hits)

  cat("\nHarmonisation status:\n")
  harm %>% count(gene, harmonise_status) %>% print()

  harm_clean <- harm %>%
    filter(!is.na(beta_out_h)) %>%
    mutate(
      beta_exp   = as.numeric(beta_exp),
      se_exp     = as.numeric(se_exp),
      beta_out_h = as.numeric(beta_out_h),
      se_out_h   = as.numeric(se_out_h),
      F_stat     = (beta_exp / se_exp)^2,
      outcome    = outcome_label
    )

  cat("\nSNPs retained per exposure × gene:\n")
  harm_clean %>% count(exposure_name, gene) %>% print()

  harm_clean
}
```

### Run Harmonisation for All Four Outcomes

```{r harmonise-all-outcomes}
#| label: harmonise-all-outcomes

harm_heelbmd <- build_harmonised(instruments_final, opengwas_outcome_std,
                                  outcome_label = "Heel BMD (Morris 2019)",
                                  outcome_kind  = "opengwas")

harm_fembmd  <- build_harmonised(instruments_final, opengwas_outcome_std,
                                  outcome_label = "Femoral neck BMD (Zheng 2015)",
                                  outcome_kind  = "opengwas")

harm_frac    <- build_harmonised(instruments_final, opengwas_outcome_std,
                                  outcome_label = "Fractures (Donertas 2021)",
                                  outcome_kind  = "opengwas")

harm_vitd    <- build_harmonised(instruments_final, vitd_gwas,
                                  outcome_label = "Vitamin D (MGI-BioVU)",
                                  outcome_kind  = "vitd")

harm_combined <- bind_rows(harm_heelbmd, harm_fembmd, harm_frac, harm_vitd)
```

### Coverage Comparison Across Outcomes

```{r coverage-comparison}
#| label: coverage-comparison

harm_combined %>%
  count(outcome, exposure_name, gene, name = "n_harmonised") %>%
  pivot_wider(names_from = outcome, values_from = n_harmonised, values_fill = 0L) %>%
  kable(caption = "Harmonised SNP counts per drug target across outcomes")
```

---

## MR Analysis

```{r mr-engine}
#| label: mr-engine

run_drug_target_mr <- function(df, exposure_col = "exposure_name") {

  ivw_re <- function(beta_x, se_x, beta_y, se_y) {
    ratio    <- beta_y / beta_x
    ratio_se <- abs(se_y / beta_x)
    w        <- 1 / ratio_se^2
    theta    <- sum(w * ratio) / sum(w)
    Q        <- sum(w * (ratio - theta)^2)
    df_Q     <- length(ratio) - 1
    Q_p      <- pchisq(Q, df = df_Q, lower.tail = FALSE)
    tau2     <- max(0, (Q - df_Q) / (sum(w) - sum(w^2) / sum(w)))
    w_re     <- 1 / (ratio_se^2 + tau2)
    theta_re <- sum(w_re * ratio) / sum(w_re)
    se_re    <- sqrt(1 / sum(w_re))
    list(estimate = theta_re, se = se_re,
         Q = Q, Q_df = df_Q, Q_p = Q_p, tau2 = tau2)
  }

  weighted_median_mr <- function(beta_x, se_x, beta_y, se_y, nboot = 1000) {
    ratio    <- beta_y / beta_x
    ratio_se <- abs(se_y / beta_x)
    w        <- 1 / ratio_se^2
    w        <- w / sum(w)
    ord      <- order(ratio)
    wm       <- ratio[ord][which(cumsum(w[ord]) >= 0.5)[1]]
    boot_est <- replicate(nboot, {
      r_b <- rnorm(length(ratio), ratio, ratio_se)
      r_b[order(r_b)][which(cumsum(sample(w)) >= 0.5)[1]]
    })
    list(estimate = wm, se = sd(boot_est))
  }

  egger_mr <- function(beta_x, beta_y, se_y) {
    flip <- beta_x < 0
    bx   <- ifelse(flip, -beta_x, beta_x)
    by   <- ifelse(flip, -beta_y,  beta_y)
    w    <- 1 / se_y^2
    fit  <- lm(by ~ bx, weights = w)
    s    <- summary(fit)
    list(
      estimate  = coef(fit)[["bx"]],
      se        = coef(s)[["bx", "Std. Error"]],
      intercept = coef(fit)[["(Intercept)"]],
      int_p     = coef(s)[["(Intercept)", "Pr(>|t|)"]]
    )
  }

  df %>%
    group_by(across(all_of(c("outcome", exposure_col, "gene")))) %>%
    group_map(~ {
      d      <- .x; key <- .y
      n_snps <- nrow(d)
      bx <- d$beta_exp; sx <- d$se_exp
      by <- d$beta_out_h; sy <- d$se_out_h

      rows <- list()

      if (n_snps == 1) {
        wr_est <- by / bx; wr_se <- abs(sy / bx)
        rows[["Wald ratio"]] <- tibble(
          method = "Wald ratio",
          estimate = wr_est, se = wr_se,
          ci_lo = wr_est - 1.96 * wr_se, ci_hi = wr_est + 1.96 * wr_se,
          p_value = 2 * pnorm(-abs(wr_est / wr_se)),
          n_snps = 1L, Q = NA_real_, Q_p = NA_real_, egger_int_p = NA_real_
        )
      } else {
        ivw <- ivw_re(bx, sx, by, sy)
        rows[["IVW-RE"]] <- tibble(
          method   = "IVW-RE",
          estimate = ivw$estimate, se = ivw$se,
          ci_lo    = ivw$estimate - 1.96 * ivw$se,
          ci_hi    = ivw$estimate + 1.96 * ivw$se,
          p_value  = 2 * pnorm(-abs(ivw$estimate / ivw$se)),
          n_snps   = n_snps, Q = round(ivw$Q, 2),
          Q_p      = round(ivw$Q_p, 4), egger_int_p = NA_real_
        )
        wm <- weighted_median_mr(bx, sx, by, sy)
        rows[["Weighted median"]] <- tibble(
          method   = "Weighted median",
          estimate = wm$estimate, se = wm$se,
          ci_lo    = wm$estimate - 1.96 * wm$se,
          ci_hi    = wm$estimate + 1.96 * wm$se,
          p_value  = 2 * pnorm(-abs(wm$estimate / wm$se)),
          n_snps   = n_snps, Q = NA_real_, Q_p = NA_real_, egger_int_p = NA_real_
        )
        if (n_snps >= 3) {
          eg <- egger_mr(bx, by, sy)
          rows[["MR-Egger"]] <- tibble(
            method      = "MR-Egger",
            estimate    = eg$estimate, se = eg$se,
            ci_lo       = eg$estimate - 1.96 * eg$se,
            ci_hi       = eg$estimate + 1.96 * eg$se,
            p_value     = 2 * pnorm(-abs(eg$estimate / eg$se)),
            n_snps      = n_snps, Q = NA_real_, Q_p = NA_real_,
            egger_int_p = round(eg$int_p, 4)
          )
        }
      }
      bind_rows(rows) %>%
        mutate(outcome  = key[["outcome"]],
               exposure = key[[exposure_col]],
               gene     = key[["gene"]])
    }, .keep = TRUE) %>%
    bind_rows()
}
```

### Running Across All Outcomes

```{r run-mr-all}
#| label: run-mr-all

set.seed(42)

mr_results <- run_drug_target_mr(harm_combined)

mr_results %>%
  mutate(
    across(c(estimate, se, ci_lo, ci_hi), ~round(.x, 4)),
    p_value  = format.pval(p_value, digits = 3, eps = 0.001),
    `95% CI` = str_glue("({ci_lo}\u2013{ci_hi})")
  ) %>%
  select(outcome, exposure, gene, method, n_snps,
         estimate, `95% CI`, p_value, Q, Q_p, egger_int_p) %>%
  arrange(exposure, gene, outcome, method) %>%
  kable(caption = paste0(
    "Drug-target MR — all outcomes; estimates are SD outcome ",
    "(or log-OR for fracture) per SD cholesterol increase"
  ))
```

### Primary Estimates Only — Summary Table

```{r primary-summary}
#| label: primary-summary

mr_results %>%
  filter(method == "IVW-RE" | method == "Wald ratio") %>%
  mutate(
    across(c(estimate, ci_lo, ci_hi), ~round(.x, 4)),
    p_value  = format.pval(p_value, digits = 3, eps = 0.001),
    `95% CI` = str_glue("({ci_lo}\u2013{ci_hi})")
  ) %>%
  select(outcome, exposure, gene, n_snps, method,
         estimate, `95% CI`, p_value) %>%
  arrange(exposure, gene, outcome) %>%
  kable(caption = "Primary MR estimates (IVW-RE or Wald ratio) across all outcomes")
```

### Heel BMD — Primary Focus

```{r heelbmd-primary}
#| label: heelbmd-primary

mr_results %>%
  filter(outcome == "Heel BMD (Morris 2019)") %>%
  mutate(
    across(c(estimate, ci_lo, ci_hi), ~round(.x, 4)),
    p_value  = format.pval(p_value, digits = 3, eps = 0.001),
    `95% CI` = str_glue("({ci_lo}\u2013{ci_hi})")
  ) %>%
  select(exposure, gene, method, n_snps,
         estimate, `95% CI`, p_value, Q, Q_p, egger_int_p) %>%
  arrange(exposure, gene, method) %>%
  kable(caption = "Heel BMD — all methods across drug targets and exposures")
```

### Leave-One-Out Sensitivity (HMGCR for Heel BMD)

The HMGCR allele score uses SNPs at `r²<0.30` so leave-one-out is the
appropriate sensitivity check for whether any single SNP drives the result.

```{r leave-one-out-hmgcr}
#| label: leave-one-out-hmgcr

loo_mr <- function(df, outcome_label, exposure_label, gene_label) {
  d <- df %>%
    filter(outcome == outcome_label,
           exposure_name == exposure_label,
           gene == gene_label)

  if (nrow(d) < 2) return(tibble())

  map_dfr(seq_len(nrow(d)), function(i) {
    d_loo <- d[-i, ]
    bx <- d_loo$beta_exp; sx <- d_loo$se_exp
    by <- d_loo$beta_out_h; sy <- d_loo$se_out_h
    ratio    <- by / bx
    ratio_se <- abs(sy / bx)
    w        <- 1 / ratio_se^2
    theta    <- sum(w * ratio) / sum(w)
    se_fe    <- sqrt(1 / sum(w))

    tibble(
      outcome     = outcome_label,
      exposure    = exposure_label,
      dropped_snp = d$rsid[i],
      estimate    = round(theta, 4),
      ci_lo       = round(theta - 1.96 * se_fe, 4),
      ci_hi       = round(theta + 1.96 * se_fe, 4),
      p_value     = round(2 * pnorm(-abs(theta / se_fe)), 4)
    )
  })
}

loo_outcomes <- c("Heel BMD (Morris 2019)",
                  "Femoral neck BMD (Zheng 2015)",
                  "Fractures (Donertas 2021)",
                  "Vitamin D (MGI-BioVU)")

loo_results <- map_dfr(loo_outcomes, function(o) {
  bind_rows(
    loo_mr(harm_combined, o, "LDL-C",             "HMGCR"),
    loo_mr(harm_combined, o, "Total cholesterol", "HMGCR")
  )
})

loo_results %>%
  group_by(outcome, exposure) %>%
  summarise(
    n_iterations = n(),
    range_lo     = round(min(estimate), 4),
    range_hi     = round(max(estimate), 4),
    `% iterations p<0.05` = round(100 * mean(p_value < 0.05), 0),
    .groups = "drop"
  ) %>%
  kable(caption = paste0(
    "HMGCR leave-one-out IVW summary across outcomes — ",
    "robustness across instruments"
  ))
```

---

## Visualisation

### Forest Plot — All Outcomes × All Targets

```{r forest-plot-all}
#| label: forest-plot-all
#| fig-width: 12
#| fig-height: 9

mr_results %>%
  filter(method %in% c("IVW-RE", "Wald ratio")) %>%
  mutate(
    label   = str_glue("{gene}\n({exposure})"),
    outcome = factor(outcome, levels = c(
      "Heel BMD (Morris 2019)",
      "Femoral neck BMD (Zheng 2015)",
      "Fractures (Donertas 2021)",
      "Vitamin D (MGI-BioVU)"
    ))
  ) %>%
  ggplot(aes(x = estimate, xmin = ci_lo, xmax = ci_hi,
             y = label, colour = outcome, shape = outcome)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(height = 0.3,
                 position = position_dodge(width = 0.7)) +
  geom_point(size = 3,
             position = position_dodge(width = 0.7)) +
  scale_colour_manual(values = c(
    "Heel BMD (Morris 2019)"        = "#00274c",
    "Femoral neck BMD (Zheng 2015)" = "#5e92a8",
    "Fractures (Donertas 2021)"     = "#ffcb05",
    "Vitamin D (MGI-BioVU)"         = "#a0a0a0"
  )) +
  scale_shape_manual(values = c(
    "Heel BMD (Morris 2019)"        = 16,
    "Femoral neck BMD (Zheng 2015)" = 17,
    "Fractures (Donertas 2021)"     = 15,
    "Vitamin D (MGI-BioVU)"         = 18
  )) +
  facet_wrap(~exposure, ncol = 1, scales = "free_y") +
  labs(
    title    = "Drug-target MR — bone and vitamin D outcomes",
    subtitle = "IVW-RE (Wald ratio for single-SNP cases)",
    x        = "Change in outcome (SD, or log-OR for fracture) per SD cholesterol increase",
    y        = NULL,
    colour   = "Outcome",
    shape    = "Outcome"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position  = "bottom",
        panel.grid.minor = element_blank())
```

### Forest Plot — Heel BMD Only (Primary)

```{r forest-plot-heelbmd}
#| label: forest-plot-heelbmd
#| fig-width: 10
#| fig-height: 5

mr_results %>%
  filter(outcome == "Heel BMD (Morris 2019)",
         method %in% c("IVW-RE", "Wald ratio")) %>%
  mutate(label = str_glue("{gene}\n({exposure})")) %>%
  ggplot(aes(x = estimate, xmin = ci_lo, xmax = ci_hi,
             y = label)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(height = 0.3, colour = color_scheme[1]) +
  geom_point(size = 3, colour = color_scheme[1]) +
  labs(
    title    = "Drug-target MR — Heel BMD (Morris 2019, n = 426,824)",
    subtitle = "IVW-RE (Wald ratio for single-SNP cases); 95% CI",
    x        = "Change in Heel BMD (SD) per SD cholesterol increase",
    y        = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

### HMGCR Scatter — Heel BMD

```{r hmgcr-scatter-heelbmd}
#| label: hmgcr-scatter-heelbmd
#| fig-width: 10
#| fig-height: 5

harm_combined %>%
  filter(outcome == "Heel BMD (Morris 2019)", gene == "HMGCR") %>%
  ggplot(aes(x = beta_exp, y = beta_out_h)) +
  geom_point(aes(size = 1 / se_out_h^2), shape = 21,
             fill = "#00274c", alpha = 0.7) +
  geom_smooth(aes(weight = 1 / se_out_h^2),
              method = "lm", se = TRUE,
              colour = "#ffcb05", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  facet_wrap(~exposure_name, scales = "free_x") +
  scale_size_continuous(guide = "none") +
  labs(
    title    = "HMGCR allele score — MR scatter (Heel BMD outcome)",
    subtitle = "Point size proportional to inverse variance; slope = IVW estimate",
    x        = "SNP effect on cholesterol exposure",
    y        = "SNP effect on Heel BMD",
    caption  = "Shaded band = 95% CI of weighted regression"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

---

## Combined Results Export

```{r write-results}
#| label: write-results

mr_results_out <- mr_results %>%
  mutate(
    across(c(estimate, se, ci_lo, ci_hi), ~round(.x, 5)),
    p_value = signif(p_value, 3)
  ) %>%
  select(outcome, exposure, gene, method, n_snps,
         estimate, se, ci_lo, ci_hi, p_value,
         Q, Q_p, egger_int_p)

mr_results_out %>%
  write_csv("MR Results - Drug Target BMD.csv")

cat("Wrote", nrow(mr_results_out),
    "rows to 'MR Results - Drug Target BMD.csv'\n")
```

---

## Session Information

```{r session-information}
sessionInfo()
```
