---
title: "Calcium and LDLC Clump Analysis from UK Biobank"
author: "Dave Bridges"
date: "2025-06-18"
editor: source
format: 
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
    fig-path: "figures-clumps/"
theme: journal
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)
library(knitr)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

This script analyses the clumped data for the calcium and LDL-C summary statistics from UK Biobank.  The clumps were 

## Data Entry

```{r datafiles}
calcium.clumps.datafile <- 'biomarkers-30680-both_sexes-irnt-results.clumps'
ldlc.clumps.datafile <- 'biomarkers-30780-both_sexes-irnt-results.clumps'
tc.clumps.datafile <- 'biomarkers-30690-both_sexes-irnt-results.clumps'

calcium.freq.file <- 'biomarkers-30680-all-snps.afreq'
ldlc.freq.file <- 'biomarkers-30780-all-snps.afreq'
tc.freq.file <- 'biomarkers-30690-all-snps.afreq'

calcium.sumstats.file <- 'biomarkers-30680-both_sexes-irnt.tsv'
ldlc.sumstats.file <- 'biomarkers-30780-both_sexes-irnt.tsv'
tc.sumstats.file <- 'biomarkers-30690-both_sexes-irnt.tsv'

calcium.outcome.gwas.file <- 'PheWeb Summary Statistics/phenocode-Ca.tsv.gz'
```

The calcium summary stats is in `r calcium.sumstats.file`, the total cholesterol in `r tc.sumstats.file` and the LDL-C summary stats file is in `r ldlc.sumstats.file`.  The clumps were generated using the following plink2 options:

```
# or 30780 for LDL-C 
plink2 \
  --pfile 1000G_EUR_merged \
  --clump biomarkers-30680-both_sexes-irnt.tsv \ 
  --clump-snp-field VARIANT \
  --clump-field P \
  --clump-p1 5e-8 \
  --clump-p2 1e-6 \
  --clump-r2 0.01 \
  --clump-kb 10000 \
  --threads 4 \
  --out biomarkers-30680-both_sexes-irnt-results 
```
  
Frequencies we then calculated from these clumped results and are present in the `r calcium.freq.file`, `r tc.freq.file` and `r ldlc.freq.file`
  

## Calcium SNPs

```{r calcium-snps}
# Read clumped SNPs and expand out clumped SNPs
calcium.clumps <- read_tsv(calcium.clumps.datafile, col_types = cols()) |>
  rename(P_clumping = P) 

# Expanded out clumps from the SP2 column
calcium.clumps.expanded <- calcium.clumps %>%
  separate_rows(SP2, sep = ",")

# Read frequency file (from PLINK --freq output)
calcium.freqs <- read_tsv(calcium.freq.file, col_types = cols())

# Read full summary stats (with BETA)
calcium.sumstats <- read_tsv(calcium.sumstats.file, col_types = cols())

# Read in the outcome GWAS summary statitiscis
calcium.outcome <- read_tsv(calcium.outcome.gwas.file, col_types = cols()) |>
  mutate(ID=paste(chrom, pos, ref,alt, sep=":"))

# Found SNPs that matched between clumped SNPs and outcome GWAS
calcium.matched.snps <-
  inner_join(calcium.clumps.expanded, calcium.outcome,
          by = c("SP2"="ID")) |>
  distinct(ID,.keep_all=TRUE)
```

## Matching Exposure Clumped SNPs to Outcome SNPs

There were only`r length(intersect(calcium.clumps$ID,calcium.outcome$ID))` SNPs in common between the calcium clumped SNPs and the calcium outcome GWAS summary statistics.  Looked up LD SNPs from each lead SNP and checked how many of those intersected and found there are now `r length(intersect(calcium.clumps.expanded$SP2,calcium.outcome$ID))` SNPS in common between the datasets (not accounting for clumping).  After comparing to the outcome GWAS there were `r nrow(calcium.matched.snps)` SNPs that are present in both the clumped SNPs and the outcome GWAS.

```{r calcium-instruments}

# Merge clumps with frequency (to get MAF for each SNP)
calcium.clumps_freq <- 
  calcium.matched.snps %>%
  select(-`#CHROM`) |>
  left_join(calcium.freqs, by = c("ID"="ID")) #missing MAF for most SNPs

# Merge with summary stats (to get beta)
calcium.clumps_freq_beta <- 
  calcium.clumps_freq %>%
  inner_join(calcium.sumstats, 
             by = c("ID"="VARIANT"))

# Filter by MAF >= 0.01
calcium.instruments <- calcium.clumps_freq_beta %>%
  filter(ALT_FREQS >= 0.01, ALT_FREQS <= 0.99)  # also exclude rare alleles > 0.99
```

We then filtered out `r nrow(calcium.clumps_freq_beta) - nrow(calcium.instruments)` SNPs with MAF < 0.01 or > 0.99, leaving `r nrow(calcium.instruments)` SNPs as instruments for calcium.

```{r calcium-instrument-strength}
n.calcium <- 385066  # replace with your actual exposure GWAS sample size

calcium.instruments <- calcium.instruments %>%
  mutate(
    R2 = 2 * ALT_FREQS * (1 - ALT_FREQS) * BETA^2,
    F = (R2 * (n.calcium - 2)) / (1 - R2)
  )

# Calculate summary metrics
calcium.summary_metrics <- 
  calcium.instruments %>%
  summarise(
    num_snps = n(),
    cumulative_R2 = sum(R2, na.rm = TRUE),
    mean_F = mean(F, na.rm=TRUE),
    median_F = median(F, na.rm=TRUE),
    mean_maf = mean(ALT_FREQS, na.rm = TRUE),
    mean_beta = mean(abs(BETA), na.rm = TRUE),
    overall_F = (cumulative_R2 * (n.calcium - num_snps - 1)) / ((1 - cumulative_R2) * num_snps)
  )

kable(calcium.summary_metrics, caption="Summary of calcium instruments")

calcium.instruments |> 
  rename(EA = alt, ,
         OA = ref,
         CHR = chrom) |>
  mutate(N_exposure=n.calcium) |>
  select(SP2,CHR,POS, EA,OA,BETA,SE,P,ALT_FREQS,N_exposure,R2, `F`) |>
  write_csv("Calcium Instruments from UKBB.csv")
```

For each SNP calculated the $R^2$ and the $F$ statistic using these formulas:

$$
R^2 = 2 \times \text{MAF} \times (1 - \text{MAF}) \times \beta^2
$$

$$
F = \frac{R^2 \times (N - k - 1)}{(1 - R^2) \times k}
$$
Where $MAF$ is the allele frequency from 1000 Genomes, EUR subset, $\beta$ is the effect size of the allele from UK Biobank, $N$ is the sample size of that trait in UK BioBank, and $k$ is the number of instruments (SNPs).  The $F$ statistic is used to assess instrument strength, with values above 10 indicating strong instruments.


## Session Information

```{r session-information}
sessionInfo()
```