---
title: "Calcium and LDLC Clump Analysis from UK Biobank"
author: "Dave Bridges"
date: "2025-06-18"
editor: source
format: 
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
    fig-path: "figures-clumps/"
theme: journal
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)
library(knitr)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

This script analyses the clumped data for the calcium and LDL-C summary statistics from UK Biobank.  The clumps were 

## Data Entry

```{r datafiles}
calcium.clumps.datafile <- 'biomarkers-30680-both_sexes-irnt-results.clumps'
ldlc.clumps.datafile <- 'biomarkers-30780-both_sexes-irnt-results.clumps'

calcium.freq.file <- 'biomarkers-30680-both_sexes-irnt-results-clumped-snps.afreq'
ldlc.freq.file <- 'biomarkers-30780-both_sexes-irnt-results-clumped-snps.afreq'

calcium.sumstats.file <- 'biomarkers-30680-both_sexes-irnt.tsv'
ldlc.sumstats.file <- 'biomarkers-30780-both_sexes-irnt.tsv'
```

The calcium summary stats is in `r calcium.sumstats.file` and the LDL-C summary stats file is in `r ldlc.sumstats.file`.  The clumps were generated using the following plink2 options:

```
# or 30780 for LDL-C 
plink2 \
  --pfile 1000G_EUR_merged \
  --clump biomarkers-30680-both_sexes-irnt.tsv \ 
  --clump-snp-field VARIANT \
  --clump-field P \
  --clump-p1 5e-8 \
  --clump-p2 1e-6 \
  --clump-r2 0.01 \
  --clump-kb 10000 \
  --threads 4 \
  --out biomarkers-30680-both_sexes-irnt-results 
```
  
Frequencies we then calculated from these clumped results and are present in the `r calcium.freq.file` and `r ldlc.freq.file`
  

## Calcium SNPs

```{r calcium-snps}
# Read clumped SNPs and expand out clumped SNPs
calcium.clumps <- read_tsv(calcium.clumps.datafile, col_types = cols()) |>
  rename(P_clumping = P) 
  
# Read frequency file (from PLINK --freq output)
calcium.freqs <- read_tsv(calcium.freq.file, col_types = cols())

# Read full summary stats (with BETA)
calcium.sumstats <- read_tsv(calcium.sumstats.file, col_types = cols())

# Inspect the column names to confirm merging keys and data availability
# head(calcium.clumps)
# head(calcium.freqs)
# head(calcium.sumstats)

# The clump file 'SNP' column should match frequency 'ID' column
# The sumstats 'VARIANT' column should match clumps 'SNP' column

# Merge clumps with frequency (to get MAF for each SNP)
calcium.clumps_freq <- 
  calcium.clumps %>%
  select(-`#CHROM`) |>
  left_join(calcium.freqs, by = c("ID"="ID")) #missing MAF for most SNPs

# Merge with summary stats (to get beta)
calcium.clumps_freq_beta <- 
  calcium.clumps_freq %>%
  inner_join(calcium.sumstats, 
             by = c("ID"="VARIANT"))

# Filter by MAF >= 0.01
calcium.instruments <- calcium.clumps_freq_beta %>%
  filter(ALT_FREQS >= 0.01, ALT_FREQS <= 0.99)  # also exclude rare alleles > 0.99

n.calcium <- 431888  # replace with your actual exposure GWAS sample size

calcium.instruments <- calcium.instruments %>%
  mutate(
    R2 = 2 * ALT_FREQS * (1 - ALT_FREQS) * BETA^2,
    F = (R2 * (n.calcium - 2)) / (1 - R2)
  )

# Calculate summary metrics
calcium.summary_metrics <- 
  calcium.instruments %>%
  summarise(
    num_snps = n(),
    cumulative_R2 = sum(R2, na.rm = TRUE),
    mean_F = mean(F, na.rm=TRUE),
    median_F = median(F, na.rm=TRUE),
    mean_maf = mean(ALT_FREQS, na.rm = TRUE),
    mean_beta = mean(abs(BETA), na.rm = TRUE),
    overall_F = (cumulative_R2 * (n.calcium - num_snps - 1)) / ((1 - cumulative_R2) * num_snps)
  )

kable(calcium.summary_metrics, caption="Summary of calcium instruments")

calcium.instruments |> 
  separate(ID, into = c("CHR", "POS", "REF", "ALT"), sep = ":", remove = FALSE) %>%
  # choose which is EA/OA – here I assume ALT is the effect allele
  rename(EA = ALT, OA = REF) |>
  mutate(N_exposure=n.calcium) |>
  select(ID,CHR,POS, EA,OA,BETA,SE,P,ALT_FREQS,N_exposure,R2, `F`) |>
  write_csv("Calcium Instruments from UKBB.csv")
```

For each SNP calculated the $R^2$ and the $F$ statistic using these formulas:

$$
R^2 = 2 \times \text{MAF} \times (1 - \text{MAF}) \times \beta^2
$$

$$
F = \frac{R^2 \times (N - k - 1)}{(1 - R^2) \times k}
$$
Where $MAF$ is the allele frequency from 1000 Genomes, EUR subset, $\beta$ is the effect size of the allele from UK Biobank, $N$ is the sample size of that trait in UK BioBank, and $k$ is the number of instruments (SNPs).  The $F$ statistic is used to assess instrument strength, with values above 10 indicating strong instruments.

## LDL-C SNPs

```{r ldlc-snps}
# Read clumped SNPs and expand out clumped SNPs
ldlc.clumps <- read_tsv(ldlc.clumps.datafile, col_types = cols())|>
  rename(P_clumping = P) |> #to separate from sumstats p-value
  mutate(SP2 = ifelse(is.na(SP2), "", SP2)) %>%     # replace NA with empty string
  separate_rows(SP2, sep = ",") %>%                 # one SNP per row
  mutate(SNP = ifelse(SP2 == "", ID, SP2)) %>%     # use SP2 if present, else lead SNP ID
  select(`#CHROM`, POS, ID, everything()) |>
  rename(LEAD_SNP=ID)

# Read frequency file (from PLINK --freq output)
ldlc.freqs <- read_tsv(ldlc.freq.file, col_types = cols())

# Read full summary stats (with BETA)
ldlc.sumstats <- read_tsv(ldlc.sumstats.file, col_types = cols())

# Inspect the column names to confirm merging keys and data availability
# head(ldlc.clumps)
# head(ldlc.freq)
# head(ldlc.sumstats)

# The clump file 'SNP' column should match frequency 'ID' column
# The sumstats 'VARIANT' column should match clumps 'SNP' column

# Merge clumps with frequency (to get MAF)
ldlc.clumps_freq <- ldlc.clumps %>%
  inner_join(ldlc.freqs %>% select(ID, ALT_FREQS), by = "SNP")

# Merge with summary stats (to get beta)
ldlc.clumps_freq_beta <- ldlc.clumps_freq %>%
  inner_join(ldlc.sumstats %>% rename(ID = VARIANT) %>% select(ID, BETA, SE, P), by = "SNP")

# Filter by MAF >= 0.01
ldlc.instruments <- ldlc.clumps_freq_beta %>%
  filter(ALT_FREQS >= 0.01, ALT_FREQS <= 0.99)  # also exclude rare alleles > 0.99

n.ldlc <- 469680   # replace with your actual exposure GWAS sample size

ldlc.instruments <- ldlc.instruments %>%
  mutate(
    R2 = 2 * ALT_FREQS * (1 - ALT_FREQS) * BETA^2,
    F = (R2 * (n.ldlc - 2)) / (1 - R2)
  )

# Calculate summary metrics
ldlc.summary_metrics <- 
  ldlc.instruments %>%
  distinct(LEAD_SNP,.keep_all = T) |> #only calculate for lead SNP
  summarise(
    num_snps = n(),
    cumulative_R2 = sum(R2, na.rm = TRUE),    
    mean_F = mean(F, na.rm=TRUE),
    median_F = median(F, na.rm=TRUE),
    mean_maf = mean(ALT_FREQS, na.rm = TRUE),
    mean_beta = mean(abs(BETA), na.rm = TRUE),
    overall_F = (cumulative_R2 * (n.ldlc - num_snps - 1)) / ((1 - cumulative_R2) * num_snps)
  )

kable(ldlc.summary_metrics, caption="Summary of LDL-C instruments")

ldlc.instruments |> 
  separate(ID, into = c("CHR", "POS", "REF", "ALT"), sep = ":", remove = FALSE) %>%
  # choose which is EA/OA – here I assume ALT is the effect allele
  rename(EA = ALT, OA = REF) |>
  mutate(N_exposure=n.ldlc) |>
  select(SNP,CHR,POS, EA,OA,BETA,SE,P,ALT_FREQS,N_exposure,R2, `F`) |>
  write_csv("LDL-C Instruments from UKBB.csv")
```

  


## Session Information

```{r session-information}
sessionInfo()
```