---
title: "Mediator Second-Leg Screen — does the mediator affect serum calcium?"
author: "Dave Bridges"
date: today
editor: source
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
    fig-path: "figures-mediator-screen/"
theme: journal
execute:
  echo: true
  warning: false
---

```{r global-options}
#| echo: false
#| message: false
library(tidyverse)
library(knitr)
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

Before building any full mediation MVMR (`cholesterol + M → calcium`), screen the
**second leg** of each candidate mediator M: does M causally affect serum
calcium at all? Conditioning the cholesterol effect on a mediator whose own
effect on calcium is null is pointless — the mediator cannot carry an indirect
effect. This script runs a univariable MR of **M → serum calcium** for every
candidate M and reports which legs are worth pursuing.

Candidates fall into two groups:

- **Bone density sites** (heel eBMD, femoral neck, total body, …). The heel
  result is already known to be null (β ≈ −0.005, p ≈ 0.76); screening the other
  sites answers the "but maybe hip/femoral-neck mediates" question directly.
- **Hormonal / turnover markers** (PTH via cis-pQTL; add CTX, P1NP, osteocalcin
  when a GWAS id is available). These index calcium *flux/regulation* rather than
  standing density and are the higher-prior candidates under the
  net-mineral-balance model.

Outcome throughout: **serum calcium**, Barton 2021 UKB, `ebi-a-GCST90025990`.

---

## Setup

```{r setup}
#| label: setup
#| message: false
library(TwoSampleMR)
library(ieugwasr)
library(knitr)
library(kableExtra)

CALCIUM_ID  <- "ebi-a-GCST90025990"   # serum calcium outcome (Barton 2021 UKB)
P_GW        <- 5e-8                    # genome-wide instrument threshold
P_CIS       <- 5e-8                    # cis-pQTL threshold (primary)
P_CIS_RELAX <- 1e-5                    # cis fallback if no genome-wide cis hit
R2_CLUMP    <- 0.001
CLUMP_KB    <- 10000

dir.create("results", showWarnings = FALSE)

# ── Candidate-mediator catalogue ─────────────────────────────────────────────
# mode = "genomewide" -> use LD-clumped genome-wide instruments (extract_instruments)
# mode = "cis"        -> use LD-clumped SNPs within `region` (for pQTLs, GRCh37)
#
# PTH gene (GRCh37/hg19) is at chr11:13,492,054-13,495,431; a ±500 kb cis window
# is used below. VERIFY the build of any pQTL dataset before trusting coordinates
# (the OpenGWAS prot-a / prot-c batches are GRCh37).
mediator_catalog <- tribble(
  ~label,                 ~gwas_id,             ~mode,        ~region,
  "Heel eBMD (Morris)",   "ebi-a-GCST006979",   "genomewide", NA_character_,
  "Femoral neck BMD",     "ieu-a-980",          "genomewide", NA_character_,
  "Total body BMD",       "ebi-a-GCST005348",   "genomewide", NA_character_,
  "PTH pQTL (INTERVAL)",  "prot-a-2431",        "cis",        "11:12992054-13995431",
  "PTH pQTL (KORA)",      "prot-c-3726_62_4",   "cis",        "11:12992054-13995431"
  # --- add when a public GWAS id is available: -------------------------------
  # "CTX (resorption)",   "<id>",               "genomewide", NA_character_,
  # "P1NP (formation)",   "<id>",               "genomewide", NA_character_,
  # "Osteocalcin pQTL",   "<prot id>",          "cis",        "<BGLAP region>",
  # "Lumbar spine BMD",   "<verify id>",        "genomewide", NA_character_,
)

kable(mediator_catalog, caption = "Candidate mediators to screen (second leg: M → serum calcium)")
```

### Confirm what each dataset actually is (trait, N)

A cheap sanity check that also settles the "are the PTH pQTLs too small?"
question — for a **cis** instrument, what matters is the cis F-statistic, not the
discovery N, but it is still worth seeing the trait label and sample size.

```{r gwasinfo}
#| label: gwasinfo
#| cache: true

info <- tryCatch(
  ieugwasr::gwasinfo(mediator_catalog$gwas_id),
  error = function(e) { message("gwasinfo failed: ", e$message); NULL }
)
if (!is.null(info)) {
  info %>%
    as_tibble() %>%
    select(any_of(c("id", "trait", "sample_size", "nsnp", "population", "author", "year"))) %>%
    kable(caption = "OpenGWAS metadata for candidate-mediator datasets")
}
```

---

## Screening function

```{r screen-fn}
#| label: screen-fn

# Build an LD-clumped cis instrument set from a region of an OpenGWAS dataset and
# return it in TwoSampleMR exposure format. Falls back to a relaxed threshold if
# nothing reaches genome-wide significance (small pQTLs can have a strong cis
# signal that still sits just under 5e-8).
extract_cis_exposure <- function(gwas_id, region, label) {
  raw <- tryCatch(
    ieugwasr::associations(variants = region, id = gwas_id, proxies = FALSE) %>% as_tibble(),
    error = function(e) { warning(label, " cis query failed: ", e$message); tibble() }
  )
  if (nrow(raw) == 0) return(NULL)

  pick <- raw %>% filter(p <= P_CIS)
  thr  <- P_CIS
  if (nrow(pick) < 1) { pick <- raw %>% filter(p <= P_CIS_RELAX); thr <- P_CIS_RELAX }
  if (nrow(pick) < 1) return(NULL)

  # LD-clump to independent cis signals (fall back to unclumped if LD ref fails)
  clumped <- if (nrow(pick) >= 2) {
    tryCatch(
      ieugwasr::ld_clump(tibble(rsid = pick$rsid, pval = pick$p, id = gwas_id),
                         clump_r2 = R2_CLUMP, clump_kb = CLUMP_KB, pop = "EUR") %>%
        inner_join(pick, by = "rsid"),
      error = function(e) { warning("clump failed (", label, "): ", e$message); pick }
    )
  } else pick

  cat("  [", label, "] cis instruments: ", nrow(clumped),
      " (threshold p<", thr, ")\n", sep = "")

  clumped %>%
    transmute(SNP = rsid, beta.exposure = beta, se.exposure = se,
              effect_allele.exposure = toupper(ea), other_allele.exposure = toupper(nea),
              eaf.exposure = eaf, pval.exposure = p,
              exposure = label, id.exposure = gwas_id,
              mr_keep.exposure = TRUE, pval_origin.exposure = "reported",
              data_source.exposure = "igd")
}

# Screen one candidate mediator's effect on serum calcium (the "b" leg).
screen_mediator <- function(label, gwas_id, mode, region = NA_character_) {
  message("\n----- screening ", label, " (", gwas_id, ") -----")

  ## 1. Mediator instruments
  exp_dat <- if (mode == "genomewide") {
    tryCatch(
      extract_instruments(gwas_id, p1 = P_GW, clump = TRUE, r2 = R2_CLUMP, kb = CLUMP_KB),
      error = function(e) { warning(label, ": ", e$message); NULL }
    )
  } else {
    extract_cis_exposure(gwas_id, region, label)
  }

  none <- tibble(mediator = label, mode = mode, n_snps = 0L, mean_F = NA_real_,
                 method = NA_character_, beta = NA_real_, se = NA_real_,
                 ci_lower = NA_real_, ci_upper = NA_real_, pval = NA_real_,
                 egger_p = NA_real_, gate = "no instruments")
  if (is.null(exp_dat) || nrow(exp_dat) < 1) return(none)

  ## 2. Effects on serum calcium + harmonise
  out <- tryCatch(extract_outcome_data(snps = exp_dat$SNP, outcomes = CALCIUM_ID),
                  error = function(e) { warning(label, " outcome: ", e$message); NULL })
  if (is.null(out) || nrow(out) < 1) return(none %>% mutate(gate = "no outcome overlap"))
  harm <- harmonise_data(exp_dat, out)
  harm <- harm %>% filter(mr_keep)
  nsnp <- nrow(harm)
  if (nsnp < 1) return(none %>% mutate(gate = "no SNPs after harmonisation"))

  ## 3. MR — method scales with instrument count (Wald ratio for a single cis SNP)
  ml <- if (nsnp >= 3) c("mr_ivw_mre", "mr_egger_regression", "mr_weighted_median") else
        if (nsnp == 2) "mr_ivw_mre" else "mr_wald_ratio"
  res     <- mr(harm, method_list = ml)
  primary <- res %>% slice(1)          # IVW-MRE (>=2 SNPs) or Wald ratio (1 SNP)
  egger_p <- tryCatch(mr_pleiotropy_test(harm)$pval, error = function(e) NA_real_)
  mean_F  <- mean((harm$beta.exposure / harm$se.exposure)^2, na.rm = TRUE)

  tibble(
    mediator = label, mode = mode, n_snps = nsnp, mean_F = mean_F,
    method   = primary$method, beta = primary$b, se = primary$se,
    ci_lower = primary$b - 1.96 * primary$se,
    ci_upper = primary$b + 1.96 * primary$se,
    pval     = primary$pval, egger_p = egger_p,
    # Gate: nominal evidence the second leg exists. A NULL here for a small pQTL
    # is inconclusive (winner's curse / low power), not proof of no effect.
    gate = dplyr::case_when(
      primary$pval < 0.05 & mean_F >= 10 ~ "carry forward",
      primary$pval < 0.05                ~ "carry forward (weak instrument — caution)",
      TRUE                               ~ "second leg null (do not condition)"
    )
  )
}
```

---

## Run the screen

```{r run-screen}
#| label: run-screen
#| cache: true

screen_results <- purrr::pmap_dfr(
  mediator_catalog,
  function(label, gwas_id, mode, region) screen_mediator(label, gwas_id, mode, region)
)

write_csv(screen_results, "results/mediator_second_leg_screen.csv")

kable(screen_results %>%
        mutate(across(c(mean_F, beta, se, ci_lower, ci_upper), ~round(.x, 4)),
               pval = signif(pval, 3), egger_p = signif(egger_p, 3)),
      caption = "Second-leg screen: causal effect of each candidate mediator on serum calcium")
```

```{r screen-plot}
#| label: screen-plot
#| fig-width: 8
#| fig-height: 4.5

plot_dat <- screen_results %>% filter(!is.na(beta))
if (nrow(plot_dat) > 0) {
  ggplot(plot_dat, aes(x = beta, y = reorder(mediator, beta))) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_pointrange(aes(xmin = ci_lower, xmax = ci_upper,
                        colour = grepl("carry", gate)), linewidth = 0.7) +
    scale_colour_manual(values = c(`TRUE` = color_scheme[1], `FALSE` = "grey60"),
                        labels = c(`TRUE` = "carry forward", `FALSE` = "null second leg"),
                        name = NULL) +
    labs(title = "Second-leg screen: mediator → serum calcium",
         subtitle = "Only mediators with a non-null effect on calcium are worth conditioning on",
         x = "Effect on serum calcium (per SD / per unit mediator)", y = NULL) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "top")
}
```

---

## Interpretation

```{r interpretation}
#| label: interpretation
#| echo: false
#| results: asis

carry <- screen_results %>% filter(grepl("carry", gate))
drop  <- screen_results %>% filter(gate == "second leg null (do not condition)")
noinst <- screen_results %>% filter(gate %in% c("no instruments", "no outcome overlap",
                                                "no SNPs after harmonisation"))

if (nrow(carry) > 0) {
  cat("**Carry forward into full mediation MVMR** (non-null M → calcium):\n\n")
  for (i in seq_len(nrow(carry))) {
    cat("- ", carry$mediator[i], ": β = ", signif(carry$beta[i], 3),
        " (p = ", signif(carry$pval[i], 2), ", ", carry$n_snps[i], " SNPs)\n", sep = "")
  }
  cat("\n")
} else {
  cat("**No candidate mediator shows a non-null effect on serum calcium.** ",
      "If this holds after adding turnover markers, it argues the cholesterol → ",
      "calcium effect is not routed through any tested bone/hormonal node.\n\n", sep = "")
}

if (nrow(drop) > 0) {
  cat("**Drop — second leg is null** (conditioning would be pointless): ",
      paste(drop$mediator, collapse = ", "), ".\n\n", sep = "")
}
if (nrow(noinst) > 0) {
  cat("**Could not test** (no usable instruments / outcome overlap): ",
      paste0(noinst$mediator, " [", noinst$gate, "]", collapse = "; "), ".\n\n", sep = "")
}
cat("Caveat: a null second leg from a small cis-pQTL (e.g. PTH) is *inconclusive* ",
    "— winner's curse and low power bias the estimate toward zero. Treat a null ",
    "PTH result as 'underpowered', not 'no effect', and revisit with a larger ",
    "pQTL (e.g. deCODE) before discarding the hormonal route.\n", sep = "")
```

---

## Session info

```{r session-info}
#| label: session-info
sessionInfo()
```
