---
title: "MR Analysis and Validation of SNPs for Calcium"
author: "Dave Bridges"
date: "April 9, 2025"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  pdf: default
knitr:
  opts_chunk:
    fig.path: "figures/"
    dev: ["png", "pdf"]  # Remove !expr, just use array syntax
    fig.keep: "all"
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

To validate SNPs for calcium GWAS using those identified using UK Biobank.  This script can be found in `r getwd()` and was most recently run on `r date()`

## Data Entry

```{r data-entry}
instruments.calcium.file <- 'Calcium Instruments from UKBB.csv'
gwas.calcium.file <- 'PheWeb Summary Statistics/phenocode-Ca.tsv.gz'
samplesize.outcome.calcium <- 46100


# loaded and renamed columns
instruments.calcium <- read_csv(instruments.calcium.file) |>
  rename(
    SNP                       = ID,
    id.exposure               = ID,  
    beta.exposure             = BETA,
    se.exposure               = SE,
    effect_allele.exposure    = EA,
    other_allele.exposure     = OA,
    pval.exposure             = P,
    eaf.exposure              = ALT_FREQS,
    samplesize.exposure       = N_exposure
  ) |>
  mutate(exposure="Calcium (UK Biobank)",
         SNP=id.exposure)


gwas.calcium <- read_tsv(gwas.calcium.file) |>
  mutate(ID=paste(chrom, pos, ref,alt, sep=":")) |>
  rename(
    SNP                        = ID,            # or ID if that’s the matching ID
    id.outcome                 = ID,               # can also just set to a string
    beta.outcome               = beta,
    se.outcome                 = sebeta,
    effect_allele.outcome      = alt,   # whichever is effect allele
    other_allele.outcome       = ref,   # whichever is other allele
    pval.outcome               = pval,
    eaf.outcome                = maf,
  ) |>
  mutate(outcome = "Calcium (Michigan GWAS)",
         SNP=id.outcome,# name of your trait)
         samplesize.outcome = samplesize.outcome.calcium)  # sample size for MGI/BioVU for calcium)
```

This presumes the sample sizes was `r as.character(samplesize.outcome.calcium)` from Table 1 of https://doi.org/10.1371/journal.pgen.1009077.

Loaded in the instruments for calcium from UK Biobank from the datafile `r instruments.calcium.file` and the GWAS summary statistics for calcium from the datafile `r gwas.calcium.file`.

```{r calcium-harmonizing-snps}
library(TwoSampleMR)

data <- harmonise_data(instruments.calcium, gwas.calcium, action = 2)
```

Harmonization results

- We used `r nrow(instruments.calcium)` SNPs as instruments for calcium from UK Biobank.
- There were `r length(intersect(instruments.calcium$SNP, gwas.calcium$SNP))` SNPs in common between the exposure and outcome datasets.
- A total of `r nrow(data)` SNPs remained for use after harmonization
- Removed `r sum(data$remove)` SNPs due to allele mismatches
- Identified `r sum(data$palindromic)` palindromic SNPs 

## Session Information

```{r session-information}
sessionInfo()
```
