---
title: "MR Analysis and Validation of SNPs for Calcium"
author: "Dave Bridges"
date: "September 9, 2025"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  pdf: default
knitr:
  opts_chunk:
    fig.path: "figures/"
    dev: ["png", "pdf"]  # Remove !expr, just use array syntax
    fig.keep: "all"
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)
library(knitr)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

To validate SNPs for calcium GWAS using those identified using UK Biobank.  This script can be found in `r getwd()` and was most recently run on `r date()`

## Data Entry

```{r data-entry}
instruments.calcium.file <- 'Calcium Instruments from UKBB.csv'
gwas.calcium.file <- 'PheWeb Summary Statistics/phenocode-Ca.tsv.gz'
samplesize.outcome.calcium <- 46100


# loaded and renamed columns
instruments.calcium <- read_csv(instruments.calcium.file) |>
  rename(
    SNP                       = SP2,
    beta.exposure             = BETA,
    se.exposure               = SE,
    effect_allele.exposure    = EA,
    other_allele.exposure     = OA,
    pval.exposure             = P,
    eaf.exposure              = ALT_FREQS,
    samplesize.exposure       = N_exposure
  ) |>
  mutate(id.exposure="Calcium (UK Biobank)",
         exposure="Calcium (UK Biobank)")


gwas.calcium <- read_tsv(gwas.calcium.file) |>
  mutate(ID=paste(chrom, pos, ref,alt, sep=":")) |>
  rename(
    SNP                        = ID,            # or ID if that’s the matching ID
    beta.outcome               = beta,
    se.outcome                 = sebeta,
    effect_allele.outcome      = alt,   # whichever is effect allele
    other_allele.outcome       = ref,   # whichever is other allele
    pval.outcome               = pval,
    eaf.outcome                = maf,
  ) |>
  mutate(id.outcome = "Calcium (Michigan GWAS)",
         outcome = "Calcium (Michigan GWAS)",
         samplesize.outcome = samplesize.outcome.calcium)  # sample size for MGI/BioVU for calcium)
```

This presumes the sample sizes was `r as.character(samplesize.outcome.calcium)` from Table 1 of https://doi.org/10.1371/journal.pgen.1009077.

Loaded in the instruments for calcium from UK Biobank from the datafile `r instruments.calcium.file` and the GWAS summary statistics for calcium from the datafile `r gwas.calcium.file`.

```{r calcium-harmonizing-snps}
library(TwoSampleMR)

data <- harmonise_data(instruments.calcium, gwas.calcium, action = 2)

table(data$mr_keep) |>
  kable(caption="Number of SNPs kept for MR analysis")

table(data$palindromic)  |>
  kable(caption="Number of palindromic SNPs")

data <- data %>%
  mutate(
    allele_match = (toupper(effect_allele.exposure) == toupper(effect_allele.outcome)) &
                  (toupper(other_allele.exposure) == toupper(other_allele.outcome)),
    allele_swapped = (toupper(effect_allele.exposure) == toupper(other_allele.outcome)) &
                    (toupper(other_allele.exposure) == toupper(effect_allele.outcome))
  )

# 2) EAF concordance checks (detect possible strand/orientation issues)
# requires eaf.exposure and eaf.outcome present
if(all(c("eaf.exposure","eaf.outcome") %in% names(data))){
  data <- data %>%
    mutate(
      eaf_diff = abs(eaf.exposure - eaf.outcome),
      eaf_flip_diff = abs(eaf.exposure - (1 - eaf.outcome)),
      suspicious_eaf = (eaf_diff > 0.2 & eaf_flip_diff > 0.2)  # very different frequencies
    )
  summary(data$eaf_diff)
  summary(data$eaf_flip_diff)
  cat("Num suspicious EAFs:", sum(data$suspicious_eaf, na.rm=TRUE), "\n")
} else {
  cat("No EAF columns present for both datasets; consider adding reference panel EAFs.\n")
}

# 3) List discordant SNPs
data <- data %>%
  mutate(sign_match = sign(beta.exposure) == sign(beta.outcome))

discordant <- data %>% filter(!sign_match) %>%
  select(SNP, beta.exposure, se.exposure, beta.outcome, se.outcome,
         effect_allele.exposure, other_allele.exposure,
         effect_allele.outcome, other_allele.outcome,
         palindromic, ambiguous, eaf.exposure, eaf.outcome)

kable(discordant |>
        arrange(beta.exposure) |>
        select(SNP,beta.exposure,se.exposure,beta.outcome,se.outcome),
      caption="Discordant SNPs where the beta coefficients directionally differ between exposure and outcome")

library(ggrepel)
ggplot(data, aes(x=beta.exposure, y=beta.outcome, color = sign_match)) +
  geom_point(size=2) +
  geom_errorbar(aes(ymin = beta.outcome - 1.96*se.outcome, ymax = beta.outcome + 1.96*se.outcome), width=0) +
  geom_text_repel(data = filter(data, !sign_match), aes(label=SNP), hjust=0, vjust=0, size=3) +
  theme_minimal() +
  labs(x="beta.exposure", y="beta.outcome", title="Exposure vs Outcome betas; discordant SNPs labeled")
```

There were `r nrow(discordant)` discordant SNPs between the exposure and outcome datasets.  These are listed above.  We can see that some of these SNPs have very small effect sizes in the outcome dataset, suggesting that the discordance may be due to noise.  These were kept in the analysis

Harmonization results

- We used `r nrow(instruments.calcium)` SNPs as instruments for calcium from UK Biobank.
- There were `r length(intersect(instruments.calcium$SNP, gwas.calcium$SNP))` SNPs in common between the exposure and outcome datasets.
- A total of `r nrow(data)` SNPs remained for use after harmonization
- Removed `r sum(data$remove)` SNPs due to allele mismatches
- Identified `r sum(data$palindromic)` palindromic SNPs 

```{r calcium-strength}

data.annot <- data %>%
  mutate(
    R2.exposure = 2 * eaf.exposure * (1 - eaf.exposure) * beta.exposure^2,
    F.exposure = (R2.exposure * (samplesize.exposure - 2)) / (1 - R2.exposure)
  )

calcium.exposure.summary <- data.annot %>%
  summarise(
    num_snps = n(),
    samplesize.exposure = first(samplesize.exposure),
    cumulative_R2 = sum(R2.exposure, na.rm = TRUE),
    mean_F = mean(F.exposure, na.rm = TRUE),
    median_F = median(F.exposure, na.rm = TRUE),
    mean_maf = mean(eaf.exposure, na.rm = TRUE),
    mean_beta = mean(abs(beta.exposure), na.rm = TRUE)
  ) |>
  mutate(overall_F = (cumulative_R2 * (samplesize.exposure - num_snps - 1)) / 
                     ((1 - cumulative_R2) * num_snps))

# For outcome (e.g., cholesterol) SNPs
outcome.summary_metrics <- data.annot %>%
  summarise(
    num_snps = n(),
    mean_beta = mean(abs(beta.outcome), na.rm = TRUE),
    mean_se = mean(se.outcome, na.rm = TRUE),
    mean_maf = mean(eaf.outcome, na.rm = TRUE)
  )

library(knitr)
kable(calcium.exposure.summary, caption="Summary of calcium instruments after harmonisation")

write_csv(outcome.summary_metrics, "Instrument Metrics - Calcium - Post-Harmonization.csv")

#write out the instruments used for calcium
data %>% filter(mr_keep==TRUE) %>% 
  mutate(Exposure = "Calcium") |>
  select(Exposure, CHR, POS, effect_allele.exposure, other_allele.exposure, beta.exposure, se.exposure, pval.exposure, eaf.exposure, R2, `F`, rsids, nearest_genes) |>
  rename(effect_allele = effect_allele.exposure,
         other_allele = other_allele.exposure,
         beta = beta.exposure,
         se = se.exposure,
         p = pval.exposure,
         eaf = eaf.exposure) |>
  write_csv("Calcium Instruments Post-Harmonization.csv")
  
kable(outcome.summary_metrics, caption="Summary of outcome summary ")
```

```{r calcium-mr-positive}
calcium.control.mr <- mr(data,
                         method_list = c(
  "mr_ivw", 
  "mr_egger_regression", 
  "mr_weighted_median", 
  "mr_weighted_mode"
))

dat_h_steiger <- steiger_filtering(data)
table(dat_h_steiger$steiger_direction) 

calcium.control.mr |> select(-starts_with('id')) |> 
  kable(caption="MR Results for Calcium Positive Control",
        digits=c(0,0,0,0,3,3,99))

ggplot(calcium.control.mr, aes(y=method,x=b)) +
  geom_point() +
  geom_errorbar(aes(xmin=b-se, xmax=b+se), width=0.2) +
  theme_classic() +
  labs(title="Mendelian Randomization Results for Calcium",
       y="",
       x="Effect Size (Beta)") +
  geom_vline(xintercept=0, linetype="dashed", color = "red") 
```   

The primary result, using the inverse variance weighted method shows a `r calcium.control.mr %>% filter(method=="Inverse variance weighted") %>% pull(b)` $\pm$ `r calcium.control.mr %>% filter(method=="Inverse variance weighted") %>% pull(se)` increase in calcium (Michigan GWAS) per 1 unit increase in calcium (UK Biobank).  This is statistically significant with a p-value of `r calcium.control.mr %>% filter(method=="Inverse variance weighted") %>% pull(pval)`.  All four MR methods (IVW, weighted median, weighted mode, MR-Egger) gave consistent, significant causal estimates, confirming instrument validity and harmonisation.

### MR-Egger Intercept

```{r calcium-mr-pleiotropy}
egger_intercept <- mr_pleiotropy_test(data)
egger_intercept|>
  select(-starts_with('id')) |> 
  kable(caption="MR Pleiotropy Results for Calcium Positive Control")
```
The MR-Egger intercept is `r egger_intercept$intercept` with a p-value of `r egger_intercept$pval`, indicating `r if(egger_intercept$pval < 0.05){"some evidence of"}else{"no evidence of"}` directional pleiotropy.  Although the p-value is small, the intercept magnitude is near zero, indicating that any pleiotropic bias is likely minor.

### Heterogeneity Statistics

```{r calcium-mr-heterogeneity}

# Heterogeneity tests for IVW and MR-Egger
heterogeneity <- mr_heterogeneity(data)
heterogeneity|>
  select(-starts_with('id')) |> 
  kable(caption="MR Heterogeneity Results for Calcium Positive Control",
        digits=c(0,0,0,3,3,99))
# Columns: method, Q, Q_df, Q_pval
# Interpretation: small Q_pval (<0.05) indicates heterogeneity among SNPs
```

This is expected with polygenic traits and does not necessarily invalidate the overall causal estimate, particularly since robust methods (weighted median, weighted mode) gave consistent results.

### Leave-one-out Analysis

Using IVW methods

```{r calcium-mr-loo}
# LOO using IVW
loo_res <- mr_leaveoneout(data)
loo_res |> 
  mutate(diff = b - filter(calcium.control.mr, method=="Inverse variance weighted")$b) |>
  arrange(-abs(diff)) |>
  head() |>
  select(SNP,diff,b,se,p) |>
  kable(caption="Leave-One-Out Results for Calcium Positive Control (IVW method) for influential SNPs",
        digits=c(0,5,5,5,5))
# Columns: SNP, nsnp, b, se, pval — gives causal estimate with each SNP removed once

# Optional: plot LOO results

ggplot(loo_res, aes(x = reorder(SNP, -b), y = b)) +
  geom_point() +
  geom_errorbar(aes(ymin = b - 1.96*se, ymax = b + 1.96*se), width = 0.1) +
  coord_flip() +
  labs(x = "SNP removed", y = "Causal estimate (IVW, leave-one-out)") +
  geom_hline(yintercept=0, linetype="dashed", color = "red") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 4)) 
```
Leave-one-out analyses suggested that two SNPs had a relatively large influence on the IVW estimate, but removal of either SNP did not qualitatively change the overall conclusion, supporting the robustness of the causal inference.”

## Session Information

```{r session-information}
sessionInfo()
```
