---
title: "MR Analysis and Validation of SNPs for Calcium"
author: "Dave Bridges"
date: "April 9, 2025"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  pdf: default
knitr:
  opts_chunk:
    fig.path: "figures/"
    dev: ["png", "pdf"]  # Remove !expr, just use array syntax
    fig.keep: "all"
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

To validate SNPs for calcium GWAS using those identified using UK Biobank.  This script can be found in `r getwd()` and was most recently run on `r date()`

## Data Entry

```{r data-entry}
instruments.calcium.file <- 'Calcium Instruments from UKBB.csv'
gwas.calcium.file <- 'PheWeb Summary Statistics/phenocode-Ca.tsv.gz'
samplesize.outcome.calcium <- 46100


# loaded and renamed columns
instruments.calcium <- read_csv(instruments.calcium.file) |>
  rename(
    SNP                       = SP2,
    id.exposure               = SP2,  
    beta.exposure             = BETA,
    se.exposure               = SE,
    effect_allele.exposure    = EA,
    other_allele.exposure     = OA,
    pval.exposure             = P,
    eaf.exposure              = ALT_FREQS,
    samplesize.exposure       = N_exposure
  ) |>
  mutate(exposure="Calcium (UK Biobank)",
         SNP=id.exposure)


gwas.calcium <- read_tsv(gwas.calcium.file) |>
  mutate(ID=paste(chrom, pos, ref,alt, sep=":")) |>
  rename(
    SNP                        = ID,            # or ID if that’s the matching ID
    id.outcome                 = ID,               # can also just set to a string
    beta.outcome               = beta,
    se.outcome                 = sebeta,
    effect_allele.outcome      = alt,   # whichever is effect allele
    other_allele.outcome       = ref,   # whichever is other allele
    pval.outcome               = pval,
    eaf.outcome                = maf,
  ) |>
  mutate(outcome = "Calcium (Michigan GWAS)",
         SNP=id.outcome,# name of your trait)
         samplesize.outcome = samplesize.outcome.calcium)  # sample size for MGI/BioVU for calcium)
```

This presumes the sample sizes was `r as.character(samplesize.outcome.calcium)` from Table 1 of https://doi.org/10.1371/journal.pgen.1009077.

Loaded in the instruments for calcium from UK Biobank from the datafile `r instruments.calcium.file` and the GWAS summary statistics for calcium from the datafile `r gwas.calcium.file`.

```{r calcium-harmonizing-snps}
library(TwoSampleMR)

data <- harmonise_data(instruments.calcium, gwas.calcium, action = 2)
```

Harmonization results

- We used `r nrow(instruments.calcium)` SNPs as instruments for calcium from UK Biobank.
- There were `r length(intersect(instruments.calcium$SNP, gwas.calcium$SNP))` SNPs in common between the exposure and outcome datasets.
- A total of `r nrow(data)` SNPs remained for use after harmonization
- Removed `r sum(data$remove)` SNPs due to allele mismatches
- Identified `r sum(data$palindromic)` palindromic SNPs 

```{r calcium-strength}
data.annot <- data |>
  mutate(R2 = 2 * eaf.outcome * (1 - eaf.outcome) * beta.outcome^2,
         F = (R2 * (samplesize.outcome.calcium - 2)) / (1 - R2))

# Calculate summary metrics
calcium.summary_metrics.outcome <- 
  data.annot %>%
  ungroup %>%
  summarise(
    num_snps = n(),
   cumulative_R2 = sum(R2, na.rm = TRUE),
    mean_F = mean(F, na.rm=TRUE),
    median_F = median(F, na.rm=TRUE),
    mean_maf = mean(eaf.outcome, na.rm = TRUE),
    mean_beta = mean(abs(beta.outcome), na.rm = TRUE),
    overall_F = (cumulative_R2 * (samplesize.outcome.calcium - n() - 1)) / ((1 - cumulative_R2) * n())
  )

library(knitr)
kable(calcium.summary_metrics.outcome, caption="Summary of calcium instruments for outcome GWAS")
```

```{r calcium-mr-positive}
calcium.control.mr <- mr(data)
summary(calcium.control.mr)
#mr_forest_plot(calcium.control.mr)

dat_h_steiger <- steiger_filtering(data)
table(dat_h_steiger$steiger_direction) 

ggplot(calcium.control.mr, aes(y=reorder(id.outcome,b),x=b)) +
  geom_point() +
  geom_errorbar(aes(xmin=b-se, xmax=b+se), width=0.2) +
  theme_classic() +
  labs(title="Mendelian Randomization Results for Calcium",
       y="Instrument",
       x="Effect Size (Beta)") +
  geom_vline(xintercept=0, linetype="dashed", color = "red") +
  theme(axis.text.y = element_text(size=2))
```   


## Session Information

```{r session-information}
sessionInfo()
```
