---
title: "MR Analysis of SNPs for Total Cholesterol and their Effect on Serum Calcium"
author: "Dave Bridges"
date: "October 14, 2025"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  pdf: default
knitr:
  opts_chunk:
    fig.path: "figures/"
    dev: ["png", "pdf"]  # Remove !expr, just use array syntax
    fig.keep: "all"
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)
library(knitr)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

To validate SNPs for total cholesterol GWAS using those identified using UK Biobank.  This script can be found in `r getwd()` and was most recently run on `r date()`

## Data Entry

```{r data-entry}
instruments.tc.file <- 'Total Cholesterol Instruments from UKBB.csv'
gwas.calcium.file <- 'PheWeb Summary Statistics/phenocode-Ca.tsv.gz'
samplesize.outcome.calcium <- 46100


# loaded and renamed columns
instruments.tc <- read_csv(instruments.tc.file) |>
  rename(
    SNP                       = SP2,
    beta.exposure             = BETA,
    se.exposure               = SE,
    effect_allele.exposure    = EA,
    other_allele.exposure     = OA,
    pval.exposure             = P,
    eaf.exposure              = ALT_FREQS,
    samplesize.exposure       = N_exposure
  ) |>
  mutate(id.exposure="Total Cholesterol (UK Biobank)",
         exposure="Total Cholesterol (UK Biobank)")


gwas.calcium <- read_tsv(gwas.calcium.file) |>
  mutate(ID=paste(chrom, pos, ref,alt, sep=":")) |>
  rename(
    SNP                        = ID,            # or ID if that’s the matching ID
    beta.outcome               = beta,
    se.outcome                 = sebeta,
    effect_allele.outcome      = alt,   # whichever is effect allele
    other_allele.outcome       = ref,   # whichever is other allele
    pval.outcome               = pval,
    eaf.outcome                = maf,
  ) |>
  mutate(id.outcome = "Calcium (MGI-BioVU LabWAS)",
         outcome = "Calcium (MGI-BioVU LabWAS)",
         samplesize.outcome = samplesize.outcome.calcium)  # sample size for MGI/BioVU for calcium)
```

This presumes the sample sizes was `r as.character(samplesize.outcome.calcium)` from Table 1 of https://doi.org/10.1371/journal.pgen.1009077.

Loaded in the instruments for total cholesterol from UK Biobank from the datafile `r instruments.tc.file` and the GWAS summary statistics for calcium from the datafile `r gwas.calcium.file`.

```{r calcium-harmonizing-snps}
library(TwoSampleMR)

data <- harmonise_data(instruments.tc, gwas.calcium, action = 2)

table(data$mr_keep) |>
  kable(caption="Number of SNPs kept for MR analysis")

table(data$palindromic)  |>
  kable(caption="Number of palindromic SNPs")

data <- data %>%
  mutate(
    allele_match = (toupper(effect_allele.exposure) == toupper(effect_allele.outcome)) &
                  (toupper(other_allele.exposure) == toupper(other_allele.outcome)),
    allele_swapped = (toupper(effect_allele.exposure) == toupper(other_allele.outcome)) &
                    (toupper(other_allele.exposure) == toupper(effect_allele.outcome))
  )

# 2) EAF concordance checks (detect possible strand/orientation issues)
# requires eaf.exposure and eaf.outcome present
if(all(c("eaf.exposure","eaf.outcome") %in% names(data))){
  data <- data %>%
    mutate(
      eaf_diff = abs(eaf.exposure - eaf.outcome),
      eaf_flip_diff = abs(eaf.exposure - (1 - eaf.outcome)),
      suspicious_eaf = (eaf_diff > 0.2 & eaf_flip_diff > 0.2)  # very different frequencies
    )
  summary(data$eaf_diff)
  summary(data$eaf_flip_diff)
  cat("Num suspicious EAFs:", sum(data$suspicious_eaf, na.rm=TRUE), "\n")
} else {
  cat("No EAF columns present for both datasets; consider adding reference panel EAFs.\n")
}

# 3) List discordant SNPs
data <- data %>%
  mutate(sign_match = sign(beta.exposure) == sign(beta.outcome))

discordant <- data %>% filter(!sign_match) %>%
  select(SNP, beta.exposure, se.exposure, beta.outcome, se.outcome,
         effect_allele.exposure, other_allele.exposure,
         effect_allele.outcome, other_allele.outcome,
         palindromic, ambiguous, eaf.exposure, eaf.outcome)

kable(discordant |>
        arrange(beta.exposure) |>
        select(SNP,beta.exposure,se.exposure,beta.outcome,se.outcome),
      caption="Discordant SNPs where the beta coefficients directionally differ between exposure and outcome")

library(ggrepel)
ggplot(data, aes(x=beta.exposure, y=beta.outcome, color = sign_match)) +
  geom_point(size=2) +
  geom_errorbar(aes(ymin = beta.outcome - 1.96*se.outcome, ymax = beta.outcome + 1.96*se.outcome), width=0) +
  geom_text_repel(data = filter(data, !sign_match), aes(label=SNP), hjust=0, vjust=0, size=3) +
  theme_minimal() +
  labs(x="beta.exposure", y="beta.outcome", title="Exposure vs Outcome betas; discordant SNPs labeled")
```

```{r tc-calcium-scatter}
ggplot(data, aes(x=beta.exposure, y=beta.outcome)) +
  geom_point(size=1) +
  geom_errorbar(aes(ymin = beta.outcome - 1.96*se.outcome,
                    ymax = beta.outcome + 1.96*se.outcome),
                alpha=0.5) +
  geom_errorbar(aes(xmin = beta.exposure - 1.96*se.exposure,
                    xmax = beta.exposure + 1.96*se.exposure),
                alpha=0.5) +
  geom_smooth(method="lm",se=F) +
  theme_classic(base_size=16) +
  labs(x="Exposure Estimate", 
       y="Outcome Estimate", 
       title="") 
```

There were `r nrow(discordant)` discordant SNPs between the exposure and outcome datasets.  These are listed above.  We can see that some of these SNPs have very small effect sizes in the outcome dataset, suggesting that the discordance may be due to noise.  These were kept in the analysis

### Steiger Filtering

```{r tc-calcium-steiger}
data_steiger <- steiger_filtering(data)

table(data_steiger$steiger_direction, useNA="ifany") |>
  kable(caption="Steiger filtering results for calcium-calcium evaluation")
```

Harmonization results

- We used `r nrow(instruments.tc)` SNPs as instruments for calcium from UK Biobank.
- There were `r length(intersect(instruments.tc$SNP, gwas.calcium$SNP))` SNPs in common between the exposure and outcome datasets.
- Removed `r sum(data$remove)` SNPs due to allele mismatches
- Identified `r sum(data$palindromic)` palindromic SNPs 
- A total of `r nrow(data)` SNPs remained for use after harmonization.  `r data |> filter(ambiguous==T) |> nrow()` SNPS were removed because the palindromic SNP is ambiguous and strand alignment could not be resolved, this variant was automatically dropped from the MR analysis to avoid mis-specified effect directions.
- After Steiger filtering, `r nrow(data_steiger %>% filter(mr_keep==T))` SNPs were retained for analysis, indicating that all SNPs had stronger associations with the exposure (calcium in UK Biobank) than the outcome (calcium in MGI/BioVU), supporting the assumed causal direction.  `r nrow(filter(data, mr_keep==T)) - nrow(filter(data_steiger, mr_keep==T))` SNPs were removed by Steiger filtering.

```{r tc-strength}

data.annot <- data_steiger %>%
  mutate(
    R2.exposure = 2 * eaf.exposure * (1 - eaf.exposure) * beta.exposure^2,
    F.exposure = (R2.exposure * (samplesize.exposure - 2)) / (1 - R2.exposure)
  )

calcium.exposure.summary <- data.annot %>%
  summarise(
    num_snps = n(),
    samplesize.exposure = first(samplesize.exposure),
    cumulative_R2 = sum(R2.exposure, na.rm = TRUE),
    mean_F = mean(F.exposure, na.rm = TRUE),
    median_F = median(F.exposure, na.rm = TRUE),
    mean_maf = mean(eaf.exposure, na.rm = TRUE),
    mean_beta = mean(abs(beta.exposure), na.rm = TRUE)
  ) |>
  mutate(overall_F = (cumulative_R2 * (samplesize.exposure - num_snps - 1)) / 
                     ((1 - cumulative_R2) * num_snps))

# For outcome (e.g., cholesterol) SNPs
outcome.summary_metrics <- data.annot %>%
  summarise(
    num_snps = n(),
    mean_beta = mean(abs(beta.outcome), na.rm = TRUE),
    mean_se = mean(se.outcome, na.rm = TRUE),
    mean_maf = mean(eaf.outcome, na.rm = TRUE)
  )

library(knitr)
kable(calcium.exposure.summary, caption="Summary of total cholesterol instruments after harmonisation")

write_csv(calcium.exposure.summary, "Instrument Metrics - Total Cholesterol - Post-Harmonization.csv")
write_csv(outcome.summary_metrics, "Outcome Metrics - Total Cholesterol - Post-Harmonization.csv")

#write out the instruments used for calcium
data_steiger %>% filter(mr_keep==TRUE) %>% 
  mutate(Exposure = "Calcium") |>
  select(Exposure, CHR, POS, effect_allele.exposure, other_allele.exposure, beta.exposure, se.exposure, pval.exposure, eaf.exposure, R2, `F`, rsids, nearest_genes) |>
  rename(effect_allele = effect_allele.exposure,
         other_allele = other_allele.exposure,
         beta = beta.exposure,
         se = se.exposure,
         p = pval.exposure,
         eaf = eaf.exposure) |>
  write_csv("Total Cholesterol Instruments Post-Harmonization.csv")
  
kable(outcome.summary_metrics, caption="Summary of total cholesterol effects after harmonisation")
```

## MR Analyses

```{r tc-calcium-mr}
tc.control.mr <- mr(data_steiger,
                         method_list = c(
  "mr_ivw_mre",
  "mr_ivw_fe",
  'mr_raps',
  "mr_egger_regression", 
  "mr_weighted_median", 
  "mr_weighted_mode"
))

#M-PRESSO has to be run separately
library(MRPRESSO)

tc.control.mr_presso_results <- mr_presso(
  BetaOutcome = "beta.outcome",      # Column name for outcome betas
  BetaExposure = "beta.exposure",    # Column name for exposure betas
  SdOutcome = "se.outcome",          # Column name for outcome SEs
  SdExposure = "se.exposure",        # Column name for exposure SEs
  data = data_steiger,                  # Your dataset
  NbDistribution = 2000,              # Number of distributions (default 1000)
  SignifThreshold = 0.05,             # Significance threshold
  OUTLIERtest = TRUE,                 # Perform outlier test
  DISTORTIONtest = TRUE,              # Perform distortion test
)

library(forcats)
tc.control.mr_presso_results$`Main MR results` |>
  select(`MR Analysis`, `Causal Estimate`, Sd, `P-value`) |>
  rename(method = `MR Analysis`,
         b = `Causal Estimate`,
         se = Sd,
         pval = `P-value`) |>
  mutate(method = fct_recode(method,
                             "MR-PRESSO (Raw)"="Raw",
                             "MR-PRESSO (Outlier-corrected)"="Outlier-corrected")) -> tc.control.mr_presso_df

tc.control.mr.mrpresso <- bind_rows(as_tibble(tc.control.mr), as_tibble(tc.control.mr_presso_df))|>
  fill(id.exposure, id.outcome, exposure, outcome,nsnp,.direction="down")

tc.control.mr.mrpresso |> select(-starts_with('id')) |> 
  kable(caption="MR Results for Total Cholesterol on Calcium Positive Control",
        digits=c(0,0,0,0,4,4,99))

tc.control.mr.mrpresso |> select(-starts_with('id')) |> 
  write_csv("MR Results - Total Cholesterol - Calciuml.csv")

ggplot(tc.control.mr.mrpresso, aes(y=method,x=b)) +
  geom_point() +
  geom_errorbar(aes(xmin=b-1.96*se, xmax=b+1.96*se), width=0.2) +
  theme_classic(base_size=16) +
  labs(title="",
       y="",
       x="Effect Size (Beta)") +
  geom_vline(xintercept=0, linetype="dashed", color = "red") 
```   

The primary result, using the inverse variance weighted (multiplicative random effects) method shows a `r tc.control.mr %>% filter(method=="Inverse variance weighted (multiplicative random effects)") %>% pull(b)` $\pm$ `r tc.control.mr %>% filter(method=="Inverse variance weighted (multiplicative random effects)") %>% pull(se)` mg/dL increase in calcium (MGI-BioVU LabWAS) per 1 SD increase in total cholesterol (UK Biobank).  This is statistically significant with a p-value of `r tc.control.mr %>% filter(method=="Inverse variance weighted (multiplicative random effects)") %>% pull(pval)`.  Three of the four MR methods (IVW, weighted median, weighted mode) gave consistent, significant causal estimates, supporting the hypothesis that total cholesterol may impact serum calcium levels in a positive direction.  The MR-Egger estimate was not statistically significant, but this method is known to have low power.

### MR-Egger Intercept

```{r tc-calcium-mr-pleiotropy}
egger_intercept <- mr_pleiotropy_test(data_steiger)
egger_intercept|>
  select(-starts_with('id')) |> 
  kable(caption="MR Pleiotropy Results for Total Cholesterol - Calcium Analysis")
```
The MR-Egger intercept is `r egger_intercept$intercept` with a p-value of `r egger_intercept$pval`, indicating `r if(egger_intercept$pval < 0.05){"some evidence of"}else{"no evidence of"}` average directional pleiotropy.  The intercept magnitude is small and near zero, indicating that any average directional pleiotropic bias is likely minor.

### Heterogeneity Statistics

```{r tc-calcium-mr-heterogeneity}
# Heterogeneity tests for IVW and MR-Egger
heterogeneity <- mr_heterogeneity(data_steiger)
heterogeneity|>
  select(-starts_with('id')) |> 
    mutate(
    I2 = pmax(0, (Q - Q_df) / Q) * 100 # 
  ) |>
  kable(caption="MR Heterogeneity Results for Total Cholesterol - Calcium Analysis",
        digits=c(0,0,0,3,3,99))
# Columns: method, Q, Q_df, Q_pval
# Interpretation: small Q_pval (<0.05) indicates heterogeneity among SNPs
```

This is expected with polygenic traits and does not necessarily invalidate the overall causal estimate, particularly since robust methods (weighted median, weighted mode) gave consistent results.

```{r tc-calcium-funnel-plot}
single_snp_results <- mr_singlesnp(data_steiger)
mr_funnel_plot(single_snp_results) 

# Get overall IVW estimate for the vertical line
ivw_beta <- tc.control.mr |> filter(method=="Inverse variance weighted (multiplicative random effects)") |> pull(b)

# Determine y-range based on your data
y_min <- 0
y_max <- max(single_snp_results$se^{-1}) * 1.1  # 10% padding above max precision

# Generate a fine grid of precision values
precision_grid <- seq(y_min, y_max, length.out = 1000)

# Compute boundaries: ivw_beta ± 1.96 / precision
lower_bound <- ivw_beta - 1.96 / precision_grid
upper_bound <- ivw_beta + 1.96 / precision_grid

# Create data frame for boundaries
bounds_df <- data.frame(precision = precision_grid, lower = lower_bound, upper = upper_bound)

# Plot
ggplot(single_snp_results, aes(x = b, y = 1/se)) +
  # Scatter points for each SNP
  geom_point(size = 1) +
  # Vertical line at IVW estimate
  geom_vline(xintercept = ivw_beta, linetype = "solid", color = "#ff7f0e", size = 1) +
  # Curved pseudo-95% CI boundaries (the cone)
  geom_line(data = bounds_df, aes(x = lower, y = precision), linetype = "dashed") +
  geom_line(data = bounds_df, aes(x = upper, y = precision), linetype = "dashed") +
  # Customize axes and labels
  labs(
    x = "Estimate (Beta-IVW)",
    y = "Precision (1/Standard Error)",
    title = ""
  ) +
  # Apply clean theme and limit y to >=0
  theme_classic(base_size = 16) +
  theme(plot.title = element_text(hjust = 0.5)) +
  coord_cartesian(ylim = c(0, y_max), xlim = c(min(single_snp_results$b), max(single_snp_results$b)))  # Adjust x-limits for visibility

```

### Global Test and Distortion Test

```{r calcium-calcium-global-test}
tc.control.mr_presso_results$`MR-PRESSO results`$`Global Test` |>
  unlist() |> as.data.frame() |>
  rename(`Global Test` = 1) |>
  kable(caption="MR-PRESSO Global Test Results for Total Cholesterol on Calcium")

tc.control.mr_presso_results$`MR-PRESSO results`$`Distortion Test` -> tc.distortion.test
```

From MR-PRESSO there were `r length(tc.distortion.test[[1]])` outliers that were removed before the distortion test.  The ratio of the IVW estimate before and after their removal was `r tc.distortion.test[[2]]`, with a p-value of `r tc.distortion.test[[3]]`, indicating `r if(tc.distortion.test[[3]] < 0.05){"significant"}else{"no significant"}` distortion due to outliers.  The global test was significant, showing evidence of the presence of horizontal pleiotropy via outliers, but their removal did not significantly alter the estimate (p=`r tc.distortion.test[[3]]`), IVW-MRE is preferred given heterogeneity (significant Q) but stable estimate after outlier removal.

### Leave-one-out Analysis

Using IVW methods

```{r tc-calcium-mr-loo}
# LOO using IVW
loo_res <- mr_leaveoneout(data_steiger)
loo_res |> 
  mutate(diff = b - filter(tc.control.mr, method=="Inverse variance weighted (multiplicative random effects)")$b) |>
  arrange(-abs(diff)) |>
  head() |>
  select(SNP,diff,b,se,p) |>
  kable(caption="Leave-One-Out Results for Total Cholesterol Analysis (IVW method) for influential SNPs",
        digits=c(0,5,5,5,5))
# Columns: SNP, nsnp, b, se, pval — gives causal estimate with each SNP removed once

# Optional: plot LOO results

ggplot(loo_res, aes(x = reorder(SNP, -b), y = b)) +
  geom_point(size=1) +
  geom_errorbar(aes(ymin = b - 1.96*se, ymax = b + 1.96*se), width = 0.01 ,alpha=0.5) +
  coord_flip() +
  labs(x = "SNP Removed", y = "Estimate (Beta-IVW; leave-one-out)") +
  geom_hline(yintercept=0, linetype="dashed", color = "red") +
  theme_classic(base_size=16) +
  theme(axis.text.y = element_text(size = 1)) 
```
Leave-one-out analyses suggested that no SNPs had a relatively large influence on the IVW estimate, because removal did not qualitatively change the overall conclusion, supporting the robustness of the causal inference.

## Session Information

```{r session-information}
sessionInfo()
```
