---
title: "Summary Tables and Figures from MR Analyses"
author: "Dave Bridges"
date: "October 17, 2025"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  pdf: default
knitr:
  opts_chunk:
    fig.path: "figures/"
    dev: ["png", "pdf"]  # Remove !expr, just use array syntax
    fig.keep: "all"
execute:
  echo: true
  warning: false
bibliography: references.bib
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)
library(knitr)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

To generate summary tables and figures from individual analyses for publicaiton

## Post-Harmonization Instrument Summary Statistics Table

```{r data-entry}
calcium.instrument.summary.file <- 'Instrument Metrics - Calcium - Post-Harmonization (Total Cholesterol).csv' #identical instruments for LDL-Cholesterol outcomes after harmonisation
ldl.instrument.summary.file <- 'Instrument Metrics - LDL Cholesterol - Post-Harmonization.csv'
tc.instrument.summary.file <- 'Instrument Metrics - Total Cholesterol - Post-Harmonization.csv'

bind_rows(read_csv(calcium.instrument.summary.file) |> mutate(Instrument = "Serum Calcium"),
      read_csv(ldl.instrument.summary.file) |> mutate(Instrument = "LDL-Cholesterol"),
      read_csv(tc.instrument.summary.file) |> mutate(Instrument = "Total Cholesterol")) |>
  relocate(Instrument, .before = everything()) |>
  relocate(overall_F, .before= mean_maf) |>
  rename(N=samplesize.exposure,
         SNPs = num_snps,
         `Cumulative R2`=cumulative_R2,
         `Mean F Statistic` = mean_F,
         `Median F Statistic`=median_F,
         `Overall F Statistic`=overall_F,
         `Mean MAF`=mean_maf,
         `Mean Beta Coefficient`=mean_beta) -> instrument.summary

instrument.summary |> 
  kable(caption="Instrument summary after harmonisation")

instrument.summary |>
  mutate(across(ends_with("R2") | ends_with("Coefficient") | ends_with('MAF'), ~ round(., 3))) %>%
  mutate(across(ends_with("Statistic"), ~ round(., 1))) %>%
  # SNPs and N are left unrounded (no action needed) |>
  write_csv("Instrument Metrics - Post-Harmonization.csv")
```

## Instrument Lists

```{r instrument-lists}
calcium.instruments.file <- 'Calcium Instruments Post-Harmonization (Total Cholesterol).csv'
ldl.instruments.file <- 'LDL Cholesterol Instruments Post-Harmonization.csv'
tc.instruments.file <- 'Total Cholesterol Instruments Post-Harmonization.csv'

bind_rows(read_csv(calcium.instruments.file) |> mutate(Instrument = "Serum Calcium"),
      read_csv(ldl.instruments.file) |> mutate(Instrument = "LDL-Cholesterol"),
      read_csv(tc.instruments.file) |> mutate(Instrument = "Total Cholesterol")) |>
  relocate(Instrument, .before = everything()) |>
  select(-Exposure) -> instrument.list

instrument.list |>
  write_csv("Instrument SNP Lists - Post-Harmonization.csv")
```

## MR Results Summary Table

```{r mr-results-summary}
calcium.tc.mr.file <- 'MR Results - Calcium - Total Cholesterol.csv'
calcium.ldl.mr.file <- 'MR Results - Calcium - LDL Cholesterol.csv'
tc.calcium.mr.file <- 'MR Results - Total Cholesterol - Calcium.csv'
ldl.calcium.instruments.file <- 'MR Results - LDL Cholesterol - Calcium.csv'

bind_rows(read_csv(calcium.tc.mr.file) |> mutate(Analysis = "Calcium -> Total Cholesterol"),
      read_csv(calcium.ldl.mr.file) |> mutate(Analysis = "Calcium -> LDL Cholesterol"),
      read_csv(tc.calcium.mr.file) |> mutate(Analysis = "Total Cholesterol -> Calcium"),
      read_csv(ldl.calcium.instruments.file) |> mutate(Analysis = "LDL Cholesterol -> Calcium")) |>
  relocate(Analysis, .before = everything()) |>
  select(-nsnp,-outcome,-exposure) -> mr.results.summary

mr.results.summary |> kable(caption="Mendelian Randomization Results Summary")

mr.results.summary |>
  filter(method=="IVW-RE") |>
  mutate(across(c(b, se), ~ round(., 4))) %>%
  mutate(across(c(pval), ~ signif(., 3))) %>%
  select(-method) %>%
  rename(`Beta Coefficient`=b,
         `Standard Error`=se,
         `P-value`=pval) %>%
  # Analysis and Method are left unrounded (no action needed) |>
  write_csv("MR Results Summary.csv")

mr.results.summary |>
  filter(method=="IVW-RE") |>
  ggplot(aes(x=Analysis, y=b)) +
  geom_point(stat="identity") +
  geom_errorbar(aes(ymin=b - 1.96*se, ymax=b + 1.96*se), width=.2) +
  coord_flip() +
  geom_hline(yintercept=0, linetype="dashed", color="red") +
  theme_classic(base_size=14)+
  labs(title="", y="Beta Coefficient", x="") +
  theme(axis.text.y = element_text(size = 8))

# Calculate x-axis limits (for beta coefficients, which will be the x-axis after coord_flip)
mr.results.summary |>
  summarize(xmin = min(b) - max(se), xmax = max(b) + max(se)) -> grid.xlims

# Create the first plot (Calcium -> Cholesterol)
calcium.mr.plot <- mr.results.summary |>
  filter(method == "IVW-RE") |>
  filter(Analysis %in% c("Calcium -> Total Cholesterol", "Calcium -> LDL Cholesterol")) |>
  ggplot(aes(x = Analysis, y = b)) +
  geom_point(stat = "identity") +
  geom_errorbar(aes(ymin = b - 1.96 * se, ymax = b + 1.96 * se), width = 0.2) +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_classic(base_size = 14) +
  labs(title = "Calcium Effects on Cholesterol", y = "Beta Coefficient", x = "") +
  # Apply limits to the y-axis (beta coefficients, which become x-axis after coord_flip)
  scale_y_continuous(limits = c(grid.xlims$xmin, grid.xlims$xmax)) 

# Create the second plot (Cholesterol -> Calcium)
cholesterol.mr.plot <- mr.results.summary |>
  filter(method == "IVW-RE") |>
  filter(Analysis %in% c("Total Cholesterol -> Calcium", "LDL Cholesterol -> Calcium")) |>
  ggplot(aes(x = Analysis, y = b)) +
  geom_point(stat = "identity") +
  geom_errorbar(aes(ymin = b - 1.96 * se, ymax = b + 1.96 * se), width = 0.2) +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_classic(base_size = 14) +
  labs(title = "Cholesterol Effects on Calcium", y = "Beta Coefficient", x = "") +
  # Apply limits to the y-axis (beta coefficients, which become x-axis after coord_flip)
  scale_y_continuous(limits = c(grid.xlims$xmin, grid.xlims$xmax)) 

# Arrange plots vertically with aligned axes
library(cowplot)
plot_grid(calcium.mr.plot, cholesterol.mr.plot, ncol = 1, align = "hv")

```

## Bayesian Inference

To come up with posterior probabilities of the four hypotheses I am interested in (focusing on total cholesterol rather than LDL-C)

```{r bayesian-inference-setup}
H1_prob <- 0.75
H2_prob <- 0.10
H3_prob <- 0.10
H4_prob <- 0.05

# pulled out relevant beta coefficents
beta1   <- mr.results.summary %>% filter(method=="IVW-RE",Analysis == "Total Cholesterol -> Calcium") |> pull(b)
se1 <- mr.results.summary %>% filter(method=="IVW-RE",Analysis == "Total Cholesterol -> Calcium") |> pull(se) 
beta2   <- mr.results.summary %>% filter(method=="IVW-RE",Analysis == "Calcium -> Total Cholesterol") |> pull(b)
se2     <- mr.results.summary %>% filter(method=="IVW-RE",Analysis == "Calcium -> Total Cholesterol") |> pull(se) 
```

* H1: Cholesterol causes increases in calcium.  I estimate a prior probability of `r H1_prob*100`% for this hypothesis
* H2: Calcium causes increases in cholesterol (`r H2_prob*100`%)
* H3: Neither causes the other (*e.g.* the observed correlation is confounded, `r H3_prob*100`%)
* H4: Both hypotheses are true (*e.g.* its bidirectional, `r H4_prob*100`%)

### Bayesian Assessment of Causal Direction

To formally compare the four competing causal hypotheses between circulating cholesterol and calcium levels, we performed Bayesian model comparison using the two bidirectional Mendelian randomization (MR) estimates (cholesterol $\rightarrow$ calcium: $\hat{\beta}_1$ =`r beta1` $\pm$ `r se1`; calcium $\rightarrow$ cholesterol: $\hat{\beta}_2$ = `r beta2` $\pm$ `r se2`). Horizontal pleiotropy was previously ruled out using MR-Egger, weighted median, and MR-PRESSO ($P > 0.05$ for all tests).

For each non-zero causal effect under H1, H2, and H4, we specified a weakly informative normal prior on the true causal effect size: $\theta \sim \mathcal{N}(0, 0.1^2)$. This corresponds to a 95% prior belief that the causal odds ratio per genetically predicted standard deviation lies between approximately 0.71 and 1.41—consistent with small-to-moderate effects typically observed for similar traits .

The marginal likelihood for each hypothesis was computed analytically under the normal approximations:

$$
\hat{\beta}_i \mid \theta_i, \sigma_i \sim \mathcal{N}(\theta_i, \sigma_i^2), \quad
\theta_i \sim 
\begin{cases}
\delta_0 & \text{(spike at zero if hypothesis assumes no effect)} \\
\mathcal{N}(0, 0.1^2) & \text{(slab if hypothesis assumes a non-zero effect)}
\end{cases}
$$

Posterior probabilities were obtained via Bayes’ theorem:

$$
P(H_k \mid \text{data}) = \frac{P(\text{data} \mid H_k) \cdot P(H_k)}{\sum_{j=1}^4 P(\text{data} \mid H_j) \cdot P(H_j)}
$$

This Bayesian framework follows the approach of [@grantBayesianApproachMendelian2024]

```{r cholesterol-calcium-bayesian-inference}

# Priors on hypotheses --------------------------------------
prior_H <- c(H1 = H1_prob, H2 = H2_prob, H3 = H3_prob, H4 = H4_prob)

# Prior on non-zero effects (small effects) -----------------
prior_sd <- 0.1   # change to 0.05–0.20 to test sensitivity
prior_sd_low <- 0.05   # change to 0.05–0.20 to test sensitivity
prior_sd_high <- 0.2  # change to 0.05–0.20 to test sensitivity

# Function to compute log marginal likelihood for a single effect
log_marg_lik <- function(beta, se, prior_sd, hypothesis_zero = FALSE) {
  if (hypothesis_zero) {
    # Spike at zero
    dnorm(beta, 0, sqrt(se^2), log = TRUE)
  } else {
    # Normal prior on effect, variance = se^2 + prior_sd^2
    post_var  <- 1 / (1/se^2 + 1/prior_sd^2)
    post_mean <- post_var * (beta / se^2)
    post_sd   <- sqrt(post_var)
    
    # log marginal likelihood under the alternative
    dnorm(beta, 0, sqrt(se^2 + prior_sd^2), log = TRUE)
  }
}

# Log marginal likelihood for each hypothesis ----------------
log_marg <- c(
  H1 = log_marg_lik(beta1, se1, prior_sd, hypothesis_zero = FALSE) +  # Chol->Ca non-zero
       log_marg_lik(beta2, se2, prior_sd, hypothesis_zero = TRUE),    # Ca->Chol zero
  
  H2 = log_marg_lik(beta1, se1, prior_sd, hypothesis_zero = TRUE) +     # Chol->Ca zero
       log_marg_lik(beta2, se2, prior_sd, hypothesis_zero = FALSE),   # Ca->Chol non-zero
  
  H3 = log_marg_lik(beta1, se1, prior_sd, hypothesis_zero = TRUE) + 
       log_marg_lik(beta2, se2, prior_sd, hypothesis_zero = TRUE),
  
  H4 = log_marg_lik(beta1, se1, prior_sd, hypothesis_zero = FALSE) + 
       log_marg_lik(beta2, se2, prior_sd, hypothesis_zero = FALSE)
)

# Posterior probabilities ------------------------------------
log_marg <- log_marg + log(prior_H)   # add log prior
log_marg <- log_marg - max(log_marg)  # stabilise
post_prob <- exp(log_marg) / sum(exp(log_marg))
post_prob |> as.data.frame() |> kable(caption="Posterior probabilities for each hypothesis")

# Sensitivities
log_marg_low <- c(
  H1 = log_marg_lik(beta1, se1, prior_sd_low, hypothesis_zero = FALSE) +  # Chol->Ca non-zero
       log_marg_lik(beta2, se2, prior_sd_low, hypothesis_zero = TRUE),    # Ca->Chol zero
  
  H2 = log_marg_lik(beta1, se1, prior_sd_low, hypothesis_zero = TRUE) +     # Chol->Ca zero
       log_marg_lik(beta2, se2, prior_sd_low, hypothesis_zero = FALSE),   # Ca->Chol non-zero
  
  H3 = log_marg_lik(beta1, se1, prior_sd_low, hypothesis_zero = TRUE) + 
       log_marg_lik(beta2, se2, prior_sd_low, hypothesis_zero = TRUE),
  
  H4 = log_marg_lik(beta1, se1, prior_sd_low, hypothesis_zero = FALSE) + 
       log_marg_lik(beta2, se2, prior_sd_low, hypothesis_zero = FALSE)
)

log_marg_low <- log_marg_low + log(prior_H)   # add log prior
log_marg_low <- log_marg_low - max(log_marg_low)  # stabilise
post_prob_low <- exp(log_marg_low) / sum(exp(log_marg_low))

log_marg_high <- c(
  H1 = log_marg_lik(beta1, se1, prior_sd_high, hypothesis_zero = FALSE) +  # Chol->Ca non-zero
       log_marg_lik(beta2, se2, prior_sd_high, hypothesis_zero = TRUE),    # Ca->Chol zero
  
  H2 = log_marg_lik(beta1, se1, prior_sd_high, hypothesis_zero = TRUE) +     # Chol->Ca zero
       log_marg_lik(beta2, se2, prior_sd_high, hypothesis_zero = FALSE),   # Ca->Chol non-zero
  
  H3 = log_marg_lik(beta1, se1, prior_sd_high, hypothesis_zero = TRUE) + 
       log_marg_lik(beta2, se2, prior_sd_high, hypothesis_zero = TRUE),
  
  H4 = log_marg_lik(beta1, se1, prior_sd_high, hypothesis_zero = FALSE) + 
       log_marg_lik(beta2, se2, prior_sd_high, hypothesis_zero = FALSE)
)

log_marg_high <- log_marg_high + log(prior_H)   # add log prior
log_marg_high <- log_marg_high - max(log_marg_high)  # stabilise
post_prob_high <- exp(log_marg_high) / sum(exp(log_marg_high))

# Create data frame
df <- data.frame(
  Hypothesis = c("H1: Cholesterol → Calcium", 
                 "H2: Calcium → Cholesterol", 
                 "H3: No causal effect", 
                 "H4: Bidirectional"),
  Probability = post_prob,
  Label = scales::percent(post_prob, accuracy = 1)
) %>% 
  arrange(desc(Probability)) %>%
  mutate(Hypothesis = factor(Hypothesis, levels = Hypothesis))
library(ggrepel)
# Beautiful bar plot
ggplot(df, aes(x = "", y = Probability, fill = Hypothesis)) +
  geom_col(width = 0.6, color = "white") +
  #geom_text_repel(aes(label = Label), 
  #          position = position_stack(vjust = 0.5), 
  #          size = 5, fontface = "bold", color = "white") +
  scale_fill_brewer(palette = "Set2") +
  coord_polar(theta = "y") +
  theme_void() +
  labs(title = "Posterior Probabilities \nof Causal Directions",
       subtitle = "Bayesian Mendelian Randomization Model Comparison",
       fill = "Hypothesis:") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 12, hjust = 0.5))

ggplot(df, aes(x = Probability, y = reorder(Hypothesis, Probability))) +
  geom_col(aes(fill = Hypothesis), width = 0.7) +
  geom_text(aes(label = Label), hjust = -0.2, size = 3, fontface = "bold") +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_fill_brewer(palette = "Set1") +
  theme_classic(base_size=16) +
  theme(legend.position = "none",
        axis.text.y = element_text(size = 12),
        plot.title = element_text(size = 14, face = "bold")) +
  labs(x = "Posterior Probability",
       y = "",
       title = "Posterior Probabilities \nof Competing Causal Hypotheses")
```

Thus, after integrating the bidirectional MR evidence with the prior belief favoring a cholesterol-to-calcium direction, the data strongly support a unidirectional causal effect of genetically predicted cholesterol on calcium levels (posterior probability = `r post_prob['H1']*100`%). Results were robust to prior sensitivity analyses varying the slab standard deviation between 0.05 and 0.20 (posterior probabilities `r round(post_prob_low['H1']*100,1)`-`r round(post_prob_high['H1']*100,1)`%).

### References

::: {#refs}
:::

## Session Information

```{r session-information}
sessionInfo()
```
