---
title: "Analysis of Diversity Outbred Strain eQTLs"
author: "Dave Bridges"
date: "May 21, 2024"
output:
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: yes
    toc: yes
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: yes
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
library(readr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)
library(broom)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```

# Purpose

# Experimental Details

This analyses the data analysed via GEMMA and provided in the various output folders.

# Data Entry

```{r data-entry}
# additive lmm data
scarb1.filename <- 'output/ENSMUSG00000037936.assoc.txt'
scarb1.additive.data <- read_tsv(scarb1.filename) %>%
  separate(rs, sep="_", into=c("chromosome","position","allele","alt"),remove=FALSE) %>%
  mutate(chromosome=factor(chromosome, c(1:19,"X"))) %>%
  mutate(position=as.integer(position))

#ENSMUSG00000005681 is apoa2
apoa2.filename <- 'output/ENSMUSG00000005681.assoc.txt'
apoa2.additive.data <- read_tsv(apoa2.filename) %>%
  separate(rs, sep="_", into=c("chromosome","position","allele","alt"),remove=FALSE) %>%
  mutate(chromosome=factor(chromosome, c(1:19,"X"))) %>%
  mutate(position=as.integer(position))

#ENSMUSG00000032083 is apoa1 on chromosome 9
apoa1.filename <- 'output/ENSMUSG00000032083.assoc.txt'
apoa1.additive.data <- read_tsv(apoa1.filename) %>%
  separate(rs, sep="_", into=c("chromosome","position","allele","alt"),remove=FALSE) %>%
  mutate(chromosome=factor(chromosome, c(1:19,"X"))) %>%
  mutate(position=as.integer(position))
```

# Additive Models

Ran the additive models using GEMMA, first using intercepts and additive covariates for diet and sex

$SNP = beta_1 SNP + \beta_2 Diet + \beta_3 Sex + \mu +\epsilon$

Where $$\epsilon$$ are the residuals and $$\mu$$ is the relationship matrix of the strains as defined by

## Scarb1

```{r scarb1-additive}
library(qqman)
qq(scarb1.additive.data$p_wald)
suggestive.pval <- 1E-5
genome.pval <- 5E-8

scarb1.additive.data %>%
  arrange(p_wald) %>% 
  filter(p_wald<genome.pval) %>%
  kable(caption="Genome-wide significant associations from mixed linear models for Scarb1 in additive model") 

scarb1.additive.data %>%
  arrange(p_wald) %>% 
  filter(p_wald<suggestive.pval) %>%
  mutate(position.start = substr(as.character(position), 1,2)) %>%
  group_by(chromosome,position.start) %>%
  summarize_all(.funs=first) %>%
  select(-position.start,-chr,-ps) %>%
  kable(caption="Suggestive genome-wide significant associations from mixed linear models for Scarb1 in additive model, clumped by first two digits of the position") 

scarb1.additive.data %>%
  arrange(p_wald) %>% 
  filter(p_wald<1E-4) %>%
  mutate(position.start = substr(as.character(position), 1,2)) %>%
  group_by(chromosome,position.start) %>%
  summarize_all(.funs=first) %>%
  select(-position.start,-chr,-ps) -> additive.snp.summary

additive.snp.summary %>%
  kable(caption="Relaxed suggestive genome-wide significant associations from mixed linear models for Scarb1 in additive model, clumped by first two digits of the position") 

library(ggmanh)
manhattan_plot(x = scarb1.additive.data, pval.colname = "p_wald", chr.colname = "chromosome", pos.colname = "position", plot.title = "eQTLs for Scarb1 (Additive Model)", y.label = "LOD Score")

snp.pos <- 118958519
library(forcats)
peak5 <- filter(scarb1.additive.data,
                  chromosome==5,
                  position>snp.pos-20000000,
                  position<snp.pos+20000000) %>%
  mutate(alt=fct_recode(as.factor(alt),
                        "C57BL/6J"="A",
                        "NZO"="B",
                        "Sv129"="C",
                        "PWK"="D",
                        "A/J"="E",
                        "NOD"="F",
                        "CAST"="G",
                        "WSB"="H"))

ggplot(data=peak5,
       aes(x=position,
       y=beta,
       col=alt,
       group=alt)) +
  geom_line() +
  geom_hline(yintercept=0,lty=2) +
  labs(title="Strain Specific Effects",
       y="Scarb1 Effect Size (Arbitrary)",
       x="Position on Chromosome 5") +
  scale_color_discrete(name="") +
  guides(col=guide_legend(ncol=2)) +
  theme_classic(base_size=12) +
  theme(legend.position=c(0.18,0.9),
        legend.text=element_text(size=8))
```

## Apoa2

```{r apoa2-additive}
qq(apoa2.additive.data$p_wald)
suggestive.pval <- 1E-5
genome.pval <- 5E-8

apoa2.additive.data %>%
  arrange(p_wald) %>% 
  filter(p_wald<genome.pval) %>%
  kable(caption="Genome-wide significant associations from mixed linear models for Apoa2 in additive model") 

apoa2.additive.data %>%
  arrange(p_wald) %>% 
  filter(p_wald<suggestive.pval) %>%
  mutate(position.start = substr(as.character(position), 1,2)) %>%
  group_by(chromosome,position.start) %>%
  summarize_all(.funs=first) %>%
  select(-position.start,-chr,-ps) %>%
  kable(caption="Suggestive genome-wide significant associations from mixed linear models for Apoa2 in additive model, clumped by first two digits of the position") 

apoa2.additive.data %>%
  arrange(p_wald) %>% 
  filter(p_wald<1E-4) %>%
  mutate(position.start = substr(as.character(position), 1,2)) %>%
  group_by(chromosome,position.start) %>%
  summarize_all(.funs=first) %>%
  select(-position.start,-chr,-ps) -> additive.snp.summary

additive.snp.summary %>%
  kable(caption="Relaxed suggestive genome-wide significant associations from mixed linear models for Apoa2 in additive model, clumped by first two digits of the position") 

manhattan_plot(x = apoa2.additive.data, pval.colname = "p_wald", chr.colname = "chromosome", pos.colname = "position", plot.title = "eQTLs for Apoa2 (Additive Model)", y.label = "LOD Score")

snp.pos <- 177024867
peak1b <- filter(apoa2.additive.data,
                  chromosome==1,
                  position>snp.pos-20000000,
                  position<snp.pos+20000000) %>%
  mutate(alt=fct_recode(as.factor(alt),
                        "C57BL/6J"="A",
                        "NZO"="B",
                        "Sv129"="C",
                        "PWK"="D",
                        "A/J"="E",
                        "NOD"="F",
                        "CAST"="G",
                        "WSB"="H"))

ggplot(data=peak1b,
       aes(x=position,
       y=beta,
       col=alt,
       group=alt)) +
  geom_line() +
  geom_hline(yintercept=0,lty=2) +
  labs(title="Strain Specific Effects",
       y="Apoa2 Effect Size (Arbitrary)",
       x="Position on Chromosome 1") +
  scale_color_discrete(name="") +
  guides(col=guide_legend(ncol=2)) +
  theme_classic(base_size=12) +
  theme(legend.position=c(0.18,0.9),
        legend.text=element_text(size=8))
```

This does not line up with the GWAS pattern, so is probably not the same hit.

## Apoa1

```{r apoa1-additive}
qq(apoa1.additive.data$p_wald)
suggestive.pval <- 1E-5
genome.pval <- 5E-8

apoa1.additive.data %>%
  arrange(p_wald) %>% 
  filter(p_wald<genome.pval) %>%
  kable(caption="Genome-wide significant associations from mixed linear models for Apoa2 in additive model") 

apoa1.additive.data %>%
  arrange(p_wald) %>% 
  filter(p_wald<suggestive.pval) %>%
  mutate(position.start = substr(as.character(position), 1,2)) %>%
  group_by(chromosome,position.start) %>%
  summarize_all(.funs=first) %>%
  select(-position.start,-chr,-ps) %>%
  kable(caption="Suggestive genome-wide significant associations from mixed linear models for Apoa2 in additive model, clumped by first two digits of the position") 

apoa1.additive.data %>%
  arrange(p_wald) %>% 
  filter(p_wald<1E-4) %>%
  mutate(position.start = substr(as.character(position), 1,2)) %>%
  group_by(chromosome,position.start) %>%
  summarize_all(.funs=first) %>%
  select(-position.start,-chr,-ps) -> additive.snp.summary

additive.snp.summary %>%
  kable(caption="Relaxed suggestive genome-wide significant associations from mixed linear models for apoa1 in additive model, clumped by first two digits of the position") 

manhattan_plot(x = apoa1.additive.data, pval.colname = "p_wald", chr.colname = "chromosome", pos.colname = "position", plot.title = "eQTLs for apoa1 (Additive Model)", y.label = "LOD Score")

snp.pos <- 45695998
peak9 <- filter(apoa1.additive.data,
                  chromosome==9,
                  position>snp.pos-20000000,
                  position<snp.pos+20000000) %>%
  mutate(alt=fct_recode(as.factor(alt),
                        "C57BL/6J"="A",
                        "NZO"="B",
                        "Sv129"="C",
                        "PWK"="D",
                        "A/J"="E",
                        "NOD"="F",
                        "CAST"="G",
                        "WSB"="H"))

ggplot(data=peak9,
       aes(x=position,
       y=beta,
       col=alt,
       group=alt)) +
  geom_line() +
  geom_hline(yintercept=0,lty=2) +
  labs(title="Strain Specific Effects",
       y="Apoa1 Effect Size (Arbitrary)",
       x="Position on Chromosome 9") +
  scale_color_discrete(name="") +
  guides(col=guide_legend(ncol=2)) +
  theme_classic(base_size=12) +
  theme(legend.position=c(0.18,0.9),
        legend.text=element_text(size=8))
```

# Session Information

```{r session-information, echo=T}
sessionInfo()
```

