---
title: "Exposure → Survival Screen"
subtitle: "Does each exposure instrument set have an edge into the survival collider?"
author: "Dave Bridges and Katie Kittell"
date: today
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  pdf: default
knitr:
  opts_chunk:
    fig.path: "figures/"
    dev: ["png", "pdf"]  # Remove !expr, just use array syntax
    fig.keep: "all"
    autodep: true        # invalidate cached chunks when upstream vars (e.g. config) change
execute:
  echo: true
  warning: false
  cache: true
---

## Purpose

In an Alzheimer's-disease MR the outcome GWAS is measured only in people who
**survived to old age to be assessed**, so the conditioning variable — the
collider — is *survival* `S`, not AD incidence. Conditioning on `S` opens the
path shown below.

```{r dag-collider, fig.width=7.5, fig.height=4.5}
library(ggdag)     # install.packages(c("ggdag", "dagitty"))
library(ggplot2)

collider_dag <- dagify(
  S  ~ G + AD + CR,     # survival is caused by the exposure, AD, and competing risks
  AD ~ U,               # AD and competing risks share etiology (latent U)
  CR ~ U,
  exposure = "G",
  outcome  = "AD",
  labels = c(
    G  = "G_exposure\n(instrument)",
    S  = "S = survival\n(conditioned on = collider)",
    AD = "AD",
    CR = "competing risks\n(IHD, stroke, cancer)",
    U  = "shared etiology"
  ),
  coords = list(
    x = c(G = 0, S = 2, AD = 4, CR = 2.9, U = 4),
    y = c(G = 0, S = 0, AD = 0, CR = 1.3, U = 1.6)
  )
)

ggdag(collider_dag, text = FALSE, use_labels = "label", node_size = 16) +
  theme_dag(base_size = 11) +
  labs(
    title    = "Survival as the collider in an AD Mendelian randomization",
    subtitle = "Conditioning on S (studying survivors) opens the G_exposure — AD path"
  )
```

The known edge is `AD → S` (AD hastens death); competing risks feed `S` and
share etiology with AD via a latent common cause `U`. Conditioning on `S`
induces a spurious `G_exposure`–AD association. **That triangle only closes
if the exposure instruments have an edge into survival** (`G → S`). This
notebook estimates that edge directly: an MR of each exposure on a
survival/longevity GWAS. A non-null effect means survival-collider bias is a
live threat for that exposure in the downstream AD analysis and must be
corrected; a near-null effect means that instrument set is comparatively safe
(with caveats at the bottom).

## Setup

You need an OpenGWAS JWT (anonymous access is gone). Get one at
<https://api.opengwas.io>, then add `OPENGWAS_JWT=<token>` to `~/.Renviron`
and restart R. Treat it like a password — don't commit it.

```{r setup, cache=FALSE}
# cache=FALSE is required: a cached setup chunk is restored WITHOUT re-running library(), so on
# re-render the packages would not be attached ("could not find function ...").
library(TwoSampleMR)   # install.packages("TwoSampleMR",
                       #   repos = c("https://mrcieu.r-universe.dev",
                       #             "https://cloud.r-project.org"))
library(ieugwasr)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(readr)

# Fail fast if the token isn't wired up
stopifnot(nchar(ieugwasr::get_opengwas_jwt()) > 0)
ieugwasr::user()   # prints your account + remaining allowance
```

## Configuration — EDIT THIS BLOCK

```{r config}
# ---- Exposures: the risk-factor / nAChR instruments you'll use against AD. ----
# Named vector: names are display labels, values are OpenGWAS ids.
# Add new instruments by appending "label" = "id" — everything downstream fans out.
exposures <- c(
  "Smoking (GSCAN, ieu-b-142)" = "ieu-b-142",  # <-- example; append more here
  "LDL-C"                    = "ieu-b-110"
  # "CHRNA5 region cis-eQTL"   = "eqtl-a-XXXX",
)

# ---- AD GWAS: the downstream outcome; here used to size the AD -> S arm. ----
# Two contrasting ascertainments so we can see if the AD->S edge is proxy-driven.
ad_gwas <- c(
  "AD — Bellenguez 2022 (proxy-incl.)" = "ebi-a-GCST90027158",  # 39k clin + 47k proxy; largest, selection-prone
  "AD — Kunkle 2019 IGAP (clinical)"   = "ieu-b-2"              # 22k clinical cases, no UKB/proxy: clean comparator
)
PRIMARY_AD <- "ieu-b-2"   # reference AD->S edge for the induced-bias direction (clean, clinical)

# ---- Longevity / survival outcomes = the collider S. ----
# All four are Timmers et al. 2019 parental-lifespan phenotypes. The combined-
# parents Martingale-residual analysis (GCST006697) is the best-powered headline
# lifespan phenotype and the PRIMARY here; the others are sensitivity swaps.
survival_outcomes <- c(
  "Parental lifespan (combined, Martingale resid)" = "ebi-a-GCST006697",  # PRIMARY: N=389,166
  "Combined parental age at death"                 = "ebi-a-GCST006702",  # N=208,118
  "Father's age at death"                          = "ebi-a-GCST006700",  # N=317,652
  "Mother's attained age"                          = "ebi-a-GCST006696"   # N=412,937
)
PRIMARY_SURVIVAL <- "ebi-a-GCST006697"   # headline longevity outcome

# ---- ROPE / Bayesian settings ----
# ROPE = region of practical equivalence, expressed in SD of the longevity outcome.
# The MR slope is the causal effect per unit exposure in per-SD-outcome units, so
# ROPE = 0.01 reads: "a shift in lifespan smaller than 0.01 SD is practically null
# for collider purposes." Deliberately small — this is a bias-relevant threshold,
# not a Cohen 'small-effect' cutoff: even a tiny survival edge opens the collider,
# and kin-proxy lifespan GWAS already ~halve per-allele effects.
# (Effects are read as per-SD of the outcome; rescale if an outcome GWAS you add
#  reports in non-SD units such as years.)
ROPE     <- 0.01    # SD of the longevity outcome
PRIOR_SD <- 0.50    # weakly-informative prior effect ~ N(0, PRIOR_SD^2); ~flat at this scale

# ---- Instrument / clumping parameters ----
P_THRESH <- 5e-8
R2       <- 0.001
KB       <- 10000
```

## Discovery — find survival/longevity ids

Run once to locate candidate datasets in the OpenGWAS catalogue, paste the ids
into `survival_outcomes` above, then set `eval: false` again.

```{r discover-survival, eval=FALSE}
ao <- TwoSampleMR::available_outcomes()   # full catalogue (large)

ao |>
  filter(grepl("lifespan|longevity|survival|age at death|parental|healthspan",
               trait, ignore.case = TRUE)) |>
  select(id, trait, sample_size, nsnp, consortium, author, year) |>
  arrange(desc(year)) |>
  print(n = 50)
```

## Screening function

One exposure × one survival outcome → one tidy row. Robust to small instrument
sets (falls back to Wald ratio) and to failed queries (returned as `NULL` by
`possibly()` in the runner).

```{r screen-fn}
screen_one <- function(exp_id, exp_label, out_id, out_label,
                       p = P_THRESH, r2 = R2, kb = KB) {

  # 1. GW-significant, LD-clumped instruments for the exposure
  inst <- extract_instruments(exp_id, p1 = p, r2 = r2, kb = kb)
  if (is.null(inst) || nrow(inst) == 0) stop("no instruments: ", exp_id)

  # 2. look those SNPs up in the survival GWAS
  out <- extract_outcome_data(snps = inst$SNP, outcomes = out_id)
  if (is.null(out) || nrow(out) == 0) stop("no overlap: ", out_id)

  # 3. harmonise (action = 2: align, drop ambiguous palindromes by frequency)
  dat <- harmonise_data(inst, out, action = 2) |> filter(mr_keep)
  if (nrow(dat) == 0) stop("nothing kept after harmonisation")

  # 4. MR — default method set adapts to the number of SNPs
  res     <- mr(dat)
  pleio_p <- tryCatch(mr_pleiotropy_test(dat)$pval, error = \(e) NA_real_)

  # pick a primary estimate by availability/priority
  primary <- res |>
    mutate(rank = match(method, c("Inverse variance weighted", "Wald ratio",
                                  "Weighted median", "MR Egger"))) |>
    arrange(rank) |> slice(1)

  # mean instrument strength (approx per-SNP F)
  mean_F <- mean((dat$beta.exposure / dat$se.exposure)^2, na.rm = TRUE)

  tibble(
    exposure    = exp_label,
    survival    = out_label,
    n_snp       = primary$nsnp,
    method      = primary$method,
    b           = primary$b,
    se          = primary$se,
    p           = primary$pval,
    mean_F      = mean_F,
    egger_int_p = pleio_p,
    b_wmedian   = res$b[match("Weighted median", res$method)],
    b_egger     = res$b[match("MR Egger",        res$method)]
  )
}
```

## The AD → longevity arm (the "known" edge)

The collider needs *both* arms. Before screening exposures, pin down the arm the
triangle cannot work without: `AD → S`. If AD-liability instruments do not move
lifespan there is no collider and nothing to correct. We MR each AD GWAS onto the
longevity outcomes; a **negative** effect (higher AD liability → shorter parental
lifespan) confirms the collider node is real. Running both AD GWAS shows whether
that edge is genuine or an artefact of Bellenguez's UK-Biobank by-proxy cases —
Kunkle/IGAP is clinically ascertained with no proxy component as the clean check.

```{r bayes-fn}
# Closed-form Bayesian posterior for one MR estimate.
# Conjugate normal: prior effect ~ N(0, tau^2); likelihood b ~ N(effect, se^2).
# Returns posterior summaries and the ROPE decision quantities.
bayes_rope <- function(b, se, rope = ROPE, tau = PRIOR_SD) {
  if (is.na(b) || is.na(se) || se <= 0)
    return(tibble(post_mean = NA_real_, post_sd = NA_real_,
                  p_out_rope = NA_real_, p_direction = NA_real_))
  prec_post <- 1 / se^2 + 1 / tau^2
  mu_post   <- (b / se^2) / prec_post
  sd_post   <- sqrt(1 / prec_post)
  # posterior mass with |effect| > ROPE (i.e. effect is practically non-null)
  p_out <- pnorm(-rope, mu_post, sd_post) +
           pnorm( rope, mu_post, sd_post, lower.tail = FALSE)
  # probability of direction: how sure we are of the sign
  p_dir <- max(pnorm(0, mu_post, sd_post),
               pnorm(0, mu_post, sd_post, lower.tail = FALSE))
  tibble(post_mean = mu_post, post_sd = sd_post,
         p_out_rope = p_out, p_direction = p_dir)
}
```

The API pull (expensive) is cached; the ROPE/Bayesian layer that reads it is
**not** cached, so retuning `ROPE` always re-computes without re-hitting OpenGWAS.

```{r ad-pull}
safe_screen <- possibly(screen_one, otherwise = NULL)

ad_grid <- expand_grid(
  tibble(exp_id = unname(ad_gwas),           exp_label = names(ad_gwas)),
  tibble(out_id = unname(survival_outcomes), out_label = names(survival_outcomes))
)

ad_raw <- pmap(ad_grid, safe_screen) |>
  compact() |>
  list_rbind()
```

```{r ad-longevity, cache=FALSE}
ad_survival <- bind_cols(
  ad_raw,
  pmap(list(ad_raw$b, ad_raw$se), bayes_rope) |> list_rbind()
)

ad_survival |>
  select(`AD GWAS` = exposure, survival, n_snp, method, b, se, p,
         mean_F, `P(|AD→S|>ROPE)` = p_out_rope) |>
  mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
  knitr::kable(caption = "AD instruments → parental lifespan: the AD → S arm.")
```

The primary contrast (both AD GWAS against the headline longevity outcome):

```{r ad-longevity-primary, cache=FALSE}
primary_label <- names(survival_outcomes)[match(PRIMARY_SURVIVAL, survival_outcomes)]

ad_survival |>
  filter(survival == primary_label) |>
  transmute(`AD GWAS` = exposure, n_snp, b = signif(b, 3), se = signif(se, 3),
            p = signif(p, 3), `P(|AD→S|>ROPE)` = round(p_out_rope, 3)) |>
  knitr::kable(caption = paste0("AD → ", primary_label,
                                " (the known collider edge)."))
```

A high `P(|AD→S| > ROPE)` with a negative `b` in both GWAS establishes the
`AD → S` arm empirically, so the collider node is live. Whether *bias* reaches
any given exposure then hinges entirely on the `E → S` arm screened next.

## Run the screen

```{r run}
grid <- expand_grid(
  tibble(exp_id = unname(exposures),         exp_label = names(exposures)),
  tibble(out_id = unname(survival_outcomes), out_label = names(survival_outcomes))
)

safe_screen <- possibly(screen_one, otherwise = NULL)

results <- pmap(grid, safe_screen) |>
  compact() |>
  list_rbind()

results |>
  select(exposure, survival, n_snp, method, b, se, p, mean_F, egger_int_p) |>
  mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
  knitr::kable(caption = "Exposure → survival screen: primary MR estimate per pair.")
```

## Flag the collider edge

```{r interpret, cache=FALSE}
results <- results |>
  mutate(
    edge = case_when(
      p < 0.05 / n() ~ "STRONG survival edge (Bonferroni)",
      p < 0.05       ~ "nominal survival edge",
      TRUE           ~ "no clear edge"),
    pleiotropy = if_else(!is.na(egger_int_p) & egger_int_p < 0.05,
                         "Egger intercept p<0.05 — directional pleiotropy", "")
  )

results |>
  select(exposure, survival, n_snp, method, b, se, p, mean_F, edge, pleiotropy) |>
  mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
  knitr::kable(caption = "Frequentist edge flags (Bonferroni / nominal) and pleiotropy check.")
```

## Forest of exposure → survival effects

Intervals clear of zero ⇒ the `G → S` edge is present ⇒ survival-collider bias
is live for that exposure in the AD MR.

```{r plot, fig.width=8, fig.height=4}
results |>
  mutate(lab = paste(exposure, survival, sep = " → ")) |>
  ggplot(aes(b, reorder(lab, b))) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
  geom_pointrange(aes(xmin = b - 1.96 * se, xmax = b + 1.96 * se)) +
  labs(
    x = "IVW/Wald effect of exposure on survival",
    y = NULL,
    title = "Step 1: does each exposure have an edge into the survival collider?"
  ) +
  theme_minimal(base_size = 12)
```

## What collider bias is present — and is the E → C arm live?

The downstream MR regresses AD liability on exposure instruments, but AD is only
ever observed in people who **survived to be assessed**. Survival `S` is a
collider on two arms: `AD → S` (established above) and, if present, `E → S`.
Selecting on survivors conditions on `S`, opening the non-causal
`G_exposure — AD` path and biasing the exposure→AD estimate — typically **toward
spurious protection** (index-event / survival bias). Because the `AD → S` arm is
already confirmed, the bias is *live for a given exposure exactly when its
`E → S` (i.e. `E → C`) arm is present*. The table below nominates that,
conditional on the collider node being real.

For each exposure×outcome we combine the frequentist IVW p-value with a
closed-form Bayesian posterior (normal-conjugate, prior `N(0, PRIOR_SD²)`) and
report the probability the effect lies **outside the ROPE** — i.e. is
practically non-null — and the probability of direction.

**Direction of the induced bias.** Existence of the `E → C` arm says the collider
is open; the *sign* of the spurious `G_exposure`–AD association it creates is set
by the **product** of the two arms measured on the *same* outcome. For a linear
collider `S = a·G_exp + c·AD`, conditioning on `S` gives
`Cov(G_exp, AD | S) ∝ −a·c`, and that sign is **invariant to how the lifespan
outcome is coded** (flipping `S` flips both `a` and `c`, leaving `a·c`
unchanged) — so we can read it straight off the betas on the primary outcome:

- `b_E · b_AD > 0` (exposure and AD push survival the *same* way) → induced
  association is **negative** → the exposure's risk alleles look *protective* for
  AD in survivors: **spurious protection** (the artifact that would fake a
  druggable protective AD signal).
- `b_E · b_AD < 0` → induced association is positive → **spurious harm**.

The reference `AD → S` edge is the clean clinical GWAS (`PRIMARY_AD`); both AD
GWAS agree in sign on the primary outcome, so the direction is robust to that
choice. The label is only meaningful where the `E → C` arm is present — magnitude
still scales with `|b_E · b_AD|`, so a within-ROPE arm induces negligible bias
regardless of sign.

```{r verdict, cache=FALSE}
primary_label <- names(survival_outcomes)[match(PRIMARY_SURVIVAL, survival_outcomes)]
ad_ref_label  <- names(ad_gwas)[match(PRIMARY_AD, ad_gwas)]

# Reference AD -> S beta on the primary outcome (sign sets the induced-bias direction)
b_ad_primary <- ad_raw |>
  filter(survival == primary_label, exposure == ad_ref_label) |>
  pull(b) |> (\(x) if (length(x)) x[1] else NA_real_)()

bayes_cols <- pmap(list(results$b, results$se), bayes_rope) |> list_rbind()

verdict <- bind_cols(results, bayes_cols) |>
  mutate(
    E_to_C = case_when(
      p_out_rope >= 0.95 & p_direction >= 0.975 ~ "PRESENT — edge into collider",
      p_out_rope <= 0.05                        ~ "absent — within ROPE (≈ null)",
      TRUE                                      ~ "indeterminate"
    ),
    # Induced-bias direction = sign of b_E * b_AD on the SAME (primary) outcome.
    # Only defined against the primary outcome; NA elsewhere.
    bias_product   = if_else(survival == primary_label, b * b_ad_primary, NA_real_),
    induced_bias = case_when(
      E_to_C == "absent — within ROPE (≈ null)" ~ "negligible (arm ≈ null)",
      is.na(bias_product)                       ~ NA_character_,
      bias_product > 0 ~ "spurious PROTECTION (mimics ↓AD)",
      bias_product < 0 ~ "spurious HARM (mimics ↑AD)",
      TRUE             ~ "—"
    )
  )

# Reported only against the primary outcome (full verdict saved below).
verdict |>
  filter(survival == primary_label) |>
  arrange(desc(p_out_rope)) |>
  transmute(
    exposure, n_snp,
    b = signif(b, 3), se = signif(se, 3), p_ivw = signif(p, 3),
    p_dir = round(p_direction, 3),
    `P(|E→S|>ROPE)` = round(p_out_rope, 3),
    `E → C arm` = E_to_C,
    `induced bias` = induced_bias
  ) |>
  knitr::kable(
    caption = paste0("Conditional collider verdict for ", primary_label,
                     " (ROPE = ±", ROPE, " SD). Induced-bias direction ",
                     "uses the reference AD→S edge (", ad_ref_label, ", b = ",
                     signif(b_ad_primary, 3), "). The collider is live where the ",
                     "E → C arm is PRESENT; its sign is the 'induced bias' column.")
  )
```

## Save

```{r save}
write_csv(results,     "exposure_survival_screen.csv")
write_csv(verdict,     "exposure_survival_verdict.csv")
write_csv(ad_survival, "ad_longevity_arm.csv")
```

## How to read this — and what it does *not* rule out

- **Non-null effect** → the exposure moves survival, the collider triangle
  closes, and you must carry a survival-bias correction into the AD analysis
  for that exposure (feed the survival GWAS as the "incidence" input to
  SlopeHunter, or estimate `b` via MR of survival→AD, or MVMR on competing
  risk). For your smoking instruments a strong edge is the expected /
  worst case, since smoking-quantity variants drive mortality hard.
- **Near-null effect** → that instrument set is comparatively insulated, which
  *strengthens* a druggable-pharmacology reading of any protective AD signal.

Three caveats before over-trusting a null:

1. **Parental-lifespan attenuation.** Kin-based lifespan GWAS measure the
   *parents'* survival through offspring genotype, so per-allele effects are
   roughly halved (and on a years-of-life / hazard scale). Treat magnitudes as
   conservative; a null here is a weaker null than it looks.
2. **Competing risk still bites.** An exposure with a small *net* survival
   effect can still bias the AD estimate if it kills through pathways sharing
   etiology with AD (IHD, stroke). A clean survival screen does not fully
   exonerate — consider MVMR conditioning on cardiovascular liability.
3. **The survival GWAS is itself measured in survivors**, an imperfect proxy
   for the exact selection your AD cohort underwent.

- **Allowance:** each exposure×outcome pair is two API calls. Keep the lists
  modest or add `Sys.sleep()`; if you hit rate limits, clump locally
  (`ieugwasr::ld_clump_local()` with a plink binary + 1000G panel) after
  `extract_instruments(..., clump = FALSE)`.
