---
title: "MR mediation: exposure → mediator → AD"
subtitle: "Direct and indirect effects via two-step MR and MVMR, with survival-collider adjustment carried through every leg"
author: "Dave Bridges and Katie Kittell"
date: today
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  pdf: default
knitr:
  opts_chunk:
    fig.path: "figures/"
    dev: ["png", "pdf"]
    fig.keep: "all"
execute:
  echo: true
  warning: false
  message: false
  # No knitr cache: knitr keys caches on chunk code only, so config edits are silently replayed
  # from stale caches (see the calcium arm's commit 034c09a). Every OpenGWAS pull is instead
  # cached to data/cache/mediation/ by cache_rds(), keyed on its full inputs.
  cache: false
---

## Purpose

Steps 1–2 estimated each exposure's **total** effect on AD and corrected it for the survival
collider. This step asks how much of a total effect runs **through a named mediator**. The
motivating triad is

$$\text{lean mass} \;\to\; \text{2-h glucose} \;\to\; \text{AD},$$

with the hypothesis that part of lean mass's protective effect on AD comes from lower
postprandial glucose. The code is written for any number of triads, including exposure →
tissue expression → AD (see [eQTL mediators](#eqtl-mediators)).

**Primary estimands:** the **indirect** effect (product of coefficients, $\beta_{X\to M}\times\beta_{M\to Y\mid X}$)
and the **direct** effect ($\beta_{X\to Y\mid M}$ from MVMR), each with a Monte Carlo CI.
Percent mediated is secondary and is reported only for triads that pass the
[pre-specified gates](#gates). Methods follow Carter et al. 2021 (*Eur J Epidemiol*,
PMID 33961203); conditional F and Q follow Sanderson, Spiller & Bowden 2021 (*Stat Med*
40:5434–52, doi:10.1002/sim.9133, PMID 34338327).

```{r dag, fig.width=7.5, fig.height=3.8}
library(ggdag)
library(ggplot2)

dagify(M ~ X, Y ~ X + M, S ~ X + Y,
       labels = c(X = "Lean mass (X)", M = "2-h glucose (M)", Y = "AD (Y)",
                  S = "Survival (collider)"),
       coords = list(x = c(X = 0, M = 2, Y = 4, S = 2), y = c(X = 0, M = 1, Y = 0, S = -1))) |>
  ggdag(text = FALSE, use_labels = "label", node_size = 14) +
  theme_dag(base_size = 11) +
  labs(title = "Mediation triad with the survival collider from Steps 1–2",
       subtitle = "direct = X → Y | M;  indirect = (X → M) × (M → Y | X)")
```

## Setup

```{r setup}
library(TwoSampleMR)
library(ieugwasr)
library(MVMR)            # conditional F and Q (Sanderson-Windmeijer)
library(MRPRESSO)
library(mr.raps)
library(dplyr)
library(tidyr)
library(purrr)
library(readr)
library(ggplot2)
# GRAPPLE (multivariable MR-RAPS) is needed only if a triad's method_MVMR is "grapple":
# remotes::install_github("jingshuw/GRAPPLE"). Without it that leg falls back to MV-IVW and the
# method_mvmr column says so.

source("R/mr_mediation_helpers.R")

# A token *string* is not enough: Step 2's naive pull silently returned nothing when the token
# had expired. user() errors on an invalid token, so this fails at the top instead.
stopifnot(nchar(ieugwasr::get_opengwas_jwt()) > 0)
invisible(ieugwasr::user())
```

## Configuration — EDIT THIS BLOCK {#config}

### Datasets

Every OpenGWAS ID here comes from the existing pipeline or the spec; none is new. The
covariate columns record **what the source paper says** about the adjustments in each GWAS,
because an exposure or mediator GWAS adjusted for a heritable covariate opens a collider path
in MR (and in MVMR the mediator's adjustment matters as much as the exposure's).

```{r config-datasets}
datasets <- tibble::tribble(
  ~id,                  ~label,                                   ~role,       ~adjusted_for,            ~adjustment_source,
  "ebi-a-GCST90000025", "Appendicular lean mass (Pei 2020, UKB)", "exposure",  "appendicular fat mass",  "Pei 2020 Commun Biol (PMID 33097823) Methods: covariates included AFM; height deliberately excluded",
  "ebi-a-GCST90002227", "2-h glucose (MAGIC, Chen 2021, EUR)",    "mediator",  "BMI",                    "Chen 2021 Nat Genet (PMID 34059833): 'FG, 2hGlu and FI analyses were adjusted for BMI'",
  "ebi-a-GCST90027158", "AD: Bellenguez 2022 (proxy + clinical)", "outcome",   NA,                       NA,
  "ieu-b-2",            "AD: Kunkle 2019 IGAP (clinical only)",   "outcome",   NA,                       NA,
  "ebi-a-GCST006697",   "Parental lifespan (Pilling, Martingale)","selection", NA,                       NA
)
LIFESPAN_ID <- "ebi-a-GCST006697"   # selection axis; MUST match the axis b_SH was fit on (Step 2)
```

::: {.callout-warning}
**The motivating triad's exposure and mediator GWAS both fail the spec's covariate
requirements.** The spec asks for lean mass with no fat-mass/BMI/height adjustment and a
BMI-*unadjusted* 2hGlu. The IDs used by the existing pipeline are:

- `ebi-a-GCST90000025`: **fat-mass-adjusted** appendicular lean mass (Pei 2020).
- `ebi-a-GCST90002227`: **BMI-adjusted** 2hGlu. Chen 2021 adjusted FG, FI and 2hGlu for BMI,
  and the GWAS Catalog lists no unadjusted European 2hGlu from that paper (`GCST90002225`–`29`
  are the per-ancestry and trans-ancestry versions of the same adjusted analysis).

Both adjustments condition on adiposity, which lean mass and glucose each affect. So the
X → M leg, and the M | X coefficient, can carry collider bias in either direction. Pei 2020
reports that its adjusted ALM has its strongest negative genetic correlation with
*BMI-adjusted* 2hGlu, which is the exact pair used here. The notebook runs as specified, but
every lean-mass triad is labelled `COLLIDER RISK` until unadjusted replacements are supplied.

A second data note: the 2hGlu file's `beta`/`standard_error` are from the **raw-trait**
(mmol/L) model, but its `p_value` is from the **inverse-normalised** model (Catalog
`README.txt`). Instruments are therefore selected on a p-value that does not come from the
reported beta. Both legs that use M are still on the same mmol/L scale, so the indirect effect's
units are internally consistent. The [p-value cross-check](#data-checks) quantifies the mismatch.
:::

### Triads

One row per **triad × outcome**. `method_XY_total` is copied from Step 2's validated
`naive_model`. `method_XM` was **not** covered by the prior assumption checks (lean mass → 2hGlu
was never run as a univariable association), so it is provisional: the
[X → M checks](#xm-checks) below exist to confirm or replace it. `method_MVMR` is the
pre-specified multivariable analog of the leg methods:
`ivw_mre → mv_ivw_mre`, `mr_presso → mv_presso`, `mr_raps → grapple`.

```{r config-triads}
triads <- tibble::tribble(
  ~triad_id,      ~exposure_id,         ~mediator_id,         ~outcome_id,          ~method_XM, ~method_XY_total, ~method_MVMR,  ~outcome_adjustment, ~mediator_type, ~inst_override, ~method_note,
  "leanmass_2hglu", "ebi-a-GCST90000025", "ebi-a-GCST90002227", "ebi-a-GCST90027158", "ivw_mre",  "ivw_mre",        "mv_ivw_mre",  "slopehunter",       "trait",        list(),         "XY: Step 2 naive_model (IVW-MRE). XM: PROVISIONAL, no prior check.",
  "leanmass_2hglu", "ebi-a-GCST90000025", "ebi-a-GCST90002227", "ieu-b-2",            "ivw_mre",  "ivw_mre",        "mv_ivw_mre",  "none",              "trait",        list(),         "XY: Step 2 naive_model (IVW-MRE). XM: PROVISIONAL, no prior check."
) |>
  mutate(
    # Pre-specified hypothesis signs (lean mass lowers glucose, glucose raises AD, lean mass
    # lowers AD). Reported against the estimates; they do not enter any gate.
    hyp_sign_xm = -1, hyp_sign_my = +1, hyp_sign_total = -1,
    # Equivalence bound for the indirect effect (log-OR per unit X). PRE-SPECIFY BEFORE RUNNING.
    # = 10% of |Step 2 naive lean mass -> AD (IVW-MRE)|, per outcome, from exposure_ad_naive.csv
    # (Bellenguez -0.1009, Kunkle -0.1299). Fixed numbers, not read from the CSV, so a re-run of
    # Step 2 cannot silently move a pre-specified bound.
    equiv_bound_indirect = case_when(
      outcome_id == "ebi-a-GCST90027158" ~ 0.10 * 0.1009,   # 0.0101
      outcome_id == "ieu-b-2"            ~ 0.10 * 0.1299,   # 0.0130
      TRUE                               ~ NA_real_)
  )

# Survival-bias adjustment per outcome follows Step 2: Bellenguez's b_SH is applied; Kunkle's
# fit is not identified (raw r = -0.0008), so Kunkle is "none" in the primary analysis.
```

### Selection slope, Monte Carlo and gates

```{r config-settings}
# b_SH per AD GWAS, exactly as Step 2 consumes it (same files, same `trusted` gate).
# `boot_rds` may point to saved SlopeHunter bootstrap replicates (fit$Bts.est). The fits CSV
# stores only b_SH and its bootstrap SE, so until those are saved the Monte Carlo draws b_SH
# from N(b_SH, se_SH^2) and every result row records that in `slope_source`.
bsh_sources <- tibble::tribble(
  ~ad_id,               ~fits_csv,                                          ~fit_set,              ~trusted, ~reason,                                                  ~boot_rds,
  "ebi-a-GCST90027158", "smoking/results/indexevent_slopehunter_fits.csv",  "excl_APOE (PRIMARY)", TRUE,     "validated in the smoking pipeline",                      NA_character_,
  "ieu-b-2",            "R/fits/kunkle_lifespan_slopehunter_fits.csv",      "excl_APOE (PRIMARY)", FALSE,    "not identified: raw r = -0.0008, sign agreement 0.499",  NA_character_
)

SETTINGS <- list(
  n_draws             = 5000,   # Monte Carlo draws per arm
  n_grid              = 21,     # b_SH grid points at which the legs are refit
  grid_sd             = 4,      # grid spans b_SH +/- 4 SE
  n_boot_cor          = 300,    # SNP-level bootstrap reps for the leg correlation matrix
  seed                = 20261001,
  cond_f_min          = 10,     # gate: conditional F for X and for M
  k_min_x_in_mediator = 10,     # eQTL mediators: skip if fewer X instruments are present
  common_snp_set      = TRUE,   # raw arm uses the same SNPs as the adjusted arm (see below)
  presso_nboot        = 1000,
  grapple_loss        = "huber",# matches TwoSampleMR's mr_raps default for the univariable legs
  gencov              = 0,      # X/M sample overlap for conditional F (Pei UKB vs MAGIC: none)
  sampler             = sample_legs_mvn   # Bayesian hook; see R/mr_mediation_helpers.R
)

# The 2 x 2 sensitivity grid: every triad x outcome under both adjustment states.
ADJUSTMENTS <- c("slopehunter", "none")

# Overrides the pre-specification check below. Leave FALSE: the equivalence bound exists to be
# chosen before the indirect effect is seen.
ALLOW_UNSET_EQUIV_BOUND <- FALSE

CACHE_DIR <- "data/cache/mediation"   # gitignored with the rest of data/cache
```

`common_snp_set = TRUE` means the **raw** arm also drops SNPs that have no lifespan
association. Raw and SlopeHunter-adjusted estimates then share an identical SNP set, so the
difference between them reflects the adjustment alone and not a change in instruments. With
`FALSE`, the raw arm reproduces the prior univariable instrument set exactly.

```{r config-checks}
# Fail before any API call if the config is internally inconsistent.
stopifnot(
  all(triads$method_XM        %in% UV_METHODS),
  all(triads$method_XY_total  %in% UV_METHODS),
  all(triads$method_MVMR      %in% MV_METHODS),
  all(triads$outcome_adjustment %in% ADJUSTMENTS),
  all(triads$mediator_type    %in% c("trait", "eqtl")),
  all(c(triads$exposure_id, triads$mediator_id, triads$outcome_id) %in% datasets$id),
  !anyDuplicated(triads[, c("triad_id", "outcome_id")])
)
# Rows sharing a triad_id must share X, M and instrument parameters (instruments are pulled once
# per triad, then reused for every outcome).
triad_keys <- triads |>
  group_by(triad_id) |>
  summarise(n_xm = n_distinct(paste(exposure_id, mediator_id)),
            n_par = n_distinct(map_chr(inst_override, \(o) paste(deparse(o), collapse = ""))))
if (any(triad_keys$n_xm > 1 | triad_keys$n_par > 1))
  stop("a triad_id maps to more than one exposure/mediator pair or instrument override")

if (any(is.na(triads$equiv_bound_indirect)) && !ALLOW_UNSET_EQUIV_BOUND)
  stop("equiv_bound_indirect is not set for: ",
       paste(unique(triads$triad_id[is.na(triads$equiv_bound_indirect)]), collapse = ", "),
       ". Pre-specify it in config-triads before running (or set ALLOW_UNSET_EQUIV_BOUND <- TRUE ",
       "to run without an equivalence test, which is then reported as 'bound not set').")
```

## Dataset checks {#data-checks}

Confirms each ID resolves to the trait we think it does, and that the mediator is
European-only. Covariate adjustment is not exposed by the API, so it is taken from the
`datasets` table above (with its source) and turned into a per-triad flag.

```{r gwasinfo}
info <- ieugwasr::gwasinfo(datasets$id)
info |>
  select(any_of(c("id", "trait", "author", "year", "pmid", "population", "sample_size",
                  "ncase", "ncontrol", "unit"))) |>
  knitr::kable(caption = "OpenGWAS metadata for every dataset in the config.")

med_pop <- info |> filter(id %in% triads$mediator_id) |> pull(population)
if (any(!grepl("European", med_pop)))
  warning("a mediator GWAS is not European-only: ", paste(med_pop, collapse = ", "))

covariate_flags <- triads |>
  distinct(triad_id, exposure_id, mediator_id) |>
  left_join(datasets |> select(exposure_id = id, x_adj = adjusted_for), by = "exposure_id") |>
  left_join(datasets |> select(mediator_id = id, m_adj = adjusted_for), by = "mediator_id") |>
  mutate(collider_risk = case_when(
    !is.na(x_adj) & !is.na(m_adj) ~ paste0("COLLIDER RISK: X adjusted for ", x_adj,
                                           "; M adjusted for ", m_adj),
    !is.na(x_adj) ~ paste0("COLLIDER RISK: X adjusted for ", x_adj),
    !is.na(m_adj) ~ paste0("COLLIDER RISK: M adjusted for ", m_adj),
    TRUE ~ "none recorded"))
covariate_flags |> knitr::kable(caption = "Covariate adjustment in the exposure and mediator GWAS.")
```

## Instruments

Selection reuses the prior univariable pipeline exactly: p < 5×10⁻⁸, clumping at
r² < 0.001 within 10 Mb on the OpenGWAS 1000G-EUR LD server, harmonisation `action = 2`
(palindromes inferred from allele frequency, ambiguous ones dropped), LD proxies at r² ≥ 0.8
(TwoSampleMR defaults), and **no** F-statistic filter. The relaxed drug-target criteria
(`smoking/config.yml: cis_drugtarget`) are not used. Any per-triad change goes through
`inst_override` and is written to the deviation log.

The MVMR instrument set is the union of X and M instruments, re-clumped with the same parameters
and extracted for both traits (`TwoSampleMR::mv_extract_exposures`).

```{r instruments}
par_tbl <- triads |>
  distinct(triad_id, exposure_id, mediator_id, .keep_all = TRUE) |>
  mutate(rp = map2(inst_override, triad_id, resolve_instrument_params))

pull_triad <- function(triad_id, exposure_id, mediator_id, rp) {
  x <- exposure_id; m <- mediator_id; par <- rp$par
  # Fail on the first pull that returns nothing, naming the call and the server's reason.
  need <- function(val, what) {
    if (is.null(val)) stop("OpenGWAS pull failed for ", triad_id, ": ", what, " -> ",
                           og_last_error(), ". A 5xx code is a server outage (retry later); ",
                           "401/403 is the token (check ieugwasr::user()).", call. = FALSE)
    val
  }
  x_inst <- need(pull_instruments(x, par, CACHE_DIR), paste("instruments for", x))
  m_inst <- need(pull_instruments(m, par, CACHE_DIR), paste("instruments for", m))
  mv_exp <- need(pull_mv_exposures(x, m, par, CACHE_DIR), "mv_extract_exposures (re-clump X u M)")
  list(x_id = x, m_id = m, x_inst = x_inst, m_inst = m_inst, mv_exp = mv_exp,
       x_in_m = pull_snps(x_inst$SNP, m, par, CACHE_DIR),
       m_in_x = pull_snps(m_inst$SNP, x, par, CACHE_DIR))
}
pulls <- par_tbl |>
  mutate(pulls = pmap(list(triad_id, exposure_id, mediator_id, rp), pull_triad))
pulls_by_triad <- setNames(pulls$pulls, pulls$triad_id)
par_by_triad   <- setNames(map(pulls$rp, "par"), pulls$triad_id)

instrument_log <- list_rbind(map(pulls$rp, "log"))

pulls |>
  transmute(triad_id,
            n_X = map_int(pulls, \(p) nrow(p$x_inst)),
            n_M = map_int(pulls, \(p) nrow(p$m_inst)),
            union_before_reclump = map_int(pulls, \(p) length(union(p$x_inst$SNP, p$m_inst$SNP))),
            mvmr_after_reclump = map_int(pulls, \(p) n_distinct(p$mv_exp$SNP)),
            X_in_M_gwas = map_int(pulls, \(p) n_distinct(p$x_in_m$SNP)),
            mean_F_X = map_dbl(pulls, \(p) mean((p$x_inst$beta.exposure / p$x_inst$se.exposure)^2)),
            mean_F_M = map_dbl(pulls, \(p) mean((p$m_inst$beta.exposure / p$m_inst$se.exposure)^2))) |>
  mutate(across(where(is.double), \(x) round(x, 1))) |>
  knitr::kable(caption = "Instrument counts per triad.")

# Log the SNPs lost at re-clumping (union -> MVMR set).
instrument_log <- bind_rows(instrument_log, imap(pulls_by_triad, \(p, t)
  log_row(t, NA, "mvmr_reclump", "dropped",
          "X or M instruments removed by re-clumping the union (or missing in the other trait)",
          setdiff(union(p$x_inst$SNP, p$m_inst$SNP), p$mv_exp$SNP))) |> bind_rows())
```

The 2hGlu file reports a raw-scale beta with an inverse-normal-model p-value. This shows how far
the reported p-values diverge from the ones implied by beta/se among the selected instruments
(the same class of mislabelling `read_gwas()` in the calcium arm checks for).

```{r pval-crosscheck}
imap(pulls_by_triad, \(p, t) bind_rows(
  p$x_inst |> mutate(trait = "X"), p$m_inst |> mutate(trait = "M")) |>
  mutate(log10p_reported = -log10(pval.exposure),
         log10p_from_beta = -(pnorm(-abs(beta.exposure / se.exposure), log.p = TRUE) + log(2)) / log(10)) |>
  group_by(triad_id = t, trait) |>
  summarise(n = n(), cor = cor(log10p_reported, log10p_from_beta),
            median_ratio = median(log10p_from_beta / log10p_reported),
            n_fail_5e8_on_beta = sum(log10p_from_beta < -log10(5e-8)), .groups = "drop")) |>
  bind_rows() |>
  mutate(across(where(is.double), \(x) signif(x, 3))) |>
  knitr::kable(caption = "Reported vs beta/se-implied -log10 p among selected instruments.")
```

## Outcome tables (built once per outcome)

For each AD outcome, the outcome effects and the parental-lifespan effects are extracted
**once** for the union of every triad's instruments, with lifespan aligned to the outcome's
effect allele. Every triad then reads from that one table, which makes the
"never mix adjusted and unadjusted data within a triad" rule hold by construction.

::: {.callout-note}
**Sign convention (from Step 2):** the lifespan axis `GCST006697` is a Martingale residual that
increases with **mortality**. `b_SH` was fit on the same axis, so the SNP-level adjustment
$\beta^{adj}_Y = \beta_Y - b_{SH}\,\beta_{life}$ is internally consistent. Do not negate either
term.
:::

```{r outcome-tables}
all_snps <- pulls_by_triad |>
  map(\(p) c(p$x_inst$SNP, p$m_inst$SNP, p$mv_exp$SNP)) |> unlist() |> unique()

# Outcome tables use the default proxy settings. A triad that overrides proxy parameters would
# not be matched by these tables, so refuse that combination rather than mix rules.
proxy_keys <- grep("^proxy", names(INSTRUMENT_DEFAULTS), value = TRUE)
if (any(map_lgl(par_by_triad, \(p) !identical(p[proxy_keys], INSTRUMENT_DEFAULTS[proxy_keys]))))
  stop("proxy settings cannot be overridden per triad; outcome tables are shared")

outcome_tables <- unique(triads$outcome_id) |>
  set_names() |>
  map(\(o) build_outcome_table(o, all_snps, LIFESPAN_ID, INSTRUMENT_DEFAULTS, CACHE_DIR))
instrument_log <- bind_rows(instrument_log, list_rbind(unname(map(outcome_tables, "log"))))

imap(outcome_tables, \(ot, o) tibble(
  outcome_id = o, union_snps = length(all_snps), with_outcome = nrow(ot$table),
  with_lifespan = sum(!is.na(ot$table$beta_life)))) |>
  bind_rows() |>
  knitr::kable(caption = "Outcome-table coverage. SNPs without lifespan are dropped from adjusted arms (and from raw arms when common_snp_set = TRUE).")
```

## Selection slope b_SH

```{r bsh}
# Same logic as Step 2's load_bsh(): the fitted value is always shown, but it is applied only
# when trusted.
load_bsh <- function(ad_id) {
  src <- bsh_sources |> filter(ad_id == !!ad_id)
  na_row <- tibble(ad_id = ad_id, b_SH = NA_real_, se_SH = NA_real_, trusted = FALSE,
                   reason = "no fits CSV", boot = list(NULL))
  if (nrow(src) == 0 || !file.exists(src$fits_csv[1])) return(na_row)
  fit <- read_csv(src$fits_csv[1], show_col_types = FALSE) |> filter(set == src$fit_set[1]) |> slice(1)
  if (nrow(fit) == 0) return(na_row)
  boot <- if (!is.na(src$boot_rds[1]) && file.exists(src$boot_rds[1])) readRDS(src$boot_rds[1]) else NULL
  tibble(ad_id = ad_id, b_SH = fit$b_SH, se_SH = fit$se_SH, trusted = src$trusted[1],
         reason = src$reason[1], boot = list(boot))
}
bsh <- map(unique(triads$outcome_id), load_bsh) |> list_rbind()

bsh |>
  transmute(outcome = ad_id, b_SH = signif(b_SH, 3), se_SH = signif(se_SH, 3),
            applied = if_else(trusted, "yes", "NO"), reason,
            draws = if_else(map_lgl(boot, is.null), "normal approx.", "bootstrap replicates")) |>
  knitr::kable(caption = "Selection slope per AD outcome (from Step 2).")
```

::: {.callout-important}
**Carried from Step 2 (caveat 0): the Bellenguez b_SH = −0.944 has not been checked against its
permutation null.** SlopeHunter returns about −0.73 on Kunkle data whose selection signal was
destroyed by permutation. Until `bsh_null_permutation()` is run on the Bellenguez merge, every
SlopeHunter-adjusted row here is **provisional**. The raw arm in the 2×2 grid is the hedge.
:::

## Estimation

Per arm (triad × outcome × adjustment):

1. **Total:** univariable X → Y on X's own instruments, using `method_XY_total`.
2. **X → M:** univariable on X's own instruments in the mediator GWAS, using `method_XM`.
3. **MVMR:** X and M jointly on Y, using the re-clumped union and `method_MVMR`. This gives
   **direct** = β(X → Y | M) and **β(M → Y | X)** with their joint covariance.
4. **Indirect** = β(X → M) × β(M → Y | X). Both glucose coefficients come from the same 2hGlu
   GWAS, so the mediator's units cancel and the indirect effect is in log-OR per unit X.
5. **Consistency:** (direct + indirect) − total, in units of the total's SE.

**Uncertainty (Monte Carlo).** The four legs are drawn jointly from a multivariate normal. The
MVMR block uses the method's own covariance. Correlations between the other legs come from a
SNP-level parametric bootstrap, because total, X → M and MVMR share SNPs and GWAS, and total and
direct share the outcome. For adjusted arms, b_SH is also redrawn. The legs are refit on a grid
of b_SH values (re-adjusting the outcome at every grid point), and each draw interpolates the
estimates and covariance at its own b_SH. This carries the shared-slope correlation between
adjusted SNP effects, which naive IVW SEs omit. MR-PRESSO outlier sets are found once at the
point estimate and held fixed across the grid. Delta-method SEs for the indirect effect are
reported alongside as a quick check.

```{r estimate}
arms <- triads |>
  select(-inst_override) |>
  crossing(adjustment = ADJUSTMENTS) |>
  mutate(primary = adjustment == outcome_adjustment)

arm_out <- pmap(arms, function(...) {
  tr <- tibble(...)
  run_triad_arm(tr, pulls_by_triad[[tr$triad_id]], outcome_tables[[tr$outcome_id]]$table,
                tr$adjustment, bsh |> filter(ad_id == tr$outcome_id),
                par_by_triad[[tr$triad_id]], SETTINGS)
})

draws <- bind_rows(map(arm_out, "draws"))
instrument_log <- bind_rows(instrument_log, list_rbind(map(arm_out, "log")))

results <- bind_rows(map(arm_out, "result")) |>
  left_join(arms |> select(triad_id, outcome_id, adjustment, primary, method_note,
                           starts_with("hyp_sign")),
            by = c("triad_id", "outcome_id", "adjustment")) |>
  left_join(covariate_flags |> select(triad_id, collider_risk), by = "triad_id") |>
  left_join(datasets |> select(outcome_id = id, outcome = label), by = "outcome_id") |>
  # FDR across triads x outcomes, within each adjustment arm (so sensitivity arms do not dilute
  # the primary one). Monte Carlo p-values, which include b_SH uncertainty.
  group_by(adjustment) |>
  mutate(across(any_of(c("total_mc_p", "direct_mc_p", "indirect_mc_p")),
                \(p) p.adjust(if_else(status == "run", p, NA_real_), "BH"),
                .names = "{.col}_fdr")) |>
  ungroup() |>
  mutate(hypothesis_consistent = sign(xm_b) == hyp_sign_xm & sign(my_b) == hyp_sign_my &
           sign(total_b) == hyp_sign_total)
```

## Results

```{r results-table}
fmt_ci <- \(b, lo, hi) ifelse(is.na(b), NA, sprintf("%.4f [%.4f, %.4f]", b, lo, hi))

results |>
  filter(primary) |>
  transmute(triad_id, outcome, adjustment, status,
            total    = fmt_ci(total_b, total_mc_lo, total_mc_hi),
            `X→M`    = fmt_ci(xm_b, xm_mc_lo, xm_mc_hi),
            `M→Y|X`  = fmt_ci(my_b, my_mc_lo, my_mc_hi),
            direct   = fmt_ci(direct_b, direct_mc_lo, direct_mc_hi),
            indirect = fmt_ci(indirect_b, indirect_mc_lo, indirect_mc_hi),
            `indirect SE (MC / delta)` = sprintf("%.4f / %.4f",
              (indirect_mc_hi - indirect_mc_lo) / (2 * 1.96), indirect_se_delta),
            `% mediated` = ifelse(is.na(proportion_b), "—",
              sprintf("%.0f%% [%.0f, %.0f]", 100 * proportion_b, 100 * proportion_mc_lo,
                      100 * proportion_mc_hi)),
            `FDR indirect` = signif(indirect_mc_p_fdr, 2),
            gates = gate_summary, collider_risk) |>
  knitr::kable(caption = paste(
    "Primary analysis (per-outcome adjustment from the config). Point estimate with 95% Monte",
    "Carlo percentile interval. % mediated is shown only when every gate passes."))
```

```{r results-note, echo=FALSE, results='asis'}
if (any(results$collider_risk != "none recorded", na.rm = TRUE))
  cat("::: {.callout-warning}\nRows marked **COLLIDER RISK** use exposure or mediator GWAS",
      "adjusted for an adiposity covariate (see [Dataset checks](#data-checks)). Treat their",
      "X → M and M → Y | X legs, and therefore the indirect effect, as potentially biased in an",
      "unknown direction.\n:::\n")
```

### Gates {#gates}

Pre-specified. A triad that fails a gate is still reported; it gets no percent mediated.

| Gate | Rule |
|---|---|
| Conditional F | > 10 for both X and M (Sanderson-Windmeijer, two-sample, recomputed on the final SNP set of each arm) |
| Total effect | 95% MC interval excludes 0 |
| Sign concordance | sign(indirect) = sign(total) |
| Direct opposite sign | flagged: direct and indirect are reported separately, no percent |
| Proportion outside [0, 1] | flagged: no percent |
| Equivalence | indirect's 90% MC interval inside ±`equiv_bound_indirect` (TOST, α = 0.05) |

```{r gates-table}
results |>
  filter(status == "run") |>
  transmute(triad_id, outcome, adjustment, primary,
            cond_F_x = round(cond_F_x, 1), cond_F_m = round(cond_F_m, 1),
            gate_condF, gate_total_ci, gate_sign_concordant, flag_direct_opposite,
            flag_prop_outside_01, equivalence, hypothesis_consistent, gate_summary) |>
  knitr::kable(caption = "Gate outcomes for every arm that ran.")
```

### Consistency: does direct + indirect reproduce total?

Total is fit on X's instruments; direct and M | X are fit on the re-clumped union, possibly with
different methods. Small gaps are expected. A gap of several total-SEs means the two-step and
MVMR decompositions disagree and the proportion should not be trusted even if the gates pass.

```{r consistency}
results |>
  filter(status == "run") |>
  transmute(triad_id, outcome, adjustment,
            total = signif(total_b, 3), `direct + indirect` = signif(direct_b + indirect_b, 3),
            `difference method (total − direct)` = signif(difference_b, 3),
            `product method (indirect)` = signif(indirect_b, 3),
            `gap / SE(total)` = round(consistency_gap_in_total_se, 2),
            `r(total, direct)` = round(r_total_direct, 2)) |>
  knitr::kable(caption = "Decomposition consistency.")
```

## Diagnostics

### Conditional F and MVMR heterogeneity

```{r diag-mvmr}
results |>
  filter(status == "run") |>
  transmute(triad_id, outcome, adjustment, n_snp_total, n_snp_xm, n_snp_mvmr,
            n_mvmr_x_inst, n_mvmr_m_inst, cond_F_x = round(cond_F_x, 1),
            cond_F_m = round(cond_F_m, 1), mvmr_Q = round(mvmr_Q, 1),
            mvmr_Q_p = signif(mvmr_Q_p, 2), method_total, method_xm, method_mvmr) |>
  knitr::kable(caption = paste(
    "Conditional F (gencov = ", SETTINGS$gencov, ") and Sanderson-Windmeijer Q, per arm.",
    "SNP counts are after SlopeHunter SNP loss. A significant Q indicates pleiotropy or",
    "heterogeneity beyond what X and M explain."))
```

### X → M assumption checks {#xm-checks}

The X → M leg is new, so the SOP checks that set `method_XM` were not done for it. If Q is
significant with no directional pleiotropy, IVW-MRE stands. If MR-PRESSO flags outliers, switch
to `mr_presso`. If F is weak, switch to `mr_raps`. The MVMR method must then follow the mapping.

```{r diag-xm}
diag <- imap(pulls_by_triad, \(p, t)
  triad_diagnostics(triads |> filter(triad_id == t) |> slice(1), p, par_by_triad[[t]],
                    SETTINGS$presso_nboot))

map(diag, "checks") |> bind_rows() |>
  select(triad_id, xm_n_snp, xm_mean_F, xm_ivw_mre_b, xm_ivw_mre_se, xm_Q_p,
         xm_egger_intercept, xm_egger_int_p, xm_presso_global_p, xm_presso_n_outlier, xm_presso_b) |>
  mutate(across(where(is.double), \(x) signif(x, 3))) |>
  knitr::kable(caption = "X → M: heterogeneity, Egger intercept and MR-PRESSO.")
```

```{r diag-xm-plot, fig.width=7, fig.height=5}
map(diag, "xm_dat") |>
  imap(\(d, t) mutate(d, triad_id = t)) |>
  bind_rows() |>
  ggplot(aes(beta.exposure, beta.outcome)) +
  geom_hline(yintercept = 0, colour = "grey70") + geom_vline(xintercept = 0, colour = "grey70") +
  geom_errorbar(aes(ymin = beta.outcome - 1.96 * se.outcome,
                    ymax = beta.outcome + 1.96 * se.outcome), colour = "grey60", width = 0) +
  geom_point(colour = "#00274c", alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ 0 + x, mapping = aes(weight = 1 / se.outcome^2),
              colour = "#ffcb05", se = FALSE) +
  facet_wrap(~triad_id, scales = "free") +
  labs(x = "SNP effect on X", y = "SNP effect on M", title = "X → M scatter (IVW slope)") +
  theme_minimal(base_size = 12)
```

### Direction: Steiger and reverse M → X

Glucose or insulin could plausibly affect lean mass. Steiger flags X instruments that explain
more variance in M than in X. The reverse MR uses M's instruments on X.

```{r diag-direction}
map(diag, "checks") |> bind_rows() |>
  transmute(triad_id, steiger_n, steiger_n_wrong,
            `X→M (all)` = sprintf("%.4f (%.4f)", xm_ivw_mre_b, xm_ivw_mre_se),
            `X→M (Steiger-filtered)` = sprintf("%.4f (%.4f)", xm_b_steiger, xm_se_steiger),
            `reverse M→X` = sprintf("%.4f (%.4f)", reverse_mx_b, reverse_mx_se),
            reverse_p = signif(reverse_mx_p, 2), reverse_n_snp = reverse_mx_n_snp) |>
  knitr::kable(caption = "Steiger filtering on X → M, and the reverse-direction estimate.")
```

### Instrument cross-lookup

How many X instruments are associated with M, and vice versa. Many shared associations at
genome-wide significance mean the two instrument sets carry common pleiotropy, which MVMR has to
separate. Compare `n_nominal` against the ~5% expected under no sharing.

```{r diag-cross}
map(diag, "cross") |> bind_rows() |>
  knitr::kable(caption = "Cross-lookup of instrument sets.")
```

## Sensitivity: outcome × adjustment (2×2)

Expectation, if lean-mass instruments carry real lifespan effects (Step 1: strong edge on two of
the four lifespan phenotypes): SlopeHunter should shift **total** and **direct** more than
**indirect**, because the selection path runs X → S and not through glucose. Shifts are in units
of the raw arm's SE. The proxy-inclusive Bellenguez betas are not on the clinical log-OR scale,
so compare outcomes on **proportions and signs**, not absolute effects.

```{r sensitivity-grid}
results |>
  select(triad_id, outcome, adjustment, status, total_b, direct_b, indirect_b,
         total_se, direct_se, indirect_se_delta, proportion_b) |>
  mutate(across(c(total_b, direct_b, indirect_b), \(x) signif(x, 3)),
         proportion = ifelse(is.na(proportion_b), "—", sprintf("%.0f%%", 100 * proportion_b))) |>
  select(-ends_with("_se"), -indirect_se_delta, -proportion_b) |>
  knitr::kable(caption = "All four cells of the grid per triad. NOT RUN cells carry their reason.")

results |>
  filter(status == "run") |>
  select(triad_id, outcome, adjustment, total_b, direct_b, indirect_b,
         total_se, direct_se, indirect_se_delta) |>
  pivot_wider(names_from = adjustment,
              values_from = c(total_b, direct_b, indirect_b, total_se, direct_se, indirect_se_delta)) |>
  filter(!is.na(total_b_slopehunter), !is.na(total_b_none)) |>
  transmute(triad_id, outcome,
            `Δ total / SE`    = round((total_b_slopehunter - total_b_none) / total_se_none, 2),
            `Δ direct / SE`   = round((direct_b_slopehunter - direct_b_none) / direct_se_none, 2),
            `Δ indirect / SE` = round((indirect_b_slopehunter - indirect_b_none) /
                                        indirect_se_delta_none, 2)) |>
  knitr::kable(caption = "Shift from SlopeHunter adjustment, per estimand.")
```

## Forest plot

```{r forest, fig.width=8.5, fig.height=5}
plot_df <- results |>
  filter(status == "run") |>
  select(triad_id, outcome, adjustment, primary,
         direct_b, direct_mc_lo, direct_mc_hi, indirect_b, indirect_mc_lo, indirect_mc_hi,
         total_b, total_mc_lo, total_mc_hi) |>
  pivot_longer(-c(triad_id, outcome, adjustment, primary),
               names_to = c("estimand", ".value"), names_pattern = "(direct|indirect|total)_(.*)") |>
  mutate(estimand = factor(estimand, c("total", "direct", "indirect")),
         arm = paste0(outcome, "\n", adjustment, if_else(primary, " (primary)", "")))

if (nrow(plot_df) == 0) {
  message("No arms ran; nothing to plot.")
} else {
  ggplot(plot_df, aes(b, arm, colour = estimand)) +
    geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
    geom_pointrange(aes(xmin = mc_lo, xmax = mc_hi), position = position_dodge(width = 0.6)) +
    scale_colour_manual(values = c(total = "grey40", direct = "#00274c", indirect = "#ffcb05")) +
    facet_wrap(~triad_id, scales = "free_x") +
    labs(x = "effect on AD (log-OR per unit exposure), 95% Monte Carlo interval", y = NULL,
         colour = NULL, title = "Total, direct and indirect effects by triad and outcome arm") +
    theme_minimal(base_size = 12)
}
```

## eQTL mediators {#eqtl-mediators}

For exposure → expression → AD triads, the X → M leg needs **trans** effects of X's instruments
on expression. Most expression resources in OpenGWAS (eQTLGen, GTEx-derived sets) are
**cis-only**, so most X instruments will be absent from the mediator dataset. Rows with
`mediator_type = "eqtl"` are therefore skipped (with the count in `status`) when fewer than
`k_min_x_in_mediator` X instruments are present. Cis-only mediators with 1–3 SNPs will also
generally fail the conditional-F gate. Those genes are better handled as a separate
drug-target analysis (expression → AD plus colocalization, as in `smoking/06_cis_drugtarget.qmd`)
than as formal mediation.

## Assumptions

- **No X × M interaction** (accepted, not tested). Product-of-coefficients and MVMR both assume it.
- **Standard IV assumptions per leg.** These were checked for X → Y in the prior univariable
  analyses. X → M is checked [above](#xm-checks) for the first time.
- **Rare-outcome approximation.** The product of a linear X → M slope and a log-OR M → Y | X
  slope is read as a log-OR. This holds approximately because AD is rare.
- **Outcome scale.** Bellenguez includes proxy cases, so its betas are not clinical log-ORs.
  Compare across outcomes on proportions and direction.
- **Sample overlap.** Lean mass (UKB) and 2hGlu (MAGIC cohorts, no UKB) do not overlap, so
  conditional F uses `gencov = 0`. Lean mass, lifespan and Bellenguez all draw on UKB.
  Step 2's robust-MR work found overlap immaterial for a comparable configuration, but it is not
  modelled here.
- **Covariate-adjusted GWAS.** See the warning under [Configuration](#config).
- **b_SH draws.** These use a normal approximation to SlopeHunter's bootstrap distribution
  until the replicates are saved (`boot_rds`).

## Save

```{r save}
write_csv(results |> select(-any_of("boot")), "mr_mediation_results.csv")
write_csv(instrument_log |> arrange(triad_id, outcome_id, stage), "mr_mediation_instrument_log.csv")
write_csv(map(diag, "checks") |> bind_rows(), "mr_mediation_diagnostics.csv")
# Draws are the input for any Bayesian or alternative summary; kept out of git with data/cache.
dir.create(CACHE_DIR, showWarnings = FALSE, recursive = TRUE)
saveRDS(draws, file.path(CACHE_DIR, "mr_mediation_draws.rds"))

instrument_log |>
  select(triad_id, outcome_id, stage, type, n, detail) |>
  knitr::kable(caption = "Instrument deviations and dropped SNPs (SNP lists are in the CSV).")
```
