---
title: "Robust univariable MR — results and diagnostics"
subtitle: "Lean mass, fat-free mass, BMI, glycemic traits → AD (targets pipeline, pipeline/)"
author: "Dave Bridges and Katie Kittell"
date: today
format:
  html:
    toc: true
    toc-location: right
    code-fold: true
    code-summary: "Show the code"
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

Everything here is read from `results/pipeline/`, written by the `targets` pipeline
(`_targets.R`, settings in `pipeline/config.yml`, design in `pipeline/README.md`). Nothing is
recomputed, so this renders in seconds and always matches the CSVs.

**Design.** Main: **MR-APSS** (models correlated pleiotropy and sample overlap). Sensitivity:
**CAUSE** (credible-interval rule). Category-2 comparison: **MRBEE** and **MR-RAPS** (assume
InSIDE; more power), reported always and flagged unstable only on process grounds. Classical
suite alongside. **Scales differ:** MR-APSS and CAUSE are per SD of exposure on per SD of outcome;
the classical methods and MRBEE are per unit of each GWAS's beta (log-OR for AD).

```{r setup}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(knitr)

res_dir <- "results/pipeline"
res   <- read_csv(file.path(res_dir, "results.csv"), show_col_types = FALSE)
# Columns added in later pipeline versions; present-but-empty if the CSV predates them.
for (col in c("stable", "p_gamma_below_0", "note", "overlap_C12", "method_handles_overlap"))
  if (!col %in% names(res)) res[[col]] <- NA
assoc <- read_csv(file.path(res_dir, "associations.csv"), show_col_types = FALSE)
ds    <- read_csv(file.path(res_dir, "datasets.csv"), show_col_types = FALSE)
bg    <- if (file.exists(f <- file.path(res_dir, "mrapss_background.csv"))) read_csv(f, show_col_types = FALSE)
gap   <- read_csv(file.path(res_dir, "step2_lifespan_gap_fill.csv"), show_col_types = FALSE)
bsh   <- read_csv(file.path(res_dir, "bsh_null_checks.csv"), show_col_types = FALSE)

labels <- setNames(ds$label, ds$key)
nice <- function(x) dplyr::coalesce(labels[x], x)
block_order <- c("exposure_to_mediator", "adiposity_to_glycemic", "to_ad", "to_lifespan",
                 "exposure_to_mediator_reverse", "adiposity_to_glycemic_reverse", "to_ad_reverse")
lat <- if (file.exists(f <- file.path(res_dir, "latent_confounding.csv"))) read_csv(f, show_col_types = FALSE)
main_methods <- c("MR-APSS", "MR-APSS (p<5e-8)", "CAUSE", "MRAID", "MRBEE", "MR-RAPS", "IVW-MRE")
```

## Datasets

```{r datasets}
ds |>
  transmute(key, Dataset = label, N = n, Type = type, Cohorts = cohorts,
            `Adjusted for (collider risk)` = coalesce(adjusted_for, "—")) |>
  kable()
```

## Main, sensitivity and category-2 estimates

One row per association. **Main = MR-APSS** (instruments p < 5×10⁻⁵), its **threshold sensitivity
at p < 5×10⁻⁸**, and **sensitivity = CAUSE** (all per SD); then **MRBEE, MR-RAPS** and **IVW-MRE**
(per unit). **MRAID** (per SD) is an extra sensitivity analysis, run only where the samples
don't overlap and there are at most 1,000 candidate instruments. Its lean mass / fat-free mass
fits were unreliable, so those show "not run (cap)". `†` marks a category-2 fit flagged unstable. Blocks ending `_reverse` are the same pairs
run in the opposite direction.

```{r main-table, results='asis'}
fmt <- function(b, lo, hi) ifelse(is.na(b), "", sprintf("%.3f [%.3f, %.3f]", b, lo, hi))
wide <- res |>
  filter(method %in% main_methods) |>
  mutate(cell = paste0(ifelse(!is.na(note) & grepl("^not run", note) & method == "MRAID", "not run (cap)",
                              fmt(b, ci_lo, ci_hi)),
                       ifelse(method != "CAUSE" & !is.na(p),
                              sprintf(" p=%.2g", p), ""),
                       ifelse(method == "CAUSE" & !is.na(p_gamma_below_0),
                              sprintf(" P(γ<0)=%.2f", p_gamma_below_0), ""),
                       ifelse(!is.na(stable) & !stable, " †", ""))) |>
  select(assoc_id, block, exposure, outcome, arm, method, cell) |>
  pivot_wider(names_from = method, values_from = cell) |>
  mutate(block = factor(block, block_order)) |>
  arrange(block, exposure, outcome, arm)

for (blk in intersect(block_order, unique(as.character(wide$block)))) {
  cat("\n")
  print(kable(wide |> filter(block == blk) |>
                transmute(Exposure = nice(exposure), Outcome = nice(outcome), Arm = arm,
                          across(any_of(main_methods))),
              caption = blk))
  cat("\n")
}
```

## Forest plots

Main and sensitivity (per SD) and the per-unit comparison methods are on separate panels, because
their scales differ.

```{r forest, fig.width=10, fig.height=11}
res |>
  filter(method %in% main_methods, is.na(stable) | stable,
         block %in% c("exposure_to_mediator", "adiposity_to_glycemic", "to_ad")) |>
  mutate(label = paste0(nice(exposure), " → ", nice(outcome),
                        ifelse(arm == "slopehunter", " [SH]", "")),
         panel = ifelse(method %in% c("MR-APSS", "MR-APSS (p<5e-8)", "CAUSE"),
                        "Main / sensitivity (SD per SD)", "Category 2 + IVW (per unit)"),
         method = factor(method, main_methods)) |>
  ggplot(aes(b, label, colour = method)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
  geom_pointrange(aes(xmin = ci_lo, xmax = ci_hi), position = position_dodge(width = 0.7), size = 0.25) +
  facet_wrap(~ panel, scales = "free_x") +
  scale_colour_manual(values = c("MR-APSS" = "#ffcb05", "MR-APSS (p<5e-8)" = "#989c97",
                                 CAUSE = "#d86018", MRBEE = "#702082",
                                 "MR-RAPS" = "#75988d", "IVW-MRE" = "#00274c")) +
  labs(x = "Estimate (95% CI)", y = NULL, colour = NULL) +
  theme_minimal(base_size = 11) + theme(legend.position = "bottom")
```

## Latent residual heritable confounding

For each pair, a latent term for **all heritable shared factors** acting on both traits,
next to the causal estimate (`latent_confounding.csv`). It is not a named confounder: explicit
confounders are handled separately through MVMR triads. Purely environmental confounding does not
bias MR and is not part of it. All values are per SD of exposure on per SD of outcome.

- **CAUSE q·η** (latent shared-factor slope, from the joint posterior of γ, η, q under the causal
  model) is the **cleaner** estimate. `frac_shared` = q·η / (γ + q·η) is the share of the naive
  slope due to shared factors. It is withheld when its denominator's interval spans 0, and flagged
  unreliable when q > 0.5.
- **MR-APSS background slope** Ω_xy/Ω_xx (block-jackknife interval) is **broader, likely an upper
  bound**: it includes residual heritable confounding but may also include causal signal carried
  by non-instrument SNPs. Sample overlap and stratification are reported separately (sample-structure
  slope, C_xy/C_xx on the β scale) and are not counted as confounding.
- **Reverse causation can load onto the latent term**, so each pair is shown in both directions.
- When an explicit confounder is modelled via MVMR, re-running CAUSE / MR-APSS and seeing how much
  q·η or the background slope shrinks approximates how much of the latent term it explains.

```{r latent, results='asis'}
if (is.null(lat)) {
  cat("_latent_confounding.csv not found: run the pipeline (CAUSE on Great Lakes) first._\n")
} else {
  ci3 <- function(e, l, h) ifelse(is.na(e), "", sprintf("%.3f [%.3f, %.3f]", e, l, h))
  tab <- lat |>
    left_join(assoc |> select(pair_id = assoc_id, block, exposure, outcome, arm), by = "pair_id") |>
    mutate(direction = ifelse(grepl("_reverse$", block), "reverse", "forward"),
           pair = paste(nice(pmin(exposure, outcome)), "↔", nice(pmax(exposure, outcome))),
           causal = ci3(causal_est, causal_lo, causal_hi),
           latent = ci3(latent_conf_est, latent_conf_lo, latent_conf_hi),
           `frac shared` = ci3(frac_shared_est, frac_shared_lo, frac_shared_hi),
           `q (causal / sharing)` = ifelse(is.na(cause_q), "", sprintf("%.2f / %.2f", cause_q, cause_q_sharing)),
           `ΔELPD (SE)` = ifelse(is.na(cause_delta_elpd), "", sprintf("%.2f (%.2f)", cause_delta_elpd, cause_delta_elpd_se)),
           `sample-structure slope` = ifelse(is.na(apss_sample_structure_slope), "", sprintf("%.4f", apss_sample_structure_slope))) |>
    arrange(pair, arm, method, direction)
  for (m in c("CAUSE", "MR-APSS")) {
    cat("\n")
    print(kable(tab |> filter(method == m) |>
                  transmute(Exposure = nice(exposure), Outcome = nice(outcome), Arm = arm, Direction = direction,
                            Causal = causal, `Latent term` = latent,
                            !!!(if (m == "CAUSE") list(`Frac shared` = quote(`frac shared`),
                                                     `q (causal / sharing)` = quote(`q (causal / sharing)`),
                                                     `ΔELPD (SE)` = quote(`ΔELPD (SE)`))
                                else list(`Sample-structure slope` = quote(`sample-structure slope`))),
                            Flags = flags),
                caption = if (m == "CAUSE") "CAUSE: q·η (latent shared-factor slope)"
                          else "MR-APSS: background (non-foreground) slope Ω_xy/Ω_xx"))
    cat("\n")
  }
}
```

## Sample overlap and genetic correlation (MR-APSS background)

`C12` is the cross-trait LDSC intercept, a direct estimate of sample overlap; `rg` is the
genetic correlation implied by the background model (the correlated-pleiotropy channel).

```{r background}
if (!is.null(bg)) bg |>
  left_join(assoc |> select(assoc_id, overlap), by = "assoc_id") |>
  transmute(assoc_id, `cohort overlap (config)` = overlap, C12 = round(C12, 4),
            `overlap flag` = overlap_flag, rg = round(rg, 3)) |>
  kable()
```

## Survival collider

```{r collider}
bsh |> mutate(across(where(is.numeric), \(x) signif(x, 3))) |> kable(caption = "b_SH identifiability and permutation null")
gap |> mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
  kable(caption = "Step 2's estimate-level correction: naive − b_SH × (E → lifespan), IVW-MRE")
```

## Diagnostics by association

Scatter (classical instruments, method lines), funnel (single-SNP Wald ratios with the IVW-MRE
line and pseudo-95% cone), leave-one-out (IVW-MRE), and the MR-APSS scatter on its own
instruments. Leave-one-out tables are next to the figures as `leave_one_out.csv`.

```{r diagnostics, results='asis'}
fig_dir <- file.path(res_dir, "figures")
figs <- c(scatter_classical = "Scatter (classical)", funnel = "Funnel",
          leave_one_out = "Leave-one-out", scatter_mrapss = "Scatter (MR-APSS)")
ids <- assoc |> mutate(block = factor(block, block_order)) |> arrange(block, exposure, outcome, arm)
for (blk in levels(ids$block)) {
  sub <- ids |> filter(block == blk)
  if (!nrow(sub)) next
  cat("\n### ", blk, "\n\n", sep = "")
  for (i in seq_len(nrow(sub))) {
    id <- sub$assoc_id[i]
    cat("\n#### ", nice(sub$exposure[i]), " → ", nice(sub$outcome[i]),
        if (sub$arm[i] == "slopehunter") " [SlopeHunter]" else "", "\n\n::: {.panel-tabset}\n\n", sep = "")
    for (f in names(figs)) {
      path <- file.path(fig_dir, id, paste0(f, ".png"))
      if (file.exists(path)) cat("## ", figs[[f]], "\n\n![](", path, "){width=80%}\n\n", sep = "")
    }
    cat(":::\n\n")
  }
}
```
