---
title: "Robust MR — MR-RAPS (uncorrelated pleiotropy, weak instruments)"
author: "Dave Bridges"
date: today
editor: source
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    embed-resources: true    # self-contained: no _files/ dir needed to view it
    code-fold: true
    code-summary: "Show the code"
    fig-path: "figures-robust-mr/"
theme: journal
execute:
  echo: true
  warning: false
---

```{r global_options}
#| echo: false
#| message: false
library(tidyverse)
library(knitr)
library(kableExtra)
.root <- local({
  d <- normalizePath(getwd(), mustWork = FALSE)
  while (!file.exists(file.path(d, "config_robust_mr.yml"))) {
    p <- dirname(d)
    if (identical(p, d)) stop("config_robust_mr.yml not found in ", getwd(),
                              " or any parent — render from the analysis folder.")
    d <- p
  }
  d
})
source(file.path(.root, "R", "robust_mr_helpers.R"))
set_project_root(.root)
cfg <- load_robust_cfg()
.inputs_hash <- tools::md5sum(c(file.path(.root, "config_robust_mr.yml"),
                                file.path(.root, "R", "robust_mr_helpers.R")))
set.seed(cfg$seed)
dir.create(pp(cfg$paths$out), recursive = TRUE, showWarnings = FALSE)
```

## Purpose

MR-RAPS (Zhao et al., *Ann. Statist.* 2020) sits deliberately between IVW and the
confounder-modelling methods. It assumes horizontal pleiotropy exists and may be
large, models it as random effects $\alpha_j \sim N(0, \tau^2)$, and — critically —
assumes those effects are **independent of the instrument–exposure effects**
(the InSIDE assumption). It does *not* posit a shared confounder.

Two properties earn it a place here:

1. **Weak-instrument correction by profile score.** This is RAPS's core
   contribution and the reason it tolerates a relaxed p-value threshold. It lets
   us ask whether the attenuation seen in MR-APSS is really about pleiotropy or
   just about instrument quality.
2. **$\hat\tau^2$ is reported.** Unlike IVW, RAPS estimates how much pleiotropy
   there is. A large $\tau^2$ with a stable $\beta$ is the signature of *balanced*
   pleiotropy — exactly the situation IVW handles badly and RAPS handles well.

::: callout-note
## Where this sits among the five methods

| | uncorrelated pleiotropy | **correlated** pleiotropy | overlap | weak IV |
|---|---|---|---|---|
| IVW | ✗ | ✗ | ✗ | ✗ |
| **MR-RAPS** | ✓ $\tau^2$ | ✗ | ✗ | ✓ |
| **MRBEE** | ✓ outlier removal | ✗ | ✓ | ✓ |
| CAUSE | ✓ | ✓ | ✓ | partial |
| MR-APSS | ✓ | ✓ | ✓ | ✓ |

RAPS is the only method here that corrects weak instruments *without* also
touching overlap or correlated pleiotropy. That isolation is the point: it tells
us what the weak-instrument correction alone is worth.
:::

## Data

Instruments come from the same harmonised substrate as MR-APSS and CAUSE
(`est_paras()$dat`), so differences between methods are attributable to the
method rather than to QC or harmonisation.

```{r load}
#| cache: true
#| cache.extra: !expr .inputs_hash
paras_path <- pp(cfg$paths$cache, "apss_paras.rds")
if (!file.exists(paras_path)) stop("Run robust_mr_apss.qmd first — missing ", paras_path)
paras <- readRDS(paras_path)

exp_of <- function(arm) if (arm == "overlapping") "glgc2021_ldl" else "willer2013_ldl"

map_dfr(paras, ~ tibble(n_snp = nrow(.x$dat)), .id = "arm") |>
  kable(caption = "Harmonised variants available per arm",
        format.args = list(big.mark = ",")) |>
  kable_styling(full_width = FALSE)
```

## Instruments at two thresholds

RAPS is run at both the conventional 5e-8 and the relaxed 5e-5 used by MR-APSS.
If the estimate is stable across thresholds, weak-instrument bias is not driving
anything.

```{r instruments}
#| cache: true
#| cache.extra: !expr .inputs_hash
thresholds <- c(`5e-8` = 5e-8, `5e-5` = cfg$mrapss$iv_threshold)

ivs <- purrr::map(names(paras), function(arm) {
  purrr::map(names(thresholds), function(th) {
    robust_iv_set(paras[[arm]]$dat, thresholds[[th]], cfg,
                  r2 = cfg$mrapss$clump_r2, kb = cfg$mrapss$clump_kb)
  }) |> setNames(names(thresholds))
}) |> setNames(names(paras))

map_dfr(names(ivs), function(arm) {
  map_dfr(names(thresholds), function(th) {
    d <- ivs[[arm]][[th]]
    tibble(arm = arm, threshold = th, n_iv = nrow(d),
           mean_F = mean((d$b.exp / d$se.exp)^2))
  })
}) |>
  kable(caption = "Instruments by arm and threshold. Mean F below ~10 at the relaxed threshold is precisely the regime RAPS is built for.",
        digits = 1) |>
  kable_styling(full_width = FALSE)
```

## Fit

Two loss functions: plain `l2` (efficient if there are no outliers) and `tukey`
(robust, downweights idiosyncratic pleiotropic outliers). `over.dispersion = TRUE`
estimates $\tau^2$ rather than assuming no pleiotropy.

```{r fit}
#| cache: true
#| cache.extra: !expr .inputs_hash
library(mr.raps)

fit_one <- function(arm, th, loss) {
  d <- ivs[[arm]][[th]]
  if (nrow(d) < 4) return(NULL)
  res <- try(fit_raps(b_exp = d$b.exp, b_out = d$b.out,
                      se_exp = d$se.exp, se_out = d$se.out,
                      loss = loss, over.dispersion = TRUE),
             silent = TRUE)
  if (inherits(res, "try-error")) {
    message("RAPS failed for ", arm, " / ", th, " / ", loss, ":\n",
            conditionMessage(attr(res, "condition")))
    return(NULL)
  }
  tidy_raps(res, arm = arm, threshold = th, loss = loss, n_iv = nrow(d))
}

raps_res <- tidyr::expand_grid(arm = names(ivs), th = names(thresholds),
                               loss = c("l2", "tukey")) |>
  purrr::pmap_dfr(function(arm, th, loss) fit_one(arm, th, loss))

if (nrow(raps_res) == 0) {
  stop("All MR-RAPS fits failed — the messages above list what each interface reported.")
}
message("MR-RAPS interface used: ", paste(unique(raps_res$interface), collapse = ", "))
write_csv(raps_res, pp(cfg$paths$out, "raps_results.csv"))

raps_res |>
  select(arm, threshold, method, n_iv, b, se, lci, uci, pval, tau2, interface) |>
  arrange(arm, threshold, method) |>
  kable(caption = "MR-RAPS. tau2 is the estimated variance of horizontal pleiotropic effects; a large value with a stable beta indicates balanced pleiotropy.",
        digits = 4) |>
  kable_styling(full_width = FALSE)
```

```{r plot}
#| fig-width: 8
#| fig-height: 4.5
if (nrow(raps_res) > 0) {
  raps_res |>
    mutate(label = paste0(method, " @ ", threshold)) |>
    robust_mr_forest(title = "MR-RAPS: LDL-C -> heel eBMD, by threshold and loss") +
    facet_wrap(~arm)
}
```

## Reading it

- **Stable across thresholds** → weak-instrument bias is not the story, and the
  relaxed threshold in MR-APSS is safe.
- **Larger $|\beta|$ at 5e-8 than 5e-5** → the extra weak instruments carry
  proportionally more pleiotropy than signal.
- **$\tau^2$ substantially above zero** → real horizontal pleiotropy, but *balanced*.
  RAPS accommodates it; IVW's standard errors would be too narrow.
- **`tukey` and `l2` disagree** → a few outlying variants are driving the estimate;
  prefer the robust fit and check which SNPs.

What RAPS *cannot* tell you is whether pleiotropy is correlated with instrument
strength. If a shared factor drives both LDL-C and eBMD, RAPS is biased and will
not warn you. That is what CAUSE and MR-APSS are for, and why the two classes are
reported side by side in `robust_mr_summary.qmd`.
