---
title: "Robust MR — MRBEE (bias-corrected estimating equations)"
author: "Dave Bridges"
date: today
editor: source
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    embed-resources: true    # self-contained: no _files/ dir needed to view it
    code-fold: true
    code-summary: "Show the code"
    fig-path: "figures-robust-mr/"
theme: journal
execute:
  echo: true
  warning: false
---

```{r global_options}
#| echo: false
#| message: false
library(tidyverse)
library(knitr)
library(kableExtra)
.root <- local({
  d <- normalizePath(getwd(), mustWork = FALSE)
  while (!file.exists(file.path(d, "config_robust_mr.yml"))) {
    p <- dirname(d)
    if (identical(p, d)) stop("config_robust_mr.yml not found in ", getwd(),
                              " or any parent — render from the analysis folder.")
    d <- p
  }
  d
})
source(file.path(.root, "R", "robust_mr_helpers.R"))
set_project_root(.root)
cfg <- load_robust_cfg()
.inputs_hash <- tools::md5sum(c(file.path(.root, "config_robust_mr.yml"),
                                file.path(.root, "R", "robust_mr_helpers.R")))
set.seed(cfg$seed)
dir.create(pp(cfg$paths$out), recursive = TRUE, showWarnings = FALSE)
```

## Purpose

MRBEE (Lorincz-Comi et al., *HGG Advances* 2024) treats MR as an
**errors-in-variables** problem. GWAS effect estimates carry measurement error;
that error biases ordinary IVW, and the direction of the bias depends on sample
overlap while its magnitude depends on instrument strength. MRBEE subtracts the
bias exactly, using the estimation-error covariance matrix $R_{xy}$.

Two things make it a genuinely different check from what we have run:

1. **$R_{xy}$ is estimated from insignificant SNPs**, and its off-diagonal is a
   direct estimate of the correlation induced by sample overlap. This is a
   **third, independent measurement** of the same quantity as MR-APSS's $C_{12}$
   and CAUSE's $\rho$ — from a third model family.
2. **Pleiotropy is removed by iterative outlier detection** (IMRP) rather than
   modelled as a distribution. Like MR-RAPS it assumes the *remaining* pleiotropy
   is uncorrelated with instrument strength; unlike MR-RAPS it also corrects for
   overlap.

::: callout-important
MRBEE handles overlap, weak instruments and *uncorrelated* pleiotropy. It does
**not** model a shared confounder. If correlated pleiotropy is present, MRBEE is
biased in the same way MR-RAPS is — the outlier test removes variants that look
anomalous, not variants that are systematically confounded.
:::

## Data and the error-covariance matrix

```{r load}
#| cache: true
#| cache.extra: !expr .inputs_hash
library(MRBEE)

paras_path <- pp(cfg$paths$cache, "apss_paras.rds")
if (!file.exists(paras_path)) stop("Run robust_mr_apss.qmd first — missing ", paras_path)
paras <- readRDS(paras_path)

# Rxy from genome-wide Z-scores; exposures first, OUTCOME LAST (MRBEE convention).
Rxy <- purrr::map(paras, ~ mrbee_error_cov(.x$dat))
saveRDS(Rxy, pp(cfg$paths$cache, "mrbee_Rxy.rds"))

rxy_tab <- imap_dfr(Rxy, ~ tibble(
  arm          = .y,
  var_exposure = .x[1, 1],
  var_outcome  = .x[2, 2],
  cov_overlap  = .x[1, 2]
))
write_csv(rxy_tab, pp(cfg$paths$out, "mrbee_error_cov.csv"))

rxy_tab |>
  kable(caption = "MRBEE estimation-error covariance. The off-diagonal is the overlap-induced correlation between the two GWAS' estimation errors — compare with MR-APSS C12 (0.023 / -0.007) and CAUSE rho (-0.011 / -0.015).",
        digits = 4) |>
  kable_styling(full_width = FALSE)
```

## Instruments

Same clumped sets as MR-RAPS, at both thresholds.

```{r instruments}
#| cache: true
#| cache.extra: !expr .inputs_hash
thresholds <- c(`5e-8` = 5e-8, `5e-5` = cfg$mrapss$iv_threshold)

ivs <- purrr::map(names(paras), function(arm) {
  purrr::map(names(thresholds), function(th) {
    robust_iv_set(paras[[arm]]$dat, thresholds[[th]], cfg,
                  r2 = cfg$mrapss$clump_r2, kb = cfg$mrapss$clump_kb)
  }) |> setNames(names(thresholds))
}) |> setNames(names(paras))

map_dfr(names(ivs), function(arm) {
  map_dfr(names(thresholds), function(th) {
    tibble(arm = arm, threshold = th, n_iv = nrow(ivs[[arm]][[th]]))
  })
}) |>
  kable(caption = "Instruments by arm and threshold") |>
  kable_styling(full_width = FALSE)
```

## Fit

```{r fit}
#| cache: true
#| cache.extra: !expr .inputs_hash
fit_one <- function(arm, th) {
  d <- ivs[[arm]][[th]]
  if (nrow(d) < 4) return(NULL)

  # NOTE the argument names. The univariable entry point takes lowercase bx/bxse
  # as VECTORS; the multivariable MRBEE.IMRP() takes bX/bXse as matrices. Passing
  # the MVMR names to the UV function fails with "unused argument".
  res <- if (!is.null(getNamespace("MRBEE")$MRBEE.IMRP.UV)) {
    try(MRBEE::MRBEE.IMRP.UV(
      by = d$b.out, bx = d$b.exp,
      byse = d$se.out, bxse = d$se.exp,
      Rxy = Rxy[[arm]]
    ), silent = TRUE)
  } else {
    try(MRBEE::MRBEE.IMRP(
      by = d$b.out, bX = matrix(d$b.exp, ncol = 1),
      byse = d$se.out, bXse = matrix(d$se.exp, ncol = 1),
      Rxy = Rxy[[arm]], var.est = "ordinal"
    ), silent = TRUE)
  }

  if (inherits(res, "try-error")) {
    message("MRBEE failed for ", arm, " / ", th, ": ",
            conditionMessage(attr(res, "condition")))
    return(NULL)
  }
  tidy_mrbee(res, arm = arm, threshold = th, n_iv = nrow(d))
}

mrbee_res <- tidyr::expand_grid(arm = names(ivs), th = names(thresholds)) |>
  purrr::pmap_dfr(function(arm, th) fit_one(arm, th))

if (nrow(mrbee_res) == 0) {
  stop("All MRBEE fits failed — see the messages above for the underlying error.")
}
write_csv(mrbee_res, pp(cfg$paths$out, "mrbee_results.csv"))

mrbee_res |>
  select(arm, threshold, n_iv, n_pleiotropic, b, se, lci, uci, pval) |>
  arrange(arm, threshold) |>
  kable(caption = "MRBEE. n_pleiotropic is the number of instruments flagged and removed by the iterative pleiotropy test.",
        digits = 4) |>
  kable_styling(full_width = FALSE)
```

```{r plot}
#| fig-width: 8
#| fig-height: 4
if (nrow(mrbee_res) > 0) {
  mrbee_res |>
    mutate(label = paste0("MRBEE @ ", threshold)) |>
    robust_mr_forest(title = "MRBEE: LDL-C -> heel eBMD") +
    facet_wrap(~arm)
}
```

## Reading it

The most valuable single number here is `cov_overlap`. We now have three
independent estimates of sample overlap, from three model families — MRBEE's
error covariance, MR-APSS's cross-trait LDSC intercept, and CAUSE's $\rho$. If all
three agree the overlap picture is settled; if MRBEE's disagrees materially, the
LDSC-based estimates deserve a second look, since they rest on assumptions
(HapMap3 restriction, LD-score model) that MRBEE's does not.

`n_pleiotropic` is the second thing to check. A large number of removed
instruments means the IMRP test is doing heavy lifting, and the estimate is
conditional on those removals being the right ones — worth a sensitivity check
against MR-RAPS's `tukey` fit, which downweights rather than deletes.
