---
title: "Correlation betwen Calcium and Cholesterol Assays from MGI Data"
author: "Dave Bridges"
date: "2026-02-18"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  gfm:
    html-math-method: webtex

theme: journal

execute:
  echo: true
  warning: false

knitr:
  opts_chunk:
    fig-path: "figures/"          # folder for all figure files
    dev: ["png", "pdf"]           # both formats for every plot
    dpi: 300                      # ← this is the correct place for resolution
    dev.args:
      png:
        type: "cairo-png"         # better anti-aliasing / font rendering (optional but recommended)
      pdf:
        family: "sans"            # font family (optional)
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)
library(knitr)
library(broom)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

Assess correlations between serum calcium and cholesterol/LDL/HDL/TG in MGI participants

## Raw Data

Describe your raw data files, including what the columns mean (and what units they are in).

```{r data-input}
library(readr) #loads the readr package
labs.filename <- "combined_data/LabResultsCleaned.csv" #input file(s)

#this loads whatever the file is into a dataframe called exp.data if it exists

lab.data <- read_csv(labs.filename)

```

These data can be found in `r getwd()` in a file named `r ifelse(labs.filename %in% dir(), labs.filename, "no file found")`.  This input file was most recently updated on `r ifelse(labs.filename %in% dir(), format(file.info(labs.filename)$mtime, "%Y-%m-%d"), "unknown")`.  This script was most recently updated on `r date()`.

There are `r distinct(lab.data, DeID_PatientID) |> nrow()` participants in this dataset with lab values.

## Analysis

```{r data-cleaning}
serum.calcium.data <- lab.data |>
  filter(test_name=="Serum Calcium")

ionized.calcium.data <- lab.data |>
  filter(test_name == "Ionized Calcium")

cholesterol.data <- lab.data |>
  filter(test_name == "Total Cholesterol")

ldl.data <- lab.data |>
  filter(test_name == "LDL-C") 

hdl.data <- lab.data |>
  filter(test_name == "HDL-C") 

tg.data <- lab.data |>
  filter(test_name == "Triglycerides") 
```

There are 

- `r distinct(serum.calcium.data, DeID_PatientID) |> nrow()` participants in this dataset with serum calcium values.
- `r distinct(ionized.calcium.data, DeID_PatientID) |> nrow()` participants in this dataset with ionized calcium values.
- `r distinct(cholesterol.data, DeID_PatientID) |> nrow()` participants in this dataset with cholesterol values.
- `r distinct(ldl.data, DeID_PatientID) |> nrow()` participants in this dataset with LDL-C values.
- `r distinct(ldl.data, DeID_PatientID) |> nrow()` participants in this dataset with HDL-C values.
- `r distinct(ldl.data, DeID_PatientID) |> nrow()` participants in this dataset with triglyceride values.

Calculated the most recent value for each participant

```{r calcium-ldlc-crude-correlation}
ldl.final.data <-
  ldl.data |>
  arrange(DeID_COLLECTION_DATE) |>
  group_by(DeID_PatientID) |>
  summarize(Last.Value = last(as.numeric(value)),
            Age = last(AgeInYears))

cholesterol.final.data <-
  cholesterol.data |>
  arrange(DeID_COLLECTION_DATE) |>
  group_by(DeID_PatientID) |>
  summarize(Last.Value = last(as.numeric(value)),
            Age = last(AgeInYears))

hdl.final.data <-
  hdl.data |>
  arrange(DeID_COLLECTION_DATE) |>
  group_by(DeID_PatientID) |>
  summarize(Last.Value = last(as.numeric(value)),
            Age = last(AgeInYears))

tg.final.data <-
  tg.data |>
  arrange(DeID_COLLECTION_DATE) |>
  group_by(DeID_PatientID) |>
  summarize(Last.Value = last(as.numeric(value)),
            Age = last(AgeInYears))

serum.calcium.final.data <-
  serum.calcium.data |>
  arrange(DeID_COLLECTION_DATE) |>
  group_by(DeID_PatientID) |>
  summarize(Last.Value = last(value),
            Age = last(AgeInYears))

ionized.calcium.final.data <-
  ionized.calcium.data |>
  arrange(DeID_COLLECTION_DATE) |>
  group_by(DeID_PatientID) |>
  summarize(Last.Value = last(value),
            Age = last(AgeInYears))
```


```{r ldl-calcium-correlations}
ca.ldl.final.data <-
  inner_join(serum.calcium.final.data,ldl.final.data,  by="DeID_PatientID", suffix=c("_ca","_ldl"))

library(ggplot2)

ggplot(ca.ldl.final.data,
       aes(y=Last.Value_ca,
           x=Last.Value_ldl)) +
  geom_smooth(method="lm") +
  geom_point(alpha=0.1) +
  theme_classic(base_size=16) +
  labs(y="Calcium (mg/dL)",
       x="LDL-C (mg/dL)")

lm(Last.Value_ca~Last.Value_ldl, data=ca.ldl.final.data) |>
  tidy() |>
  kable()

#for ionized calcium
ionized.ca.ldl.final.data <-
  inner_join(ionized.calcium.final.data,ldl.final.data,  by="DeID_PatientID", suffix=c("_ca","_ldl"))


ggplot(ionized.ca.ldl.final.data,
       aes(y=Last.Value_ca,
           x=Last.Value_ldl)) +
  geom_smooth(method="lm") +
  geom_point(alpha=0.1) +
  theme_classic(base_size=16) +
  labs(y="Ionized Calcium (mg/dL)",
       x="LDL-C (mg/dL)")

ionized.lm.crude <- lm(Last.Value_ca~Last.Value_ldl, data=ionized.ca.ldl.final.data) 

ionized.lm.crude |>  
  tidy() |>
  kable()

library(lindia)
library(ggfortify)
autoplot(ionized.lm.crude)
```

#### Selecting an Appropriate Distribution for Ionized Calcium

Due to the poor model fit with a gaussian distribution, checked several other options for modelling calcium

```{r ionized-calcium-distribution-analysis}
library(fitdistrplus)
library(gamlss) #for generalized gamma
library(gamlss.dist)

# Fit individual distributions
fit_norm  <- fitdist(ionized.calcium.final.data$Last.Value, "norm")
fit_gamma  <- fitdist(ionized.calcium.final.data$Last.Value, "gamma")
fit_lnorm  <- fitdist(ionized.calcium.final.data$Last.Value, "lnorm")
fit_weibull <- fitdist(ionized.calcium.final.data$Last.Value, "weibull")

gofstat(list(fit_norm, fit_gamma, fit_lnorm, fit_weibull),
        fitnames = c("Normal", "Gamma", "Log-Normal", "Weibull"))


qqcomp(list(fit_norm, fit_gamma, fit_lnorm), 
       legendtext = c("Normal", "Gamma", "Log-Normal"),
       main = "Q-Q: Gamma vs Log-Normal")
```

### Crude Association

Based on these data I should use a Gamma distribution for further modelling.  The Gamma Beta distribution was a better fit, but did not converge with my data


```{r ionized-calcium-glm}

ionized.glm.crude <- glm(Last.Value_ca ~ Last.Value_ldl, data=ionized.ca.ldl.final.data, 
                            family=Gamma(link="log"))  

ionized.glm.crude |>  
  tidy(conf.int = T) |>
  kable(caption="GLM for LDL-C on Ionized Calcium (Gamma Distribution)")
```

##### Bayesian Analysis for Crude Association

```{r ionized-calcium-brms}
library(brms)
library(broom.mixed)

ionized.priors <- c(
  prior(normal(0, 0.01),   class = b, coef=Last.Value_ldl),          # predictor coefficients
  prior(normal(1, 0.5), class = Intercept),  # log-scale intercept
  prior(exponential(1), class = shape)       # shape must be positive
)


ionized.brm.crude <- brm(Last.Value_ca ~ Last.Value_ldl, 
           data = ionized.ca.ldl.final.data, 
           family = Gamma(link="log"),
           prior = ionized.priors)

prior_summary(ionized.brm.crude) %>% kable(caption="Prior summary for ionized calcium - LDL-C bayesian fit")
tidy(ionized.brm.crude) %>% kable(caption="Summary of model fit for ionized calcium - LDL-C bayesian fit")
plot(ionized.brm.crude)
hypothesis(ionized.brm.crude, "Last.Value_ldl>0") # testing for whether there is a beta coefficient greater than zero

as_draws_df(ionized.brm.crude) %>%
  ggplot(aes(x=b_Last.Value_ldl)) +
  geom_density(fill=color_scheme[1]) +
  geom_vline(xintercept=0,color="red",lty=2) +
  labs(y="",
       x="Beta coefficient for LDL-C on ionized calcium",
       title="Posterior probabilities") +
  theme_classic(base_size=16)

```


### Interpretation Notes

- The Q-Q plot still shows heavy tails, particularly the upper tail with observations 192, 943, 980 appearing as extreme outliers. This suggests the Gamma may still not be fully capturing the tail behavior, though it's likely better than the Gaussian was
- Observations 192 and 1343 are consistently flagged across all four panels — these are your most influential points and worth pulling out to inspect clinically. Are they plausible values or potential data quality issues?

### Multivariable Analyses

```{r demographic-inputs}
demographic.file <- 'combined_data/DemographicInfo.csv'
ancestry.file <- 'combined_data/MGI_Ancestry.csv'

demographic.data <- read_csv(demographic.file) 
ancestry.data <- read_csv(ancestry.file)|>
  mutate(MajorityAncestry = relevel(as.factor(MajorityAncestry),ref="EUR"))
  
age.breaks <- c(0, 18, 45, 55, 65, 75, Inf)
age.labels <- c("<18", "18-44", "45-54", "55-64", "65-74", "75+")

ionized.ca.ldl.final.data.annot <-
  ionized.ca.ldl.final.data |>
  left_join(demographic.data, by="DeID_PatientID") |>
  left_join(ancestry.data, by="DeID_PatientID") |>
  mutate(Age.Group = cut(Age_ldl, breaks = age.breaks, labels = age.labels, right = FALSE))

#stepwise modeling
m1 <- glm(Last.Value_ca ~ Last.Value_ldl, data=ionized.ca.ldl.final.data.annot, family=Gamma(link="log"))
m5 <- glm(Last.Value_ca ~ Last.Value_ldl+GenderCode+Age_ca+MajorityAncestry, data=ionized.ca.ldl.final.data.annot, family=Gamma(link="log"))

m5 |>
  anova() |>
  tidy () |>
  kable(caption="ANOVA for GLM adjusting for age, gender and ancestry")

m5 |>
  tidy(conf.int=T) |>
  kable(caption="Multivariable GLM adjusting for age, gender and ancestry")

library(effectsize)
omega_squared(m5, partial = TRUE, ci = 0.95, alternative = "two.sided") 

library(performance)
r2(m5)  # returns several R² variants - Nagelkerke is most commonly reported
```

## Summary of All Lipid Associations with Ionized Calcium

```{r all-lipid-associations}
combined.data <- ionized.calcium.final.data |>
  left_join(ldl.final.data,         by = "DeID_PatientID", suffix = c("_ca", "_ldl")) |>
  left_join(cholesterol.final.data,  by = "DeID_PatientID", suffix = c("", "_chol")) |>
  left_join(hdl.final.data,          by = "DeID_PatientID", suffix = c("", "_hdl")) |>
  left_join(tg.final.data,           by = "DeID_PatientID", suffix = c("", "_tg"))|>
  left_join(demographic.data, by="DeID_PatientID") |>
  left_join(ancestry.data, by="DeID_PatientID") |>
  mutate(Age.Group = cut(Age_ldl, breaks = age.breaks, labels = age.labels, right = FALSE))


library(broom)

# define predictors and adjustment covariates
predictors <- c("Last.Value_ldl", "Last.Value_hdl", "Last.Value_tg") #not enough data for Last.Value_chol
covariates <- "GenderCode + MajorityAncestry + Age.Group"

# function to run both models and extract results
run_models <- function(predictor, data) {
  
  crude.formula    <- as.formula(paste("Last.Value_ca ~", predictor))
  adjusted.formula <- as.formula(paste("Last.Value_ca ~", predictor, "+", covariates))
  
  crude.model    <- glm(crude.formula,    data = data, family = Gamma(link = "log"))
  adjusted.model <- glm(adjusted.formula, data = data, family = Gamma(link = "log"))
  
  crude.est    <- tidy(crude.model)    |> filter(term == predictor)
  adjusted.est <- tidy(adjusted.model) |> filter(term == predictor)
  
  tibble(
    Predictor       = predictor,
    Crude.Beta      = crude.est$estimate,
    Crude.SE        = crude.est$std.error,
    Crude.P         = crude.est$p.value,
    Adjusted.Beta   = adjusted.est$estimate,
    Adjusted.SE     = adjusted.est$std.error,
    Adjusted.P      = adjusted.est$p.value
  )
}

results <- map_dfr(predictors, run_models, data = combined.data)
kable(results, caption="Summary of glm's for each lipid as a predictor of ionized calcium")

 library(ggplot2)

results |>
  mutate(
    Crude.CI.low    = Crude.Beta    - 1.96 * Crude.SE,
    Crude.CI.high   = Crude.Beta    + 1.96 * Crude.SE,
    Adjusted.CI.low  = Adjusted.Beta - 1.96 * Adjusted.SE,
    Adjusted.CI.high = Adjusted.Beta + 1.96 * Adjusted.SE,
    Predictor = recode(Predictor,
      "Last.Value_ldl"  = "LDL-C",
      "Last.Value_chol" = "Total Cholesterol",
      "Last.Value_hdl"  = "HDL-C",
      "Last.Value_tg"   = "Triglycerides")
  ) |>
  pivot_longer(
    cols = c(Crude.Beta, Adjusted.Beta),
    names_to = "Model",
    values_to = "Beta"
  ) |>
  mutate(
    CI.low  = if_else(Model == "Crude.Beta", Crude.CI.low,  Adjusted.CI.low),
    CI.high = if_else(Model == "Crude.Beta", Crude.CI.high, Adjusted.CI.high),
    Model   = recode(Model, "Crude.Beta" = "Crude", "Adjusted.Beta" = "Adjusted")
  ) |>
  ggplot(aes(x = Beta, y = Predictor, color = Model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = CI.low, xmax = CI.high),
                  position = position_dodge(width = 0.5)) +
  scale_color_manual(values = c("Crude" = color_scheme[1], "Adjusted" = color_scheme[2])) +
  labs(
    x = "Beta (95% CI)",
    y = NULL,
    title = "Association of Lipids with Ionized Calcium",
    color = NULL
  ) +
  theme_classic(base_size=16) +
  theme(legend.position = "bottom") 
```



## Session Information

```{r session-information}
sessionInfo()
```
