---
title: "Combining of Raw Data Files"
author: "Dave Bridges"
date: "2026-03-24"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  gfm:
    html-math-method: webtex

theme: journal

execute:
  echo: true
  warning: false

knitr:
  opts_chunk:
    fig-path: "figures/"          # folder for all figure files
    dev: ["png", "pdf"]           # both formats for every plot
    dpi: 300                      # ← this is the correct place for resolution
    dev.args:
      png:
        type: "cairo-png"         # better anti-aliasing / font rendering (optional but recommended)
      pdf:
        family: "sans"            # font family (optional)
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)
library(knitr)
library(broom)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

Combine the raw data files.  These had to be pulled in three separate batches each with t heir own folder `males`, `females_umarried` and `females_married.`  This script will take each csv file in each of those three subfolders, combine them and then write them into the `combined_data` folder.  There should be the same files in each subfolder.

## Combining Data

```{r combining}
library(tidyverse)


base_dir   <- "."    # folder containing males/, females_unmarried/, females_married/
batch_dirs <- c("males", "females_unmarried", "females_married")
output_dir <- file.path(base_dir, "combined_data")

dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

### Identifying Files

```{r identifying-files}

# Use first batch folder as reference for expected filenames
csv_files <- file.path(base_dir, batch_dirs[1]) %>%
  list.files(pattern = "\\.csv$", ignore.case = TRUE, full.names = FALSE)

cat("CSV files found:", length(csv_files), "\n")
walk(csv_files, ~cat(" ", .x, "\n"))

# Check all three folders have the same files
walk(batch_dirs, function(batch) {
  files_here <- file.path(base_dir, batch) %>%
    list.files(pattern = "\\.csv$", ignore.case = TRUE, full.names = FALSE)
  missing <- setdiff(csv_files, files_here)
  extra   <- setdiff(files_here, csv_files)
  cat(
    batch, "—", length(files_here), "files |",
    if (length(missing) == 0) "no missing" else paste("MISSING:", str_c(missing, collapse = ", ")), "|",
    if (length(extra)   == 0) "no extra"   else paste("EXTRA:",   str_c(extra,   collapse = ", ")),
    "\n"
  )
})
```

### Read Combine and Deduplicate

```{r read-combine-deduplicate}
audit_log <- map_dfr(csv_files, function(fname) {

  cat("\nProcessing:", fname, "\n")

  # Read all three batches into a named list
  batch_list <- batch_dirs %>%
    set_names() %>%
    map(function(batch) {
      fpath <- file.path(base_dir, batch, fname)
      if (file.exists(fpath)) {
        df <- read_csv(fpath, show_col_types = FALSE,
                       col_types = cols(.default = col_character()))
        cat(" ", batch, "—", nrow(df), "rows\n")
        df
      } else {
        cat(" ", batch, "— not found, skipping\n")
        NULL
      }
    }) %>%
    compact()

  if (length(batch_list) == 0) {
    cat("  No data found — skipping\n")
    return(NULL)
  }

  # Warn if column names differ across batches
  col_sets <- map(batch_list, names)
  if (n_distinct(map(col_sets, sort)) > 1) {
    cat("  *** WARNING: column names differ across batches — check output\n")
  }

  # Bind all batches
  combined <- bind_rows(batch_list)
  cat("  Combined rows:", nrow(combined), "\n")

  # Detect PatientID column
  id_col <- names(combined) %>%
    str_subset(regex("patientid|patient_id", ignore_case = TRUE)) %>%
    first()

  # Deduplicate if patient-level file (rows ≈ n_patients)
  n_dups <- 0L
  if (!is.na(id_col)) {
    n_unique <- n_distinct(combined[[id_col]])
    cat("  Unique patients:", n_unique, "\n")

    # If total rows ≈ unique patients × n_batches, it's a patient-level file
    if (nrow(combined) <= n_unique * length(batch_list)) {
      combined <- combined %>%
        distinct(across(all_of(id_col)), .keep_all = TRUE)
      n_dups <- nrow(combined) - n_unique
      if (n_dups > 0)
        cat("  Deduplicated — rows after:", nrow(combined), "\n")
    }
  }

  # Write combined file
  out_path <- file.path(output_dir, fname)
  write_csv(combined, out_path)
  cat("  Written to:", out_path, "\n")
  
   # Free memory before next file
  rm(batch_list, combined)
  gc(verbose = FALSE)
})
```


## Session Information

```{r session-information}
sessionInfo()
```
