---
title: "Cleaning of Diagnoses Data"
author: "Dave Bridges"
date: "2026-03-09"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  gfm:
    html-math-method: webtex

theme: journal

execute:
  echo: true
  warning: false

knitr:
  opts_chunk:
    fig-path: "figures/"          # folder for all figure files
    dev: ["png", "pdf"]           # both formats for every plot
    dpi: 300                      # ← this is the correct place for resolution
    dev.args:
      png:
        type: "cairo-png"         # better anti-aliasing / font rendering (optional but recommended)
      pdf:
        family: "sans"            # font family (optional)
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)
library(knitr)
library(broom)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

Clean the data for various diagnoses, for use in other scripts

## Raw Data

```{r data-input}
library(readr) #loads the readr package
diagnosis.filename <- "combined_data/DiagnosesComprehensiveAll.csv" #input file(s)
diagnosis.data <- read_csv(diagnosis.filename) |>
  mutate(DeID_ProblemObservationDate = mdy_hm(DeID_ProblemObservationDate))

#encounter.datafile <- 'combined_data/EncounterAll.csv'
#encounter.anthro.datafile <- 'combined_data/EncounterAnthropometricsBMI.csv'

#load in encounter metadata
#encounter.data <- read_csv(encounter.datafile)
#encounter.anthro.data <- read_csv(encounter.anthro.datafile)
```

These data can be found in `r getwd()` in a file named `r ifelse(diagnosis.filename %in% dir(), diagnosis.filename, "no file found")`.  This input file was most recently updated on `r ifelse(diagnosis.filename %in% dir(), format(file.info(diagnosis.filename)$mtime, "%Y-%m-%d"), "unknown")`.  This script was most recently updated on `r date()`.

There are `r distinct(diagnosis.data, DeID_PatientID) |> nrow()` participants in this dataset with any diagnoses values.

## Data Cleaning

Will first clean the crude diagnosis data and then add in the elixhauser and charson data.

### MACE Data

```{r mace-data-cleaning}
icd_pattern_mace <- paste0(
  "^410",          # ICD-9 MI
  "|^433\\.[0-9]1",# ICD-9 Occlusion of precerebral arteries (5th digit = 1)
  "|^434",         # ICD-9 Occlusion of cerebral arteries
  "|^436",         # ICD-9 Acute but ill-defined CVA
  "|^430",         # ICD-9 Subarachnoid haemorrhage
  "|^431",         # ICD-9 Intracerebral haemorrhage
  "|^I21",         # ICD-10 Acute MI
  "|^I22",         # ICD-10 Subsequent MI
  "|^I63",         # ICD-10 Cerebral infarction
  "|^I60",         # ICD-10 Subarachnoid haemorrhage
  "|^I61",         # ICD-10 Intracerebral haemorrhage
  "|^I46"          # ICD-10 Cardiac arrest
)

diagnosis.data.mace <- diagnosis.data %>%
  filter(grepl(icd_pattern_mace, TermCodeMapped, ignore.case = FALSE)) 

mace.diagnoses <-
  diagnosis.data.mace |>
  group_by(DeID_PatientID) |>
  arrange(DeID_ProblemObservationDate) |>
  summarize(MACE.onset = first(DeID_ProblemObservationDate)) |>
  mutate(MACE = TRUE)

library(ggplot2)
ggplot(mace.diagnoses,
       aes(x=MACE.onset)) +
  geom_histogram() +
  theme_classic(base_size=16) +
  labs(x="Year of Incident MACE")
```


After cleaning are `r distinct(diagnosis.data.mace, DeID_PatientID) |> nrow()` participants in this dataset with MACE.


### Osteoporosis Data

```{r osteoporosis-data-cleaning}
icd_pattern_osteoporosis <- paste0(
  "^M80",       # ICD-10 Osteoporosis with pathological fracture
  "|^M81",      # ICD-10 Osteoporosis without pathological fracture
  "|^733\\.0"   # ICD-9 733.0x (covers 733.00–733.09)
)

diagnosis.data.osteoporosis <- diagnosis.data %>%
  filter(grepl(icd_pattern_osteoporosis, TermCodeMapped, ignore.case = FALSE)) 

osteoporosis.diagnoses <-
  diagnosis.data.osteoporosis|>
  group_by(DeID_PatientID) |>
  arrange(DeID_ProblemObservationDate) |>
  summarize(Osteoporosis.onset = first(DeID_ProblemObservationDate)) |>
  mutate(Osteoporosis = TRUE)

library(ggplot2)
ggplot(osteoporosis.diagnoses,
       aes(x=Osteoporosis.onset)) +
  geom_histogram() +
  theme_classic(base_size=16) +
  labs(x="Year of Incident Osteoporosis")
```

After cleaning are `r distinct(diagnosis.data.osteoporosis, DeID_PatientID) |> nrow()` participants in this dataset with Osteoporosis

## Writing Out Cleaned Files

```{r writing-cleaned}
combined.diagnoses <-
  full_join(osteoporosis.diagnoses,mace.diagnoses, by='DeID_PatientID')

cleaned.diagnoses.file <- 'combined_data/DiagnosesCleaned.csv'
write_csv(combined.diagnoses, file=cleaned.diagnoses.file)
```

Wrote this out to `r cleaned.diagnoses.file`

## Session Information

```{r session-information}
sessionInfo()
```
