---
title: "Cleaning of Diagnoses Data"
author: "Dave Bridges"
date: "2026-03-09"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  gfm:
    html-math-method: webtex

theme: journal

execute:
  echo: true
  warning: false

knitr:
  opts_chunk:
    fig-path: "figures/"          # folder for all figure files
    dev: ["png", "pdf"]           # both formats for every plot
    dpi: 300                      # ← this is the correct place for resolution
    dev.args:
      png:
        type: "cairo-png"         # better anti-aliasing / font rendering (optional but recommended)
      pdf:
        family: "sans"            # font family (optional)
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)
library(knitr)
library(broom)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

Clean the data for various diagnoses, for use in other scripts

## Raw Data

```{r data-input}
library(readr) #loads the readr package
diagnosis.filename <- "combined_data/DiagnosesComprehensiveAll.csv" #input file(s)
diagnosis.data <- read_csv(diagnosis.filename) |>
  mutate(DeID_ProblemObservationDate = mdy_hm(DeID_ProblemObservationDate))
```

These data can be found in `r getwd()` in a file named `r ifelse(diagnosis.filename %in% dir(), diagnosis.filename, "no file found")`.  This input file was most recently updated on `r ifelse(diagnosis.filename %in% dir(), format(file.info(diagnosis.filename)$mtime, "%Y-%m-%d"), "unknown")`.  This script was most recently updated on `r date()`.

There are `r distinct(diagnosis.data, DeID_PatientID) |> nrow()` participants in this dataset with any diagnoses values.

## Data Cleaning

Will first clean the crude diagnosis data and then add in the elixhauser and charson data.

### MACE Data

```{r mace-data-cleaning}
icd_pattern_mace <- paste0(
  "^410",          # ICD-9 MI
  "|^433\\.[0-9]1",# ICD-9 Occlusion of precerebral arteries (5th digit = 1)
  "|^434",         # ICD-9 Occlusion of cerebral arteries
  "|^436",         # ICD-9 Acute but ill-defined CVA
  "|^430",         # ICD-9 Subarachnoid haemorrhage
  "|^431",         # ICD-9 Intracerebral haemorrhage
  "|^I21",         # ICD-10 Acute MI
  "|^I22",         # ICD-10 Subsequent MI
  "|^I63",         # ICD-10 Cerebral infarction
  "|^I60",         # ICD-10 Subarachnoid haemorrhage
  "|^I61",         # ICD-10 Intracerebral haemorrhage
  "|^I46"          # ICD-10 Cardiac arrest
)

diagnosis.data.mace <- diagnosis.data %>%
  filter(grepl(icd_pattern_mace, TermCodeMapped, ignore.case = FALSE)) 

mace.diagnoses <-
  diagnosis.data.mace |>
  group_by(DeID_PatientID) |>
  arrange(DeID_ProblemObservationDate) |>
  summarize(MACE.onset = first(DeID_ProblemObservationDate)) |>
  mutate(MACE = TRUE)

library(ggplot2)
ggplot(mace.diagnoses,
       aes(x=MACE.onset)) +
  geom_histogram() +
  theme_classic(base_size=16) +
  labs(x="Year of Incident MACE")
```


After cleaning are `r distinct(diagnosis.data.mace, DeID_PatientID) |> nrow()` participants in this dataset with MACE.


### Osteoporosis Data

```{r osteoporosis-data-cleaning}
icd_pattern_osteoporosis <- paste0(
  "^M80",       # ICD-10 Osteoporosis with pathological fracture
  "|^M81",      # ICD-10 Osteoporosis without pathological fracture
  "|^733\\.0"   # ICD-9 733.0x (covers 733.00–733.09)
)

diagnosis.data.osteoporosis <- diagnosis.data %>%
  filter(grepl(icd_pattern_osteoporosis, TermCodeMapped, ignore.case = FALSE)) 

osteoporosis.diagnoses <-
  diagnosis.data.osteoporosis|>
  group_by(DeID_PatientID) |>
  arrange(DeID_ProblemObservationDate) |>
  summarize(Osteoporosis.onset = first(DeID_ProblemObservationDate)) |>
  mutate(Osteoporosis = TRUE)

library(ggplot2)
ggplot(osteoporosis.diagnoses,
       aes(x=Osteoporosis.onset)) +
  geom_histogram() +
  theme_classic(base_size=16) +
  labs(x="Year of Incident Osteoporosis")
```

After cleaning are `r distinct(diagnosis.data.osteoporosis, DeID_PatientID) |> nrow()` participants in this dataset with Osteoporosis

### Comorbidity Data

```{r comorbidity-data-cleaning-loading-files}
library(tidyverse)
library(lubridate)

elixhauser.data <- read_csv(
  "combined_data/ComorbiditiesElixhauserComprehensive.csv",
  show_col_types = FALSE
) %>%
  mutate(
    DeID_PatientID   = as.character(DeID_PatientID),
    DeID_EncounterID = as.character(DeID_EncounterID)
  )

charlson.data <- read_csv(
  "combined_data/ComorbiditiesCharlsonComprehensive.csv",
  show_col_types = FALSE
) %>%
  mutate(
    DeID_PatientID   = as.character(DeID_PatientID),
    DeID_EncounterID = as.character(DeID_EncounterID)
  )

encounter.data <- read_csv(
  "combined_data/EncounterAll.csv",
  show_col_types = FALSE
) %>%
  mutate(
    DeID_PatientID   = as.character(DeID_PatientID),
    DeID_EncounterID = as.character(DeID_EncounterID),
    encounter_date   = mdy_hm(DeID_AdmitDate)
  ) %>%
  filter(!is.na(encounter_date)) %>%
  select(DeID_PatientID, DeID_EncounterID, encounter_date)

cat("Elixhauser rows:", nrow(elixhauser.data), "\n")
cat("Charlson rows:",   nrow(charlson.data),   "\n")
cat("Unique patients (Elixhauser):",
    n_distinct(elixhauser.data$DeID_PatientID), "\n")
cat("Unique patients (Charlson):",
    n_distinct(charlson.data$DeID_PatientID), "\n")
```

```{r collapse-and-date}
# --- Elixhauser: collapse paired conditions, join encounter dates ---
elix_dated <- elixhauser.data %>%
  mutate(
    diabetes      = pmax(DiabetesComplicated, DiabetesUncomplicated,
                         na.rm = TRUE),
    hypertension  = pmax(HypertensionComplicated, HypertensionUncomplicated,
                         na.rm = TRUE),
    alcohol_abuse = AlcoholAbuse,
    arrhythmia    = CardiacArrhythmias,
    chf           = CongestiveHeartFailure,
    obesity       = Obesity,
    pulm_circ     = PulmonaryCirculationDisorders,
    renal_failure = RenalFailure,
    elix_score    = TotalScore
  ) %>%
  select(DeID_PatientID, DeID_EncounterID,
         diabetes, hypertension, alcohol_abuse, arrhythmia,
         chf, obesity, pulm_circ, renal_failure, elix_score) %>%
  inner_join(
    encounter.data %>% select(DeID_EncounterID, encounter_date),
    by = "DeID_EncounterID"
  ) %>%
  arrange(DeID_PatientID, encounter_date)

# --- Charlson: collapse paired diabetes, join encounter dates ---
charlson_dated <- charlson.data %>%
  mutate(
    charlson_diabetes = pmax(DiabetesWithChronicComplication,
                             DiabetesWithoutChronicComplication,
                             na.rm = TRUE),
    charlson_chf      = CongestiveHeartFailure,
    charlson_mi       = MyocardialInfarction,
    charlson_pvd      = PeripheralVascularDisease,
    charlson_renal    = RenalDisease,
    charlson_score    = TotalScore,
    charlson_score_age = TotalScoreAdjustedForAge
  ) %>%
  select(DeID_PatientID, DeID_EncounterID,
         charlson_diabetes, charlson_chf, charlson_mi,
         charlson_pvd, charlson_renal,
         charlson_score, charlson_score_age) %>%
  inner_join(
    encounter.data %>% select(DeID_EncounterID, encounter_date),
    by = "DeID_EncounterID"
  ) %>%
  arrange(DeID_PatientID, encounter_date)

cat("Elixhauser rows after date join:", nrow(elix_dated), "\n")
cat("Charlson rows after date join:",   nrow(charlson_dated), "\n")
```

```{r reversal-diagnostic}
# Confirm that 0 after 1 means "not coded this visit" not "resolved"
# Expected: high reversal rate confirming 0s should be ignored

check_reversals <- function(df, col) {
  df %>%
    group_by(DeID_PatientID) %>%
    arrange(encounter_date) %>%
    summarise(
      ever_positive = any(.data[[col]] == 1, na.rm = TRUE),
      reversals     = {
        x    <- .data[[col]]
        ever <- cummax(replace_na(x, 0L))
        sum(x == 0 & ever == 1, na.rm = TRUE)
      },
      .groups = "drop"
    ) %>%
    summarise(
      condition             = col,
      n_ever_positive       = sum(ever_positive),
      n_with_reversals      = sum(reversals > 0 & ever_positive),
      pct_reversal          = round(100 * n_with_reversals / n_ever_positive, 1)
    )
}

reversal_summary <- bind_rows(
  check_reversals(elix_dated, "diabetes"),
  check_reversals(elix_dated, "hypertension"),
  check_reversals(elix_dated, "chf"),
  check_reversals(elix_dated, "renal_failure")
)

reversal_summary %>% kable(caption = "Reversal diagnostic — confirms 0 = not coded, not resolved")
```

```{r extract-onsets}
# For chronic conditions: onset = first encounter where flag == 1
# 0s after a 1 are ignored — condition treated as permanent from onset

first_onset <- function(df, col) {
  df %>%
    filter(.data[[col]] == 1) %>%
    group_by(DeID_PatientID) %>%
    slice_min(encounter_date, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    select(DeID_PatientID, onset_date = encounter_date) %>%
    rename("{col}_onset" := onset_date)
}

# --- Elixhauser onsets ---
elix_onsets <- c("diabetes", "hypertension", "alcohol_abuse",
                 "arrhythmia", "chf", "obesity",
                 "pulm_circ", "renal_failure") %>%
  map(~ first_onset(elix_dated, .x)) %>%
  reduce(full_join, by = "DeID_PatientID")

# --- Charlson onsets ---
charlson_onsets <- c("charlson_diabetes", "charlson_chf",
                     "charlson_mi", "charlson_pvd", "charlson_renal") %>%
  map(~ first_onset(charlson_dated, .x)) %>%
  reduce(full_join, by = "DeID_PatientID")

# --- Charlson composite scores ---
# Baseline: score at first ever encounter for that patient
# Maximum: highest score ever recorded (most complete coding episode)
charlson_scores <- charlson_dated %>%
  group_by(DeID_PatientID) %>%
  summarise(
    charlson_score_baseline     = first(charlson_score),
    charlson_score_max          = max(charlson_score,     na.rm = TRUE),
    charlson_score_age_baseline = first(charlson_score_age),
    charlson_score_age_max      = max(charlson_score_age, na.rm = TRUE),
    .groups = "drop"
  )

# --- Elixhauser composite scores ---
elix_scores <- elix_dated %>%
  group_by(DeID_PatientID) %>%
  summarise(
    elix_score_baseline = first(elix_score),
    elix_score_max      = max(elix_score, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r combine-and-save}
# One row per patient, all onset dates and scores in wide format
all_patients <- bind_rows(
  elixhauser.data %>% distinct(DeID_PatientID),
  charlson.data   %>% distinct(DeID_PatientID)
) %>% distinct(DeID_PatientID)

comorbidity_onset <- all_patients %>%
  left_join(elix_onsets,     by = "DeID_PatientID") %>%
  left_join(charlson_onsets, by = "DeID_PatientID") %>%
  left_join(charlson_scores, by = "DeID_PatientID") %>%
  left_join(elix_scores,     by = "DeID_PatientID")

# --- Onset summary ---
comorbidity_onset %>%
  select(ends_with("_onset")) %>%
  summarise(across(everything(),
                   list(N   = ~ sum(!is.na(.x)),
                        Pct = ~ round(100 * mean(!is.na(.x)), 1)))) %>%
  pivot_longer(everything(),
               names_to  = c("Condition", ".value"),
               names_sep = "_(?=N$|Pct$)") %>%
  arrange(desc(N)) %>%
  kable(caption = "Comorbidity prevalence — patients with at least one coded encounter")

# --- Score summaries ---
comorbidity_onset %>%
  select(starts_with("charlson_score"), starts_with("elix_score")) %>%
  summary() %>%
  print()
```

## Writing Out Cleaned Files

```{r writing-cleaned}
combined.diagnoses <-
  full_join(osteoporosis.diagnoses,mace.diagnoses, by='DeID_PatientID')

cleaned.diagnoses.file <- 'combined_data/DiagnosesCleaned.csv'
write_csv(combined.diagnoses, file=cleaned.diagnoses.file)

comorbidity.outfile <- "combined_data/ComorbiditiesOnset.csv"
write_csv(comorbidity_onset, comorbidity.outfile)
```

Wrote this out to `r cleaned.diagnoses.file` and `r comorbidity.outfile`.

## Session Information

```{r session-information}
sessionInfo()
```
