---
title: "Fracture Adverse Events in Lipid-Lowering Outcome Trials (ClinicalTrials.gov)"
author: "Dave Bridges"
date: today
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  gfm:
    html-math-method: webtex
theme: journal
execute:
  echo: true
  warning: false
knitr:
  opts_chunk:
    fig-path: "figures-fracture-ae/"
    dev: ["png", "pdf"]
    dpi: 300
    dev.args:
      png:
        type: "cairo-png"
      pdf:
        family: "sans"
---

```{r setup}
#| label: setup
#| message: false

library(httr2)
library(tidyverse)   # dplyr, purrr, stringr, tidyr, ggplot2
library(broom)
library(metafor)     # random-effects meta-analysis
library(knitr)
```

## Motivation

This is a **preliminary, hypothesis-generating** scan of fracture-related adverse
events (AEs) reported on ClinicalTrials.gov for large lipid-lowering cardiovascular
outcome trials. The motivating question is whether lowering LDL-C — and by extension
the cholesterol/calcium axis we have been studying in the EHR cohort — is associated
with a signal in fracture risk. We pull the structured `adverseEventsModule` from the
v2 API, filter to fracture terms, and compare active vs. control arms within each
trial before pooling.

Two drug classes are included:

- **Non-statin LDL-lowering** (PCSK9 inhibitors, ezetimibe) — the original target set.
- **Statins** — added here for context, since they are the dominant LDL-lowering class
  and carry their own (mixed) literature on bone/fracture effects.

::: {.callout-warning}
## Caveats up front
- AE modules report **all-cause** fracture counts from safety surveillance, not
  adjudicated efficacy endpoints. Ascertainment differs across trials.
- Denominators (`numAtRisk`) are the safety-population arm sizes, not person-time, so
  these are **crude risk ratios**, not incidence-rate ratios. Differential follow-up
  between arms is *not* accounted for.
- Many landmark statin trials (4S, WOSCOPS, HPS, LIPID, CARDS) predate mandatory
  results posting and have **no machine-readable AE module** — they will silently
  return empty. Treat the statin set as a convenience sample of registry-era trials.
- A wrong NCT fails silently: it returns either an empty result *or — worse — a real
  but unrelated trial*. In an earlier draft `NCT00831441` (APPRAISE-2, apixaban) and
  `NCT00153231` (a vaginal-prolapse surgery study) had been mislabeled as statin
  trials and one slipped into the pool. Every NCT and **every arm title** below has now
  been confirmed against a live registry pull, and arm roles are assigned by an explicit
  hand-checked map rather than parsed from free-text arm titles.
:::

## Trial registry

```{r trials}
#| label: trials

# class:      drug class for grouping
# comparator: "placebo" or "active" (more- vs less-intensive LDL lowering)
# has_ae:     TRUE = confirmed posted AE module on a live pull (2026-06); FALSE = no module
trials <- tribble(
  ~nct,          ~trial,             ~drug,          ~class,        ~comparator, ~has_ae,
  # --- Non-statin LDL-lowering (original set) ---
  "NCT01764633", "FOURIER",          "evolocumab",   "PCSK9i",      "placebo",   TRUE,
  "NCT01663402", "ODYSSEY OUTCOMES", "alirocumab",   "PCSK9i",      "placebo",   TRUE,
  "NCT00202878", "IMPROVE-IT",       "ezetimibe",    "ezetimibe",   "active",    TRUE,   # eze+simva vs simva
  "NCT00125593", "SHARP",            "eze+simva",    "ezetimibe",   "placebo",   TRUE,   # vs placebo (confounded)
  "NCT02993406", "CLEAR Outcomes",   "bempedoic",    "ATP-citrate", "placebo",   TRUE,
  # --- Statins ---
  "NCT00239681", "JUPITER",          "rosuvastatin", "statin",      "placebo",   TRUE,   # rosuva 20 vs placebo
  # Registry-era statin trials kept for the coverage table but with NO posted AE
  # module (confirmed empty on a live pull), so they contribute nothing downstream:
  "NCT00147602", "SPARCL",           "atorvastatin", "statin",      "placebo",   FALSE,  # atorva 80 vs placebo
  "NCT00327691", "SATURN",           "rosuvastatin", "statin",      "active",    FALSE,  # rosuva vs atorva
  "NCT00468923", "HOPE-3",           "rosuvastatin", "statin",      "placebo",   FALSE   # rosuva vs placebo (corrected NCT)
  # Dropped: NCT00831441 = APPRAISE-2 (apixaban), NCT00153231 = prolapse surgery.
  # Both were wrong NCTs masquerading as statin trials in an earlier draft.
)

trials |> arrange(class, trial) |> kable()
```

Arm roles are assigned by an **explicit map** keyed on the registry's stable event-group
id (`EG000` / `EG001`), confirmed against a live pull of each trial's `eventGroups`. This
replaces the old free-text title heuristic, which silently mislabeled the lone-`Simvastatin`
control arm in IMPROVE-IT and an apixaban arm as "active". Only trials with a posted AE
module appear here.

```{r arm-map}
#| label: arm-map

# role: "treatment" (LDL-lowering / more-intensive arm) vs "control"
arm_roles <- tribble(
  ~nct,          ~group_id, ~role,
  "NCT01764633", "EG000",   "control",     # FOURIER     Placebo
  "NCT01764633", "EG001",   "treatment",   # FOURIER     Evolocumab
  "NCT01663402", "EG000",   "control",     # ODYSSEY     Placebo
  "NCT01663402", "EG001",   "treatment",   # ODYSSEY     Alirocumab
  "NCT00202878", "EG000",   "treatment",   # IMPROVE-IT  Ezetimibe/Simvastatin
  "NCT00202878", "EG001",   "control",     # IMPROVE-IT  Simvastatin (active comparator)
  "NCT00125593", "EG000",   "treatment",   # SHARP       Simvastatin Plus Ezetimibe
  "NCT00125593", "EG001",   "control",     # SHARP       Placebo
  "NCT02993406", "EG000",   "treatment",   # CLEAR       Bempedoic Acid 180 mg
  "NCT02993406", "EG001",   "control",     # CLEAR       Placebo Comparator
  "NCT00239681", "EG000",   "control",     # JUPITER     Placebo
  "NCT00239681", "EG001",   "treatment"    # JUPITER     Rosuvastatin 20 mg
)
```

## Pull adverse-event modules

```{r fetch-ae}
#| label: fetch-ae
#| cache: true

fetch_ae <- function(nct) {
  aem <- request(paste0("https://clinicaltrials.gov/api/v2/studies/", nct)) |>
    req_url_query(format = "json",
                  fields = "resultsSection.adverseEventsModule") |>
    req_user_agent("fracture-AE-pull (research; davebrid@umich.edu)") |>
    req_retry(max_tries = 3) |>
    req_perform() |>
    resp_body_json() |>
    pluck("resultsSection", "adverseEventsModule")

  if (is.null(aem)) return(tibble(nct = nct))

  groups <- map_dfr(aem$eventGroups,
                    \(g) tibble(group_id = g$id, arm = g$title))

  parse_events <- function(events, category) {
    map_dfr(events %||% list(), \(e) {
      map_dfr(e$stats, \(s) tibble(
        category     = category,
        term         = e$term,
        organ        = e$organSystem  %||% NA_character_,
        group_id     = s$groupId,
        num_affected = s$numAffected  %||% NA_integer_,
        num_at_risk  = s$numAtRisk    %||% NA_integer_
      ))
    })
  }

  bind_rows(parse_events(aem$seriousEvents, "serious"),
            parse_events(aem$otherEvents,   "other")) |>
    left_join(groups, by = "group_id") |>
    mutate(nct = nct, .before = 1)
}

# Be polite to the API; a small pause between calls.
ae_raw <- map_dfr(trials$nct, \(x) {
  out <- fetch_ae(x)
  Sys.sleep(0.5)
  out
})

# Did the live pull match what the registry table claims (has_ae)?
ae_coverage <- ae_raw |>
  group_by(nct) |>
  summarise(pulled_ae = any(!is.na(category)), .groups = "drop") |>
  right_join(trials, by = "nct") |>
  transmute(trial, drug, class, comparator,
            expected_ae = has_ae,
            pulled_ae = coalesce(pulled_ae, FALSE))

ae_coverage |> arrange(desc(pulled_ae), class, trial) |> kable()
```

## Filter to fracture terms

```{r fracture-terms}
#| label: fracture-terms

# Broad fracture regex. We exclude "fracture" appearing only as part of unrelated
# MedDRA terms is unlikely, but inspect the distinct terms to be safe.
fracture <- ae_raw |>
  filter(str_detect(term, regex("fracture", ignore_case = TRUE))) |>
  left_join(trials, by = "nct")

# Sanity-check exactly which MedDRA terms were captured.
fracture |>
  distinct(term, organ) |>
  arrange(term) |>
  kable(caption = "Distinct fracture MedDRA terms captured")
```

```{r fracture-by-arm}
#| label: fracture-by-arm

# Collapse to trial x arm x (serious/other). num_at_risk is constant within an
# event group, so max() recovers the denominator. Sum across the (possibly several)
# distinct fracture terms within an arm.
# Keep nct + group_id so we can join the explicit arm-role map (group_id is the
# registry's stable key; arm titles are free text and not trustworthy as a key).
fracture_by_arm <- fracture |>
  group_by(nct, trial, drug, class, comparator, group_id, arm, category) |>
  summarise(affected  = sum(num_affected, na.rm = TRUE),
            n_at_risk = max(num_at_risk,  na.rm = TRUE),
            .groups = "drop") |>
  mutate(pct = 100 * affected / n_at_risk)

fracture_by_arm |>
  arrange(class, trial, category, desc(arm)) |>
  select(trial, drug, class, arm, category, affected, n_at_risk, pct) |>
  kable(digits = 2, caption = "Fracture AE counts by trial, arm, and seriousness")
```

## Assign arm roles (treatment vs. control)

Each event group is labeled by joining the explicit `arm_roles` map (keyed on the stable
`group_id`). The table below pairs each role with the **actual registry arm title** so you
can eyeball that "treatment" and "control" are not flipped — particularly for the
active-comparator trial (IMPROVE-IT) and the confounded SHARP comparison.

```{r arm-roles}
#| label: arm-roles

fracture_roles <- fracture_by_arm |>
  inner_join(arm_roles, by = c("nct", "group_id"))

# Show the actual arm title behind each role so flips are easy to catch.
fracture_roles |>
  distinct(trial, class, comparator, group_id, arm, role) |>
  arrange(class, trial, role) |>
  kable(caption = "Arm-role assignment with registry arm titles — verify nothing is flipped")
```

::: {.callout-note}
**IMPROVE-IT** is an *active-comparator* trial (ezetimibe+simvastatin vs simvastatin
alone), so its "treatment vs control" contrast is *more- vs less-intensive* LDL lowering,
not drug-vs-placebo. **SHARP** randomized eze+simva vs placebo, so its effect cannot be
attributed to ezetimibe alone. Both are flagged `comparator` in the registry and IMPROVE-IT
is excluded from the placebo-controlled pool below.
:::

## Per-trial risk ratios

```{r rr-helper}
#| label: rr-helper

# Shared 2x2 -> risk ratio with Wald CI on the log scale, Haldane–Anscombe 0.5
# continuity correction when any cell is zero. Expects columns a, n1, c, n0
# (a/c = treatment/control events; n1/n0 = treatment/control at risk).
compute_rr <- function(df) {
  df |>
    mutate(
      cc = if_else(a == 0 | c == 0, 0.5, 0),
      rr      = ((a + cc) / (n1 + 2 * cc)) / ((c + cc) / (n0 + 2 * cc)),
      log_rr  = log(rr),
      se_log  = sqrt(1 / (a + cc) - 1 / (n1 + 2 * cc) +
                     1 / (c + cc) - 1 / (n0 + 2 * cc)),
      lci = exp(log_rr - 1.96 * se_log),
      uci = exp(log_rr + 1.96 * se_log)
    )
}
```

```{r risk-ratios}
#| label: risk-ratios

# 2x2 per trial x seriousness: treatment vs control. Keep the arm titles so the
# RR table is self-documenting. Drop rows lacking a clean two-arm contrast.
rr_tab <- fracture_roles |>
  group_by(trial, drug, class, comparator, category, role) |>
  summarise(affected = sum(affected), n_at_risk = max(n_at_risk),
            arm = first(arm), .groups = "drop") |>
  pivot_wider(names_from = role,
              values_from = c(affected, n_at_risk, arm)) |>
  filter(!is.na(affected_treatment), !is.na(affected_control),
         n_at_risk_treatment > 0, n_at_risk_control > 0) |>
  transmute(trial, drug, class, comparator, category,
            a  = affected_treatment, n1 = n_at_risk_treatment,
            c  = affected_control,   n0 = n_at_risk_control,
            treatment_arm = arm_treatment, control_arm = arm_control,
            source = "registry") |>
  compute_rr()

rr_tab |>
  transmute(trial, drug, class, comparator, category,
            `treatment arm` = str_glue("{treatment_arm} ({a}/{n1})"),
            `control arm`   = str_glue("{control_arm} ({c}/{n0})"),
            rr = round(rr, 2),
            `95% CI` = str_glue("{round(lci,2)}–{round(uci,2)}")) |>
  arrange(class, category, trial) |>
  kable(caption = "Crude fracture risk ratio (treatment vs control) by trial, with arm labels")
```

## Manual evidence: pre-registry trials

The landmark statin trials (4S, WOSCOPS, HPS, LIPID, CARDS) predate the FDAAA 2007
results-posting mandate, so they have **no machine-readable AE module** to pull. If/when
you hand-extract fracture counts from a publication, FDA review, or Clinical Study Report,
enter them below and they flow into the same pooled analysis with a `source` flag so the
forest plot shows provenance. Rows with `NA` event counts are skipped automatically, so
this block is safe to leave partially filled.

::: {.callout-important}
Many of these CV trials likely **never recorded fracture as an adverse event** — confirm
the count exists before trusting a blank. Record *where* each number came from in
`source_detail` (e.g. "FDA Stat Review Tbl 12", "Lancet 1994 appendix", "Vivli CSR p.214")
so the provenance is auditable. Use the safety/ITT population denominators and match the
`category` (serious vs other) to how the source classified the event.
:::

```{r manual-ae}
#| label: manual-ae

# Fill a/n1 (treatment events/at-risk) and c/n0 (control events/at-risk) as you find
# them. Leave NA until then. comparator: "placebo" or "active". category: "serious"
# or "other". source_detail: free-text citation for auditability.
manual_ae <- tribble(
  ~trial,    ~drug,          ~class,  ~comparator, ~category, ~a, ~n1,  ~c, ~n0, ~source_detail,
  "4S",      "simvastatin",  "statin", "placebo",  "serious", NA, NA,   NA, NA,  "",   # Merck; Lancet 1994 / FDA review
  "WOSCOPS", "pravastatin",  "statin", "placebo",  "serious", NA, NA,   NA, NA,  "",   # BMS/Sankyo; NEJM 1995
  "HPS",     "simvastatin",  "statin", "placebo",  "serious", NA, NA,   NA, NA,  "",   # Merck/CTSU; Lancet 2002 appendix
  "LIPID",   "pravastatin",  "statin", "placebo",  "serious", NA, NA,   NA, NA,  "",   # BMS; NEJM 1998
  "CARDS",   "atorvastatin", "statin", "placebo",  "serious", NA, NA,   NA, NA,  ""    # Pfizer; Lancet 2004
)

manual_rr <- manual_ae |>
  filter(!is.na(a), !is.na(n1), !is.na(c), !is.na(n0), n1 > 0, n0 > 0) |>
  mutate(treatment_arm = drug, control_arm = "control",
         source = "manual") |>
  compute_rr()

if (nrow(manual_rr) > 0) {
  manual_rr |>
    transmute(trial, drug, class, comparator, category,
              `trt (a/n1)` = str_glue("{a}/{n1}"),
              `ctl (c/n0)` = str_glue("{c}/{n0}"),
              rr = round(rr, 2),
              `95% CI` = str_glue("{round(lci,2)}–{round(uci,2)}"),
              source_detail) |>
    arrange(trial) |>
    kable(caption = "Hand-extracted fracture risk ratios (pre-registry trials)")
} else {
  message("No manual fracture counts entered yet — `manual_ae` is still all NA.")
}

# Combine registry + manual into one evidence base for pooling. Keep only the
# columns both share.
keep <- c("trial", "drug", "class", "comparator", "category",
          "a", "n1", "c", "n0", "rr", "lci", "uci", "source")
rr_all <- bind_rows(select(rr_tab, any_of(keep)),
                    if (nrow(manual_rr) > 0) select(manual_rr, any_of(keep)))
```

## Forest plot and random-effects pooling

We pool **serious** fracture events across placebo-controlled trials with a
random-effects model (REML). Active-comparator trials are excluded from the pooled
estimate but shown in the per-trial table above.

```{r meta}
#| label: meta
#| fig-width: 8
#| fig-height: 6

# Restrict to serious events and to placebo-controlled trials (per the `comparator`
# flag — no hardcoded trial list). Active-comparator trials (IMPROVE-IT) are excluded
# automatically. Pulls from rr_all, so any hand-entered manual rows join the pool.
meta_dat <- rr_all |>
  filter(category == "serious", comparator == "placebo") |>
  arrange(source, trial)

message("Pooling ", nrow(meta_dat), " placebo-controlled trials (",
        sum(meta_dat$source == "registry"), " registry / ",
        sum(meta_dat$source == "manual"), " manual): ",
        paste(meta_dat$trial, collapse = ", "))

if (nrow(meta_dat) >= 2) {
  es <- escalc(measure = "RR",
               ai = a, n1i = n1,    # treatment arm: events, n
               ci = c, n2i = n0,    # control arm: events, n
               data = meta_dat,
               slab = trial)

  res <- rma(yi, vi, data = es, method = "REML")

  forest(res, atransf = exp,
         at = log(c(0.25, 0.5, 1, 2, 4)),
         xlab = "Risk ratio (serious fracture, treatment vs control)",
         ilab = es$source, ilab.xpos = -3.2,    # show provenance column
         mlab = "RE model (REML)",
         header = c("Trial", "RR [95% CI]"))
  text(-3.2, res$k + 2, "Source", font = 2)

  res
} else {
  message("Not enough trials with serious fracture data to pool — ",
          "add manual rows or verify NCTs and re-run.")
}
```

```{r meta-by-class}
#| label: meta-by-class

# Subgroup pooling by drug class, if enough classes are represented.
if (exists("res") && nrow(meta_dat) >= 2 && n_distinct(meta_dat$class) >= 2) {
  es <- escalc(measure = "RR", ai = a, n1i = n1, ci = c, n2i = n0,
               data = meta_dat, slab = trial)
  res_class <- rma(yi, vi, mods = ~ class, data = es, method = "REML")
  res_class
}
```

## Session info

```{r session}
#| label: session
sessionInfo()
```

## Next steps

- **Verify the `FALSE` NCTs** in the registry table — several statin entries are
  best-guesses and may lack a posted AE module (they'll show `has_ae = FALSE` above).
- Swap crude risk ratios for **incidence-rate ratios** where person-time is recoverable
  (some modules expose `frequencyThreshold` / time-frame metadata).
- Hand-correct arm roles for active-comparator trials and decide whether to keep them.
- If a real signal emerges, cross-reference against the EHR-derived osteoporosis /
  fracture competing-risks analyses in this project for triangulation.
```

