---
title: "Cleaning of Lab Result Data"
author: "Dave Bridges"
date: "2026-03-03"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  gfm:
    html-math-method: webtex

theme: journal

execute:
  echo: true
  warning: false

knitr:
  opts_chunk:
    fig-path: "figures/"          # folder for all figure files
    dev: ["png", "pdf"]           # both formats for every plot
    dpi: 300                      # ← this is the correct place for resolution
    dev.args:
      png:
        type: "cairo-png"         # better anti-aliasing / font rendering (optional but recommended)
      pdf:
        family: "sans"            # font family (optional)
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)
library(knitr)
library(broom)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

Clean the data for various lab results, for use in other scrupts

## Raw Data

```{r data-input}
library(readr) #loads the readr package
labs.filename <- "combined_data/LabResults.csv" #input file(s)

lab.data <- read_csv(labs.filename)
```

These data can be found in `r getwd()` in a file named `r ifelse(labs.filename %in% dir(), labs.filename, "no file found")`.  This input file was most recently updated on `r ifelse(labs.filename %in% dir(), format(file.info(labs.filename)$mtime, "%Y-%m-%d"), "unknown")`.  This script was most recently updated on `r date()`.

There are `r distinct(lab.data, DeID_PatientID) |> nrow()` participants in this dataset with lab values.

## Data Cleaning

```{r data-cleaning}
#uncomment to see list of all labs
lab.data |>
  group_by(ORDER_NAME) |>
  count() |>
  arrange(-n) -> lab.summary.table

lab.summary.table |>
  kable(caption="All labs by test name")

#lab.summary.table |>
  #filter(grepl("ca", ORDER_NAME, ignore.case = TRUE)) #case insensitive has a ca in it
  #filter(grepl("cho", ORDER_NAME, ignore.case = TRUE)) #case insensitive has a ca in it
  #filter(grepl("tri", ORDER_NAME, ignore.case = TRUE)) 

ionized.calcium.labs <- c('CALCIUM, IONIZED','Ionized Calcium EX')
serum.calcium.labs <- c('Calcium EX','CALCIUM LEVEL')
calcium.labs <- c(serum.calcium.labs,ionized.calcium.labs)
tg.labs <- c('TRIGLYCERIDES','TRIGLYCERIDES, FLUID','Triglycerides EX')
ldl.labs <- c('LDL CHOLESTEROL, DIRECT','LOW DENSITY LIPOPROTEIN CHOL')
hdl.labs <- c('HDL EX','HIGH DENSITY LIPOPROTEIN CHOL')
cholesterol.labs <- c('CHOLESTEROL','Cholesterol EX','CHOLESTEROL, TOTAL and HDL')#includes both total and HDL, so duplicate measures, total has to be definitively the higher of the two, they have the same DeID_EncounterID.  need to filter the HDLs into a different category

lab.data.cleaned <-
  lab.data |>
  mutate(VALUE= as.numeric(VALUE)) |> #first make all the values into 
  mutate(test_name = case_when(ORDER_NAME %in% ionized.calcium.labs ~ 'Ionized Calcium',
                               ORDER_NAME %in% serum.calcium.labs ~ 'Serum Calcium',
                               ORDER_NAME %in% tg.labs ~ 'Triglycerides',
                               ORDER_NAME %in% hdl.labs ~ 'HDL-C',
                               ORDER_NAME %in% ldl.labs ~ 'LDL-C',
                               ORDER_NAME %in% cholesterol.labs ~ 'chol_temp' )) %>% #temporary placeholder for cholesterol
  group_by(DeID_EncounterID, test_name) |>
  mutate(test_name = case_when(test_name == 'chol_temp' ~ if_else(VALUE == max(VALUE, na.rm = TRUE),'Total Cholesterol','HDL-C'),
    TRUE ~ test_name
  )) |>
  ungroup() |>
  mutate(value = case_when(UNIT=="mmol/L"&ORDER_NAME %in% calcium.labs ~ VALUE*4.01,
                           UNIT=="MMOL/L"&ORDER_NAME %in% calcium.labs ~ VALUE*4.01,
                          UNIT=="mg/dL"&ORDER_NAME %in% calcium.labs ~ VALUE,
                          ORDER_NAME %in% c(cholesterol.labs,tg.labs,hdl.labs,ldl.labs) ~ VALUE)) |> #always given in mg/dL for calcium
  filter(!is.na(value)) |> #removes tests without values
  filter(!is.na(test_name))  #removes non-cleaned tests

# testing to count how many different units are present now for each test
#lab.data.cleaned |>
#  group_by(test_name, unit_upper = toupper(UNIT)) |>
#  summarise(n = n(), .groups = "drop") |>
#  arrange(test_name, desc(n)) |>
#  kable(caption="Count of the units in each data type")

lab.data.cleaned |>
  group_by(test_name) |>
  summarize(n = length(!is.na(value)),
            mean=mean(value,na.rm=T),
            sd = sd(value,na.rm=T),
            min = min(value,na.rm=T),
            max = max(value,na.rm=T)) |>
  arrange(-n) |>
  kable(caption="Lab results after cleaning and combining")
```

```{r merging-with-encounter-data}
encounter.datafile <- 'combined_data/EncounterAll.csv'
encounter.anthro.datafile <- 'combined_data/EncounterAnthropometricsBMI.csv'

#load in encounter metadata
encounter.data <- read_csv(encounter.datafile)
encounter.anthro.data <- read_csv(encounter.anthro.datafile)

merged.lab.data <-
  lab.data.cleaned |>
  left_join(encounter.data, by=c("DeID_PatientID","DeID_EncounterID")) |>
  left_join(encounter.anthro.data, by=c("DeID_PatientID","DeID_EncounterID"))

output.file <- "combined_data/LabResultsCleaned.csv"
merged.lab.data |>
  write_csv(file=output.file)
```

After cleaning are `r distinct(lab.data.cleaned, DeID_PatientID) |> nrow()` participants in this dataset with lab values.

The cleaned results are output at **`r output.file`**.


## Session Information

```{r session-information}
sessionInfo()
```
