---
title: "Baseline LDL-C and MACE — Fixed Exposure Cox Model"
author: "Dave Bridges"
date: "2026-03-27"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  gfm:
    html-math-method: webtex

theme: journal

execute:
  echo: true
  warning: false

knitr:
  opts_chunk:
    fig-path: "figures-baseline/"
    dev: ["png", "pdf"]
    dpi: 300
    dev.args:
      png:
        type: "cairo-png"
      pdf:
        family: "sans"
---

```{r global_options}
#| echo: false
#| message: false

se <- function(x) sd(x, na.rm = TRUE) / sqrt(length(x))

library(tidyverse)
library(knitr)
library(broom)
library(lubridate)
library(survival)
library(survminer)

color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

This script implements a **baseline LDL-C exposure model** as a methodological companion to the primary time-varying LDL-C analysis (`ldlc_mace_analysis.qmd`). The two scripts answer related but distinct scientific questions:

- **Primary script (time-varying)**: Does a patient's *current* LDL-C level predict near-term MACE risk? LDL-C is updated at each measurement visit using last-observation-carried-forward (LOCF).
- **This script (baseline)**: Does a patient's LDL-C level *at the time they enter care* predict their long-term MACE risk? LDL-C is fixed at the first measurement and does not change over follow-up.

The baseline approach is methodologically closer to landmark observational studies (Framingham, INTERHEART) that measured LDL-C once at enrollment. It avoids the treatment dilution problem inherent in time-varying models — post-treatment LDL-C values, which reflect statin response rather than underlying atherogenic burden, do not enter the exposure definition. The tradeoff is that baseline LDL-C may not represent the patient's lipid burden throughout the full follow-up period, particularly in a cohort with 14-year median follow-up.

Because the exposure is fixed at time zero, statin adjustment requires a different approach than the primary model. We stratify on **baseline statin status** — whether the patient was already on a statin at the time of their first LDL-C measurement — rather than using a time-varying statin covariate. This is the most internally consistent approach: the exposure (LDL-C) and the primary confounder adjustment (statin use) are both anchored at the same time point. It cleanly separates patients whose baseline LDL-C reflects pre-treatment lipid biology from those already on treatment.

The cohort derivation, exclusion criteria, and secondary covariates (diabetes, hypertension, age stratification, follow-up period) are identical to the primary script to ensure comparability.

---

## Raw Data

All raw data files are loaded independently — this script is fully self-contained and does not depend on objects from the primary analysis script.

```{r data-input}
#| message: false

# --- Full study population ---
demographic.data <- read_csv("combined_data/DemographicInfo.csv",
                             show_col_types = FALSE) %>%
  mutate(DeID_PatientID = as.character(DeID_PatientID))

# --- Diagnoses (MACE and Osteoporosis only) ---
diagnosis.data <- read_csv("combined_data/DiagnosesCleaned.csv",
                           show_col_types = FALSE) %>%
  mutate(
    DeID_PatientID = as.character(DeID_PatientID),
    MACE.onset     = ymd(MACE.onset)
  )

# --- Labs ---
lab.data <- read_csv("combined_data/LabResultsCleaned.csv",
                     show_col_types = FALSE) %>%
  mutate(
    DeID_PatientID = as.character(DeID_PatientID),
    lab_date = coalesce(
      parse_date_time(DeID_COLLECTION_DATE,
                      orders = c("ymd", "mdy", "ymd HMS", "mdy HM"),
                      quiet  = TRUE),
      parse_date_time(DeID_AdmitDate,
                      orders = c("ymd", "mdy", "ymd HMS", "mdy HM"),
                      quiet  = TRUE)
    ))

ldlc.data <- lab.data %>% filter(test_name == "LDL-C")
```

The full demographic file contains `r scales::comma(n_distinct(demographic.data$DeID_PatientID))` unique patients.

---

## Cohort Derivation

The cohort derivation follows the identical pipeline as the primary script. Patients are included if they have at least one valid LDL-C measurement after age 18, a valid follow-up period, and a classifiable MACE outcome.

```{r cohort-derivation}
#| message: false

# --- Build full cohort with MACE flag ---
full_cohort <- demographic.data %>%
  left_join(
    diagnosis.data %>%
      select(DeID_PatientID, MACE, MACE.onset,
             Osteoporosis, Osteoporosis.onset),
    by = "DeID_PatientID"
  ) %>%
  mutate(MACE_flag = if_else(MACE == TRUE, 1L, 0L, missing = 0L))

ids_demo <- unique(full_cohort$DeID_PatientID)
ids_ldlc <- unique(ldlc.data$DeID_PatientID)

# --- Clean LDL-C data ---
ldlc_clean <- ldlc.data %>%
  filter(!is.na(lab_date), !is.na(AgeInYears)) %>%
  filter(value > 0, value <= 400) %>%
  mutate(BMI = if_else(BMI < 10 | BMI > 80, NA_real_, BMI)) %>%
  filter(AgeInYears >= 18) %>%
  group_by(DeID_PatientID, lab_date) %>%
  summarise(
    LDL_value  = mean(value,      na.rm = TRUE),
    AgeInYears = mean(AgeInYears, na.rm = TRUE),
    BMI        = mean(BMI,        na.rm = TRUE),
    .groups    = "drop"
  ) %>%
  arrange(DeID_PatientID, lab_date)

# --- Clean diagnosis data ---
diag_clean <- full_cohort %>%
  filter(!(MACE_flag == 1L & is.na(MACE.onset))) %>%
  select(DeID_PatientID, MACE.onset, MACE_flag)

# --- Restrict to shared patients ---
shared_ids  <- intersect(diag_clean$DeID_PatientID,
                         ldlc_clean$DeID_PatientID)
diag_cohort <- diag_clean  %>% filter(DeID_PatientID %in% shared_ids)
ldlc_cohort <- ldlc_clean  %>% filter(DeID_PatientID %in% shared_ids)

# --- Encounter data for censoring ---
encounter.data <- read_csv("combined_data/EncounterAll.csv",
                           show_col_types = FALSE) %>%
  mutate(
    DeID_PatientID = as.character(DeID_PatientID),
    EncounterDate  = mdy_hm(DeID_AdmitDate)
  )

last_encounter <- encounter.data %>%
  filter(!is.na(EncounterDate), DeID_PatientID %in% shared_ids) %>%
  group_by(DeID_PatientID) %>%
  slice_max(EncounterDate, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(DeID_PatientID, last_encounter_date = EncounterDate)

# --- First LDL-C measurement = time zero and baseline exposure ---
first_ldlc <- ldlc_cohort %>%
  group_by(DeID_PatientID) %>%
  slice_min(lab_date, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(DeID_PatientID, t0 = lab_date,
         baseline_LDL   = LDL_value,
         baseline_age   = AgeInYears,
         baseline_BMI   = BMI)

# --- Build base cohort ---
base_cohort_full <- diag_cohort %>%
  left_join(first_ldlc,     by = "DeID_PatientID") %>%
  left_join(last_encounter, by = "DeID_PatientID") %>%
  mutate(
    t_end = case_when(
      MACE_flag == 1              ~ MACE.onset,
      !is.na(last_encounter_date) ~ last_encounter_date,
      TRUE                        ~ NA_POSIXct_
    ),
    follow_up_days = as.numeric(difftime(t_end, t0, units = "days"))
  )

base_cohort <- base_cohort_full %>%
  filter(follow_up_days > 0) %>%
  mutate(follow_up_days = pmin(follow_up_days, 14610))

# --- CONSORT flow ---
tibble(
  Step = c(
    "Full demographic cohort",
    "No LDL-C data",
    "MACE=TRUE with missing onset date",
    "Negative or zero follow-up (prevalent MACE + invalid controls)",
    "Final analytic cohort"
  ),
  N = c(
    n_distinct(demographic.data$DeID_PatientID),
    -length(setdiff(ids_demo, ids_ldlc)),
    -(nrow(full_cohort %>% filter(MACE_flag == 1 & is.na(MACE.onset)))),
    -(nrow(base_cohort_full %>% filter(follow_up_days <= 0))),
    nrow(base_cohort)
  )
) %>% kable(caption = "CONSORT flow — cohort derivation")
```

The final analytic cohort contains **`r nrow(base_cohort)` patients** (`r sum(base_cohort$MACE_flag)` incident MACE cases and `r sum(base_cohort$MACE_flag == 0)` controls), identical to the primary script.

---

## Baseline LDL-C Characterisation

Unlike the primary script, the exposure in this analysis is a single value per patient — the LDL-C measured at their first visit. It is important to characterise this value and understand how it relates to statin use at baseline, since statin-treated patients will have systematically lower measured LDL-C than statin-naive patients.

```{r baseline-ldlc-characterisation}
baseline_summary <- base_cohort %>%
  summarise(
    `N patients`              = n(),
    `Min baseline LDL-C`      = round(min(baseline_LDL, na.rm = TRUE), 1),
    `25th percentile`         = round(quantile(baseline_LDL, 0.25, na.rm = TRUE), 1),
    `Median baseline LDL-C`   = round(median(baseline_LDL, na.rm = TRUE), 1),
    `Mean baseline LDL-C`     = round(mean(baseline_LDL, na.rm = TRUE), 1),
    `75th percentile`         = round(quantile(baseline_LDL, 0.75, na.rm = TRUE), 1),
    `Max baseline LDL-C`      = round(max(baseline_LDL, na.rm = TRUE), 1),
    `Missing baseline LDL-C`  = sum(is.na(baseline_LDL))
  )

baseline_summary %>%
  pivot_longer(everything(), names_to = "Metric", values_to = "Value") %>%
  kable(caption = "Baseline LDL-C distribution across analytic cohort")

# Distribution by MACE status
base_cohort %>%
  group_by(MACE_flag) %>%
  summarise(
    N      = n(),
    Mean   = round(mean(baseline_LDL, na.rm = TRUE), 1),
    Median = round(median(baseline_LDL, na.rm = TRUE), 1),
    SD     = round(sd(baseline_LDL, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  mutate(MACE_flag = if_else(MACE_flag == 1, "Cases (MACE=1)",
                             "Controls (MACE=0)")) %>%
  rename(Group = MACE_flag) %>%
  kable(caption = "Baseline LDL-C by case/control status")

# Histogram
ggplot(base_cohort, aes(x = baseline_LDL,
                         fill = factor(MACE_flag,
                                       labels = c("Control", "MACE")))) +
  geom_histogram(binwidth = 10, alpha = 0.7, position = "identity",
                 colour = "white") +
  scale_fill_manual(values = color_scheme) +
  labs(title    = "Baseline LDL-C distribution by MACE status",
       subtitle = "LDL-C at first measurement (time zero)",
       x = "Baseline LDL-C (mg/dL)", y = "Count", fill = NULL) +
  theme_classic() +
  theme(legend.position = "bottom")
```

---

## Baseline Statin Status

The primary confounder adjustment for this model is baseline statin status — whether the patient was already on a statin at the time of their first LDL-C measurement. This is ascertained by checking whether any statin prescription interval was active at time zero.

Patients already on a statin at baseline will have lower measured LDL-C that reflects treatment response rather than underlying atherogenic burden. Stratifying on this variable allows the model to estimate the LDL-C/MACE association separately within statin-naive and statin-treated patients, avoiding the assumption that the relationship is the same in both groups.

```{r baseline-statin-status}
#| message: false

statin_intervals <- read_csv(
  "combined_data/MedicationOrdersCleanedStatins.csv",
  show_col_types = FALSE
) %>%
  mutate(
    DeID_PatientID = as.character(DeID_PatientID),
    period_start   = as_date(period_start),
    period_end     = as_date(period_end),
    intensity      = factor(intensity, levels = c("low", "moderate", "high"))
  )

# Convert statin intervals to days-since-t0
statin_tv <- statin_intervals %>%
  inner_join(first_ldlc %>% select(DeID_PatientID, t0),
             by = "DeID_PatientID") %>%
  mutate(
    tstart_statin = as.numeric(period_start - as_date(t0)),
    tstop_statin  = as.numeric(period_end   - as_date(t0))
  ) %>%
  filter(tstop_statin > 0) %>%
  mutate(tstart_statin = pmax(tstart_statin, 0)) %>%
  filter(tstop_statin > tstart_statin) %>%
  select(DeID_PatientID, tstart_statin, tstop_statin, intensity)

# Baseline statin status: was the patient on a statin at t=0?
baseline_statin <- statin_intervals %>%
  inner_join(first_ldlc %>% select(DeID_PatientID, t0),
             by = "DeID_PatientID") %>%
  mutate(
    tstart_statin  = as.numeric(period_start - as_date(t0)),
    tstop_statin   = as.numeric(period_end   - as_date(t0)),
    on_at_baseline = tstart_statin <= 0 & tstop_statin > 0
  ) %>%
  group_by(DeID_PatientID) %>%
  summarise(
    baseline_statin    = as.integer(any(on_at_baseline, na.rm = TRUE)),
    baseline_intensity = {
      active <- as.character(intensity[on_at_baseline])
      active <- active[!is.na(active)]
      if (length(active) == 0) NA_character_
      else {
        lvls  <- c("low", "moderate", "high")
        ranks <- match(active, lvls)
        active[which.max(ranks)]
      }
    },
    .groups = "drop"
  )

# Join baseline statin status onto base_cohort
base_cohort <- base_cohort %>%
  left_join(baseline_statin, by = "DeID_PatientID") %>%
  mutate(
    baseline_statin    = coalesce(baseline_statin, 0L),
    baseline_intensity = factor(baseline_intensity,
                                levels = c("low", "moderate", "high")),
    baseline_LDL_40    = baseline_LDL / 40
  )

# Statin status summary
base_cohort %>%
  group_by(baseline_statin) %>%
  summarise(
    N                      = n(),
    `Mean LDL-C (mg/dL)`   = round(mean(baseline_LDL, na.rm = TRUE), 1),
    `Median LDL-C (mg/dL)` = round(median(baseline_LDL, na.rm = TRUE), 1),
    `MACE events`          = sum(MACE_flag),
    `MACE rate (%)`        = round(100 * mean(MACE_flag), 1),
    .groups = "drop"
  ) %>%
  mutate(baseline_statin = if_else(baseline_statin == 1,
                                   "On statin at baseline",
                                   "Statin-naive at baseline")) %>%
  rename(`Statin status` = baseline_statin) %>%
  kable(caption = "Baseline LDL-C and MACE rates by baseline statin status")

# Intensity distribution among baseline statin users
base_cohort %>%
  filter(baseline_statin == 1) %>%
  count(baseline_intensity) %>%
  kable(caption = "Statin intensity distribution among baseline statin users")
```

The difference in mean LDL-C between statin-naive and statin-treated patients at baseline quantifies the treatment dilution effect — the degree to which statin use at baseline depresses the measured LDL-C relative to underlying atherogenic burden.

---

## Comorbidity Data

Comorbidity onset dates are loaded from the pre-processed file and converted to a baseline flag — whether each condition had been diagnosed *before or at time zero*. This is appropriate for the baseline model since we want to characterise the patient's comorbidity burden at the time of the first LDL-C measurement, not over follow-up.

For consistency with the primary model, diabetes and hypertension are also modelled as time-varying covariates (turning on at first diagnosis date, whether before or after time zero). This avoids discarding post-baseline incident diagnoses.

```{r load-comorbidities}
#| message: false

comorbidity_onset <- read_csv(
  "combined_data/ComorbiditiesOnset.csv",
  show_col_types = FALSE
) %>%
  mutate(
    DeID_PatientID      = as.character(DeID_PatientID),
    diabetes_onset      = as_datetime(diabetes_onset),
    hypertension_onset  = as_datetime(hypertension_onset),
    obesity_onset       = as_datetime(obesity_onset),
    renal_failure_onset = as_datetime(renal_failure_onset)
  )

# Time-varying comorbidity covariates (days since t0, same as primary model)
comorbidity_tv <- comorbidity_onset %>%
  inner_join(first_ldlc %>% select(DeID_PatientID, t0),
             by = "DeID_PatientID") %>%
  mutate(
    t_diabetes     = as.numeric(difftime(diabetes_onset,
                                          t0, units = "days")),
    t_hypertension = as.numeric(difftime(hypertension_onset,
                                          t0, units = "days")),
    t_renal        = as.numeric(difftime(renal_failure_onset,
                                          t0, units = "days")),
    t_obesity      = as.numeric(difftime(obesity_onset,
                                          t0, units = "days"))
  ) %>%
  select(DeID_PatientID, t_diabetes, t_hypertension, t_renal, t_obesity)

# Baseline comorbidity flags (diagnosed at or before t0)
baseline_comorbidities <- comorbidity_onset %>%
  inner_join(first_ldlc %>% select(DeID_PatientID, t0),
             by = "DeID_PatientID") %>%
  mutate(
    baseline_diabetes     = as.integer(!is.na(diabetes_onset) &
                                         diabetes_onset <= t0),
    baseline_hypertension = as.integer(!is.na(hypertension_onset) &
                                         hypertension_onset <= t0),
    baseline_obesity      = as.integer(!is.na(obesity_onset) &
                                         obesity_onset <= t0),
    baseline_renal        = as.integer(!is.na(renal_failure_onset) &
                                         renal_failure_onset <= t0)
  ) %>%
  select(DeID_PatientID, baseline_diabetes, baseline_hypertension,
         baseline_obesity, baseline_renal)

base_cohort <- base_cohort %>%
  left_join(baseline_comorbidities, by = "DeID_PatientID") %>%
  mutate(across(starts_with("baseline_d") | starts_with("baseline_h") |
                  starts_with("baseline_o") | starts_with("baseline_r"),
                ~ replace_na(.x, 0L)))

# Baseline comorbidity prevalence
tibble(
  Condition = c("Diabetes", "Hypertension", "Obesity", "Renal failure"),
  `N at baseline` = c(
    sum(base_cohort$baseline_diabetes),
    sum(base_cohort$baseline_hypertension),
    sum(base_cohort$baseline_obesity),
    sum(base_cohort$baseline_renal)
  ),
  `% of cohort` = round(100 * c(
    mean(base_cohort$baseline_diabetes),
    mean(base_cohort$baseline_hypertension),
    mean(base_cohort$baseline_obesity),
    mean(base_cohort$baseline_renal)
  ), 1)
) %>% kable(caption = "Comorbidity prevalence at baseline (time zero)")
```

---

## Building the Cox Model Dataset

The counting process structure is retained for consistency with the primary model and to allow time-varying comorbidity covariates to be included. The key difference is that `baseline_LDL` is joined as a fixed covariate — it does not change across intervals for a given patient.

```{r tmerge-construction}
surv_base <- base_cohort %>%
  mutate(
    DeID_PatientID = as.character(DeID_PatientID),
    tstart_days    = 0,
    tstop_days     = as.numeric(follow_up_days)
  )

# Build tmerge structure — baseline LDL is fixed, no LOCF needed
surv.data <- tmerge(
  data1 = surv_base, data2 = surv_base,
  id    = DeID_PatientID,
  MACE  = event(tstop_days, MACE_flag)
) %>%
  filter(tstop - tstart >= 0.5)

# Verify
tibble(
  Metric = c("Rows", "Patients", "MACE events",
             "Patients with >1 event (must be 0)"),
  Value  = c(
    nrow(surv.data),
    n_distinct(surv.data$DeID_PatientID),
    sum(surv.data$MACE),
    surv.data %>% group_by(DeID_PatientID) %>%
      summarise(e = sum(MACE)) %>% filter(e > 1) %>% nrow()
  )
) %>% kable(caption = "tmerge dataset structure")
```

### Adding Time-Varying Comorbidity Covariates

Diabetes and hypertension are added as time-varying covariates (turning on permanently at first diagnosis, whether that occurred before or after time zero) to match the primary model's covariate structure. This ensures the LDL-C HR is estimated under the same covariate conditions in both models.

```{r add-tvc}
comorbidity_tv_cohort <- comorbidity_tv %>%
  filter(DeID_PatientID %in% base_cohort$DeID_PatientID)

surv.data <- surv.data %>%
  # Time-varying comorbidity onset dates only — baseline_LDL, baseline_statin
  # etc. are already in surv.data carried through from surv_base via tmerge
  left_join(comorbidity_tv_cohort, by = "DeID_PatientID") %>%
  mutate(
    on_diabetes     = as.integer(!is.na(t_diabetes)     & tstart >= t_diabetes),
    on_hypertension = as.integer(!is.na(t_hypertension) & tstart >= t_hypertension),
    on_renal        = as.integer(!is.na(t_renal)        & tstart >= t_renal),
    on_obesity      = as.integer(!is.na(t_obesity)      & tstart >= t_obesity),
    period = factor(
      case_when(
        tstart < 1825 ~ "0-5yr",
        tstart < 3650 ~ "5-10yr",
        TRUE          ~ "10yr+"
      ),
      levels = c("0-5yr", "5-10yr", "10yr+")
    ),
    age_cat = cut(baseline_age,
                  breaks = c(18, 45, 55, 65, 75, Inf),
                  labels = c("<45", "45-54", "55-64", "65-74", "75+"),
                  right  = FALSE),
    cardiomet = factor(
      case_when(
        on_diabetes == 1 & on_hypertension == 1 ~ "DM+HTN",
        on_diabetes == 1 & on_hypertension == 0 ~ "DM only",
        on_diabetes == 0 & on_hypertension == 1 ~ "HTN only",
        TRUE                                     ~ "Neither"
      ),
      levels = c("Neither", "HTN only", "DM only", "DM+HTN")
    )
  ) %>%
  select(-t_diabetes, -t_hypertension, -t_renal, -t_obesity)

# Verify key columns present
cat("Key columns present:\n")
print(c("baseline_LDL", "baseline_LDL_40", "baseline_statin",
        "cardiomet", "period", "age_cat") %in% names(surv.data))

surv.data %>%
  summarise(
    `% intervals with diabetes Dx`  = round(100 * mean(on_diabetes == 1), 1),
    `% intervals with HTN Dx`       = round(100 * mean(on_hypertension == 1), 1),
    `% intervals statin at baseline` = round(100 * mean(baseline_statin == 1), 1)
  ) %>%
  pivot_longer(everything(),
               names_to = "Covariate", values_to = "% of intervals") %>%
  kable(caption = "Covariate coverage across intervals")

surv.data %>%
  filter(MACE == 1) %>%
  count(period, baseline_statin, age_cat, cardiomet) %>%
  arrange(n) %>%
  head(20) %>%
  kable(caption = "MACE events per stratum")
```

## Cox Proportional Hazards Models

### Unadjusted Baseline LDL-C Model

The unadjusted model estimates the crude association between baseline LDL-C and MACE risk with no covariate adjustment. As in the primary script, this is expected to show a paradoxical inverse association driven by confounding by indication — statin users have lower LDL-C and higher underlying MACE risk.

```{r unadjusted-model}
fit_unadj <- coxph(
  Surv(tstart, tstop, MACE) ~ baseline_LDL,
  data = surv.data, ties = "efron", id = DeID_PatientID
)

broom::tidy(fit_unadj, exponentiate = TRUE, conf.int = TRUE) %>%
  select(Term = term, HR = estimate,
         `95% CI low` = conf.low, `95% CI high` = conf.high,
         `p-value` = p.value) %>%
  mutate(across(where(is.numeric), ~round(.x, 4))) %>%
  kable(caption = "Unadjusted model — baseline LDL-C effect on MACE (per 1 mg/dL)")
```

### Proportional Hazards — Unadjusted

```{r ph-unadjusted}
ph_unadj <- cox.zph(fit_unadj)

tibble(
  Term      = rownames(ph_unadj$table),
  `Chi-sq`  = round(ph_unadj$table[, "chisq"], 3),
  df        = ph_unadj$table[, "df"],
  `p-value` = round(ph_unadj$table[, "p"], 4)
) %>% kable(caption = "Schoenfeld residuals — unadjusted model")

ggcoxzph(ph_unadj, font.main = 12, ggtheme = theme_minimal())
```

### Primary Baseline Model — Stratified Cox

The primary model stratifies on baseline statin status, age category, and follow-up period, and adjusts for diabetes and hypertension as time-varying regression terms. This is directly comparable to the primary time-varying LDL-C model in `ldlc_mace_analysis.qmd`.

**Caveat on baseline statin stratification**: This approach estimates the LDL-C effect *within* statin-naive and statin-treated subgroups, which is appropriate for the baseline framing. It does not account for statin initiation during follow-up. Patients who were statin-naive at baseline and later initiated statins contribute person-time to the statin-naive stratum throughout their follow-up, which may attenuate the LDL-C association in that stratum if late statin initiation is common. This is an inherent limitation of the fixed-exposure design and is discussed in the comparison with the time-varying model.
```{r primary-model}
# Diabetes and hypertension combined into composite cardiomet stratum
# to avoid sparse cells. Stratifying separately gives 120 strata
# with multiple cells having <5 events; composite gives 60 strata.

fit_primary <- coxph(
  Surv(tstart, tstop, MACE) ~ baseline_LDL +
    strata(period) + strata(baseline_statin) +
    strata(age_cat) + strata(cardiomet),
  data = surv.data, ties = "efron", id = DeID_PatientID
)

fit_primary_40 <- coxph(
  Surv(tstart, tstop, MACE) ~ baseline_LDL_40 +
    strata(period) + strata(baseline_statin) +
    strata(age_cat) + strata(cardiomet),
  data = surv.data, ties = "efron", id = DeID_PatientID
)

# All terms (per 1 mg/dL)
broom::tidy(fit_primary, exponentiate = TRUE, conf.int = TRUE) %>%
  select(Term = term, HR = estimate,
         `95% CI low` = conf.low, `95% CI high` = conf.high,
         `p-value` = p.value) %>%
  mutate(across(where(is.numeric), ~round(.x, 4))) %>%
  kable(caption = "Primary baseline model — LDL-C HR per 1 mg/dL")

# Per 40 mg/dL
broom::tidy(fit_primary_40, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term == "baseline_LDL_40") %>%
  select(Term = term, HR = estimate,
         `95% CI low` = conf.low, `95% CI high` = conf.high,
         `p-value` = p.value) %>%
  mutate(across(where(is.numeric), ~round(.x, 4))) %>%
  kable(caption = "Primary baseline model — LDL-C HR per 40 mg/dL")
```

### Proportional Hazards — Primary Baseline Model

```{r ph-primary}
ph_primary <- cox.zph(fit_primary)

tibble(
  Term      = rownames(ph_primary$table),
  `Chi-sq`  = round(ph_primary$table[, "chisq"], 3),
  df        = ph_primary$table[, "df"],
  `p-value` = round(ph_primary$table[, "p"], 4)
) %>% kable(caption = "Schoenfeld residuals — primary baseline model")

ggcoxzph(ph_primary, font.main = 12, ggtheme = theme_minimal())
```

`r if(ph_primary$table["GLOBAL", "p"] > 0.05) "The proportional hazards assumption holds in the primary baseline model (global p > 0.05)." else paste0("**Warning:** PH assumption violated: global p = ", round(ph_primary$table['GLOBAL','p'], 4), ". Review individual terms.")`

### Sensitivity Analyses

```{r sensitivity-analyses}
fit_no_comorbid <- coxph(
  Surv(tstart, tstop, MACE) ~ baseline_LDL +
    strata(period) + strata(baseline_statin) + strata(age_cat),
  data = surv.data, ties = "efron", id = DeID_PatientID
)

fit_full <- coxph(
  Surv(tstart, tstop, MACE) ~ baseline_LDL +
    on_renal + on_obesity +
    strata(period) + strata(baseline_statin) +
    strata(age_cat) + strata(cardiomet),
  data = surv.data, ties = "efron", id = DeID_PatientID
)

# Statin-naive: baseline_statin drops from strata (constant = 0)
fit_statin_naive <- coxph(
  Surv(tstart, tstop, MACE) ~ baseline_LDL +
    strata(period) + strata(age_cat) + strata(cardiomet),
  data = surv.data %>% filter(baseline_statin == 0),
  ties = "efron", id = DeID_PatientID
)

fit_statin_naive_40 <- coxph(
  Surv(tstart, tstop, MACE) ~ baseline_LDL_40 +
    strata(period) + strata(age_cat) + strata(cardiomet),
  data = surv.data %>% filter(baseline_statin == 0),
  ties = "efron", id = DeID_PatientID
)

# Statin-treated: baseline_statin drops from strata (constant = 1)
fit_statin_treated <- coxph(
  Surv(tstart, tstop, MACE) ~ baseline_LDL +
    strata(period) + strata(age_cat) + strata(cardiomet),
  data = surv.data %>% filter(baseline_statin == 1),
  ties = "efron", id = DeID_PatientID
)

extract_hr <- function(fit, label, term = "baseline_LDL") {
  broom::tidy(fit, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(.data$term == .env$term) %>%
    mutate(
      Model      = label,
      HR         = round(estimate, 4),
      `95% CI`   = paste0("(", round(conf.low, 4),
                          "–", round(conf.high, 4), ")"),
      `p-value`  = round(p.value, 4),
      `N obs`    = fit$n,
      `N events` = fit$nevent
    ) %>%
    select(Model, HR, `95% CI`, `p-value`, `N obs`, `N events`)
}

bind_rows(
  extract_hr(fit_unadj,          "Unadjusted"),
  extract_hr(fit_no_comorbid,    "Stratified, no comorbidity"),
  extract_hr(fit_primary,        "Stratified + cardiomet — PRIMARY"),
  extract_hr(fit_full,           "Stratified + cardiomet + renal + obesity"),
  extract_hr(fit_statin_naive,   "Statin-naive at baseline only"),
  extract_hr(fit_statin_treated, "Statin-treated at baseline only")
) %>%
  kable(caption = "Baseline LDL-C HR per 1 mg/dL across model specifications")

broom::tidy(fit_primary_40, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term == "baseline_LDL_40") %>%
  mutate(across(where(is.numeric), ~round(.x, 4))) %>%
  select(HR = estimate, `95% CI low` = conf.low,
         `95% CI high` = conf.high, `p-value` = p.value) %>%
  kable(caption = "Primary baseline model — LDL-C HR per 40 mg/dL")
```

The statin-naive subgroup analysis (Sensitivity 3) is the most methodologically comparable to landmark studies such as Framingham, which enrolled largely untreated populations. If the LDL-C effect is strongest in this subgroup, it supports the interpretation that treatment dilution explains the null result in the full cohort.

---

## Comparison with Primary Time-Varying Model

This section compiles the key results from both analytical approaches side by side. Differences in the LDL-C HR between the two models reflect the methodological distinction: baseline LDL-C captures pre-treatment lipid biology at a single point, while time-varying LDL-C reflects the current treated level throughout follow-up.

```{r model-comparison}
# Load primary model results for comparison
# These values are taken from ldlc_mace_analysis.qmd output
# Update these manually if the primary script is re-run

primary_tvc_results <- tibble(
  Model     = c("Time-varying LDL-C — primary (730d)",
                "Time-varying LDL-C — sensitivity (1825d)"),
  Approach  = "Time-varying (LOCF)",
  HR        = c(0.9816, 0.9776),
  CI_low    = c(0.9454, 0.9415),
  CI_high   = c(1.0191, 1.0150),
  `p-value` = c(0.3316, 0.2367)
)

# Extract baseline model results (per 40 mg/dL)
baseline_results <- broom::tidy(fit_primary_40,
                                 exponentiate = TRUE,
                                 conf.int     = TRUE) %>%
  filter(term == "baseline_LDL_40") %>%
  transmute(
    Model     = "Baseline LDL-C — primary",
    Approach  = "Fixed baseline",
    HR        = round(estimate, 4),
    CI_low    = round(conf.low, 4),
    CI_high   = round(conf.high, 4),
    `p-value` = round(p.value, 4)
  )

# Statin-naive subgroup (per 40 mg/dL)
fit_statin_naive_40 <- coxph(
  Surv(tstart, tstop, MACE) ~ I(baseline_LDL / 40) +
    on_diabetes + on_hypertension +
    strata(period) + strata(age_cat),
  data = surv.data %>% filter(baseline_statin == 0),
  ties = "efron", id = DeID_PatientID
)

naive_results <- broom::tidy(fit_statin_naive_40,
                              exponentiate = TRUE,
                              conf.int     = TRUE) %>%
  filter(str_detect(term, "baseline_LDL")) %>%
  transmute(
    Model     = "Baseline LDL-C — statin-naive subgroup",
    Approach  = "Fixed baseline",
    HR        = round(estimate, 4),
    CI_low    = round(conf.low, 4),
    CI_high   = round(conf.high, 4),
    `p-value` = round(p.value, 4)
  )

bind_rows(primary_tvc_results, baseline_results, naive_results) %>%
  mutate(`95% CI` = paste0("(", CI_low, "–", CI_high, ")")) %>%
  select(Model, Approach, HR, `95% CI`, `p-value`) %>%
  kable(caption = "LDL-C effect on MACE — per 40 mg/dL across analytical approaches")
```

---

## Forest Plot

```{r forest-plot}
#| fig-width: 10
#| fig-height: 6

all_results <- bind_rows(
  tibble(
    Model    = c("Time-varying: full cohort (730d)",
                 "Time-varying: full cohort (1825d)"),
    Approach = "Time-varying LDL-C",
    HR       = c(0.9816, 0.9776),
    CI_low   = c(0.9454, 0.9415),
    CI_high  = c(1.0191, 1.0150)
  ),
  broom::tidy(fit_primary_40, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term == "baseline_LDL_40") %>%
    transmute(
      Model    = "Baseline: full cohort (primary)",
      Approach = "Fixed baseline LDL-C",
      HR       = round(estimate, 4),
      CI_low   = round(conf.low, 4),
      CI_high  = round(conf.high, 4)
    ),
  broom::tidy(fit_statin_naive_40, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(str_detect(term, "baseline_LDL")) %>%
    transmute(
      Model    = "Baseline: statin-naive subgroup",
      Approach = "Fixed baseline LDL-C",
      HR       = round(estimate, 4),
      CI_low   = round(conf.low, 4),
      CI_high  = round(conf.high, 4)
    )
) %>%
  # Factor with levels in display order (bottom to top on y axis)
  mutate(Model = factor(Model, levels = c(
    "Time-varying: full cohort (1825d)",
    "Time-varying: full cohort (730d)",
    "Baseline: full cohort (primary)",
    "Baseline: statin-naive subgroup"
  )))

ggplot(all_results,
       aes(x = HR, xmin = CI_low, xmax = CI_high, y = Model,
           colour = Approach, shape = Approach)) +
  geom_vline(xintercept = 1,    linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 1.20, linetype = "dotted", colour = "red",
             alpha = 0.6) +
  annotate("text", x = 1.21, y = 4.3,
           label = "RCT expected\n(~1.20/40mg/dL)",
           size = 3, colour = "red", hjust = 0) +
  geom_errorbarh(height = 0.2) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("Time-varying LDL-C"   = "#888888",
                                  "Fixed baseline LDL-C" = "#00274c")) +
  scale_shape_manual(values  = c("Time-varying LDL-C"   = 16,
                                  "Fixed baseline LDL-C" = 17)) +
  scale_y_discrete() +
  labs(
    title    = "LDL-C effect on MACE — per 40 mg/dL",
    subtitle = paste0("Stratified Cox models; red dotted line = ",
                      "RCT-expected effect (~HR 1.20 per 40 mg/dL)"),
    x = "Hazard Ratio (95% CI)", y = NULL,
    colour = "Analytical approach", shape = "Analytical approach"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        legend.position  = "bottom")
```

The red dotted line at HR = 1.20 marks the effect size expected from RCT evidence (CTT Collaboration meta-analyses). Observational estimates falling substantially below this line suggest treatment dilution or other forms of confounding specific to real-world LDL-C measurement.

---

## Session Information

```{r session-information}
sessionInfo()
```
