---
title: "Cleaning of Medication Data"
author: "Dave Bridges"
date: "2026-03-25"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  gfm:
    html-math-method: webtex

theme: journal

execute:
  echo: true
  warning: false

knitr:
  opts_chunk:
    fig-path: "figures/"          # folder for all figure files
    dev: ["png", "pdf"]           # both formats for every plot
    dpi: 300                      # ← this is the correct place for resolution
    dev.args:
      png:
        type: "cairo-png"         # better anti-aliasing / font rendering (optional but recommended)
      pdf:
        family: "sans"            # font family (optional)
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)
library(knitr)
library(broom)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

Clean the data for various medication data, for use in other scrupts

## Raw Data

```{r data-input}
library(readr) #loads the readr package
meds.filename <- "combined_data/MedicationOrdersComprehensive.csv" #input file(s)

meds_raw <- read_csv(meds.filename) |>
  mutate(DeID_PatientID = as.character(DeID_PatientID)) 

cat("Total medication rows loaded:", nrow(meds_raw), "\n")
```

These data can be found in `r getwd()` in a file named `r ifelse(meds.filename %in% dir(), meds.filename, "no file found")`.  This input file was most recently updated on `r ifelse(meds.filename %in% dir(), format(file.info(meds.filename)$mtime, "%Y-%m-%d"), "unknown")`.  This script was most recently updated on `r date()`.

There are `r distinct(meds_raw, DeID_PatientID) |> nrow()` participants in this dataset with meds values.

## Data Cleaning

### Defining Statin Use

```{r statin-use}
# Match on known statin drug names ONLY — avoids nystatin, bystatin, etc.
statin_pattern <- regex(
  "\\b(atorvastatin|simvastatin|rosuvastatin|pravastatin|
       lovastatin|fluvastatin|pitavastatin|cerivastatin)\\b",
  ignore_case = TRUE
)

meds_statins_raw <- meds_raw |>
  filter(str_detect(MedicationName, statin_pattern))

cat("Statin rows identified:", nrow(meds_statins_raw), "\n")
cat("Unique patients with any statin order:",
    n_distinct(meds_statins_raw$DeID_PatientID), "\n")

# Quick check — confirm nystatin not included
cat("Nystatin rows (should be 0):",
    sum(str_detect(meds_statins_raw$MedicationName,
                   regex("nystatin", ignore_case = TRUE))), "\n")
```

### Parsing Name and Dose

```{r name-dose-parsing}
meds_statins <- meds_statins_raw |>
  mutate(
    # Standardised drug name (lowercase, no brand name parenthetical)
    drug_name = str_extract(MedicationName,
                            statin_pattern) |> str_to_lower(),

    # Dose in mg — extract first number preceding "mg" (case-insensitive)
    dose_mg = str_extract(MedicationName,
                          regex("([0-9]+(?:\\.[0-9]+)?)\\s*mg",
                                ignore_case = TRUE)) |>
      str_extract("[0-9]+(?:\\.[0-9]+)?") |>
      as.numeric(),

    # Parse order date
    order_date = mdy_hm(DeID_OrderStart) |> as_date()
  )

# ============================================================
# BLOCK 1b: Patch dose_mg for the NA rows
# ============================================================

meds_statins <- meds_statins |>
  mutate(
    dose_mg = case_when(

      # Already parsed — keep as-is
      !is.na(dose_mg) ~ dose_mg,

      # Vytorin / ezetimibe-simvastatin combos:
      # "10/40" or "10-40" → second number is the simvastatin dose
      str_detect(MedicationName,
                 regex("ezetimibe.simvastatin|vytorin",
                       ignore_case = TRUE)) ~
        str_extract(MedicationName,
                    regex("(?:[0-9]+[/\\-])([0-9]+)",
                          ignore_case = TRUE)) |>
        str_extract("[0-9]+$") |>
        as.numeric(),

      # Niacin-simvastatin — MinDose is ambiguous (could be niacin dose)
      # Flag as NA and keep a marker column
      str_detect(MedicationName,
                 regex("niacin", ignore_case = TRUE)) ~ NA_real_,

      # Bulk/misc formulations — not classifiable
      str_detect(MedicationName,
                 regex("bulk|misc", ignore_case = TRUE)) ~ NA_real_,

      # Everything else with NA dose: try MinDose
      # (simple "SIMVASTATIN ORAL", salt forms, amlodipine combo, etc.)
      !is.na(MinDose) ~ as.numeric(MinDose),

      # Still NA — leave as NA
      TRUE ~ NA_real_
    ),

    # Flag combos where dose attribution is uncertain
    dose_source = case_when(
      str_detect(MedicationName, regex("ezetimibe.simvastatin|vytorin",
                                       ignore_case = TRUE)) ~ "vytorin_name",
      str_detect(MedicationName, regex("niacin",
                                       ignore_case = TRUE))  ~ "niacin_ambiguous",
      str_detect(MedicationName, regex("amlodipine",
                                       ignore_case = TRUE))  ~ "mindose",
      is.na(str_extract(MedicationName,
                        regex("[0-9]+(?:\\.[0-9]+)?\\s*mg",
                              ignore_case = TRUE)))           ~ "mindose",
      TRUE                                                    ~ "name_parsed"
    )
  )

# Check results
cat("Remaining NA doses after patch:\n")
meds_statins |>
  filter(is.na(dose_mg)) |>
  count(MedicationName, dose_source, sort = TRUE) |>
  print(n = 20)

cat("\nDose source breakdown:\n")
meds_statins |> count(dose_source, sort = TRUE) |> print()

cat("\nRe-run intensity classification check (spot-check combos):\n")
meds_statins |>
  filter(dose_source != "name_parsed") |>
  count(drug_name, dose_mg, dose_source, sort = TRUE) |>
  print(n = 30)
```

### Quality of Dose Extraction

```{r dose-extraction-quality}
cat("\n--- Dose extraction summary ---\n")
meds_statins |>
  count(drug_name, dose_mg, sort = TRUE) |>
  print(n = 30)

cat("\nRows with unparseable dose (NA):",
    sum(is.na(meds_statins$dose_mg)), "\n")
```

### Statin Dose Intensity Quantification

```{r statin-intensity}
# ACC/AHA 2013 statin intensity categories
# High:     reduces LDL ≥50%
# Moderate: reduces LDL 30-49%
# Low:      reduces LDL <30%

classify_intensity <- function(drug, dose) {
  case_when(
    # High intensity
    drug == "atorvastatin"  & dose >= 40              ~ "high",
    drug == "rosuvastatin"  & dose >= 20              ~ "high",
    # Moderate intensity
    drug == "atorvastatin"  & dose %in% c(10, 20)    ~ "moderate",
    drug == "rosuvastatin"  & dose %in% c(5, 10)     ~ "moderate",
    drug == "simvastatin"   & dose %in% c(20, 40)    ~ "moderate",
    drug == "pravastatin"   & dose >= 40              ~ "moderate",
    drug == "lovastatin"    & dose >= 40              ~ "moderate",
    drug == "fluvastatin"   & dose >= 80              ~ "moderate",
    drug == "pitavastatin"  & dose %in% c(2, 4)      ~ "moderate",
    # Low intensity
    drug == "simvastatin"   & dose <= 10              ~ "low",
    drug == "pravastatin"   & dose <= 20              ~ "low",
    drug == "lovastatin"    & dose <= 20              ~ "low",
    drug == "fluvastatin"   & dose <= 40              ~ "low",
    drug == "pitavastatin"  & dose == 1               ~ "low",
    # Cerivastatin (withdrawn 2001 — rare, treat as moderate if present)
    drug == "cerivastatin"                            ~ "moderate",
    # If dose is NA or doesn't match above, flag for review
    TRUE                                              ~ NA_character_
  )
}

meds_statins <- meds_statins |>
  mutate(
    intensity = classify_intensity(drug_name, dose_mg),
    intensity = factor(intensity, levels = c("low", "moderate", "high"))
  )

cat("Intensity classification:\n")
meds_statins |> count(intensity, drug_name, dose_mg) |> print(n = 40)
cat("\nUnclassified (NA intensity):", sum(is.na(meds_statins$intensity)), "\n")

# --- Status cleaning ---
# Keep:    Sent, Dispensed, Verified → prescription start events
# End:     Discontinued, Suspend     → prescription stop events
# Exclude: Canceled                  → never filled
# Exclude: Unknown                   → uninterpretable
# Exclude: Completed                 → too rare (n=1304) and ambiguous

meds_statins_clean <- meds_statins |>
  filter(!Status %in% c("Canceled", "Unknown", "Completed")) |>
  mutate(
    event_type = case_when(
      Status %in% c("Sent", "Dispensed", "Verified") ~ "start",
      Status %in% c("Discontinued", "Suspend")       ~ "stop"
    )
  ) |>
  filter(!is.na(order_date)) |>
  arrange(DeID_PatientID, drug_name, order_date)

cat("\nRows after status cleaning:", nrow(meds_statins_clean), "\n")
cat("Start events:", sum(meds_statins_clean$event_type == "start"), "\n")
cat("Stop events: ", sum(meds_statins_clean$event_type == "stop"), "\n")
```

### Patient-Level On/Off Intervals for Statins

```{r on-off-intervals}

SUPPLY_DAYS <- 90

# Use integer levels for safe comparison inside the loop
INTENSITY_LEVELS <- c("low" = 1L, "moderate" = 2L, "high" = 3L)

build_statin_intervals <- function(df) {

  starts <- df |>
    filter(event_type == "start") |>
    mutate(
      intensity_int = INTENSITY_LEVELS[as.character(intensity)]
    ) |>
    select(order_date, drug_name, intensity_int)

  stops <- df |>
    filter(event_type == "stop") |>
    pull(order_date)

  if (nrow(starts) == 0) return(tibble())

  intervals <- starts |>
    mutate(
      end_provisional = order_date + days(SUPPLY_DAYS),
      stop_date = map(order_date, \(s) {
        valid_stops <- stops[stops > s]
        if (length(valid_stops) == 0) NA_Date_ else min(valid_stops)
      }) |> map_vec(~ if (is.null(.x)) NA_Date_ else .x),
      period_end = pmin(end_provisional,
                        coalesce(stop_date, end_provisional))
    ) |>
    select(period_start = order_date, period_end, intensity_int)

  # Collapse overlapping / abutting intervals
  intervals <- intervals |> arrange(period_start)

  merged    <- vector("list", nrow(intervals))
  current   <- as.list(intervals[1, ])
  merge_idx <- 1L

  for (i in seq_len(nrow(intervals))[-1]) {
    row <- intervals[i, ]
    if (row$period_start <= current$period_end + days(1)) {
      current$period_end    <- max(current$period_end, row$period_end)
      current$intensity_int <- max(current$intensity_int,
                                   row$intensity_int, na.rm = TRUE)
    } else {
      merged[[merge_idx]] <- as_tibble(current)
      merge_idx <- merge_idx + 1L
      current   <- as.list(row)
    }
  }
  merged[[merge_idx]] <- as_tibble(current)

  bind_rows(merged[seq_len(merge_idx)]) |>
    mutate(
      intensity = factor(
        names(INTENSITY_LEVELS)[intensity_int],
        levels = names(INTENSITY_LEVELS)
      )
    ) |>
    select(period_start, period_end, intensity)
}

statin_intervals <- meds_statins_clean |>
  group_by(DeID_PatientID) |>
  group_modify(~ build_statin_intervals(.x)) |>
  ungroup()

cat("Statin interval rows built:", nrow(statin_intervals), "\n")
cat("Patients with >=1 statin interval:",
    n_distinct(statin_intervals$DeID_PatientID), "\n")

cat("\nInterval duration summary (days):\n")
statin_intervals |>
  mutate(duration = as.numeric(period_end - period_start)) |>
  summarise(
    min    = min(duration),
    median = median(duration),
    mean   = round(mean(duration), 1),
    p90    = quantile(duration, 0.9),
    max    = max(duration)
  ) |> print()

cat("\nIntensity distribution across intervals:\n")
statin_intervals |> count(intensity) |> print()
```


```{r output-writing}
output.file <- 'combined_data/MedicationOrdersCleanedStatins.csv'
write_csv(statin_intervals, output.file)
```

The cleaned results are output at **`r output.file`**.


## Session Information

```{r session-information}
sessionInfo()
```
